---
title: "Backtesting and scoring: which specification, for what"
subtitle: "An expanding-window pseudo-out-of-sample comparison of stage 1, stage 2b, stage 2c, stage 3 and a naive benchmark"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/evaluation.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir_eval <- file.path("data", "cache", "evaluation")
a <- readRDS(file.path(dir_eval, "analysis.rds"))
```

## A note on what this backtest is, and is not

**Every origin in this backtest trains on, and is scored against, the same
current EA-MD/QD and Eurostat data vintage.** There is no historical-vintage
archive in this project, so a 2012Q1 origin "knows" the 2012 data exactly as
it stands revised today, not as it was published in 2012. This is not a true
real-time evaluation, and it biases measured accuracy **upward** relative to
what a real-time forecaster would actually have achieved, for two reasons:
the actuals a forecast is scored against already reflect every subsequent
revision, and the training data feeding each origin's estimation is the
*revised* series, not the vintage that was actually available at the time.
The bias is largest for GDP components (heavily and repeatedly revised) and
smallest for financial-market series (policy rates, spreads), which are
essentially never revised after publication. Read every number below as an
upper bound on real-time performance, not an estimate of it.

## 1. Methodology

- **Expanding window**, origins from 2010Q1 to 2024Q1 (the latest origin that
  still leaves a full 8-quarter horizon inside the panel, which ends
  2026Q1), horizons 1-8 quarters, **re-estimating at every origin** -- this
  is not a fixed-coefficient rolling forecast.
- **Five specifications**: a closed-form naive AR(1) (growth-rate concepts) /
  driftless random-walk (rate/none concepts) benchmark, dispatched by each
  series' own tag; Stage 1 (eleven independent per-country models); Stage 2b
  (the eleven-country linked core system); **Stage 2c (the refined core --
  the same eleven countries with the four refinements
  `reports/stage2c_refined_core.qmd` adopted)**; Stage 3 (the DE-only extended
  system -- labour, external, fiscal and financial blocks -- the same
  configuration `reports/stage3b_external_fiscal_financial.qmd` and
  `reports/stage3c_rollout.qmd` established; the full eleven-country rollout
  was found infeasible there and is not re-attempted here).
- **Stage 2c is the one comparison in this report with no confound.** Its
  refinements are all `k`-neutral, so it has *exactly* the same `k` as stage
  2b at every origin, hence the same feasibility gate, the same usable
  origins, and the same sample. Every other pairwise comparison here differs
  in sample size, model capacity, or both. See §6.
- **Scoring space**: point and density scores (RMSE, MAE, CRPS, log score)
  are computed in **rate space** -- each quarter's growth-rate forecast
  against that quarter's realised growth -- because compounding posterior
  draws into levels over an 8-quarter horizon is exactly the pathology the
  spillover work (`reports/stage3c_rollout.qmd` §4) already found:
  explosive draws blow up multiplicatively in level space long before they
  would be flagged in rate space. `score_forecast()` also records a
  level-space RMSE/MAE for every row; it is not used below for the reason
  just given, but is available in the cached results for anyone who wants
  it.
- **Origin feasibility, and a stricter usability finding on top of it**:
  Stage 2 and Stage 3 self-skip any origin where `k >= T` (the same
  constraint documented throughout `CLAUDE.md`), logging why rather than
  silently omitting it. That gate turned out to be necessary but not
  sufficient -- see §2.
- **Re-estimation frequency**: Stage 1 and Stage 2 each ran at both
  quarterly and annual spacing, to measure the actual cost/precision
  trade-off rather than assert one -- see §3. Stage 3's feasible window is
  capacity-bound, not frequency-bound (see §2), so it only ran annual.
- **No per-origin tuning**: every origin within a spec reuses that spec's
  existing default Gibbs settings (2000 draws) and, for Stage 2/3, the
  already-tuned production `tau` -- consistent with how this project has
  handled the acceptance-rate band elsewhere (reported, not silently
  re-tuned away).
- **COVID handling is asymmetric across specs, by necessity.** Stage 1 has
  no dummy mechanism in this project -- its only COVID handling anywhere is
  the bespoke conditional-fill gap used for the one production window,
  which does not generalise to a rolling-origin backtest. So once a Stage 1
  origin passes 2020, the collapse enters its estimation sample as ordinary
  data. Stage 2/3 reuse the existing `covid_2020q1`-style dummies,
  restricted per origin to quarters that have actually occurred
  (`stage2b_dummies_through()`) so a not-yet-reached dummy does not hand
  koma an all-zero exogenous column. Separately, **forecast horizons whose
  target quarter falls in 2020Q1-2021Q4 are reported apart from calm-period
  horizons** throughout §4, so one shock quarter does not swamp fifty
  ordinary ones in a pooled average.

## 2. Which origins are actually usable

```{r feasibility}
kable(a$feasibility_summary, row.names = FALSE,
      caption = "Origins requested, passing the k < T gate ('feasible'), and actually producing a scoreable forecast ('usable').")
```

`feasible` (`k < T`) and `usable` (did `koma::forecast()` actually return a
non-degenerate result) are **not the same thing**, and the gap between them
is not noise:

```{r unusable}
kable(a$unusable, row.names = FALSE,
      caption = "Every origin that passed the k < T gate but failed at the forecast stage anyway.")
```

Every one of these has `df` of 1, 2 or 3. Every origin in the backtest with
`df >= 4` produced a usable forecast; every one with `df < 4` did not. This
is an exact, clean cutoff across 92 attempted joint-system estimations (77
Stage 2 + 15 Stage 3), not a rough tendency. Mechanically it is the same
failure this project has already documented at these boundary values --
`riwish(T - k, .)` drawing from a near-degenerate Wishart at `df` this small
throws `"v must be >= dimension of S in rwish()"` on a large share of draws,
and `reports/stage3c_rollout.qmd`'s own language ("no usable degrees of
freedom" at `df = 1`) anticipated exactly this. The practical upshot:
**`origin_feasible()`'s `k < T` check is the right first gate but not a
sufficient one** -- a caller backtesting this project's stage 2/3 systems
should treat `df < 4` as infeasible in practice, not just `df <= 0`.

The consequence for Stage 3 is severe: of fifteen requested annual origins,
only **one** -- 2024Q1, `df = 5` -- ever produces a scoreable forecast. Every
result in §4-5 involving Stage 3 rests on that single origin. It is reported
descriptively, not tested.

## 3. The re-estimation-frequency trade-off

```{r frequency-cost}
kable(a$freq_cost, digits = 1, row.names = FALSE,
      caption = "Measured cost of quarterly vs. annual re-estimation. 'origins_usable' per the df >= 4 finding in section 2.")
```

Quarterly buys real density for Stage 1 (57 usable origins vs. 15, at 3.9x
the cost) and for Stage 2 (20 usable origins vs. 5, at 4.0x the cost) --
almost exactly proportional, since Stage 1 has no feasibility gate at all
and Stage 2's usable window (2019Q2 onward) is long enough that quarterly
spacing multiplies it by roughly the frequency ratio. **The frequency knob
does nothing for Stage 3**: its usable window is one origin regardless of
spacing, because it is capacity-bound (df collapses too fast as the window
narrows) rather than frequency-bound. Quarterly re-estimation is worth
running for Stage 1 and Stage 2 if the extra evidence is worth ~40-70
minutes of compute; it cannot rescue Stage 3.

## 4. Head-to-head

### 4.1 Naive vs. Stage 1, quarterly, the full 2010Q1-2024Q1 window

This is the strongest comparison in the report: 57 origins, both specs
usable at every one, 1,254 (origin x country x concept) observations per
horizon.

```{r hh-quarterly-calm}
kable(a$lb_quarterly_calm[order(a$lb_quarterly_calm$horizon, a$lb_quarterly_calm$variant), ],
      row.names = FALSE, digits = 4, caption = "Calm-period horizons (target quarter outside 2020Q1-2021Q4).")
```

```{r hh-quarterly-covid}
kable(a$lb_quarterly_covid[order(a$lb_quarterly_covid$horizon, a$lb_quarterly_covid$variant), ],
      row.names = FALSE, digits = 4, caption = "COVID-era horizons (target quarter in 2020Q1-2021Q4) -- reported separately, not pooled in.")
```

RMSE tells an inconsistent-looking story (Stage 1 sometimes far ahead,
sometimes far behind) that MAE resolves: RMSE is dominated by a handful of
extreme misses on **both** sides, not a real difference in typical
accuracy -- see §5.3. MAE is the number to read here.

### 4.2 Naive vs. Stage 1 vs. Stage 2b vs. Stage 2c, annual, common window

Restricted to the five annual origins (2020-2024) where Stage 2 actually
produces a forecast (§2), so this is a fair, identical-origin comparison
across all four -- not Stage 2's full requested range.

```{r hh-annual-common}
kable(a$lb_annual_common[order(a$lb_annual_common$horizon, a$lb_annual_common$variant), ],
      row.names = FALSE, digits = 4, caption = "n = 110 (5 origins x 11 countries x 2 concepts) per row.")
```

Stage 2b is worse than both Stage 1 and naive on every metric at every
horizon here -- roughly 1.5-2x the RMSE, 3-5x the CRPS. This is a much
smaller, more volatile sample (5 origins, all post-2019) than §4.1, so the
margin should not be over-read quantitatively, but the direction is
unambiguous and consistent across all eight horizons. Stage 2c's position
relative to stage 2b is the subject of §6, on the larger quarterly sample.

### 4.3 Stage 3's one usable origin

```{r hh-stage3}
kable(a$lb_stage3_origin[order(a$lb_stage3_origin$horizon, a$lb_stage3_origin$variant), ],
      row.names = FALSE, digits = 4, caption = "2024Q1 only, n = 22 (11 countries x 2 concepts) per row -- descriptive, not tested.")
```

At this single origin, Stage 3 is not competitive with any of the other
three specs at any horizon -- by horizon 8 its RMSE (13.3) is roughly 7x
naive's (1.8) and its CRPS (7.9) is 12x naive's (0.6). One origin cannot
support a general claim, but it is directly consistent with everything
already established about this system: `reports/stage3c_rollout.qmd` found
its in-sample fit was not better than stage 2b's and its degrees of freedom
too thin to validate out of sample at all. This is that validation, run
once, and it confirms the worry rather than contradicting it.

## 5. Diebold-Mariano tests: naive vs. Stage 1, quarterly

**Nested-model caveat, stated up front.** Stage 1 is not literally nested in
Stage 2 in the classical parameter-restriction sense, but Stage 2 extends
five of Stage 1's six equations with additional cross-country terms and
adds a monetary block on top, and Stage 3 extends Stage 2 further still --
each specification is a structural superset of the one before it. Clark &
McCracken (2001) show the classical DM test's asymptotic normality is
invalid under the null of equal predictive ability when one model nests the
other, because the population loss differential degenerates to zero rather
than to a well-behaved limiting distribution. This is stated, not corrected
-- exactly the caveat the brief asked for, not a request to implement the
Clark-West correction. It does **not** apply to naive vs. Stage 1: the
closed-form AR(1)/RW benchmark is not nested in the structural system, so
this is the one clean DM comparison the backtest has real power for (the
Stage 2/3 samples are far too thin for a DM test to mean anything regardless
of nesting -- 5 and 1 origins respectively).

**A second, mechanical caveat.** `diebold_mariano()`'s HAC correction sums
autocovariances of the loss differential at lags `1..h-1`, computed on the
vector *in the order it is given*. A pooled panel has no single natural order,
so the statistic at `h > 1` depends on how the rows happen to be sorted --
which is a property of the test applied to pooled data, not of the data. The
table below fixes the order explicitly: sorted by `(variable, origin)`, so each
series' own origin sequence is contiguous and in time order, which is the only
ordering under which those autocovariance terms mean what the test assumes. At
`h = 1` there is no lag term and the statistic is order-free.

```{r dm-table}
kable(a$dm_table, row.names = FALSE, digits = c(0, 0, 4, 4, 3, 4, 3, 4, 3, 4),
      caption = "Naive vs. Stage 1, quarterly, n = 1254 per horizon. Negative dm_stat favours naive; positive favours Stage 1.")
```

### 5.1 What the RMSE-loss DM test says

**Nothing significant, in either direction, at any horizon** (all
`p_value_rmse > 0.2`). The point estimate favours Stage 1 by a widening
margin as horizon grows, but the DM statistic barely clears 1 even at
`mean_diff` in the tens of thousands (horizon 8) -- variance, not
direction, is the problem. §5.3 diagnoses why.

### 5.2 What the CRPS-loss DM test says

**Naive is significantly better at horizon 1** (`dm_stat = -3.84`,
`p < 0.001`) -- the one genuinely significant result in this report, and it
runs against the structural system, not for it. From horizon 2 onward the
CRPS difference is not significant either way (`p > 0.2` throughout).

Read together with §5.1's mean-squared-error point estimates favouring
Stage 1 at longer horizons, this is a specific, defensible claim: **Stage
1's point median catches up to and passes naive's from horizon 2 onward
(§5.3's MAE numbers make this precise), but nowhere in the eight horizons
tested does Stage 1's predictive *distribution* significantly beat naive's,
and at horizon 1 naive's distribution is significantly better.** A
structural system that cannot beat a random walk's density forecast even
once it starts beating its point forecast is not a system whose forecasts
should be trusted as calibrated, whatever the point-forecast horse race
says.

### 5.3 Why RMSE and MAE disagree, and the naive benchmark's own failure mode

```{r naive-explosion}
kable(a$naive_explosion, digits = c(0, 4, 1),
      caption = "Naive AR(1)/RW forecast: median vs. maximum squared error, by horizon. The median is essentially flat; the maximum is not.")
```

The naive benchmark's own AR(1) recursion is not immune to the instability
this project has repeatedly documented for koma's Gibbs sampler --
`naive_ar1_forecast()` is a deliberately plain textbook closed-form
recursion (`mean_h = mu + phi^h*(y_T - mu)`) with no stability guard, so a
country/concept whose estimated `phi` lands at or above 1 at some origin
produces a forecast that compounds geometrically over 8 quarters. The
median squared error stays close to 0.13-0.15 across every horizon; the
maximum reaches **42 million** at horizon 8. A handful of explosive AR(1)
fits, not a systematic difference in typical accuracy, is what drives
§5.1's noisy RMSE picture and the huge point-estimate swings in §4.1's RMSE
column. This was a deliberate design choice (§1: the naive benchmark is the
textbook closed-form, not clamped or winsorised) rather than an oversight,
but it means **MAE, not RMSE, is the point-forecast metric to trust** in
this report -- it is far less sensitive to the rare cases where either
model's forecast blows up, and it is the metric §5.2's mean_diff_mae
figures (naive 0.94 to 7.15 growing with horizon vs. Stage 1's flat
0.86-0.97) actually support the "Stage 1 pulls ahead on point forecasts
from horizon 2" reading above.

## 6. Stage 2b vs Stage 2c: the one clean comparison

Every other head-to-head in this report is confounded. Stage 1 vs Stage 2
changes the sample (57 usable origins vs 20) *and* the model. Stage 2 vs
Stage 3 changes the sample dramatically (20 vs 1) *and* the degrees of
freedom (`df` 22 vs 8). Stage 2b vs Stage 2c changes **neither**: every
stage-2c refinement is `k`-neutral, so the two systems have the same `k`,
the same `T`, the same `df`, the same feasibility gate and the same usable
origins at every point in the backtest.

```{r k-identity}
if (!is.null(a$k_identity_quarterly)) {
  ki <- a$k_identity_quarterly
  kable(data.frame(
    check = c("origins compared", "k identical at every origin",
              "df identical at every origin", "feasibility verdict identical"),
    result = c(nrow(ki), all(ki$k_2b == ki$k_2c), all(ki$df_2b == ki$df_2c),
               all(ki$feasible_2b == ki$feasible_2c))
  ), row.names = FALSE, caption = "Quarterly origins. The claim that makes this comparison clean, checked rather than asserted.")
}
```

### 6.1 Head-to-head on the common quarterly window

`r if (!is.null(a$stage2c_quarterly_keys)) sprintf("The %d quarterly origins where both systems produce a scoreable forecast -- four times the annual sample of §4.2.", length(a$stage2c_quarterly_keys)) else ""`

```{r hh-2b-2c}
if (!is.null(a$lb_quarterly_2b_2c)) {
  z <- a$lb_quarterly_2b_2c
  kable(z[order(as.integer(z$horizon), z$variant), ], row.names = FALSE, digits = 4,
        caption = "All horizons pooled across origins, countries and both concepts.")
}
```

```{r hh-2b-2c-calm}
if (!is.null(a$lb_quarterly_2b_2c_calm)) {
  z <- a$lb_quarterly_2b_2c_calm
  kable(z[order(as.integer(z$horizon), z$variant), ], row.names = FALSE, digits = 4,
        caption = "Calm-period horizons only (target quarter outside 2020Q1-2021Q4).")
}
```

```{r ratio-2b-2c}
if (!is.null(a$lb_quarterly_2b_2c)) {
  z <- a$lb_quarterly_2b_2c
  w <- merge(z[z$variant == "stage2", c("horizon", "rmse", "mae", "crps", "log_score")],
             z[z$variant == "stage2c", c("horizon", "rmse", "mae", "crps", "log_score")],
             by = "horizon", suffixes = c("_2b", "_2c"))
  w <- w[order(as.integer(w$horizon)), ]
  out <- data.frame(horizon = w$horizon,
                    rmse_ratio = w$rmse_2c / w$rmse_2b,
                    mae_ratio = w$mae_2c / w$mae_2b,
                    crps_ratio = w$crps_2c / w$crps_2b)
  kable(out, row.names = FALSE, digits = 4,
        caption = "Stage 2c relative to stage 2b. Below 1 favours stage 2c. MAE is the metric to read -- see 5.3.")
}
```

The headline is that **stage 2c forecasts worse than stage 2b**, and the gap
widens with horizon: the CRPS ratio rises monotonically from 0.97 at horizon 1
to 1.31 at horizon 8. The MAE ratio is close to 1 for most horizons and 1.36 at
horizon 8. Only at horizon 1 is stage 2c ahead on all three metrics.

Note that this **contradicts the five-origin annual sample** of §4.2, where
stage 2c looked *better* than stage 2b at long horizons (h=8 MAE 1.33 vs 1.61).
Twenty origins and four hundred and forty observations per horizon is the
sample to believe; five origins is not enough to sign this comparison, and the
fact that it signs it the other way is a useful calibration on how much weight
§4.2's Stage 2 numbers can bear.

### 6.2 The entire degradation is in prices

```{r by-concept}
if (!is.null(a$lb_quarterly_by_concept)) {
  z <- a$lb_quarterly_by_concept
  kable(z[order(z$concept, as.integer(z$horizon), z$variant),
          c("concept", "variant", "horizon", "rmse", "mae", "crps", "log_score", "n")],
        row.names = FALSE, digits = 4,
        caption = "Stage 2b vs stage 2c, split by concept.")
}
```

Split by concept, the aggregate picture resolves into two opposite ones.

**GDP forecasts are no worse and arguably better.** Stage 2c's CRPS is lower at
every horizon (4.07 vs 4.67 at horizon 1, 6.48 vs 7.16 at 4, 7.69 vs 8.33 at 8)
and its MAE is better at horizon 1.

**Price forecasts are dramatically worse**, and worsen with horizon: CRPS 0.91
vs 0.48 at horizon 1, 2.63 vs 0.62 at horizon 4, and 3.99 vs 0.62 at horizon 8
— a factor of six. RMSE at horizon 8 is 3.90 against 1.04.

### 6.3 Why: the Phillips curve exposed prices to the explosive draws

This is not a mystery, and it is not a bug. It follows mechanically from the
one term stage 2c added to the price equation.

```{r explosive-concept}
if (!is.null(a$explosive_by_concept)) {
  kable(a$explosive_by_concept, row.names = FALSE, digits = 4,
        col.names = c("Concept", "Horizon", "Stage 2b", "Stage 2c"),
        caption = "Share of posterior draws flagged explosive, averaged over origins and countries.")
}
```

In **stage 2b the price equation had no contemporaneous endogenous regressor at
all** — own lag, COVID dummies, `oil_price` and `eur_usd`, every one of them
exogenous. That made it effectively a univariate AR with cost-push terms, and
it was therefore **structurally immune** to the explosive-draw population that
this project has documented throughout: its explosive share is 0.0000 at
horizon 1 and 0.0026 at horizon 8. Essentially none.

Stage 2c's Phillips curve couples prices to `cc_gdp` — whose own explosive
share reaches 0.76 by horizon 8 — and prices inherit it wholesale: **0.016 at
horizon 1 rising to 0.634 at horizon 8**. GDP's own share is unchanged (0.764
vs 0.752). The price forecast did not get worse because the Phillips curve is
mis-specified; it got worse because prices are now connected to the rest of the
system, and the rest of the system has heavy-tailed draws.

That is the same term, and the same coupling, that closes the monetary loop.
`reports/stage2c_refined_core.qmd` §2 reports the benefit —
`ea_policy_rate` goes from reaching no price variable to reaching all twelve.
This is the bill for it. **One cannot be had without the other**: making prices
reachable from the policy rate necessarily makes them reachable from everything
else that reaches GDP, explosive draws included.

### 6.4 Diebold-Mariano, stage 2b vs stage 2c

The nesting caveat of §5 applies **less** here than anywhere else in this
report, which is worth stating precisely. Stage 2c is *not* a structural
superset of stage 2b: it removes eleven long-rate regressions and replaces them
with identities, while adding contemporaneous terms elsewhere. Neither model
nests the other, so the classical DM asymptotics are not invalidated by the
Clark-McCracken degeneracy that makes the Stage 1/2/3 comparisons unusable. The
sample is also the largest any stage-2 comparison in this report has.

The same explicit-ordering caveat from §5 applies to the `h > 1` rows.

```{r dm-2b-2c}
if (!is.null(a$dm_2b_vs_2c)) {
  kable(a$dm_2b_vs_2c, row.names = FALSE, digits = c(0, 0, 4, 4, 3, 4, 3, 4, 3, 4),
        caption = "Stage 2b vs stage 2c. Negative dm_stat favours stage 2b; positive favours stage 2c.")
}
```

**On CRPS the result is significant and consistent: stage 2b is better**, at
`p < 0.02` at every horizon except 2, and at `p < 0.002` from horizon 3 onward.
This is the only comparison in the entire report with both a large sample and a
defensible non-nested DM test, and it goes against the newer specification.

On MAE nothing is significant at conventional levels (the closest are horizons
4 and 8, `p = 0.058` and `p = 0.080`, both favouring stage 2b). So the point
forecast is roughly a draw and the **predictive distribution** is where stage
2c loses — which is exactly what a jump from 0.3% to 63% explosive draws
predicts.

## 7. Verdict: forecasting vs. scenario analysis

**Stage 2c is a better model of the system and a worse forecaster than Stage
2b, and §6.3 shows those are the same fact.** Its refinements close the
monetary loop, fix the spillover matrix's largest anomaly, and make that matrix
three times more reproducible across seeds
(`reports/stage2c_refined_core.qmd`, `reports/stage2c_spillovers.qmd`). The
mechanism that delivers all of that -- coupling prices to GDP -- is also what
takes the price forecast's explosive-draw share from 0.3% to 63% at horizon 8,
and loses the CRPS comparison at `p < 0.002`. This is the cleanest
structure-versus-forecast trade-off in the project, and the cleanest comparison
in this report: identical `k`, identical `df`, identical usable origins, so
nothing else can be blamed for it. **Use Stage 2c for mechanism and Stage 2b
for a price fan chart**, and do not pretend one specification does both.

**For forecasting, use Stage 1 -- with a real caveat, not an endorsement.**
It is the only specification with a fully powered, favourable comparison
(§4.1, 57 origins): its point forecasts (MAE) pull ahead of naive from
horizon 2 and the gap widens, and it is the cheapest of the three
structural specs to run (§3). But §5.2 is the qualifier that keeps this
from being a clean win: naive's *density* forecast is significantly better
at horizon 1 and not significantly worse at any other horizon, so Stage 1's
advantage is in the point median, not in a demonstrably better-calibrated
predictive distribution. Neither Stage 2 variant clears this bar -- both are
worse than naive and Stage 1 on every metric at every horizon in the annual
window (§4.2), and Stage 2c is worse than Stage 2b on the larger quarterly
window (§6) -- and Stage 3's single usable origin (§4.3) is worse still. None
should be preferred over Stage 1 for pure forecasting on this evidence, thin as
the Stage 3 evidence in particular is.

**For scenario analysis, the calculus is different, and this backtest does
not overturn it.** Stage 2's and Stage 3's entire reason to exist is
mechanisms Stage 1 cannot represent at all: cross-country trade and
monetary spillovers (Stage 2b/2c), and the wage-price loop, competitiveness
channel and debt-to-spread link (Stage 3, per `reports/stage3c_rollout.qmd`
§4.5's placebo-controlled finding that Stage 3's own-country amplification
is real). Within that use, **Stage 2c is the better instrument despite §6**:
a monetary scenario run on Stage 2b is being run on a system where the policy
rate provably cannot reach any price at all, which is a worse defect for
scenario work than a wider price fan is. A single-country model has no spillover to ask about and no
fiscal-financial loop to stress-test, regardless of how well it forecasts.
That said, this backtest adds a genuinely new, unfavourable data point for
Stage 3 specifically: its one testable origin was not just uncompetitive but
dramatically worse (§4.3), which is a second independent confirmation (the
first being `reports/stage3c_rollout.qmd`'s in-sample/out-of-sample-capacity
findings) that the extended system's structural richness has not yet been
shown to buy *any* predictive value, not even on the one variable set and
window it could be tested on. Use Stage 3 to reason about mechanisms it
alone can represent; do not report its point forecasts or fan charts as if
they carried the same evidentiary weight as Stage 1's.

**On the question the brief specifically asked**: at horizon 1, the
structural system does not beat the random-walk-style benchmark -- on CRPS
it loses significantly, and on RMSE/MAE the difference is not significant
in either direction. It only pulls ahead, and only on point forecasts, from
horizon 2 onward.
