---
title: "A multi-country Bayesian macro model"
subtitle: "Structure, results and forecasts across five estimated systems"
format:
  # Two outputs from one source, so the deck and the handout cannot drift.
  # Both formats emit .html, so each needs an explicit `output-file` or the
  # second render silently overwrites the first.
  revealjs:
    output-file: overview_presentation.html
    theme: simple
    slide-number: true
    embed-resources: true
    width: 1250
    height: 760
    fig-align: center
    code-line-numbers: false
  html:
    output-file: overview_handout.html
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/overview_presentation.qmd`.
# Reads only small cached artefacts -- shapes and templates from
# scratch/overview_build.R, forecasts from scratch/forecasts_build.R, and the
# stage-2d spillover battery. No fitted system is loaded here.
devtools::load_all(".", quiet = TRUE)
library(knitr)
library(ggplot2)
source("reports/_forecast_helpers.R")

shapes <- readRDS(file.path("data", "cache", "overview", "shapes.rds"))
templates <- readRDS(file.path("data", "cache", "overview", "templates.rds"))
fc <- stats::setNames(lapply(shapes$stage, forecast_artefact), shapes$stage)

dir_sp <- file.path("data", "cache", "stage2d_spillovers")
mat_a <- readRDS(file.path(dir_sp, "matrix_a.rds"))
mat_b <- readRDS(file.path(dir_sp, "matrix_b.rds"))
sanity <- readRDS(file.path(dir_sp, "sanity_a.rds"))
mon_a <- readRDS(file.path(dir_sp, "monetary_a.rds"))
oil_a <- readRDS(file.path(dir_sp, "oil_a.rds"))

sh <- function(stage, col) shapes[[col]][shapes$stage == stage]

# Presentation-friendly rendering of a koma equation string: spaces around the
# operators, and the four COVID dummies collapsed to one placeholder so the
# structure is legible from the back of a room.
pretty_eq <- function(x) {
  x <- gsub("\\+covid_[0-9]{4}q[1-4]", "", x)
  x <- sub("~constant\\+", " ~ ", x)
  x <- sub("~constant", " ~ ", x)
  x <- gsub("\\+", " + ", x)
  x <- gsub("(?<![=<>])-(?!-)", " - ", x, perl = TRUE)
  x <- gsub("==", " == ", x)
  gsub(" +", " ", x)
}
show_eqs <- function(stage, which = c("stochastic", "identity")) {
  which <- match.arg(which)
  e <- templates[[stage]]
  e <- if (which == "stochastic") grep("==", e, invert = TRUE, value = TRUE) else grep("==", e, value = TRUE)
  cat(paste(pretty_eq(e), collapse = "\n"))
}

# Explosive share by stage and concept -- the cross-stage chart.
expl <- do.call(rbind, lapply(shapes$stage, function(s) {
  d <- fc[[s]]$paths
  d <- d[d$kind == "forecast", ]
  d$concept <- sub("^[a-z]{2,3}_", "", d$variable)
  do.call(rbind, lapply(c("gdp", "prices"), function(cn) {
    x <- d[d$concept == cn, ]
    if (!nrow(x)) return(NULL)
    data.frame(stage = s, label = sh(s, "label"), df = sh(s, "df"), concept = cn,
               h8 = mean(x$explosive_frac[x$horizon == max(x$horizon)]),
               stringsAsFactors = FALSE)
  }))
}))

placebo_src <- c("de", "us", "cn", "reu")
pa <- mat_a[placebo_src, , drop = FALSE]; pb <- mat_b[placebo_src, , drop = FALSE]
off <- outer(placebo_src, colnames(mat_a), function(i, j) i != j)
placebo_mean <- mean(abs(pa - pb)[off]); placebo_max <- max(abs(pa - pb)[off])
placebo_flips <- sum(sign(pa[off]) != sign(pb[off])); placebo_n <- sum(off)

cum_by <- function(d, concept) {
  vars <- country_var(stage2d_countries(), concept)
  x <- d[d$variable %in% vars, ]
  x$country <- toupper(sub(paste0("_", concept, "$"), "", x$variable))
  vapply(split(x, x$country), function(z) sum(z$median_diff), numeric(1))
}
peak_by <- function(d, concept) {
  vars <- country_var(stage2d_countries(), concept)
  x <- d[d$variable %in% vars, ]
  x$country <- toupper(sub(paste0("_", concept, "$"), "", x$variable))
  vapply(split(x, x$country), function(z) z$median_diff[which.max(abs(z$median_diff))], numeric(1))
}
```

## What this is

::: {.content-visible unless-format="revealjs"}
*This is the handout form of `overview_presentation.qmd`. The same source also
renders as a slide deck (`overview_presentation.html`); this version is meant to
be read end to end rather than presented. Every figure is derived from the
cached fits — see the Reference section for which report argues each point in
full.*
:::

A **multi-country Bayesian simultaneous-equations macro model**, built on the
`koma` package.

::: {.incremental}
- **11 euro-area economies + the US**, quarterly, 2000–2024
- Every country's behavioural equations and accounting identities estimated in
  **one system**, so cross-country effects are jointly identified rather than
  bolted on afterwards
- Five estimated systems, each a deliberate step: **1 → 2a → 2b → 2c → 2d**,
  plus two extension blocks (3a, 3b)
- Everything below comes from cached fits and is reproducible from the repo
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

**The one thing that governs all of it:** every system spends observations on
parameters. `df = T − k`. Nearly every result in this deck is downstream of
that number.

## The country template — behavioural equations

::: {.content-visible unless-format="revealjs"}
The model is written once, for a generic country `cc`, and applied to every
economy. That is what makes the equation count manageable: stage 2c has 68
stochastic equations but only **eight distinct forms**, six of which are these,
applied identically eleven times. The forms below are stage 2c's and stage 2d's,
which are the same; earlier stages differ in the ways set out later.
:::

Six equations per country, identical in form for every economy:

```{r tmpl-stoch}
#| results: asis
cat("```\n"); show_eqs("stage2d", "stochastic"); cat("\n```\n")
```

::: {style="font-size: 0.75em"}
- Estimated in **growth-rate space** — coefficients on volumes are elasticities
- `long_rate`, `spread` and the policy rates are **rates**: koma leaves them
  untransformed, so their coefficients are pp per pp
- Four COVID impulse dummies (2020Q1–Q3, 2021Q2) enter the five real-side
  equations; suppressed above for readability
:::

## The country template — identities

Enforced **during** estimation, not fitted. Weights are fixed shares:

```{r tmpl-ident}
#| results: asis
cat("```\n"); show_eqs("stage2d", "identity"); cat("\n```\n")
```

::: {style="font-size: 0.75em"}
- **GDP** — expenditure decomposition, exact only up to the statistical discrepancy
- **Domestic demand** — absorption; government is out of the system entirely
- **Foreign demand** — the trade-weighted external demand index exports load on.
  **This is what links the countries.** The weights shown are Germany's; the
  partner set and shares are country-specific by construction
- **Long rate** — a sovereign yield *is* the policy path plus a premium; both
  weights imposed at exactly 1
:::

## Plus, once per system

```{r tmpl-shared}
#| results: asis
cat("```\n")
cat("ea_policy_rate ~ ea_prices + ea_gdp + ea_policy_rate.L(1)\n")
cat("us_policy_rate ~ us_prices + us_gdp + us_policy_rate.L(1)   (stage 2b onward)\n")
cat("cn_policy_rate ~ cn_prices + cn_gdp + cn_policy_rate.L(1)   (stage 2d only)\n\n")
cat("ea_gdp    == sum_i g_i * <member i>_gdp\n")
cat("ea_prices == sum_i g_i * <member i>_prices\n")
cat("```\n")
```

There is exactly **one** `ea_policy_rate` equation however many euro-area
countries the system contains — that is what makes it a genuinely common
instrument rather than ten separate ones.

## The stages at a glance

::: {.content-visible unless-format="revealjs"}
Each stage is a deliberate step rather than an iteration on the last. Stages 2a
through 2d change the *specification* or the *country partition* — never both at
once, which is what keeps any pair of them comparable. Stages 3a and 3b add
blocks to one country, Germany, and are where the degrees-of-freedom budget
finally binds.
:::

```{r shapes-table}
kable(
  data.frame(
    Stage = shapes$label,
    Scope = shapes$scope,
    Eqs = paste0(shapes$stochastic, " + ", shapes$identities),
    k = shapes$k, T = shapes$t_obs, df = shapes$df,
    Window = paste0(shapes$est_start, "–", shapes$est_end)
  ),
  row.names = FALSE, align = "llcrrrl"
)
```

::: {style="font-size: 0.8em"}
Stage 1's `k`/`T`/`df` are **per country** — its eleven satellites never share a
design matrix. Everything from 2a on is one joint system.
:::

## Stages 1 and 2a — the mechanism

::: {.columns}
::: {.column width="50%"}
**Stage 1** — each country its own small system, world variables exogenous.

**Stage 2a** — DE + FR in *one* system. Three things change, and they are the
whole point:

- exports driven by **partner GDP**, not an exogenous world aggregate
- `ea_policy_rate` becomes **endogenous**
- `long_rate` enters **investment**
:::
::: {.column width="50%"}
**The finding that set the agenda:**

Making the policy rate endogenous did **not** give it a route to output.
Without `long_rate` in investment it reached only the two long-rate
equations — which are terminal.

> "A rate rise cannot move GDP."

Structure has to be *traced*, not assumed.
:::
:::

## Stage 2b — all eleven, and an open loop

`r sh("stage2b","stochastic")` stochastic equations, `r sh("stage2b","identities")` identities,
`k = `r sh("stage2b","k")``, `df = `r sh("stage2b","df")``.

::: {.content-visible when-format="revealjs"}
. . .
:::

Tracing what `ea_policy_rate` can reach within a quarter:

::: {style="font-size: 1.1em"}
**91 of 103** endogenous variables — including every country's GDP, and itself.

**0 of 11** price variables.
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

The Taylor rule responds to an inflation rate that **nothing it does can
influence**. The monetary block was an open loop, and no amount of estimation
would have revealed it — only a structural trace did.

## Stage 2c — closing the loop, for free

Four refinements, applied to every country. **All four cost zero `k`** — same
`k = `r sh("stage2c","k")``, same `df = `r sh("stage2c","df")`` as stage 2b, so the two are
directly comparable.

```{r refinements}
kable(data.frame(
  `#` = 1:4,
  Refinement = c("`prices` gains `gdp`", "`foreign_demand` over partner **imports**",
                 "model the **spread**, `long_rate` becomes an identity",
                 "`consumption` gains `long_rate`"),
  Why = c("the Phillips curve — closes the monetary loop",
          "partner GDP nets imports out, so it had the wrong sign",
          "policy-rate loading straddled zero in all 11",
          "a second monetary channel"),
  check.names = FALSE
), row.names = FALSE)
```

::: {.content-visible when-format="revealjs"}
. . .
:::

**Result:** the policy rate now reaches **all 114** endogenous variables,
including every price. A fifth refinement was estimated and **rejected** — it
halved in-sample RMSE while gutting the demand elasticity in 9 of 11 countries.
*Improved fit was the symptom, not the reassurance.*

## Stage 2c — and what closing the loop cost

```{r price-explosive}
#| fig-width: 9
#| fig-height: 3.6
d <- expl[expl$concept == "prices" & expl$stage %in% c("stage2b", "stage2c", "stage2d"), ]
d$label <- factor(d$label, levels = d$label)
ggplot(d, aes(label, h8)) +
  geom_col(fill = "#4477AA", width = 0.55) +
  geom_text(aes(label = sprintf("%.3f", h8)), vjust = -0.4, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "share of price draws exploding by h8") +
  theme_minimal(base_size = 13)
```

In stage 2b `prices` had **no** contemporaneous endogenous regressor — it was
effectively a univariate AR, and therefore **structurally immune** to explosive
draws. Coupling it to GDP makes it reachable from the policy rate *and* from
everything else that reaches GDP.

**Use stage 2c for mechanism, stage 2b for a price fan chart.**

## Stage 2d — the first stage that *buys* degrees of freedom

Stage 2c's equations, a **re-partitioned world**: `DE`, `FR`, `IT`, `US`
separate, the other seven euro-area economies collapsed into a **`REU` bloc**,
and **China** added.

```{r df-table}
kable(data.frame(
  Stage = c("Stage 2c", "Stage 3a", "Stage 3b", "**Stage 2d**"),
  Change = c("11 countries, refined", "+ German labour block", "+ external/fiscal/financial",
             "**re-partition to 6 entities**"),
  k = c(sh("stage2c","k"), sh("stage3a","k"), sh("stage3b","k"), sh("stage2d","k")),
  df = c(sh("stage2c","df"), sh("stage3a","df"), sh("stage3b","df"), sh("stage2d","df"))
), row.names = FALSE, align = "llrr")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

Every earlier structural change traded `df` for detail: **22 → 15 → 8**.
Collapsing seven countries removes enough equations that `k` falls 76 → 46 —
so `df` rises to **`r sh("stage2d","df")`** even after giving up eighteen quarters of sample.

## Stage 2d — China, and what the data does not have

::: {.columns}
::: {.column width="52%"}
**What exists** (free, keyless sources):

- real GDP, SA — World Bank GEM
- CPI — World Bank GEM
- merchandise trade volumes — GEM value ÷ price
- policy rate — OECD `IRSTCI`
- trade weights — **observed IMF DOTS**
:::
::: {.column width="48%"}
**What does not exist, anywhere:**

**Quarterly consumption and investment.**

Verified absent from OECD QNA, OECD Economic Outlook, IMF IFS, World Bank GEM,
FRED and the NBS itself.

→ the two equations **merge** into one estimated `cn_domestic_demand`, rather
than being invented from indicator proxies.
:::
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

::: {style="font-size: 0.85em"}
Two things stated rather than buried: `cn_long_rate` is **spliced over 45%** of
the window, and `reu_investment` is **not identified** — Ireland's investment sd
is 43.0 and it carries 10.9% of the bloc, so the seven-country average is *more*
volatile than any single member.
:::

## Stage 3a / 3b — the wall

Germany gets a labour/wage–price block, then external, fiscal and financial
blocks on top.

```{r stage3}
kable(data.frame(
  Stage = c("Stage 2b", "Stage 3a", "Stage 3b"),
  Adds = c("—", "labour + disaggregated prices (DE only)", "+ external, fiscal, financial (DE only)"),
  k = c(sh("stage2b","k"), sh("stage3a","k"), sh("stage3b","k")),
  df = c(sh("stage2b","df"), sh("stage3a","df"), sh("stage3b","df"))
), row.names = FALSE, align = "llrr")
```

::: {.incremental}
- **It cannot be rolled out.** Eleven countries with the full block needs
  `k = 142` against `T = 98`.
- At `df = 8`, `de_long_rate`'s own lag crossed to **1.004** — explosive — the
  moment debt became its fifth regressor.
- A quiet loop is not a stable one: the fiscal–financial cycle showed no
  explosive draws *because one of its links was ~0*.
:::

## Forecasts — how they are produced

::: {.content-visible unless-format="revealjs"}
Adding forecasts was not just a reporting exercise: running every system forward
is what turned the degrees-of-freedom argument from a claim about estimation
into a measurable property of the output. The caveats below are not boilerplate
— each one is a trap that produced a wrong answer before it was handled.
:::

Every system is run **8 quarters** forward, 1000 posterior draws.

::: {.incremental}
- koma **silently shortens** a horizon whose exogenous series run out — they are
  extended first, and the delivered horizon is checked
- koma constrains no draw to be stationary, so **explosive draws are filtered**
  and the discarded share is reported
- Charts show an **inner 25–75%** band solid and an outer 10–90% band clipped to
  the panel edge; the untruncated bounds are tabulated
- Stages 1 and 2a forecast from **2023Q1** (window already observed — a genuine
  out-of-sample check); stages 2b onward from **2025Q1**
:::

## Forecast — stage 2d GDP

```{r fan-2d}
#| fig-width: 11
#| fig-height: 4.9
v <- fc_vars(fc$stage2d$paths, stage2d_countries(), "gdp")
plot_forecast_fan(fc$stage2d$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  subtitle = "inner 25-75%, outer 10-90% clipped; grey = realised")
```

## Forecast — stage 1, against the outturn

```{r fan-1}
#| fig-width: 11
#| fig-height: 4.9
v <- fc_vars(fc$stage1$paths, c("de", "fr", "it", "es", "nl", "us"), "gdp")
plot_forecast_fan(fc$stage1$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  subtitle = "forecast from 2023Q1 -- the whole window is observed")
```

The realised path sits inside the band for every country, and the fan widens in
exactly the countries you would expect.

## The finding the forecasts made visible

```{r df-vs-explosive}
#| fig-width: 10
#| fig-height: 4.6
d <- expl[expl$concept == "gdp", ]
# Stage 1's df is per country, not comparable with a joint system's, so it is
# labelled but excluded from the fitted line.
d$joint <- d$stage != "stage1"
ggplot(d, aes(df, h8)) +
  geom_smooth(data = d[d$joint, ], method = "lm", se = FALSE,
              colour = "grey70", linewidth = 0.5, formula = y ~ x) +
  geom_point(aes(shape = joint), size = 3.5, colour = "#11336B") +
  # 2b and 2c sit on the same df, so one label goes below the point.
  geom_text(aes(label = label, vjust = ifelse(stage == "stage2c", 1.9, -1.1)), size = 4.2) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 1), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.16))) +
  scale_x_continuous(expand = expansion(mult = 0.09)) +
  labs(x = "degrees of freedom (T - k)",
       y = "share of GDP draws exploding by horizon 8",
       caption = "hollow point = stage 1, whose df is per country and not comparable") +
  theme_minimal(base_size = 13)
```

Degrees of freedom are not an abstraction. They are **exactly** how much of the
predictive distribution survives to horizon 8.

## Spillovers — stage 2d

Cumulative 8-quarter GDP effect of a **+1pp demand shock**; rows shocked,
columns responding.

```{r spill-matrix}
kable(round(mat_a, 3), align = "r")
```

::: {style="font-size: 0.8em"}
Own effect dominates **6 of 6**. China's row is an order of magnitude smaller
than Europe's — as its trade weights say it should be, now that they come from
observed DOTS rather than the reciprocal approximation that would have put 56%
of Chinese trade inside the euro area.
:::

## Spillovers — is any of it real?

::: {.content-visible unless-format="revealjs"}
koma has no impulse-response function, so a spillover is the difference between
two forecasts — one unconditional, one restricted — paired by common random
numbers so the difference isolates the shock rather than koma's own
call-to-call noise. That pairing is exact. What it does **not** remove is
sampling error in the level of the estimand itself, and the only way to find out
how large that is, is to re-run the identical battery at a different seed.
:::

Same fit, same battery, **only the seed changed**:

```{r placebo}
kable(data.frame(
  System = c("Stage 2b (df 22)", "Stage 2c (df 22)", "**Stage 2d (df 32)**"),
  `Mean |A − B|` = c("0.350", "0.119", sprintf("**%.3f**", placebo_mean)),
  Max = c("1.850", "0.356", sprintf("**%.3f**", placebo_max)),
  `Sign flips` = c("7–10 of 30", "1 of 44", sprintf("**%d of %d**", placebo_flips, placebo_n)),
  check.names = FALSE
), row.names = FALSE, align = "lrrr")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

For stage 2b, **most of what anyone wanted to conclude was smaller than the
matrix's own noise.**

Stage 2d is the first system where no off-diagonal cell changes sign when only
the seed changes — the concrete payoff of those ten extra degrees of freedom.

## What does not work

```{r fails}
kable(data.frame(
  Check = c("Monetary → GDP (euro area)", "Monetary → prices",
            "Monetary shock on China's own rule", "Oil → prices",
            "Spillover ~ bilateral trade weight"),
  Result = c(
    sprintf("correctly signed, %d/6 peak negative", sum(peak_by(mon_a, "gdp") < 0)),
    sprintf("**wrong sign** — only %d/6 peak negative", sum(peak_by(mon_a, "prices") < 0)),
    "**does not work** — a horizon-1 adjustment in the rule, not transmission",
    sprintf("correctly signed **%d/6**", sum(cum_by(oil_a, "prices") > 0)),
    "**fails** — Spearman 0.138 against a 0.3 bar")
), row.names = FALSE, align = "ll")
```

::: {style="font-size: 0.85em"}
The oil shock is the control: the price block responds sensibly to a **cost**
shock and perversely to a **policy rate**. That localises the fault to the
policy rate's route in — not to the price equations.
:::

## Which system for which question

::: {.content-visible unless-format="revealjs"}
No single system dominates. The table below is the practical summary; the reason
each row is what it is has been given above.
:::

```{r verdict}
kable(data.frame(
  `If you want to…` = c(
    "understand a mechanism / trace transmission",
    "produce a price fan chart",
    "study EA–US–China spillovers",
    "say anything about Austria, Spain, Ireland…",
    "model German wages, debt or credit",
    "run a monetary experiment on China"),
  `Use` = c("**Stage 2c**", "**Stage 2b**", "**Stage 2d**", "Stage 2b / 2c",
            "Stage 3a / 3b — Germany only", "**nothing here yet**"),
  check.names = FALSE
), row.names = FALSE, align = "ll")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

::: {style="font-size: 0.9em"}
**The through-line:** every result in this deck that survived did so because it
was checked against its own noise floor — a seed placebo, a sign rule, a
structural trace, a second specification. The ones that did not survive are in
the deck too.
:::

## Reference

::: {style="font-size: 0.8em"}
| Report | What it covers |
|---|---|
| `stage1_diagnostics.qmd` | per-country satellites, diagnostics |
| `stage2a_pilot.qmd` / `stage2b_full_system.qmd` | the linkage mechanism, then all eleven |
| `stage2c_refined_core.qmd` / `stage2c_equations.qmd` | the four refinements, every equation |
| `stage2d_regional_core.qmd` / `stage2d_equations.qmd` | the re-partition and China |
| `stage2d_spillovers.qmd` | shock battery + seed placebo |
| `stage3a_labour_prices.qmd` / `stage3b_*.qmd` | the extension blocks |
| `evaluation.qmd` | expanding-window backtest, four specifications |

Artefacts: `scratch/overview_build.R`, `scratch/forecasts_build.R`,
`scratch/stage2d_build.R`, `scratch/stage2d_spillovers_build.R`.
:::
