---
title: "A multi-country Bayesian macro model"
subtitle: "Structure, results and forecasts across eight estimated systems"
format:
  # Two outputs from one source, so the deck and the handout cannot drift.
  # Both formats emit .html, so each needs an explicit `output-file` or the
  # second render silently overwrites the first.
  revealjs:
    output-file: overview_presentation.html
    theme: simple
    slide-number: true
    embed-resources: true
    width: 1250
    height: 760
    fig-align: center
    code-line-numbers: false
  html:
    output-file: overview_handout.html
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/overview_presentation.qmd`.
# Reads only small cached artefacts -- shapes and templates from
# scratch/overview_build.R, forecasts from scratch/forecasts_build.R, and the
# stage-2d spillover battery. No fitted system is loaded here.
devtools::load_all(".", quiet = TRUE)
library(knitr)
library(ggplot2)
source("reports/_forecast_helpers.R")

shapes <- readRDS(file.path("data", "cache", "overview", "shapes.rds"))
templates <- readRDS(file.path("data", "cache", "overview", "templates.rds"))
fc <- stats::setNames(lapply(shapes$stage, forecast_artefact), shapes$stage)

dir_sp <- file.path("data", "cache", "stage2d_spillovers")
mat_a <- readRDS(file.path(dir_sp, "matrix_a.rds"))
mat_b <- readRDS(file.path(dir_sp, "matrix_b.rds"))
sanity <- readRDS(file.path(dir_sp, "sanity_a.rds"))
mon_a <- readRDS(file.path(dir_sp, "monetary_a.rds"))
oil_a <- readRDS(file.path(dir_sp, "oil_a.rds"))

sh <- function(stage, col) shapes[[col]][shapes$stage == stage]

# Presentation-friendly rendering of a koma equation string: spaces around the
# operators, and the four COVID dummies collapsed to one placeholder so the
# structure is legible from the back of a room.
pretty_eq <- function(x) {
  x <- gsub("\\+covid_[0-9]{4}q[1-4]", "", x)
  x <- sub("~constant\\+", " ~ ", x)
  x <- sub("~constant", " ~ ", x)
  x <- gsub("\\+", " + ", x)
  x <- gsub("(?<![=<>])-(?!-)", " - ", x, perl = TRUE)
  x <- gsub("==", " == ", x)
  gsub(" +", " ", x)
}
show_eqs <- function(stage, which = c("stochastic", "identity")) {
  which <- match.arg(which)
  e <- templates[[stage]]
  e <- if (which == "stochastic") grep("==", e, invert = TRUE, value = TRUE) else grep("==", e, value = TRUE)
  cat(paste(pretty_eq(e), collapse = "\n"))
}

# Explosive share by stage and concept -- the cross-stage chart.
expl <- do.call(rbind, lapply(shapes$stage, function(s) {
  d <- fc[[s]]$paths
  d <- d[d$kind == "forecast", ]
  d$concept <- sub("^[a-z]{2,3}_", "", d$variable)
  do.call(rbind, lapply(c("gdp", "prices"), function(cn) {
    x <- d[d$concept == cn, ]
    if (!nrow(x)) return(NULL)
    data.frame(stage = s, label = sh(s, "label"), df = sh(s, "df"), concept = cn,
               h8 = mean(x$explosive_frac[x$horizon == max(x$horizon)]),
               stringsAsFactors = FALSE)
  }))
}))

placebo_src <- c("de", "us", "cn", "reu")
pa <- mat_a[placebo_src, , drop = FALSE]; pb <- mat_b[placebo_src, , drop = FALSE]
off <- outer(placebo_src, colnames(mat_a), function(i, j) i != j)
placebo_mean <- mean(abs(pa - pb)[off]); placebo_max <- max(abs(pa - pb)[off])
placebo_flips <- sum(sign(pa[off]) != sign(pb[off])); placebo_n <- sum(off)

cum_by <- function(d, concept) {
  vars <- country_var(stage2d_countries(), concept)
  x <- d[d$variable %in% vars, ]
  x$country <- toupper(sub(paste0("_", concept, "$"), "", x$variable))
  vapply(split(x, x$country), function(z) sum(z$median_diff), numeric(1))
}
peak_by <- function(d, concept) {
  vars <- country_var(stage2d_countries(), concept)
  x <- d[d$variable %in% vars, ]
  x$country <- toupper(sub(paste0("_", concept, "$"), "", x$variable))
  vapply(split(x, x$country), function(z) z$median_diff[which.max(abs(z$median_diff))], numeric(1))
}

# ---------------------------------------------------------------------------
# The handout carries a stage-by-stage template walk-through that the deck does
# not: eight stages' worth of equations is a reference section, not something
# to put in front of a room. Rather than wrap `##` headings in a
# `content-visible` div -- which leaves empty slides behind in revealjs -- the
# whole section is emitted from R and simply produces nothing for the deck.
is_deck <- identical(knitr::opts_knit$get("rmarkdown.pandoc.to"), "revealjs")

# Split a koma equation into its left-hand side, its operator and its
# regressor names, with the intercept, the COVID dummies and any identity
# WEIGHT stripped. Comparing on names rather than on the rendered string is
# what lets the stage-to-stage diff below distinguish "this equation gained a
# regressor" from "this identity's weights were re-estimated on a new sample".
eq_parts <- function(e) {
  op <- if (grepl("==", e, fixed = TRUE)) "==" else "~"
  lhs <- sub("\\s*(~|==).*$", "", e)
  rhs <- sub("^.*?(~|==)", "", e)
  tm <- strsplit(gsub("-", "+-", rhs), "+", fixed = TRUE)[[1]]
  tm <- sub("^[-0-9.]+\\*", "", sub("^-", "", tm[nzchar(tm)]))
  list(lhs = lhs, op = op,
       terms = unique(tm[tm != "constant" & !grepl("^covid_", tm)]))
}

# What changed between one stage's template and its parent's. Derived, never
# transcribed -- the same rule the rest of this report follows, and the reason
# a stage cannot be described as doing something it does not do.
template_diff <- function(prev, cur) {
  idx <- function(e) { l <- lapply(e, eq_parts); stats::setNames(l, vapply(l, `[[`, "", "lhs")) }
  P <- idx(prev); C <- idx(cur)
  bits <- character(0)
  new <- setdiff(names(C), names(P))
  gone <- setdiff(names(P), names(C))
  if (length(new)) bits <- c(bits, sprintf("**new:** %s", paste0("`", new, "`", collapse = ", ")))
  if (length(gone)) bits <- c(bits, sprintf("**dropped:** %s", paste0("`", gone, "`", collapse = ", ")))
  for (v in intersect(names(P), names(C))) {
    a <- setdiff(C[[v]]$terms, P[[v]]$terms)
    r <- setdiff(P[[v]]$terms, C[[v]]$terms)
    flip <- if (P[[v]]$op != C[[v]]$op) {
      if (C[[v]]$op == "==") " — **now an identity**" else " — **now estimated**"
    } else ""
    # A `foreign_demand` identity's partner list is the country PARTITION, not
    # a specification change, so listing eleven added and six removed partners
    # would drown the one thing worth reading.
    if (grepl("_foreign_demand$", v) && (length(a) > 3 || length(r) > 3)) {
      bits <- c(bits, sprintf("`%s` — partner set follows the partition", v))
      next
    }
    if (!length(a) && !length(r) && !nzchar(flip)) next
    ch <- c(if (length(a)) paste0("+", a), if (length(r)) paste0("\u2212", r))
    bits <- c(bits, sprintf("`%s`%s%s", v, flip,
                            if (length(ch)) paste0(": ", paste0("`", ch, "`", collapse = ", ")) else ""))
  }
  if (!length(bits)) "no change to the template — the *partition* is what differs" else bits
}

# Each stage's template is a change to ONE earlier stage, and it is not always
# the previous row of the table: 3a extends 2b, 3d puts 3a's block on 2d.
# Getting this wrong would describe stage 3d as inventing a labour block.
tmpl_parent <- c(stage1 = NA, stage2a = "stage1", stage2b = "stage2a",
                 stage2c = "stage2b", stage2d = "stage2c", stage3a = "stage2b",
                 stage3b = "stage3a", stage3d = "stage2d")

# Forecast-error comparison. Two groups, never pooled: stages 1 and 2a forecast
# from 2023Q1 into a fully observed window, stages 2b onward from 2025Q1 where
# only five quarters have happened.
err_concepts <- c("gdp", "prices", "consumption", "investment", "exports", "imports")
# `de`/`fr`/`it`/`us` are the four entities every joint stage carries under its
# own name. `ea_gdp` is deliberately excluded even though it exists everywhere:
# it aggregates eleven countries in stage 2b and four in stage 2d, so scoring it
# across stages would compare two different objects.
err_entities <- c("de", "fr", "it", "us")
err_vars <- function(fcl) {
  v <- fc_common_vars(fcl, err_concepts)
  v[sub("_.*$", "", v) %in% err_entities]
}
late <- c("stage2b", "stage2c", "stage2d", "stage3a", "stage3b", "stage3d")
early <- c("stage1", "stage2a")
fc_late <- fc[late]; fc_early <- fc[early]
err_late <- fc_error_by_horizon(fc_late, err_vars(fc_late))
err_early <- fc_error_by_horizon(fc_early, err_vars(fc_early))
err_late$short <- sub(" \\(.*$", "", err_late$label)
err_early$short <- sub(" \\(.*$", "", err_early$label)

# The multi-origin backtest, for the section that says what a single origin
# cannot settle. Built by reports/evaluation.qmd's own pipeline.
evaluation <- readRDS(file.path("data", "cache", "evaluation", "analysis.rds"))
bt <- evaluation$lb_quarterly_2b_2c
# The leaderboard stores `horizon` as a character key, which a continuous scale
# cannot train on.
bt$horizon <- as.integer(bt$horizon)
bt$variant <- factor(bt$variant, levels = c("naive", "stage1", "stage2", "stage2c"),
                     labels = c("naive AR(1)", "Stage 1", "Stage 2b", "Stage 2c"))
```

## What this is

::: {.content-visible unless-format="revealjs"}
*This is the handout form of `overview_presentation.qmd`. The same source also
renders as a slide deck (`overview_presentation.html`); this version is meant to
be read end to end rather than presented. Every figure is derived from the
cached fits — see the Reference section for which report argues each point in
full.*
:::

A **multi-country Bayesian simultaneous-equations macro model**, built on the
`koma` package.

::: {.incremental}
- **11 euro-area economies + the US**, quarterly, 2000–2024
- Every country's behavioural equations and accounting identities estimated in
  **one system**, so cross-country effects are jointly identified rather than
  bolted on afterwards
- Five estimated systems, each a deliberate step: **1 → 2a → 2b → 2c → 2d**,
  plus two extension blocks (3a, 3b)
- Everything below comes from cached fits and is reproducible from the repo
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

**The one thing that governs all of it:** every system spends observations on
parameters. `df = T − k`. Nearly every result in this deck is downstream of
that number.

```{r stage-template-walk}
#| results: asis
if (!is_deck) {
  cat("## The equation template, stage by stage\n\n")
  cat(paste(
    "Every stage is written once for a generic country `cc` and applied to each",
    "economy, so a system with 68 stochastic equations still has only a handful",
    "of distinct *forms*. That is what makes the stages comparable at all, and",
    "it is why this section is the right place to start: read the eight blocks",
    "below and you have read the whole model.\n\n"))
  cat(paste(
    "The equations are **extracted from the fitted systems**, not transcribed --",
    "`scratch/overview_build.R` takes one representative entity's equations off",
    "each `sys_eq` and replaces its own prefix with `cc`. A block below therefore",
    "cannot describe a specification that was not estimated. The intercept and the",
    "four COVID impulse dummies are suppressed throughout for readability; `.L(1)`",
    "is a one-quarter lag; `==` marks an identity, imposed during estimation rather",
    "than fitted.\n\n"))
  cat(paste(
    "Each block ends with what changed **relative to the stage it descends from**,",
    "which is not always the previous row: stage 3a extends stage 2b, and stage 3d",
    "puts stage 3a's block on stage 2d's partition. The diffs are computed from the",
    "two templates rather than written by hand. They compare regressor *names*,",
    "so a change of identity weight alone -- an expenditure share re-derived on a",
    "different sample, say -- does not appear as a change; the weights are printed",
    "in full above each diff for exactly that reason.\n\n"))
  for (st in shapes$stage) {
    row <- shapes[shapes$stage == st, ]
    cat(sprintf("### %s — %s\n\n", row$label, row$scope))
    cat(sprintf("`k = %d`, `T = %d`, **`df = %d`** · %s–%s · %d stochastic + %d identities\n\n",
                row$k, row$t_obs, row$df, row$est_start, row$est_end,
                row$stochastic, row$identities))
    stoch <- grep("==", templates[[st]], invert = TRUE, value = TRUE)
    ident <- grep("==", templates[[st]], value = TRUE)
    cat("```\n", paste(pretty_eq(stoch), collapse = "\n"), "\n```\n\n", sep = "")
    if (length(ident)) {
      cat("*Identities:*\n\n")
      cat("```\n", paste(pretty_eq(ident), collapse = "\n"), "\n```\n\n", sep = "")
    }
    par <- tmpl_parent[[st]]
    if (!is.na(par)) {
      cat(sprintf("**Changed from %s:**\n\n", shapes$label[shapes$stage == par]))
      cat(paste0("- ", template_diff(templates[[par]], templates[[st]]), collapse = "\n"), "\n\n")
    }
  }
  cat(paste(
    "Two things are visible in that sequence and in no single block. The first is",
    "that the **behavioural** forms barely move after stage 2c — 2d changes the",
    "partition and not the specification, and 3d adds 3a's block unchanged — so",
    "the differences between those systems are differences of *sample and",
    "capacity*, not of theory. The second is that every step that adds detail",
    "spends `df`, and the `df` column above falls monotonically until stage 2d",
    "re-partitions and buys it back.\n\n"))
}
```

## The country template — behavioural equations

::: {.content-visible unless-format="revealjs"}
The model is written once, for a generic country `cc`, and applied to every
economy. That is what makes the equation count manageable: stage 2c has 68
stochastic equations but only **eight distinct forms**, six of which are these,
applied identically eleven times. The forms below are stage 2c's and stage 2d's,
which are the same; earlier stages differ in the ways set out later.
:::

Six equations per country, identical in form for every economy:

```{r tmpl-stoch}
#| results: asis
cat("```\n"); show_eqs("stage2d", "stochastic"); cat("\n```\n")
```

::: {style="font-size: 0.75em"}
- Estimated in **growth-rate space** — coefficients on volumes are elasticities
- `long_rate`, `spread` and the policy rates are **rates**: koma leaves them
  untransformed, so their coefficients are pp per pp
- Four COVID impulse dummies (2020Q1–Q3, 2021Q2) enter the five real-side
  equations; suppressed above for readability
:::

## The country template — identities

Enforced **during** estimation, not fitted. Weights are fixed shares:

```{r tmpl-ident}
#| results: asis
cat("```\n"); show_eqs("stage2d", "identity"); cat("\n```\n")
```

::: {style="font-size: 0.75em"}
- **GDP** — expenditure decomposition, exact only up to the statistical discrepancy
- **Domestic demand** — absorption; government is out of the system entirely
- **Foreign demand** — the trade-weighted external demand index exports load on.
  **This is what links the countries.** The weights shown are Germany's; the
  partner set and shares are country-specific by construction
- **Long rate** — a sovereign yield *is* the policy path plus a premium; both
  weights imposed at exactly 1
:::

## Plus, once per system

```{r tmpl-shared}
#| results: asis
cat("```\n")
cat("ea_policy_rate ~ ea_prices + ea_gdp + ea_policy_rate.L(1)\n")
cat("us_policy_rate ~ us_prices + us_gdp + us_policy_rate.L(1)   (stage 2b onward)\n")
cat("cn_policy_rate ~ cn_prices + cn_gdp + cn_policy_rate.L(1)   (stage 2d only)\n\n")
cat("ea_gdp    == sum_i g_i * <member i>_gdp\n")
cat("ea_prices == sum_i g_i * <member i>_prices\n")
cat("```\n")
```

There is exactly **one** `ea_policy_rate` equation however many euro-area
countries the system contains — that is what makes it a genuinely common
instrument rather than ten separate ones.

## The stages at a glance

::: {.content-visible unless-format="revealjs"}
Each stage is a deliberate step rather than an iteration on the last. Stages 2a
through 2d change the *specification* or the *country partition* — never both at
once, which is what keeps any pair of them comparable. Stages 3a and 3b add
blocks to one country, Germany, and are where the degrees-of-freedom budget
finally binds.
:::

```{r shapes-table}
kable(
  data.frame(
    Stage = shapes$label,
    Scope = shapes$scope,
    Eqs = paste0(shapes$stochastic, " + ", shapes$identities),
    k = shapes$k, T = shapes$t_obs, df = shapes$df,
    Window = paste0(shapes$est_start, "–", shapes$est_end)
  ),
  row.names = FALSE, align = "llcrrrl"
)
```

::: {style="font-size: 0.8em"}
Stage 1's `k`/`T`/`df` are **per country** — its eleven satellites never share a
design matrix. Everything from 2a on is one joint system.
:::

## Stages 1 and 2a — the mechanism

::: {.columns}
::: {.column width="50%"}
**Stage 1** — each country its own small system, world variables exogenous.

**Stage 2a** — DE + FR in *one* system. Three things change, and they are the
whole point:

- exports driven by **partner GDP**, not an exogenous world aggregate
- `ea_policy_rate` becomes **endogenous**
- `long_rate` enters **investment**
:::
::: {.column width="50%"}
**The finding that set the agenda:**

Making the policy rate endogenous did **not** give it a route to output.
Without `long_rate` in investment it reached only the two long-rate
equations — which are terminal.

> "A rate rise cannot move GDP."

Structure has to be *traced*, not assumed.
:::
:::

## Stage 2b — all eleven, and an open loop

`r sh("stage2b","stochastic")` stochastic equations, `r sh("stage2b","identities")` identities,
`k = `r sh("stage2b","k")``, `df = `r sh("stage2b","df")``.

::: {.content-visible when-format="revealjs"}
. . .
:::

Tracing what `ea_policy_rate` can reach within a quarter:

::: {style="font-size: 1.1em"}
**91 of 103** endogenous variables — including every country's GDP, and itself.

**0 of 11** price variables.
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

The Taylor rule responds to an inflation rate that **nothing it does can
influence**. The monetary block was an open loop, and no amount of estimation
would have revealed it — only a structural trace did.

## Stage 2c — closing the loop, for free

Four refinements, applied to every country. **All four cost zero `k`** — same
`k = `r sh("stage2c","k")``, same `df = `r sh("stage2c","df")`` as stage 2b, so the two are
directly comparable.

```{r refinements}
kable(data.frame(
  `#` = 1:4,
  Refinement = c("`prices` gains `gdp`", "`foreign_demand` over partner **imports**",
                 "model the **spread**, `long_rate` becomes an identity",
                 "`consumption` gains `long_rate`"),
  Why = c("the Phillips curve — closes the monetary loop",
          "partner GDP nets imports out, so it had the wrong sign",
          "policy-rate loading straddled zero in all 11",
          "a second monetary channel"),
  check.names = FALSE
), row.names = FALSE)
```

::: {.content-visible when-format="revealjs"}
. . .
:::

**Result:** the policy rate now reaches **all 114** endogenous variables,
including every price. A fifth refinement was estimated and **rejected** — it
halved in-sample RMSE while gutting the demand elasticity in 9 of 11 countries.
*Improved fit was the symptom, not the reassurance.*

## Stage 2c — and what closing the loop cost

```{r price-explosive}
#| fig-width: 9
#| fig-height: 3.6
d <- expl[expl$concept == "prices" & expl$stage %in% c("stage2b", "stage2c", "stage2d"), ]
d$label <- factor(d$label, levels = d$label)
ggplot(d, aes(label, h8)) +
  geom_col(fill = "#4477AA", width = 0.55) +
  geom_text(aes(label = sprintf("%.3f", h8)), vjust = -0.4, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "share of price draws exploding by h8") +
  theme_minimal(base_size = 13)
```

In stage 2b `prices` had **no** contemporaneous endogenous regressor — it was
effectively a univariate AR, and therefore **structurally immune** to explosive
draws. Coupling it to GDP makes it reachable from the policy rate *and* from
everything else that reaches GDP.

**Use stage 2c for mechanism, stage 2b for a price fan chart.**

## Stage 2d — the first stage that *buys* degrees of freedom

Stage 2c's equations, a **re-partitioned world**: `DE`, `FR`, `IT`, `US`
separate, the other seven euro-area economies collapsed into a **`REU` bloc**,
and **China** added.

```{r df-table}
kable(data.frame(
  Stage = c("Stage 2c", "Stage 3a", "Stage 3b", "**Stage 2d**"),
  Change = c("11 countries, refined", "+ German labour block", "+ external/fiscal/financial",
             "**re-partition to 6 entities**"),
  k = c(sh("stage2c","k"), sh("stage3a","k"), sh("stage3b","k"), sh("stage2d","k")),
  df = c(sh("stage2c","df"), sh("stage3a","df"), sh("stage3b","df"), sh("stage2d","df"))
), row.names = FALSE, align = "llrr")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

Every earlier structural change traded `df` for detail: **22 → 15 → 8**.
Collapsing seven countries removes enough equations that `k` falls 76 → 46 —
so `df` rises to **`r sh("stage2d","df")`** even after giving up eighteen quarters of sample.

## Stage 2d — China, and what the data does not have

::: {.columns}
::: {.column width="52%"}
**What exists** (free, keyless sources):

- real GDP, SA — World Bank GEM
- CPI — World Bank GEM
- merchandise trade volumes — GEM value ÷ price
- policy rate — OECD `IRSTCI`
- trade weights — **observed IMF DOTS**
:::
::: {.column width="48%"}
**What does not exist, anywhere:**

**Quarterly consumption and investment.**

Verified absent from OECD QNA, OECD Economic Outlook, IMF IFS, World Bank GEM,
FRED and the NBS itself.

→ the two equations **merge** into one estimated `cn_domestic_demand`, rather
than being invented from indicator proxies.
:::
:::

::: {.content-visible when-format="revealjs"}
. . .
:::

::: {style="font-size: 0.85em"}
Two things stated rather than buried: `cn_long_rate` is **spliced over 45%** of
the window, and `reu_investment` is **not identified** — Ireland's investment sd
is 43.0 and it carries 10.9% of the bloc, so the seven-country average is *more*
volatile than any single member.
:::

## Stage 3a / 3b — the wall

Germany gets a labour/wage–price block, then external, fiscal and financial
blocks on top.

```{r stage3}
kable(data.frame(
  Stage = c("Stage 2b", "Stage 3a", "Stage 3b"),
  Adds = c("—", "labour + disaggregated prices (DE only)", "+ external, fiscal, financial (DE only)"),
  k = c(sh("stage2b","k"), sh("stage3a","k"), sh("stage3b","k")),
  df = c(sh("stage2b","df"), sh("stage3a","df"), sh("stage3b","df"))
), row.names = FALSE, align = "llrr")
```

::: {.incremental}
- **It cannot be rolled out** — on eleven countries. The full block for all of
  them needs `k = 142` against `T = 98`.
- At `df = 8`, `de_long_rate`'s own lag crossed to **1.004** — explosive — the
  moment debt became its fifth regressor.
- A quiet loop is not a stable one: the fiscal–financial cycle showed no
  explosive draws *because one of its links was ~0*.
:::

::: {.content-visible unless-format="revealjs"}
Stage 3c set out to roll these blocks across the euro area and concluded the
degrees-of-freedom budget forbade it. That conclusion held for the
eleven-country partition; the next slide is what happened when the partition
changed.
:::

## Stage 3d — the rollout, on the regional core

Stage 3c's verdict was that further structure needed **a longer sample or a
smaller system**. Stage 2d built the smaller system; stage 3d puts the labour
block on it.

```{r stage3d}
kable(data.frame(
  Configuration = c("DE labour only", "DE + FR labour", "**DE + REU labour**", "DE + REU + FR + IT"),
  Partition = c("11 countries", "11 countries", "**6 entities**", "6 entities"),
  k = c(83, 90, sh("stage3d", "k"), 74),
  df = c(15, 8, sh("stage3d", "df"), 4),
  `Economies covered` = c(1, 2, 8, 10),
  check.names = FALSE
), row.names = FALSE, align = "llrrr")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

**The block still costs seven columns per entity.** What changed is what a
column buys: a bloc covers **seven economies for one entity's worth of `k`**.

::: {style="font-size: 0.85em"}
And the aggregation is not only a cost. The **wage Phillips curve is identified
on the bloc** (−0.176, interval excluding zero) and not on Germany (−0.105,
straddling zero) — the bloc's unemployment rate carries the crisis-era
variation Germany's does not. What is lost is the *institutional* coefficients:
wage indexation and ULC pass-through both go wrong-signed on the bloc. The bloc
has the variation but not the institutions; Germany the reverse.
:::

## Forecasts — how they are produced

::: {.content-visible unless-format="revealjs"}
Adding forecasts was not just a reporting exercise: running every system forward
is what turned the degrees-of-freedom argument from a claim about estimation
into a measurable property of the output. The caveats below are not boilerplate
— each one is a trap that produced a wrong answer before it was handled.
:::

Every system is run **8 quarters** forward, 1000 posterior draws.

::: {.incremental}
- koma **silently shortens** a horizon whose exogenous series run out — they are
  extended first, and the delivered horizon is checked
- koma constrains no draw to be stationary, so **explosive draws are filtered**
  and the discarded share is reported
- Charts show an **inner 25–75%** band solid and an outer 10–90% band clipped to
  the panel edge; the untruncated bounds are tabulated
- Stages 1 and 2a forecast from **2023Q1** (window already observed — a genuine
  out-of-sample check); stages 2b onward from **2025Q1**
:::

## Forecast — stage 2d GDP

```{r fan-2d}
#| fig-width: 11
#| fig-height: 4.9
v <- fc_vars(fc$stage2d$paths, stage2d_countries(), "gdp")
plot_forecast_fan(fc$stage2d$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  subtitle = "inner 25-75%, outer 10-90% clipped; grey = realised")
```

## Forecast — stage 1, against the outturn

```{r fan-1}
#| fig-width: 11
#| fig-height: 4.9
v <- fc_vars(fc$stage1$paths, c("de", "fr", "it", "es", "nl", "us"), "gdp")
plot_forecast_fan(fc$stage1$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  subtitle = "forecast from 2023Q1 -- the whole window is observed")
```

The realised path sits inside the band for every country, and the fan widens in
exactly the countries you would expect.

## Forecast — German GDP, every stage side by side

::: {.content-visible unless-format="revealjs"}
The charts above show one stage each. This one holds the *variable* fixed and
varies the *system*, which is the comparison that actually separates the stages:
same country, same origin, same eight quarters, same 1000 draws — only the model
changes. Stages 1 and 2a are absent because they forecast from 2023Q1, a
different origin, and putting them on this axis would compare two different
questions.
:::

```{r fan-cross-gdp}
#| fig-width: 11
#| fig-height: 4.6
cmp <- fc_stage_compare(fc_late, "de_gdp")
plot_forecast_fan(cmp$paths, cmp$vars, space = "rate", ncol = 3,
                  labeller = cmp$labeller,
                  subtitle = "de_gdp, all six systems that forecast from 2025Q1; grey = realised")
```

```{r cross-gdp-note}
#| results: asis
w <- do.call(rbind, lapply(names(fc_late), function(s) {
  d <- fc_late[[s]]$paths
  d <- d[d$kind == "forecast" & d$variable == "de_gdp" & d$horizon == 8, ]
  data.frame(stage = s, df = sh(s, "df"),
             width = d$rate_upper - d$rate_lower, median = d$rate_median)
}))
cat(sprintf(paste(
  "The **medians barely differ** -- %.2f to %.2fpp at horizon 8, a spread of",
  "%.1fpp. What differs is the **width of the band**: %.0f to %.0fpp across the",
  "same six systems, and it tracks `df` at a rank correlation of **%.2f**. The",
  "systems disagree far less about where growth is going than about how much they",
  "know.\n"),
  min(w$median), max(w$median), diff(range(w$median)),
  min(w$width), max(w$width),
  stats::cor(w$df, w$width, method = "spearman")))
```

## Forecast — German prices, every stage side by side

```{r fan-cross-prices}
#| fig-width: 11
#| fig-height: 4.6
cmpp <- fc_stage_compare(fc_late, "de_prices")
plot_forecast_fan(cmpp$paths, cmpp$vars, space = "rate", ncol = 3,
                  labeller = cmpp$labeller,
                  subtitle = "de_prices, same six systems; grey = realised")
```

::: {.content-visible unless-format="revealjs"}
This is the stage-2c price finding as a picture. Stage 2b's `cc_prices` has no
contemporaneous endogenous regressor, so it is effectively a univariate AR with
exogenous cost-push terms and is *structurally immune* to the explosive-draw
population. Stage 2c's Phillips curve couples it to `cc_gdp`, and prices inherit
GDP's explosive share — which is the price of making prices reachable from the
policy rate at all.
:::

```{r cross-prices-expl}
# forecast_explosive_table() filters to kind == "forecast" itself; indexing
# `paths` on horizon alone would also pick up the history rows.
expl_de_prices <- do.call(rbind, lapply(names(fc_late), function(s) {
  e <- forecast_explosive_table(fc_late[[s]]$paths, "de_prices")
  data.frame(Stage = sub(" \\(.*$", "", fc_late[[s]]$label),
             `Explosive share of draws, h1` = sprintf("%.3f", e$h1),
             `h8` = sprintf("%.3f", e$h_last),
             check.names = FALSE, stringsAsFactors = FALSE)
}))
kable(expl_de_prices, row.names = FALSE, align = "lrr")
```

## Forecast — stage 2d, prices across all six entities

```{r fan-2d-prices}
#| fig-width: 11
#| fig-height: 4.9
v <- fc_vars(fc$stage2d$paths, stage2d_countries(), "prices")
plot_forecast_fan(fc$stage2d$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  subtitle = "inner 25-75%, outer 10-90% clipped; grey = realised")
```

::: {.content-visible unless-format="revealjs"}
## Forecast — the wider grids

The two grids below are handout-only: they carry more panels than a slide can
hold, and are the reference version of the two charts above.

```{r fan-2b-grid}
#| fig-width: 10
#| fig-height: 7.5
v <- fc_vars(fc$stage2b$paths,
             c("de", "fr", "it", "es", "nl", "be", "at", "pt", "gr", "ie", "us"), "gdp")
plot_forecast_fan(fc$stage2b$paths, v, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(v),
                  title = "Stage 2b — GDP, all eleven economies",
                  subtitle = "forecast from 2025Q1; grey = realised where it exists")
```

```{r fan-3d-grid}
#| fig-width: 10
#| fig-height: 5.2
v <- c(fc_vars(fc$stage3d$paths, c("de", "reu"), "wages"),
       fc_vars(fc$stage3d$paths, c("de", "reu"), "unemployment"),
       fc_vars(fc$stage3d$paths, c("de", "reu"), "nonenergy_prices"))
plot_forecast_fan(fc$stage3d$paths, v, space = "rate", ncol = 3,
                  labeller = fc_full_labels(v),
                  title = "Stage 3d — the labour block's own variables",
                  subtitle = "the wage-price channel, on Germany and the REU bloc")
```
:::

## Forecast errors — the stages against the outturn

::: {.content-visible unless-format="revealjs"}
Everything above is about the *shape* of the predictive distribution. This is
about where its centre actually landed. Five quarters of 2025 and 2026 have now
been published, so the six systems that forecast from 2025Q1 can be scored on
horizons 1 to 5 against realised data none of them saw.

The comparison is only meaningful on a **shared variable set**: every stage is
scored on the same `r length(err_vars(fc_late))` series — Germany, France, Italy
and the US, on GDP, prices, consumption, investment, exports and imports.
`ea_gdp` is excluded even though every stage carries it, because it aggregates
eleven countries in stage 2b and four in stage 2d and would compare two
different objects. Errors are in **rate space**, quarterly growth in percentage
points; a level error would compound horizon 1's mistake into every later
horizon and would additionally be incomparable for stages 1 and 2a, whose level
paths are anchored on filled rather than observed values.
:::

Same `r length(err_vars(fc_late))` series, same origin, same outturn:

```{r err-curve}
#| fig-width: 10
#| fig-height: 4.4
ggplot(err_late, aes(horizon, mae, colour = short)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(err_late$horizon)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "horizon (quarters ahead of 2025Q1)",
       y = "mean absolute error, pp of quarterly growth",
       caption = sprintf("%d variable-quarters behind each point", err_late$n[1])) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Forecast errors — the numbers

```{r err-table}
w <- reshape(err_late[, c("short", "horizon", "mae")], idvar = "short",
             timevar = "horizon", direction = "wide")
names(w) <- c("Stage", paste0("h", sort(unique(err_late$horizon))))
w[-1] <- lapply(w[-1], function(x) sprintf("%.2f", x))
w$Mean <- sprintf("%.2f", vapply(split(err_late$mae, err_late$short),
                                 mean, numeric(1))[w$Stage])
kable(w, row.names = FALSE, align = c("l", rep("r", ncol(w) - 1)))
```

```{r err-rank}
#| results: asis
# Derived, not transcribed: a hand-written ranking here went stale the first
# time the artefacts were rebuilt.
m <- vapply(split(err_late$mae, err_late$short), mean, numeric(1))
m <- sort(m)
cat(sprintf(paste(
  "::: {style=\"font-size: 0.85em\"}\nThe spread across stages is **small relative to",
  "the level of the error** -- every system is between %.2f and %.2fpp at every",
  "horizon, and the ordering changes from one horizon to the next. Averaged over",
  "the five horizons **%s** is lowest at %.2f and **%s** highest at %.2f, a gap of",
  "%.0f%%. On five quarters and a single origin that ordering is a sanity check,",
  "not a ranking.\n:::\n"),
  min(err_late$mae), max(err_late$mae),
  names(m)[1], m[1], names(m)[length(m)], m[length(m)],
  100 * (m[length(m)] / m[1] - 1)))
```

## Forecast errors — what one origin cannot settle

::: {.content-visible unless-format="revealjs"}
The previous two slides are one draw of the world. This project has already been
burned by exactly that: on a five-origin annual sample stage 2c looked *better*
than stage 2b at long horizons, and on a twenty-origin quarterly sample it is
clearly worse. A single origin cannot sign a comparison between systems this
close together, and the table above should be read as a sanity check that none
of them is broken — not as a ranking.

The honest version is an expanding-window backtest that re-estimates every
system at every origin. `reports/evaluation.qmd` runs it: 92 joint-system
estimations, twenty usable quarterly origins, four specifications. It predates
stages 2d and 3d, so those are absent here.
:::

**Twenty origins, re-estimated at each one** — the comparison that can carry weight:

```{r backtest-curve}
#| fig-width: 10
#| fig-height: 4.2
ggplot(bt, aes(horizon, mae, colour = variant)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:8) +
  # Log scale, because the unclamped naive benchmark reaches 18.8pp at horizon 8
  # while the models sit near 2 -- on a linear axis the four model lines would
  # be indistinguishable from each other and from the floor.
  scale_y_log10() +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(x = "horizon (quarters)", y = "mean absolute error, pp (log scale)",
       caption = sprintf("%s variable-quarters per point; expanding window, re-estimated at each origin", bt$n[1])) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

::: {style="font-size: 0.85em"}
The naive AR(1) benchmark **explodes** past horizon 6 — deliberately unclamped,
which is the honest textbook comparison, but it is why RMSE is not readable here
and MAE is. Stage 1 beats the joint systems on MAE at every horizon; the joint
systems buy cross-country structure, not point accuracy.
:::

## Forecast errors — stages 1 and 2a, on their own origin

::: {.content-visible unless-format="revealjs"}
Stages 1 and 2a forecast from 2023Q1 into a window that is now entirely
observed, so they get all eight horizons scored — but against different quarters
of history than the six systems above, which is why they are not on the same
chart. Their common set is Germany and France only, `r length(err_vars(fc_early))`
series.
:::

```{r err-early}
kable(
  {
    w2 <- reshape(err_early[, c("short", "horizon", "mae")], idvar = "short",
                  timevar = "horizon", direction = "wide")
    names(w2) <- c("Stage", paste0("h", sort(unique(err_early$horizon))))
    w2[-1] <- lapply(w2[-1], function(x) sprintf("%.2f", x))
    w2
  }, row.names = FALSE, align = c("l", rep("r", 8)))
```

Adding the linkage mechanism (2a) neither helped nor hurt the point forecast at
this origin — which is what stage 2a was for. It was a test that the *mechanism*
worked, not a bid for accuracy.

## The finding the forecasts made visible

```{r df-vs-explosive}
#| fig-width: 10
#| fig-height: 4.6
d <- expl[expl$concept == "gdp", ]
# Stage 1's df is per country, not comparable with a joint system's, so it is
# labelled but excluded from the fitted line.
d$joint <- d$stage != "stage1"
ggplot(d, aes(df, h8)) +
  geom_smooth(data = d[d$joint, ], method = "lm", se = FALSE,
              colour = "grey70", linewidth = 0.5, formula = y ~ x) +
  geom_point(aes(shape = joint), size = 3.5, colour = "#11336B") +
  # 2b and 2c sit on the same df, so one label goes below the point.
  geom_text(aes(label = label, vjust = ifelse(stage == "stage2c", 1.9, -1.1)), size = 4.2) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 1), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.16))) +
  scale_x_continuous(expand = expansion(mult = 0.09)) +
  labs(x = "degrees of freedom (T - k)",
       y = "share of GDP draws exploding by horizon 8",
       caption = "hollow point = stage 1, whose df is per country and not comparable") +
  theme_minimal(base_size = 13)
```

Degrees of freedom are not an abstraction. They are **exactly** how much of the
predictive distribution survives to horizon 8.

## Spillovers — stage 2d

Cumulative 8-quarter GDP effect of a **+1pp demand shock**; rows shocked,
columns responding.

```{r spill-matrix}
kable(round(mat_a, 3), align = "r")
```

::: {style="font-size: 0.8em"}
Own effect dominates **6 of 6**. China's row is an order of magnitude smaller
than Europe's — as its trade weights say it should be, now that they come from
observed DOTS rather than the reciprocal approximation that would have put 56%
of Chinese trade inside the euro area.
:::

## Spillovers — is any of it real?

::: {.content-visible unless-format="revealjs"}
koma has no impulse-response function, so a spillover is the difference between
two forecasts — one unconditional, one restricted — paired by common random
numbers so the difference isolates the shock rather than koma's own
call-to-call noise. That pairing is exact. What it does **not** remove is
sampling error in the level of the estimand itself, and the only way to find out
how large that is, is to re-run the identical battery at a different seed.
:::

Same fit, same battery, **only the seed changed**:

```{r placebo}
kable(data.frame(
  System = c("Stage 2b (df 22)", "Stage 2c (df 22)", "**Stage 2d (df 32)**"),
  `Mean |A − B|` = c("0.350", "0.119", sprintf("**%.3f**", placebo_mean)),
  Max = c("1.850", "0.356", sprintf("**%.3f**", placebo_max)),
  `Sign flips` = c("7–10 of 30", "1 of 44", sprintf("**%d of %d**", placebo_flips, placebo_n)),
  check.names = FALSE
), row.names = FALSE, align = "lrrr")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

For stage 2b, **most of what anyone wanted to conclude was smaller than the
matrix's own noise.**

Stage 2d is the first system where no off-diagonal cell changes sign when only
the seed changes — the concrete payoff of those ten extra degrees of freedom.

## What does not work

```{r fails}
kable(data.frame(
  Check = c("Monetary → GDP (euro area)", "Monetary → prices",
            "Monetary shock on China's own rule", "Oil → prices",
            "Spillover ~ bilateral trade weight"),
  Result = c(
    sprintf("correctly signed, %d/6 peak negative", sum(peak_by(mon_a, "gdp") < 0)),
    sprintf("**wrong sign** — only %d/6 peak negative", sum(peak_by(mon_a, "prices") < 0)),
    "**does not work** — a horizon-1 adjustment in the rule, not transmission",
    sprintf("correctly signed **%d/6**", sum(cum_by(oil_a, "prices") > 0)),
    "**fails** — Spearman 0.138 against a 0.3 bar")
), row.names = FALSE, align = "ll")
```

::: {style="font-size: 0.85em"}
The oil shock is the control: the price block responds sensibly to a **cost**
shock and perversely to a **policy rate**. That localises the fault to the
policy rate's route in — not to the price equations.
:::

## Which system for which question

::: {.content-visible unless-format="revealjs"}
No single system dominates. The table below is the practical summary; the reason
each row is what it is has been given above.
:::

```{r verdict}
kable(data.frame(
  `If you want to…` = c(
    "understand a mechanism / trace transmission",
    "produce a price fan chart",
    "study EA–US–China spillovers",
    "say anything about Austria, Spain, Ireland…",
    "model German wages, debt or credit",
    "run a monetary experiment on China"),
  `Use` = c("**Stage 2c**", "**Stage 2b**", "**Stage 2d**", "Stage 2b / 2c",
            "Stage 3a / 3b — Germany only", "**nothing here yet**"),
  check.names = FALSE
), row.names = FALSE, align = "ll")
```

::: {.content-visible when-format="revealjs"}
. . .
:::

::: {style="font-size: 0.9em"}
**The through-line:** every result in this deck that survived did so because it
was checked against its own noise floor — a seed placebo, a sign rule, a
structural trace, a second specification. The ones that did not survive are in
the deck too.
:::

## Reference

::: {style="font-size: 0.8em"}
| Report | What it covers |
|---|---|
| `stage1_diagnostics.qmd` | per-country satellites, diagnostics |
| `stage2a_pilot.qmd` / `stage2b_full_system.qmd` | the linkage mechanism, then all eleven |
| `stage2c_refined_core.qmd` / `stage2c_equations.qmd` | the four refinements, every equation |
| `stage2d_regional_core.qmd` / `stage2d_equations.qmd` | the re-partition and China |
| `stage2d_spillovers.qmd` | shock battery + seed placebo |
| `stage3a_labour_prices.qmd` / `stage3b_*.qmd` | the extension blocks |
| `stage3c_rollout.qmd` / `stage3d_regional_blocks.qmd` | why the rollout stalled, and the partition that unstalled it |
| `evaluation.qmd` | expanding-window backtest, four specifications |

Artefacts: `scratch/overview_build.R`, `scratch/forecasts_build.R`,
`scratch/stage2d_build.R`, `scratch/stage2d_spillovers_build.R`,
`scratch/stage3d_build.R`.
:::
