---
title: "Stage 2c: the refined linked core"
subtitle: "Four refinements at zero cost in degrees of freedom — closing the monetary loop, and the fifth refinement that had to be thrown away"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2c_refined_core.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir2c  <- file.path("data", "cache", "stage2c")
dir_2b <- file.path("data", "cache", "stage2b")

# `rec` is the recommended stage-2c system (four refinements). `with_ic` also
# carries refinement 5, the import-content term this report rejects. Cache file
# names predate the decision: `full_*` is the five-refinement run.
rec     <- readRDS(file.path(dir2c, "noic_full_tuned.rds"))
with_ic <- readRDS(file.path(dir2c, "full_tuned.rds"))
b2b     <- readRDS(file.path(dir_2b, "full_tuned.rds"))

shape        <- readRDS(file.path(dir2c, "noic_shape.rds"))
acc_rec      <- readRDS(file.path(dir2c, "noic_acceptance.rds"))
acc_before   <- readRDS(file.path(dir2c, "acceptance_untuned.rds"))
lag_rec      <- readRDS(file.path(dir2c, "noic_lag_stability.rds"))
coefs_rec    <- readRDS(file.path(dir2c, "noic_coef.rds"))
coefs_ic     <- readRDS(file.path(dir2c, "coef_full.rds"))
signs_rec    <- readRDS(file.path(dir2c, "noic_signs.rds"))
signs_ic     <- readRDS(file.path(dir2c, "sign_checks.rds"))
reach        <- readRDS(file.path(dir2c, "noic_reachability.rds"))
loops        <- readRDS(file.path(dir2c, "noic_monetary_loop.rds"))
rates        <- readRDS(file.path(dir2c, "noic_rate_equations.rds"))
elas         <- readRDS(file.path(dir2c, "noic_import_elasticity.rds"))
moved        <- readRDS(file.path(dir2c, "noic_coef_movement.rds"))
moved_ic     <- readRDS(file.path(dir2c, "coef_movement.rds"))
insample     <- readRDS(file.path(dir2c, "noic_insample.rds"))
insample_ic  <- readRDS(file.path(dir2c, "insample_rmse.rds"))
fd           <- readRDS(file.path(dir2c, "foreign_demand_basis.rds"))
tuning_history <- readRDS(file.path(dir2c, "tuning_history.rds"))

# --- the spillover battery, run on the recommended system ---------------
dir_sp   <- file.path("data", "cache", "stage2c_spillovers")
dir_sp2b <- file.path("data", "cache", "spillovers")
mon_2c    <- readRDS(file.path(dir_sp, "noic_monetary_a.rds"))
mon_2b    <- readRDS(file.path(dir_sp2b, "diffs_monetary.rds"))
oil_2c    <- readRDS(file.path(dir_sp, "noic_oil_a.rds"))
mat_2c    <- readRDS(file.path(dir_sp, "noic_matrix_a.rds"))
mat_2b    <- readRDS(file.path(dir_sp2b, "spillover_matrix.rds"))
sanity_2c <- readRDS(file.path(dir_sp, "noic_sanity_a.rds"))
sanity_2b <- readRDS(file.path(dir_sp2b, "sanity.rds"))
mon_failed <- readRDS(file.path(dir_sp, "noic_monetary_failed_draws.rds"))
explosive  <- readRDS(file.path(dir_sp, "noic_baseline_explosive.rds"))
mat_2c_b  <- readRDS(file.path(dir_sp, "noic_matrix_b.rds"))
mon_2c_b  <- readRDS(file.path(dir_sp, "noic_monetary_b.rds"))
PLACEBO_SRC <- c("de", "fr", "it", "nl")

response <- function(d, concept) {
  vars <- country_var(countries, concept)
  x <- d[d$variable %in% vars, ]
  x$country <- sub(paste0("_", concept, "$"), "", x$variable)
  x[order(x$country, x$horizon), ]
}
peak_table <- function(d, concept) {
  x <- response(d, concept)
  do.call(rbind, lapply(split(x, x$country), function(z) data.frame(
    country = z$country[1],
    peak_horizon = z$horizon[which.max(abs(z$median_diff))],
    peak = z$median_diff[which.max(abs(z$median_diff))],
    cumulative = sum(z$median_diff), stringsAsFactors = FALSE)))
}

countries <- modelled_countries
sys_eq <- rec$sys_eq

# Substantive coefficients only. A constant or COVID impulse dummy that was near
# zero in stage 2b makes any relative-change ranking meaningless.
behavioural <- moved[!grepl("^covid_|^constant$", moved$term), ]
touched <- country_var(countries, c("prices", "consumption"))
elasticity <- function(tab, eq_concept, term_concept) {
  x <- tab[tab$equation %in% country_var(countries, eq_concept) &
           tab$term %in% country_var(countries, term_concept), ]
  x[order(x$equation), ]
}
imports_is <- insample$equation %in% country_var(countries, "imports")
imports_is_ic <- insample_ic$equation %in% country_var(countries, "imports")
```

## Summary

Stage 2c refines stage 2b's eleven-country system. Its defining property is
that **the refinements are free**: `k` stays at 76 against `T = 98`, so
`df = 22` in both systems. Every stage-3 block in this project bought structure
with degrees of freedom — stage 3a went to `df = 15`, then 6; stage 3b to 8;
stage 3c could not be validated at all. Stage 2c is the one extension that can
be compared against its predecessor with no size caveat attached.

Five refinements were estimated. **Four are adopted; one is rejected.**

| # | Refinement | Verdict |
|---|---|---|
| 1 | `prices` gains `<iso2>_gdp` (Phillips curve) | **Adopted.** The monetary loop closes: `ea_policy_rate` goes from reaching **no price variable at all** to reaching every one. 10 of 11 correctly signed. |
| 2 | `foreign_demand` over partner **imports** | **Adopted.** Largely a rescaling — partner imports are ~1.8x as volatile as partner GDP — though the US index genuinely changes. |
| 3 | Model the **spread**; `long_rate` becomes an identity | **Adopted.** Motivation confirmed emphatically: all eleven stage-2b policy-rate loadings straddle zero. One stated benefit is not delivered. |
| 4 | `consumption` gains `<iso2>_long_rate` | **Adopted, weak.** 8 of 11 correctly signed, and the loadings are small. |
| 5 | `imports` gains `<iso2>_exports` (import content) | **Rejected.** Collapses the domestic-demand elasticity in **9 of 11** countries while halving the imports equations' RMSE. |

The recommended system is clean on every sampler diagnostic:
`r sum(acc_rec$flagged, na.rm = TRUE)` acceptance-rate flags of
`r sum(acc_rec$has_mh_step)`, `r sum(lag_rec$flagged, na.rm = TRUE)`
lag-stability flags, and `r sum(!signs_rec$ok)` sign-check failures of
`r nrow(signs_rec)`.

**The most useful thing this stage produced is the rejection.** Refinement 5
improves in-sample fit dramatically and breaks the model, and it took three
separate diagnostics — an elasticity comparison, a re-estimation without it, and
a sign-check count corrected for absent terms — to see that. Section 4 is the
substance of this report.

## 1. Four refinements, none of which costs a degree of freedom

```{r shape}
kable(shape, row.names = FALSE,
      caption = "All three systems are the same size. Only the identity count moves.")
```

The equation *count* is unchanged at 68 stochastic, because the long-rate level
equation is **swapped** for a spread equation rather than added to. Identities go
35 → 46: eleven new `<iso2>_long_rate == 1*<iso2>_spread + 1*<policy rate>`.

Why each refinement is free:

- **1 and 4** add only *contemporaneous endogenous* regressors. `k = 1 + (one
  lag per stochastic equation) + (exogenous)`, and a contemporaneous endogenous
  term is none of those three.
- **2** re-weights an existing identity; the weights are fixed trade shares.
- **3** is `k`-neutral by construction: the long rate loses its own lag column
  and the spread gains one.

The second `k` constraint is slack too. `draw_omega_j()` draws `riwish(T - k, .)`
on a matrix that is `(endogenous regressors in that equation + 1)` square, so the
binding requirement here is `df >= 3` against `df = 22`.

### 1.1 A silent identity violation, found and closed off

Stage 2c is the first configuration to put the **US** in `spread_countries`, and
that exposed a live trap. `<iso2>_long_rate == <iso2>_spread + <policy rate>` is
written in two places that must agree: the identity, and the `<iso2>_spread`
series the panel builder constructs. The identity resolved `us_policy_rate` for
the US under `policy_rule = TRUE`; the panel builder's scalar default subtracted
`ea_policy_rate` for everyone. **The US identity was violated by the entire
EA–US policy-rate gap — up to 3.25 percentage points — and koma said nothing**,
because koma has no identity-consistency check at all.

Both sides now resolve through one `policy_rate_map()`, and
`stage2_preflight()` gained a ninth check, `identity_consistency()`, which does
the arithmetic koma does not: it recomputes every identity from the panel and
reports its worst absolute error, in level space for the `rate`/`none` rate
family and in rate space for everything else. On the stage-2c panel, 23 exact
identities hold to a worst error of 1.05e-13. `<iso2>_gdp` and
`<iso2>_domestic_demand` are reported but never flagged — the first holds only
up to the statistical discrepancy, the second is deliberately inexact under
`include_government = FALSE`.

## 2. The loop that was open

`CLAUDE.md` has recorded since stage 1 that "making a variable endogenous does
not give it a transmission channel". Stage 2b gave `ea_policy_rate` its own
Taylor rule; `reports/stage2_spillovers.qmd` §4 then found no monetary
transmission at all, and §7.2 recorded the sanity check failing as a genuine
property of the fitted system rather than a code defect.

`contemporaneous_reachability()` says exactly where the break was.

```{r reach}
kable(reach, row.names = FALSE,
      caption = "Endogenous variables reachable from ea_policy_rate through contemporaneous links.")
```

Stage 2b's policy rate reaches 91 of 103 endogenous variables — **and not one of
them is a price**. It is even reachable from *itself*, through the Taylor rule's
`ea_gdp` term, so a naive "is the loop closed?" test would have passed. The
rule's **output** term was closed all along; its **inflation** term was the open
one, responding to a variable nothing the rule did could influence.

The price equation was the break. In stage 2b its contemporaneous endogenous
regressor set was empty — own lag, COVID dummies, `eur_usd` and `oil_price`, all
exogenous. Adding `<iso2>_gdp` closes it, and stage 2c's policy rate reaches
**all 114** endogenous variables, all 12 price variables included.

There is a second, mechanical confirmation. An equation with no contemporaneous
endogenous regressor gets no Metropolis step, so koma reports
`count_accepted = NA` for it:

```{r mh}
mh_2b <- check_acceptance_rates(b2b$fit)
kable(data.frame(
  system = c("stage 2b", "stage 2c"),
  with_metropolis_step = c(sum(mh_2b$has_mh_step), sum(acc_rec$has_mh_step)),
  without = c(sum(!mh_2b$has_mh_step), sum(!acc_rec$has_mh_step)),
  which_without = c(paste(mh_2b$equation[!mh_2b$has_mh_step], collapse = ", "), "—")
), caption = "The eleven stage-2b equations with no Metropolis step were exactly the price equations.")
```

**Structure is not magnitude.** Reachability says a path exists, not that
anything travels along it — §4.5 and §8 put numbers on that.

## 3. Estimation and the sampler

```{r tuning}
kable(tuning_history, row.names = FALSE,
      caption = "Equations outside the 20-60% acceptance band at each tune_tau_system() round.")
```

Both configurations converge, the recommended one in two rounds. Two things
about that table are worth stating rather than glossing.

First, the five-refinement run's iteration 0 flags **14** where a separately
estimated untuned fit of the same system flagged
`r sum(acc_before$flagged, na.rm = TRUE)`. That is not a change:
`model_identification()` consumes RNG, so two estimations of the same system give
slightly different acceptance rates, and a dozen equations sit within a
percentage point of the 60% boundary. Reading 9 → 14 as a finding would be
reading noise.

Second, the tuned recommended system has
`r sum(acc_rec$flagged, na.rm = TRUE)` equations outside the band and
`r sum(lag_rec$flagged, na.rm = TRUE)` lag-stability flags. The sampler is not
the explanation for anything in what follows.

## 4. Refinement 5, and why it had to go

This is the substance of the report, so it is worth stating the shape of the
argument before the tables. The import-content term — giving `<iso2>_imports` a
contemporaneous `<iso2>_exports` regressor, on the sound reasoning that exports
require imported intermediates — passes its own sign check in **11 of 11
countries** and improves the imports equations' in-sample fit **dramatically**.
It is still wrong, and three separate diagnostics were needed to establish that.

### 4.1 The sign check saw one failure; nine elasticities were being gutted

```{r imports-elas}
kable(elas[, c("equation", "stage2b", "stage2c_full", "stage2c_noIC")],
      row.names = FALSE, digits = 3, col.names =
        c("equation", "stage 2b", "with import content", "recommended"),
      caption = "imports ~ domestic_demand. The elasticity the equation exists to measure.")
```

With the term in, nine of eleven elasticities fall and two are destroyed:
Germany from 0.382 to **0.049**, Austria from 0.450 to **−0.104**. Only Austria
crosses zero, so `sign_checks()` — which looks at one coefficient at a time —
reported exactly one failure while most of the block was being hollowed out.

`CLAUDE.md` already contains this lesson in the abstract ("a wrong sign is often
not the first symptom — a neighbouring coefficient moving is") and this failure
in the concrete: four attempts at putting an import *price* into this same
equation, of which the contemporaneous one "collapses the domestic-demand
elasticity from 0.38 to 0.08". Substituting `exports` for `import_prices` lands
German demand elasticity at 0.049 — the same number, by the same mechanism.
Exports and imports share the trade cycle, so a contemporaneous export term is an
excellent predictor of imports and a direct competitor to domestic demand for the
same variance.

### 4.2 The improved fit is the symptom, not the reassurance

```{r imports-fit}
imp_fit <- insample_ic[imports_is_ic, ]
kable(imp_fit[order(imp_fit$ratio), c("equation", "rmse_stage2b", "rmse_stage2c", "ratio")],
      row.names = FALSE, digits = 4, col.names =
        c("equation", "stage 2b RMSE", "with import content", "ratio"),
      caption = "In-sample RMSE of the imports equations. Ratio below 1 is a better fit.")
```

Belgium's imports equation fits **2.8x better** and the Netherlands' **2.0x
better**, while their structural elasticities fall. That is what a
contemporaneous regressor correlated with the dependent variable does, whether or
not it belongs there. Had this stage been judged on in-sample fit — the
five-refinement system fits better in
`r round(100 * mean(insample_ic$better == "stage2c"))`% of shared equations — the
term would have been adopted.

### 4.3 Removing it recovers every elasticity, and repairs equations it never touched

```{r recovered}
kable(data.frame(
  check = c("import_elasticity_positive", "phillips_curve_positive",
            "consumption_rate_channel_negative", "spread_loads_on_prices_positive",
            "mpc_in_0_1", "TOTAL (of 55 comparable checks)"),
  with_import_content = c(
    sum(!signs_ic$ok & signs_ic$check == "import_elasticity_positive"),
    sum(!signs_ic$ok & signs_ic$check == "phillips_curve_positive"),
    sum(!signs_ic$ok & signs_ic$check == "consumption_rate_channel_negative"),
    sum(!signs_ic$ok & signs_ic$check == "spread_loads_on_prices_positive"),
    sum(!signs_ic$ok & signs_ic$check == "mpc_in_0_1"),
    sum(!signs_ic$ok)),
  recommended = c(
    sum(!signs_rec$ok & signs_rec$check == "import_elasticity_positive"),
    sum(!signs_rec$ok & signs_rec$check == "phillips_curve_positive"),
    sum(!signs_rec$ok & signs_rec$check == "consumption_rate_channel_negative"),
    sum(!signs_rec$ok & signs_rec$check == "spread_loads_on_prices_positive"),
    sum(!signs_rec$ok & signs_rec$check == "mpc_in_0_1"),
    sum(!signs_rec$ok))
), row.names = FALSE, caption = "Sign-check failures. Both columns are over the same 55 checks.")
```

All eleven elasticities move back toward their stage-2b values (Germany 0.320
against 0.382, France 1.215 against 1.245, the US 1.782 against 1.798) — the same
recovery stage 3a saw when the import-price term was lagged, "0.386 vs 0.382".

The wider effect is the interesting one. Removing a term from the **imports**
equation repairs two Phillips-curve sign checks and two consumption-rate sign
checks, in equations it does not appear in. In a simultaneous system with a
system-wide residual covariance, one badly specified equation is not a local
problem — which is the other half of a lesson `CLAUDE.md` already carries from
stage 3a phase B: "a bad specification in one equation can masquerade as a
systemic problem". Here it masqueraded as two.

The one thing that gets worse is Ireland's MPC, which crosses from +0.001 to
−0.005. Ireland's consumption–GDP loading is ~0 in all three systems (0.017 in
stage 2b), so this is a coefficient that is not identified rather than one that
broke; it is reported here rather than explained away.

### 4.4 A reporting trap this exposed

Read naively, `sign_checks(stage2c = TRUE)` reports **19 failures of 66** for the
recommended system against **12 of 66** for the rejected one — the opposite of
the truth. Eleven of those nineteen are `import_content_positive` scoring
`NA`/`FALSE` because the term is not in the model.

That default is right for the stage-3 blocks it was designed for ("a check that
could not be evaluated is not a check that passed") and wrong for a deliberately
dropped refinement. `sign_checks(stage2c = )` now takes a subset of
`stage2c_refinements()`, and `stage2c_config()` returns the subset it carries.

### 4.5 The monetary loop is closed, and nearly inert

```{r loops}
kable(loops[, c("iso2", "median", "q05", "q95", "share_ge_1",
                "consumption_long_rate", "phillips", "taylor")],
      row.names = FALSE, digits = 4,
      caption = "monetary_loop(): policy rate -> long rate -> consumption -> GDP -> prices -> ea_prices -> policy rate.")
```

No draw in any country reaches a round-trip gain of 1, so the loop stage 2c adds
is not a stability risk. `CLAUDE.md` warns against reading that as reassurance —
"a quiet loop is not necessarily a stable one" — so the per-link means are in the
table, and they are the point:

- The **Taylor rule's** inflation loading is `r sprintf("%.3f", loops$taylor[1])`,
  positive and sensible, shared by every country.
- The **Phillips curve** link is positive in 9 of the 10 euro-area countries, up
  to `r sprintf("%.3f", max(loops$phillips))`. Germany is the exception, at
  `r sprintf("%.3f", loops$phillips[loops$iso2 == "de"])`.
- The **consumption rate channel** is the weak link: correctly signed in most
  countries but tiny, and positive in Austria and France.

A six-link chain of coefficients this small multiplies out to a contemporaneous
gain around 1e-5. That is a statement about *within-quarter feedback*, not about
whether a monetary shock has any effect over a horizon — the dynamic response
accumulates through the lag structure, which is what §8 measures.

## 5. The rate block: from a level equation to a spread equation

```{r rates}
kable(rates, row.names = FALSE, digits = 3,
      caption = "The stage-2b long-rate equation against the stage-2c spread equation.")
```

The motivation for refinement 3 is confirmed, and more strongly than the design
note claimed. **All eleven** stage-2b policy-rate loadings have 90% credible
intervals straddling zero, while the long rate's own lag sits at 0.910–0.964. The
level equation was an AR(1) with decoration, and imposing the pass-through at 1
is strictly better than estimating it into insignificance.

One of the four stated benefits does not survive contact with the data. The
design note argued that "a spread is stationary where a rate level is not, so the
own lag has no unit root to run to". The stage-2c spread own lags run
**0.933–0.983** — as persistent as the levels they replaced, and for Spain
(0.961 → 0.983) and Portugal (0.964 → 0.979) more so. The other three arguments
are unaffected. The roxygen for `financial_block()` has been corrected to say so.

## 6. What happened to the equations that were already there

```{r moved}
kable(data.frame(
  set = c("touched equations (prices, consumption)", "untouched equations"),
  behavioural_terms = c(sum(behavioural$equation %in% touched),
                        sum(!behavioural$equation %in% touched)),
  sign_flips = c(sum(behavioural$sign_flip[behavioural$equation %in% touched]),
                 sum(behavioural$sign_flip[!behavioural$equation %in% touched]))
), row.names = FALSE,
caption = "Behavioural coefficients only -- constants and COVID dummies excluded, since a near-zero stage-2b value makes any relative comparison meaningless.")
```

```{r stable}
kable(rbind(
  cbind(block = "MPC (consumption ~ gdp)",
        elasticity(behavioural, "consumption", "gdp")[, c("equation", "estimate_2b", "estimate_2c", "delta")]),
  cbind(block = "investment ~ long_rate",
        elasticity(behavioural, "investment", "long_rate")[, c("equation", "estimate_2b", "estimate_2c", "delta")])
), row.names = FALSE, digits = 3,
caption = "Two behavioural blocks stage 2c does not modify.")
```

The MPC is stable and stays inside `(0, 1)` in ten of eleven (Ireland's is the
unidentified one from §4.3), the price equations' own lags move by at most 0.036,
and the investment channel is essentially unchanged. The refinements are not
disruptive once refinement 5 is out.

## 7. In-sample fit

```{r insample-summary}
kable(data.frame(
  measure = c("equations where stage 2c fits better",
              "median RMSE ratio, all shared equations",
              "median RMSE ratio, imports equations",
              "median RMSE ratio, everything except imports"),
  recommended = c(sprintf("%d of %d (%.0f%%)", sum(insample$ratio < 1), nrow(insample),
                          100 * mean(insample$ratio < 1)),
                  sprintf("%.4f", stats::median(insample$ratio)),
                  sprintf("%.4f", stats::median(insample$ratio[imports_is])),
                  sprintf("%.4f", stats::median(insample$ratio[!imports_is]))),
  with_import_content = c(
    sprintf("%d of %d (%.0f%%)", sum(insample_ic$ratio < 1), nrow(insample_ic),
            100 * mean(insample_ic$ratio < 1)),
    sprintf("%.4f", stats::median(insample_ic$ratio)),
    sprintf("%.4f", stats::median(insample_ic$ratio[imports_is_ic])),
    sprintf("%.4f", stats::median(insample_ic$ratio[!imports_is_ic])))
), row.names = FALSE,
caption = "Against stage 2b. Both systems estimated on the identical 2000Q1-2024Q4 window.")
```

Read the two columns together and the case against refinement 5 completes itself.
The rejected system's headline is better on every summary line — until the
imports equations are separated out, at which point **its entire advantage is
there and nowhere else**: every other equation's median ratio is ~1.000.

The recommended system's improvement is the opposite shape: small, and spread
across the equations that are *not* imports (median ratio
`r sprintf("%.4f", stats::median(insample$ratio[!imports_is]))` against
`r sprintf("%.4f", stats::median(insample$ratio[imports_is]))` for imports, which
is unchanged because nothing in refinement 1–4 touches that equation). That is
what a genuine structural improvement looks like: modest, diffuse, and not
concentrated in the equation that gained a regressor.

```{r best-worst}
kable(rbind(utils::head(insample[, c("equation", "rmse_2b", "rmse_noIC", "ratio")], 5),
            utils::tail(insample[, c("equation", "rmse_2b", "rmse_noIC", "ratio")], 3)),
      row.names = FALSE, digits = 4,
      caption = "Recommended system: the five largest improvements and three largest deteriorations.")
```

## 8. The monetary shock, re-run

The whole point of refinement 1 was `reports/stage2_spillovers.qmd` §4: a
+100bp/4-quarter `ea_policy_rate` shock produced "no evidence of a real
transmission channel" on the stage-2b system, and §7.2 recorded the sanity
check failing as a genuine property of the fitted linkage. Stage 2c closes the
structural gap. This section asks whether that changed the numbers.

The battery is run exactly as stage 2b's was — same 8-quarter horizon, same
shock definitions, same `scenario_diff()` common-random-numbers pairing.

### 8.1 The failing draws, recovered rather than worked around

The sustained 4-quarter restriction makes koma's conditional solve singular for
`r attr(mon_failed, "n_failed")` of `r attr(mon_failed, "n_draws")` draws:
restricting four horizons at once requires inverting `R %*% Omega %*% t(R)`, and
for a numerically rank-deficient draw no valid conditional innovation exists.
koma drops those by subsetting the list, so `scenario_diff()` aborts on the
draw-count mismatch rather than silently mispair the two legs.

`failed_restriction_draws()` re-derives the indices by tracing
`koma:::forecast_draw()` and asking, per draw, whether it returned normally. Two
verifications before using them:

- The verification stage 2b applied holds here too: **all
  `r attr(mon_failed, "n_failed")` are also flagged explosive** in the
  unconditional baseline (`r length(explosive)` of 1000 draws exceed
  `|de_gdp| > 100` at horizon 8). The same badly-behaved population surfaces as
  an outright failure under a demanding restriction and as a blow-up without
  one — not a new failure mode.
- The failing set is **identical at both seeds** — the same 40 indices, not
  merely the same count. That is the prediction the mechanism makes (failure
  depends on the draw's posterior `Omega`, which carries no RNG; the innovation
  draw is what the seed controls), and it holds on the production system rather
  than only on a test fixture.

### 8.2 GDP: a real channel, barely

```{r mon-gdp}
g2b <- peak_table(mon_2b, "gdp"); g2c <- peak_table(mon_2c, "gdp")
p2b <- peak_table(mon_2b, "prices"); p2c <- peak_table(mon_2c, "prices")
g <- merge(g2b, g2c, by = "country", suffixes = c("_2b", "_2c"))
kable(g[, c("country", "peak_horizon_2b", "peak_2b", "cumulative_2b",
            "peak_horizon_2c", "peak_2c", "cumulative_2c")],
      row.names = FALSE, digits = 3,
      caption = "GDP response to +100bp for 4 quarters. Cumulative is the sum of median_diff over 8 quarters.")
```

```{r mon-gdp-summary}
kable(data.frame(
  measure = c("peaks at horizon 1", "peak is negative", "cumulative 8q effect is negative"),
  stage2b = c(sum(g$peak_horizon_2b == 1), sum(g$peak_2b < 0), sum(g$cumulative_2b < 0)),
  stage2c = c(sum(g$peak_horizon_2c == 1), sum(g$peak_2c < 0), sum(g$cumulative_2c < 0))
), row.names = FALSE, caption = "Out of 11 countries.")
```

Read on its own, this looks like progress: cumulative GDP moves from negative in
`r sum(g$cumulative_2b < 0)` of 11 to `r sum(g$cumulative_2c < 0)` of 11, and
Germany, Greece and Ireland now peak *negative* at horizons 4, 6 and 2 rather
than showing a positive horizon-1 spike and nothing after. Greece's path is the
most transmission-shaped thing in the battery: −0.15, −0.18, −0.18, −0.23,
−0.27, −0.18, −0.27 from horizon 2 on — persistent, negative, not decaying.

**Most of that is the seed.** `CLAUDE.md` requires a placebo before reporting
any such change, and here it removes the headline outright.

```{r mon-placebo}
gb <- peak_table(mon_2c_b, "gdp")
kable(data.frame(
  measure = c("GDP cumulative negative", "GDP peak negative", "price peak negative"),
  stage2b = c(sum(g2b$cumulative < 0), sum(g2b$peak < 0), sum(p2b$peak < 0)),
  stage2c_seedA = c(sum(g2c$cumulative < 0), sum(g2c$peak < 0), sum(p2c$peak < 0)),
  stage2c_seedB = c(sum(gb$cumulative < 0), sum(gb$peak < 0),
                    sum(peak_table(mon_2c_b, "prices")$peak < 0))
), row.names = FALSE,
caption = "Out of 11. Seeds A and B are the SAME stage-2c fit -- any gap between those two columns is noise.")
```

The count of countries with a negative cumulative GDP response is
`r sum(g2c$cumulative < 0)` at seed A and `r sum(gb$cumulative < 0)` at seed B —
and `r sum(g2b$cumulative < 0)` in stage 2b. The stage-2c "improvement" is
smaller than the distance between two runs of stage 2c itself, so **it is not a
finding**.

Per country, the same comparison:

```{r mon-placebo-country}
cmp <- merge(g2c[, c("country", "cumulative")], gb[, c("country", "cumulative")],
             by = "country", suffixes = c("_seedA", "_seedB"))
cmp <- merge(cmp, g2b[, c("country", "cumulative")], by = "country")
names(cmp)[4] <- "cumulative_2b"
cmp$seed_gap <- abs(cmp$cumulative_seedA - cmp$cumulative_seedB)
cmp$system_gap <- abs(cmp$cumulative_seedA - cmp$cumulative_2b)
cmp$survives <- cmp$system_gap > 2 * cmp$seed_gap
kable(cmp, row.names = FALSE, digits = 3,
      caption = "A change survives only if the stage-2b-to-2c gap clearly exceeds the seed-to-seed spread on the same fit.")
```

Mean seed-to-seed gap `r sprintf("%.3f", mean(cmp$seed_gap))` against a mean
cross-system gap of `r sprintf("%.3f", mean(cmp$system_gap))` — the same order of
magnitude. For `r sum(cmp$system_gap < cmp$seed_gap)` of the eleven the
cross-system gap is *smaller* than the seed spread, and
`r sum(sign(cmp$cumulative_seedA) != sign(cmp$cumulative_seedB))` cumulative
responses flip sign between the two seeds.

**What does survive** is `r paste(cmp$country[cmp$survives], collapse = ", ")`.
Two of those are substantive: Greece's cross-system gap is 1.78 against a seed
spread of 0.19, and Ireland's 1.05 against 0.32, both moving strongly more
contractionary under stage 2c — and both are among the three countries whose
response peaks at a later horizon. Portugal clears the same 2x bar only because
its seed spread happens to be unusually small (0.030); its actual movement is
0.17pp and in the *opposite* direction, slightly less contractionary, so it is
better read as a country the placebo cannot resolve than as a third finding.
Germany's gap (0.43) is not clearly above its seed spread (0.37) and should not
be claimed at all.

What has **not** changed at all is the horizon-1 echo. Eight of eleven still
peak at horizon 1 with a positive value, which §1.3 of the stage-2b report
diagnosed as a same-quarter simultaneity artefact of imposing an exact-value
restriction, not a transmission effect.

### 8.3 Prices: the loop is closed and the sign is wrong

```{r mon-prices}
pp <- merge(p2b, p2c, by = "country", suffixes = c("_2b", "_2c"))
kable(pp[, c("country", "peak_horizon_2b", "peak_2b", "cumulative_2b",
             "peak_horizon_2c", "peak_2c", "cumulative_2c")],
      row.names = FALSE, digits = 4,
      caption = "Price response to the same shock.")
```

**Every one of the eleven price peaks is positive**, in both systems — a
monetary tightening appearing to raise prices. Stage 2c did not fix this; if
anything it is marginally worse (`r sum(pp$peak_2b < 0)` negative peaks in stage
2b, `r sum(pp$peak_2c < 0)` in stage 2c).

The mechanism is legible from §4.5's per-link means, and it is not that the
price block is broken. The Phillips-curve link is positive and correctly signed
in 10 of 11 countries: prices rise when GDP rises. But the policy shock's effect
on GDP is *positive at horizon 1* — the simultaneity echo — and the price
response is almost entirely a horizon-1 effect that decays to zero rather than
turning negative. Prices are faithfully tracking the wrong part of the GDP path.

The cost-push channel, by contrast, is correctly signed and substantial:

```{r oil}
kable(peak_table(oil_2c, "prices"), row.names = FALSE, digits = 3,
      caption = "Price response to a sustained +50% oil price level (stage 2c).")
```

All eleven positive, cumulative 0.8–2.3pp. So the price equation responds
sensibly to a genuine cost shock and perversely to a policy rate. That
localises the remaining problem: it is not the price block, it is that the
policy rate's route into GDP is dominated by a within-quarter artefact rather
than by a lagged demand channel.

## 9. The spillover matrix, and the seed placebo

```{r matrix}
kable(round(mat_2c, 3), caption = "Stage 2c: cumulative 8-quarter GDP response, source (row) x receiver (column).")
```

```{r sanity}
sc <- data.frame(check = sanity_2b$check,
                 stage2b = paste0(ifelse(sanity_2b$ok, "PASS", "FAIL"), " — ", sanity_2b$detail),
                 stage2c = paste0(ifelse(sanity_2c$ok, "PASS", "FAIL"), " — ", sanity_2c$detail))
kable(sc, row.names = FALSE, caption = "The three spillover sanity checks, both systems.")
```

**One of the three checks goes from fail to pass, and it is the structural one.**
Stage 2b's own-effect check failed for four countries: a French, German or
Belgian demand shock moved *Austrian* GDP more than it moved the source's own,
an indirect two-hop path through Germany that §7.1 of that report analysed at
length and concluded was a legitimate property of a simultaneous system rather
than a defect. Under stage 2c the own effect dominates in **11 of 11**, and the
diagonals are far more uniform (0.955–1.553 against stage 2b's 0.782–1.662).

The trade-weight correlation more than doubles, from ρ = 0.141 to ρ = 0.292,
though it still misses the 0.3 threshold.

That number would not exist at all without a fix made for this stage.
`spillover_sanity_checks()` looked each source up as `<source>_gdp` in the
receiver's `foreign_demand` weights. Stage 2c's identity loads partner
**imports**, so every lookup missed, the weight column came out all zeros, and
the reported correlation was computed against a constant — a meaningless verdict
rather than an error. The lookup now matches on the ISO-2 prefix, which also
repairs the same latent bug under `ireland_proxy`.

### 9.1 Placebo: how much of this is the seed?

`CLAUDE.md` is emphatic that an individual spillover cell is not a reliable
number at 1000 draws — re-running the same battery on the *same* fit with only
the seed changed moves an off-diagonal cell by 0.35pp on average and flips the
sign of 7–10 of 30. So before reading anything above as a stage-2c effect, the
same four source countries were re-run at a second seed on the same fit.

```{r placebo}
d_seed <- abs(mat_2c[PLACEBO_SRC, ] - mat_2c_b[PLACEBO_SRC, ])
d_sys  <- abs(mat_2c[PLACEBO_SRC, ] - mat_2b[PLACEBO_SRC, ])
kable(data.frame(
  comparison = c("seed A vs seed B, same stage-2c fit (noise)",
                 "stage 2c vs stage 2b, same seed (the claim)"),
  mean_abs_diff = c(mean(d_seed), mean(d_sys)),
  max_abs_diff = c(max(d_seed), max(d_sys)),
  sign_flips = c(sum(sign(mat_2c[PLACEBO_SRC, ]) != sign(mat_2c_b[PLACEBO_SRC, ])),
                 sum(sign(mat_2c[PLACEBO_SRC, ]) != sign(mat_2b[PLACEBO_SRC, ])))
), row.names = FALSE, digits = 3,
caption = "Four source countries (de, fr, it, nl) x 11 receivers = 44 cells.")
```

Unlike §8.2, the matrix survives its placebo. The cross-system difference
(mean `r sprintf("%.3f", mean(d_sys))`) is **more than three times** the
seed-to-seed spread (`r sprintf("%.3f", mean(d_seed))`), and
`r sum(d_sys > 2 * d_seed)` of the 44 cells move by more than twice the noise.

There is a second, more interesting reading of that same table.
`CLAUDE.md` records the stage-2b matrix's own seed sensitivity: re-running it at
a second seed moved an off-diagonal cell by **0.35pp on average, up to 1.85, and
flipped the sign of 7–10 of 30**. The equivalent numbers for stage 2c are
`r sprintf("%.3f", mean(d_seed))`, `r sprintf("%.3f", max(d_seed))`, and
**`r sum(sign(mat_2c[PLACEBO_SRC, ]) != sign(mat_2c_b[PLACEBO_SRC, ]))` of 44**.
The stage-2c spillover matrix is roughly **three times more reproducible** than
stage 2b's. That is not a claim that can itself be dismissed as noise — it is a
measurement *of* the noise, and it is the strongest single result in this
report.

```{r placebo-diag}
kable(data.frame(
  country = PLACEBO_SRC,
  own_2b = diag(mat_2b)[PLACEBO_SRC],
  own_2c_seedA = diag(mat_2c)[PLACEBO_SRC],
  own_2c_seedB = diag(mat_2c_b)[PLACEBO_SRC],
  seed_spread = abs(diag(mat_2c)[PLACEBO_SRC] - diag(mat_2c_b)[PLACEBO_SRC]),
  system_gap = abs(diag(mat_2c)[PLACEBO_SRC] - diag(mat_2b)[PLACEBO_SRC])
), row.names = FALSE, digits = 3,
caption = "Own diagonals: is the stage-2b-to-2c gap larger than the seed-to-seed spread?")
```

Cell by cell the diagonals behave as the placebo would demand: France, Italy and
the Netherlands move by ~10x their seed spread and can be reported; **Germany's
diagonal moves by less than its own seed spread (0.103 against 0.183) and must
not be**. That is the same discipline the stage-3c report applied, and it lands
on a different country this time.

## 10. Verdict

**Adopt stage 2c, with four refinements, not five.** It is the first extension
in this project that costs nothing: `k = 76` and `df = 22`, identical to stage
2b, so nothing here trades structure against degrees of freedom. `stage2c_config()`
and the `_targets.R` stage-2c chain build the recommended system.

What it demonstrably buys:

1. **A structurally closed monetary loop.** `ea_policy_rate` reaches all 114
   endogenous variables including all 12 prices, against 91 of 103 and *zero*
   prices in stage 2b. This is exact and verifiable, not estimated.
2. **A materially better spillover structure.** The own-effect sanity check goes
   from 7/11 (fail) to 11/11 (pass); the stage-2b Austria artefact is gone; the
   trade-weight correlation more than doubles (0.141 → 0.292, still short of the
   0.3 threshold).
3. **A three-times more reproducible spillover matrix** — mean seed-to-seed
   movement 0.119 against stage 2b's 0.35, and 1 sign flip in 44 cells against
   7–10 in 30. For a battery whose individual cells `CLAUDE.md` warns should
   never be read one at a time, this is the most useful improvement of the
   three.
4. **A tidier rate block.** All eleven stage-2b policy-rate loadings straddled
   zero; imposing the pass-through at 1 and modelling the spread removes eleven
   coefficients that were never identified.

What it does **not** buy, stated plainly:

- **A working monetary transmission.** The price response to a +100bp tightening
  is positive at peak in all eleven countries — the wrong sign — and the GDP
  response still concentrates in a horizon-1 simultaneity echo. The apparent
  improvement in cumulative GDP sign (6/11 → 8/11) **does not survive the seed
  placebo**: the same fit gives 6/11 at a second seed. Only Greece and Ireland
  show a change larger than seed noise. Closing a loop structurally is necessary
  and, on this evidence, nowhere near sufficient.
- **A better in-sample fit.** Median RMSE ratio
  `r sprintf("%.4f", stats::median(insample$ratio))`; the improvement is real but
  small and diffuse.
- **A stationary spread.** Refinement 3's persistence argument is contradicted by
  its own estimates (own lags 0.933–0.983 against 0.910–0.964).

**And the most valuable thing the stage produced is a rejection.** Refinement 5
— the import-content term — improves in-sample fit in the imports equations by
up to 2.8x and passes its own sign check 11 of 11, while collapsing the
domestic-demand elasticity in 9 of 11 countries and corrupting four sign checks
in equations it never touches. Every diagnostic that looks at *the term itself*
approves it. Only comparing it against the specification without it exposes it.
That is the third time this project has met the same failure in the same
equation, and the standing conclusion holds: **the imports volume equation takes
domestic demand and its own lag, and nothing contemporaneous.**


## Eight-quarter forecast

```{r fc-load}
#| include: false
source("reports/_forecast_helpers.R")
library(ggplot2)
fc <- forecast_artefact("stage2c")
fc_entities <- modelled_countries
gdp_vars <- fc_vars(fc$paths, fc_entities, "gdp")
price_vars <- fc_vars(fc$paths, fc_entities, "prices")
```

`r fc_window_note(fc)`

Every fan below is the **filtered** predictive interval: koma constrains no
draw to be stationary, so a draw whose quarterly growth exceeds 100% in
absolute value is dropped before the quantiles are taken, exactly as
`score_forecast()` does when it scores a backtest. The solid band is the
25th-75th percentile of what survives, the faint band the 10th-90th, the dark
line their median, and the grey line the realised path where one exists.
`r fc_explosive_note(fc, c(gdp_vars, price_vars))`
A variable whose discarded share climbs steeply is one whose fan
**understates** the model's true uncertainty, so read the tail of a wide fan
as an ordering rather than a number.

**The panels are clipped, and that is worth saying plainly.** Eight quarters
out, the 10-90 interval on *quarterly* growth spans tens of percentage points
in every stage of this project -- an honest number, and one that makes the
median invisible if the axis is scaled to contain it. Each panel is therefore
scaled to its inner band, and the outer band runs to the panel edge rather
than being drawn in full. Nothing is concealed: the table at the end of this
section reports the untruncated bounds.

`r fc_anker_note(fc)`

### GDP

```{r fc-gdp}
#| fig-width: 9
#| fig-height: 7.5
plot_forecast_fan(fc$paths, gdp_vars, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(gdp_vars),
                  title = "Real GDP growth, quarterly %",
                  subtitle = "inner 25-75%, outer 10-90% clipped at the panel edge; grey = realised")
```

```{r fc-gdp-table}
kable(fc_median_table(fc, gdp_vars, space = "rate"),
      row.names = FALSE,
      caption = "Median GDP growth forecast by quarter, percent.")
```

### Prices

```{r fc-prices}
#| fig-width: 9
#| fig-height: 7.5
plot_forecast_fan(fc$paths, price_vars, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(price_vars),
                  title = "Consumer price growth, quarterly %",
                  subtitle = "inner 25-75%, outer 10-90% clipped at the panel edge; grey = realised")
```

```{r fc-prices-table}
kable(fc_median_table(fc, price_vars, space = "rate"),
      row.names = FALSE,
      caption = "Median price growth forecast by quarter, percent.")
```

### Policy rates

A `method = "none"` series passes through koma untouched, so for a policy rate
the forecast **is** the level -- there is no rate-versus-level distinction to
make, and no anchoring question either.

```{r fc-rates}
#| fig-width: 9
#| fig-height: 3.2
rate_vars <- intersect(c("ea_policy_rate", "us_policy_rate"), unique(fc$paths$variable))
plot_forecast_fan(fc$paths, rate_vars, space = "level", ncol = 2,
                  labeller = fc_full_labels(rate_vars),
                  title = "Policy rates, percent",
                  subtitle = "inner 25-75%, outer 10-90%; grey = realised")
```

```{r fc-rates-table}
kable(forecast_table(fc$paths, rate_vars, space = "level"), row.names = FALSE,
      caption = "Median policy-rate path, percent.")
```

### What was discarded, and what the exogenous path assumes

```{r fc-explosive}
kable(
  transform(forecast_explosive_table(fc$paths, c(gdp_vars, price_vars)),
            h1 = round(h1, 3), h_last = round(h_last, 3), max = round(max, 3)),
  row.names = FALSE,
  col.names = c("variable", "explosive share, h1", paste0("h", fc$horizon), "max"),
  caption = "Share of posterior draws discarded as explosive, by variable."
)
```

```{r fc-uncertainty}
kable(forecast_uncertainty_table(fc$paths, c(gdp_vars, price_vars), space = "rate"),
      row.names = FALSE,
      caption = paste("The untruncated predictive band at horizon", fc$horizon,
                      "-- quarterly growth, percent. This is what the charts clip."))
```

A forecast this long needs the **exogenous** series to reach the end of the
horizon, and they do not: every one of them stops at the fit's own native
forecast end. `extend_forecast_horizon()` carries each one forward on its
trailing four-quarter average growth -- a momentum continuation, not a view on
oil markets or world trade -- and koma would otherwise have silently returned a
shorter horizon than the one requested.

```{r fc-extension}
kable(fc$extension, row.names = FALSE,
      caption = "How each exogenous series was carried past the fit's native forecast end.")
```
