---
title: "Stage 2b: the full eleven-country system"
subtitle: "103 equations in one system_of_equations() — scaling, the degrees-of-freedom wall, and the Ireland and trade-weight decisions"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2b_full_system.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir2b <- file.path("data", "cache", "stage2b")
bench <- readRDS(file.path(dir2b, "benchmark.rds"))
projection <- readRDS(file.path(dir2b, "projection.rds"))
tuned <- readRDS(file.path(dir2b, "full_tuned.rds"))
base_fit <- readRDS(file.path(dir2b, "10_EA_US.rds"))
tuning <- readRDS(file.path(dir2b, "tuning.rds"))
stability <- readRDS(file.path(dir2b, "stability.rds"))
coefs <- readRDS(file.path(dir2b, "coef_full.rds"))
signs <- readRDS(file.path(dir2b, "sign_checks.rds"))
ireland <- readRDS(file.path(dir2b, "ireland_ab.rds"))
sys_eq <- tuned$sys_eq
```

## 1. The wall we hit first

The eleven-country system **cannot be estimated on the window stage 1 and
stage 2a use**, and the reason is structural rather than a matter of patience.

koma's sampler projects *every* equation on the **full** `k`-column
`x_matrix` on *every* draw — `construct_pi_hat_0()` and
`construct_theta_hat_j()` both compute `Matrix::solve(t(x) %*% x)` — and it
draws the covariance from `riwish(T - k, .)`. Both need `k < T`. At eleven
countries on the 2019Q4 window that is `k = 83` against `T = 78`, which fails
twice over:

```
Error: system is computationally singular: reciprocal condition number = 1.97e-19
Error: v must be >= dimension of S in rwish()      # Wishart df = T - k = -4
ESTIMATION FAILED: All equations failed to estimate.
```

Residual degrees of freedom `T - k` across the options:

| estimation end | T | 4 EA | 6 EA | 8 EA | 10 EA | 10 EA + US |
|---|---|---|---|---|---|---|
| 2019Q4 (stage 1 / 2a) | 78 | 45 | 31 | 17 | **3** | **fails (−5)** |
| 2024Q4 (**stage 2b**) | 98 | 65 | 51 | 37 | 23 | 15 |
| 2026Q1 (panel end) | 103 | 70 | 56 | 42 | 28 | 20 |

Even ten EA countries on the old window leaves 3 df — invertible, but the
reduced-form projection would be very nearly interpolating the data.

**What stage 2b changes to buy the room**, and what each costs:

- **Estimation window 2000Q1–2024Q4** (T = 98) instead of 2019Q4. This gives
  up the conditional-fill trick that neutralised COVID in stages 1 and 2a, so
  COVID is controlled with dummies instead of excluded.
- **`<iso2>_government` leaves the model.** It appears only in the
  domestic-demand identity, so dropping it there removes it entirely and
  returns 11 exogenous columns. The identity becomes
  `dd == a*consumption + b*investment` with the shares **renormalised to sum
  to 1** — left raw they sum to ~0.8 and would systematically under-predict
  domestic-demand growth. The cost is real and worth stating: government is
  ~19.7% of GDP, the omitted term `share_G × g(G)` has median sd 0.222pp
  against `g(dd)` median sd 2.005pp, so **the identity now explains about 89%
  of domestic-demand growth variation rather than ~100%**, and the model has
  no fiscal channel at all.
- **Four COVID dummies**, `2020Q1/Q2/Q3` and `2021Q2` — the only quarters
  whose cross-country mean absolute GDP growth exceeds 3× the non-COVID
  baseline of 0.81pp (they run 3.16, 12.19, 10.68, 2.48). They enter the five
  real-side equations only; `long_rate` and the policy rules show no
  mechanical COVID break.

Net: **k = 76, T = 98, df = 22.**

An alternative bought the same 11 columns with no accounting damage — drop
the `imports` own-lag, insignificant in 7 of 11 stage-1 fits — but it leaves
government exogenous and so breaks the three-variable exogenous spec. It is
recorded here as the option not taken.

## 2. The system

```{r shape}
kable(data.frame(
  quantity = c("stochastic equations", "identities", "endogenous variables",
               "exogenous variables", "predetermined (lags)", "k = 1 + lags + exogenous",
               "T (2000Q3–2024Q4)", "residual df"),
  value = c(length(sys_eq$stochastic_equations), length(sys_eq$identities),
            length(sys_eq$endogenous_variables), length(sys_eq$exogenous_variables),
            length(sys_eq$predetermined_variables), length(sys_eq$total_exogenous_variables),
            98, 98 - length(sys_eq$total_exogenous_variables))
), caption = "The eleven-country system.")
```

```{r exog}
#| results: asis
cat("Exogenous (", length(sys_eq$exogenous_variables), "): `",
    paste(sort(sys_eq$exogenous_variables), collapse = "`, `"), "`.\n\n", sep = "")
```

Per EA country: `consumption`, `investment` (with `<cc>_long_rate`),
`exports` (on `<cc>_foreign_demand`), `imports`, `prices`, `long_rate` (on
the shared `ea_policy_rate`), plus `gdp`, `domestic_demand` and
`foreign_demand` identities. The US adds `us_policy_rate ~ us_prices +
us_gdp + us_policy_rate.L(1)` and loads its own `long_rate` on that instead.
There is exactly **one** `ea_policy_rate` equation however many countries the
system contains. Per the three-variable exogenous spec the US `prices`
equation uses `eur_usd`, not the `us_exchange_rate` stage 1 and 2a gave it —
which breaks continuity with the US block in both earlier stages.

Nothing here is hand-written: `build_system(countries, blocks, weights)` in
`R/equations.R` assembles blocks (`country_block()`, `monetary_block()`,
`ea_aggregate_block()`), and stage 3 extends it by adding a block rather than
editing the country loop.

```{r preflight}
kable(tuned$preflight, caption = "Pre-flight. The k < T row is new in stage 2b and exists because of §1.")
```

## 3. Scaling

All four steps ran under the identical stage-2b configuration and 8 workers,
so only the country count varies. The stage-2a fit is **not** a point on this
curve — different window, different `k`.

```{r bench}
kable(bench[, c("step", "countries", "stochastic", "identities", "equations",
                "k", "df", "seconds", "peak_mem_mb", "fit_mb",
                "min_acceptance", "max_acceptance", "n_flagged")],
      row.names = FALSE,
      caption = "Benchmark. Acceptance rates are at the default tau = 1.1, before tuning.")
```

```{r scaling}
#| results: asis
final <- project_stage2_runtime(bench, stochastic = 68)
actual <- bench$seconds[bench$stochastic == 68]
cat(sprintf(
"Fitting `log(seconds) ~ log(stochastic equations)` on the **first three** steps gave an exponent of **%.3f** (R² %.4f) — clearly superlinear — and projected the full system at **%.0f s**. It actually took **%.1f s**, an error of **%+.1f%%**. Across all four steps the exponent is **%.3f** (R² %.4f).\n\n",
projection$exponent, projection$r_squared, projection$projected_seconds,
actual, 100 * (actual - projection$projected_seconds) / projection$projected_seconds,
final$exponent, final$r_squared))
```

The exponent lands almost exactly on koma's own documented `n^1.4`. It is
superlinear because `k` grows with the system and `solve(t(x) %*% x)` is
recomputed on every draw of every equation — cost scales with the equation
count *and* with the per-equation work.

### The BLAS trap, which cost more than the scaling did

The first benchmark attempt was ~30× slower than these numbers and had to be
killed. This R links a **pthread OpenBLAS** (`libopenblasp-r0.3.34.so`), and
each process it runs in claims every core for its own thread pool. With
`future::multicore` forking 8 workers, that is 8 × 16 threads on 16 cores:

- **load average 116** on a 16-core machine, each worker burning ~180% CPU;
- the 25-equation step **did not finish in twenty minutes**, against **41.5 s**
  once fixed.

The fix is to pin BLAS to one thread per process *before R starts*, so the
parallelism is purely across equations:

```sh
OMP_NUM_THREADS=1 OPENBLAS_NUM_THREADS=1 Rscript ...
```

It must be the environment rather than `Sys.setenv()` in-session, because
OpenBLAS sizes its pool at library initialisation. Pinning costs nothing
here — koma's matrices are at most 76 × 76, far too small for threaded BLAS
to repay its own synchronisation. `warn_if_blas_threaded()` now warns when
`workers > 1` and the variables are unset.

## 4. Diagnostics

### Acceptance rates

```{r acceptance}
#| results: asis
cat(sprintf("At the default `tau = 1.1`, **%d of 68** equations sat outside the 20–60%% band, all marginally over the ceiling. One round at `tau = 2.2` cleared every one, leaving the band at **[%.3f, %.3f]**.\n\n",
  sum(tuning$flagged_before), min(tuning$acceptance_rate_after, na.rm = TRUE),
  max(tuning$acceptance_rate_after, na.rm = TRUE)))
untouched <- tuning[tuning$tau == 1.1 & !is.na(tuning$acceptance_rate_before), ]
cat(sprintf("All **%d of %d** untouched equations came back bit-identical, confirming again that a per-equation `[tau = ]` override does not leak into the rest of the system.\n\n",
  sum(untouched$acceptance_rate_before == untouched$acceptance_rate_after), nrow(untouched)))
```

```{r tuning-table}
t <- tuning[tuning$tau == 2.2, c("equation", "acceptance_rate_before", "acceptance_rate_after")]
kable(t[order(t$equation), ], row.names = FALSE, digits = 4,
      caption = "The tuned equations, before and after.")
```

### Convergence

```{r stability}
#| results: asis
cat(sprintf("`check_running_mean_stability()` examined **%d** coefficients across all 68 equations and flagged **%d**. No chain is still drifting in its final third.\n\n",
  nrow(stability), sum(stability$flagged)))
```

### Coefficient plausibility

```{r signs}
tbl <- as.data.frame.matrix(table(check = signs$check, ok = signs$ok))
tbl$check <- rownames(tbl)
kable(tbl[, c("check", if ("FALSE" %in% names(tbl)) "FALSE", "TRUE")], row.names = FALSE,
      caption = "Sign checks across all eleven countries.")
```

```{r sign-fails}
fails <- signs[!signs$ok, c("country", "check", "equation", "term", "estimate")]
if (nrow(fails) > 0) {
  kable(fails, row.names = FALSE, digits = 4, caption = "Violations, reported not corrected.")
}
```

This is a **marked improvement on stage 1**, which failed six of these checks.
`mpc_in_0_1` and `import_elasticity_positive` now pass 11/11 — including
Spain, whose stage-1 MPC was an implausible 1.20, and Ireland, whose was
effectively zero. Only `long_rate_loads_on_policy_rate` still fails, in
Greece (−0.049) and Portugal (−0.013), both near-zero rather than
meaningfully wrong-signed; stage 1 failed it in four countries.

### The linkage

```{r linkage}
fd <- coefs[grepl("_foreign_demand$", coefs$term), c("equation", "estimate", "ci_low", "ci_high")]
fd$excludes_zero <- fd$ci_low > 0 | fd$ci_high < 0
kable(fd[order(-fd$estimate), ], row.names = FALSE, digits = 3,
      caption = "Export elasticity on trade-weighted foreign demand.")
```

Every one of the eleven is **positive**, and 7 of 11 have intervals excluding
zero. This is the substantive payoff over stage 2a, where the DE–FR loop gain
was ~0.001 and moved nothing: with all eleven economies present, each
country's foreign demand is built from real partners rather than a `row_gdp`
residual, and the elasticities are both larger and much better identified
(Germany 1.84 against 0.96 in the pilot).

The US is the exception at 0.02 — see §5.

### The monetary block

```{r monetary}
kable(coefs[coefs$equation %in% c("ea_policy_rate", "us_policy_rate"), ],
      row.names = FALSE, digits = 3, caption = "The two policy rules.")
```

The euro-area Taylor rule loads **+0.253 on `ea_prices`** with an interval
excluding zero, but only **+0.006 on `ea_gdp`** — it responds to inflation and
essentially not to output. Both rules are close to a unit root in their own
lag (0.98, 0.96), as policy rates are.

```{r rate-channel}
rc <- coefs[grepl("_long_rate$", coefs$term) & grepl("_investment$", coefs$equation),
            c("equation", "estimate", "ci_low", "ci_high")]
kable(rc, row.names = FALSE, digits = 3,
      caption = "The rate -> investment channel that gives policy a route to GDP.")
```

Nine of eleven are correctly signed (negative), but only Portugal's interval
excludes zero. The channel exists structurally — which is what stage 2a
established it must, for a policy shock to reach GDP at all — but it is not
sharply estimated.

## 5. Ireland

The brief proposed swapping Irish GDP for modified domestic demand. **On this
panel that would make things four times worse**, because the multinational
distortion lives in `ie_investment`, which is *inside* domestic demand:

| series | sd | max abs | quarters >10pp |
|---|---|---|---|
| `ie_gdp` | 4.00 | 22.1 | 4 |
| `ie_domestic_demand` (proposed) | **17.39** | **80.0** | **21** |
| `ie_investment` (the culprit) | 38.01 | 148.5 | — |
| `ie_consumption` (**used**) | 2.54 | 13.3 | 2 |
| `ie_consumption + ie_government` | 1.76 | 9.5 | 0 |

The CSO's real Modified Domestic Demand strips exactly the IP and
aircraft-leasing components; we have no such series. `ie_consumption` is the
best-behaved Irish demand series that is **already endogenous**, which is what
makes the A/B clean: swapping it in changes no equation count, no `k` and no
degrees of freedom, so the two fits are directly comparable. `C+G` would be
marginally better still but needs its own identity plus `ie_government` back
as an exogenous, costing a column and breaking that comparability.

```{r ireland}
kable(ireland[order(-ireland$ie_weight),
              c("equation", "ie_weight", "estimate_gdp", "estimate_proxy", "delta")],
      row.names = FALSE, digits = 3,
      caption = "Export elasticity with partners loading on ie_gdp vs ie_consumption. Both fits at tau = 1.1.")
```

```{r ireland-summary}
#| results: asis
others <- ireland$delta[ireland$equation != "us_exports"]
cat(sprintf("**The contamination is almost entirely one cell.** The median absolute change across the ten non-US countries is **%.3f**; for `us_exports` it is **%+.3f** — twenty times larger.\n\n",
  median(abs(others)), ireland$delta[ireland$equation == "us_exports"]))
```

That is exactly where theory said to look: Ireland's weight in the US row is
**0.247**, an order of magnitude above its 0.016–0.039 in every EA partner,
and it is precisely the `W_trade["us","ie"]` cell CLAUDE.md already flags as
unreliable — inflated by the same multinational redomiciliation that distorts
Irish GDP, and reached through the reciprocal approximation used because the
ECB publishes no US-reporter series.

The effect is economically decisive for that equation: the US export
elasticity moves from **0.02** (implausibly flat) to **0.98**. Acceptance
rates are identical across the two variants (min 0.291, max 0.616, 14 flagged
each), confirming the comparison isolates the swapped variable.

**Recommendation: use the proxy.** Not because Ireland pollutes everyone —
it does not, at a median delta of 0.05 — but because it badly distorts the
one partner whose weight on it is already known to be wrong.

## 6. Trade weights and Greece

### Threshold: fold below 1% into `row`

Applied. It drops 18 of 110 bilateral cells (16.4%) carrying **1.6%** of total
bilateral mass; 92 cells survive, and the folded weight goes into `row_gdp` so
every row still sums to 1.

**One correction to the brief's rationale: this does not estimate faster.**
koma samples only *stochastic* equation columns, and `foreign_demand` is an
identity — identity columns are never sampled, so their density has
essentially no effect on Gibbs cost. The real arguments are identification
(every zeroed cell is an extra exclusion restriction, and the rank condition
is what binds in a 103-equation system), interpretability, and a sparser,
better-conditioned `Gamma` for the forecast solve.

### Greece: not ragged, no truncation needed

The premise does not hold for this vintage. Checked in both the aligned panel
and the raw `global_panel` target: **every EA country, Greece included, starts
2000Q1 and runs to 2026Q1 with zero NAs.** The only late starter anywhere is
`es_core_prices` (2001Q1), a concept this system does not use.

The real Greek issue is already handled elsewhere: `gr_long_rate` has one
*internal* gap at 2015Q3, when capital controls shut the bond market, and
`fill_internal_gaps()` interpolates it with a warning.

On the general question — koma handles a ragged panel at the **edges** only.
`fill_ragged_edge()` conditionally fills leading and trailing gaps during
`estimate()`; an internal hole fails with an opaque error from inside
`level()`. Note that `fill_ragged_edge()`'s `readline()` is **not**
`interactive()`-guarded, so under `Rscript` or `targets` a ragged edge inside
the estimation window is silently *not* filled — which is why the pre-flight
asserts the window is gap-free rather than trusting the fill.

## 7. Verdict

The eleven-country system estimates, converges and is identified. 68
stochastic equations and 35 identities in one `system_of_equations()`, 230s on
8 workers, all acceptance rates in band after one tuning round, zero
unstable chains, and export elasticities that are positive everywhere and
significant in most places — the linkage now does real work, which it did not
in the two-country pilot.

Three things to carry forward:

- **The df budget is now the binding constraint on the model.** At k = 76
  against T = 98 there are 22 residual degrees of freedom, and *every*
  additional equation or exogenous costs one. Stage 3 cannot add much to this
  system without either a longer sample or something coming out.
- **Government is gone, and with it any fiscal channel.** The
  domestic-demand identity explains ~89% of its own growth variation rather
  than ~100%. If fiscal policy matters for stage 3, this is the thing to
  revisit first.
- **The `W_trade["us", ]` row remains the weakest input.** The Ireland result
  is really a symptom: the US row is a reciprocal approximation, and the one
  cell where it is most distorted is the one that moved the most.
