---
title: "Stage 3c: rolling the extended blocks across the euro area"
subtitle: "Why the rollout cannot happen, what the data would have allowed, and whether the extended system is better or merely bigger"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3c_rollout.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir3c <- file.path("data", "cache", "stage3c")
dir3b <- file.path("data", "cache", "stage3b")

avail <- readRDS(file.path(dir3c, "availability.rds"))
frontier <- readRDS(file.path(dir3c, "k_frontier.rds"))
room <- readRDS(file.path(dir3c, "holdout_room.rds"))
insample <- readRDS(file.path(dir3c, "insample_rmse.rds"))
bench <- readRDS(file.path(dir3c, "bench_summary.rds"))
```

## Summary

The brief was to roll the stage-3 blocks across all eleven economies, expecting
data availability to be the constraint — Greece and Ireland were flagged as
likely to fall short.

**Data availability is not the constraint.** Most countries can support most
blocks. The constraint is the degrees-of-freedom budget, and it binds by
roughly a factor of two: the system is already at `df = 8` with **one** country
carrying the blocks, and a second fully-loaded country would need `k = 104`
against `T = 98`.

The rollout therefore does not happen, and this report is about what was
learned instead: a country × block availability matrix that would drive the
rollout if it were affordable, the exact affordability frontier, and — most
usefully — an answer to whether the extended system is actually better than
stage 2b, or merely bigger.

It is merely bigger.

## 1. Data availability: country × block

Built by auditing every source series per country: EA-MD/QD codes from the
vintage on disk, and four Eurostat datasets fetched and checked directly. The
column that matters most is not "does the series exist" but **"when does it
start"** — `T` is system-wide, so one late-starting series moves the estimation
start for every equation in the system. That is what forced the fiscal block's
redesign in stage 3b.

```{r availability}
kable(avail[, c("iso2", "labour", "external", "fiscal", "financial")],
      row.names = FALSE, caption = "Which blocks each country's data can support.")
```

```{r availability-reasons}
kable(avail[avail$blocker != "-", c("iso2", "blocker")], row.names = FALSE,
      caption = "Why a country falls short.")
```

Reading the matrix:

- **The fiscal block is universally available** — `gov_10q_ggdebt` starts 2000Q1
  or earlier for all ten euro-area countries with no internal gaps. The 2002Q1
  problem that forced stage 3b's redesign was specific to `gov_10q_ggnfa`
  (revenue and expenditure), which the reduced design does not use. Ten of
  eleven.
- **The external block is the real data constraint**, not Greece and Ireland
  alone: Belgium's and the Netherlands' current accounts start 2003, Greece's
  and Ireland's 2002. Switching the block on for all four would move the whole
  system's start to 2004Q1 and cost every country sixteen quarters.
- **The financial block is 7 of 11.** Greece and Portugal have no BIS house
  price series; Ireland has none of the four loan components used to build
  private credit. Ireland's EA-MD/QD file is the thinnest of the eleven.
- **Spain is the one price-block exception**, and cheaply — its non-energy HICP
  starts 2001Q1, a four-quarter cost.
- **The United States supports nothing beyond the core.** No FRED series for
  employment, wages, energy CPI, credit or house prices is currently fetched.

So the brief's expectation was half right: Greece and Ireland do fall short, but
so do Portugal, Spain, Belgium, the Netherlands and the US, in different blocks
and for different reasons — and *none of that is what stops the rollout*.

## 2. What stops the rollout: the degrees-of-freedom budget

```{r frontier}
kable(frontier[, c("configuration", "k", "df", "feasible")], row.names = FALSE,
      caption = "Measured against the real specification, not estimated. T = 98 throughout.")
```

Three rows carry the argument:

- **`DE + FR, all blocks` needs `k = 104` against `T = 98`.** A second fully
  loaded country is not affordable. Each country costs about fourteen columns
  (labour 7, external 1, fiscal 4, financial 2), so eleven would need roughly
  `k = 230` — more than twice the observations available.
- **`DE full + 10 partner export prices` needs `k = 100`.** This is the
  *cheapest possible* way to satisfy the brief's requirement that the
  competitiveness channel be active for every country, and it overshoots by two
  columns.
- The largest configurations that do fit are `DE full + 7 partner export prices`
  (`df = 1`) and `DE + FR labour blocks` (`df = 8`) — neither of which is a
  rollout, and the first of which has no usable degrees of freedom.

### 2.1 The bidirectional price channel degenerates before it becomes affordable

The brief asks for `<iso2>_foreign_prices` as a trade-weighted identity over
partners' export prices, making the channel two-way. That requires each partner
to have an *endogenous* export price, i.e. its own labour block. With only a
subset of countries carrying one, the "trade-weighted index" is renormalised
over whoever is left:

| labour countries | partners in DE's index | share of DE's modelled trade weight captured |
|---|---|---|
| DE, FR | 1 | 15.2% — the index *is* France |
| DE, FR, IT | 2 | 27.3% |
| DE, FR, IT, NL | 3 | 44.6% (but `df = -2`) |

And Germany's *modelled* trade weight already excludes a 57% rest-of-world
residual, so a two-country index captures roughly **6% of Germany's actual
trade**. The channel becomes affordable only in a form too degenerate to mean
anything, and meaningful only in a form that does not fit.

The one-way channel — `de_foreign_prices` exogenous, constructed from all eight
partners' observed export prices — is what stage 3a already built, and it does
capture the full modelled weight. That remains the honest version of the
competitiveness channel in this system.

## 3. Is the extended system better, or just bigger?

### 3.1 It cannot be validated out of sample

```{r holdout}
kable(room, row.names = FALSE,
      caption = "How much data each system can hold out before k >= T. 'df at that point' is what remains at the maximum hold-out.")
```

The intended test was a hold-out: estimate both systems on a truncated window,
score the held-out quarters. **The extended system cannot sit that test.** At a
2000Q1–2021Q4 estimation window (`T = 86`) stage 2b has `df = 10` and estimates
fine; the extended system needs `k = 90` and has `df = -4`.

Its maximum possible hold-out is seven quarters, and only by running at `df = 1`.
A model with no degrees of freedom to spare for validation is a model whose
generalisation cannot be checked — and that is a property of the extended system
that stage 2b does not share.

### 3.2 In-sample, it is not better either

Both systems were estimated on the identical 2000Q1–2024Q4 window, so their
in-sample RMSEs are directly comparable on the equations they share.

```{r insample}
de <- insample[grepl("^de_", insample$equation), ]
kable(de[order(de$ratio), c("equation", "rmse_stage2b", "rmse_extended", "ratio")],
      row.names = FALSE, digits = 4,
      caption = "German equations -- the only ones the blocks touch. Ratio below 1 favours the extended system.")
```

```{r insample-summary}
non <- insample[!grepl("^de_", insample$equation), ]
kable(data.frame(
  measure = c("equations where extended fits better", "median RMSE ratio (all shared equations)",
              "median ratio, non-German equations"),
  value = c(sprintf("%d of %d (%.0f%%)", sum(insample$better == "extended"), nrow(insample),
                    100 * mean(insample$better == "extended")),
            sprintf("%.4f", stats::median(insample$ratio)),
            sprintf("%.4f", stats::median(non$ratio)))
), caption = "In-sample fit across all shared equations.")
```

This is the damning number: the extended system adds nineteen equations and
fourteen predetermined columns, and the **median equation fits marginally
worse**. In-sample fit almost always improves when parameters are added — that
is what parameters do — so a median ratio slightly *above* one, with only 21% of
equations improving, says the extra structure is not buying fit even where it
cannot be penalised for complexity.

What it does buy is confined and small: German exports fit about 5% better and
consumption 2.6% better, partly offset by investment fitting 2% worse. Every
non-German equation is unchanged, as expected — the blocks do not touch them.

### 3.3 What it costs

```{r bench}
kable(bench, row.names = FALSE, digits = 4, caption = "System comparison.")
```

Fourteen degrees of freedom, nineteen equations, and about 36% more runtime.
Acceptance rates are clean in both after tuning, so the sampler is not the
problem — the budget is.

### 3.4 The verdict

**Merely bigger.** The extended blocks are worth building as *mechanisms* — the
wage–price loop, the competitiveness channel and the debt-to-spread link are
real structures that stage 2b cannot represent at all, and several of them come
back correctly signed. But as a forecasting system the extension does not pay:
negligible in-sample gain, a two-percent regression in German investment,
36% more runtime, and no way to check out-of-sample generalisation because the
degrees of freedom to do so no longer exist.

The right reading is that this project has reached the limit of what 98
quarterly observations will support in a single joint system. Further structure
needs either a longer sample or a smaller system — not a better sampler, and
not more equations.
