---
title: "Cross-country spillovers from the eleven-country system"
subtitle: "Demand, monetary and oil shocks via matched conditional forecasts -- koma has no impulse-response helper, so spillovers are built as forecast differences"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2_spillovers.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
library(ggplot2)

dir_sp <- file.path("data", "cache", "spillovers")
dir_2b <- file.path("data", "cache", "stage2b")

extension <- readRDS(file.path(dir_sp, "extension.rds"))
diffs_by_country <- readRDS(file.path(dir_sp, "diffs_by_country.rds"))
diffs_monetary <- readRDS(file.path(dir_sp, "diffs_monetary.rds"))
diffs_monetary_1q <- readRDS(file.path(dir_sp, "diffs_monetary_1q.rds"))
diffs_oil <- readRDS(file.path(dir_sp, "diffs_oil.rds"))
mat <- readRDS(file.path(dir_sp, "spillover_matrix.rds"))
sanity <- readRDS(file.path(dir_sp, "sanity.rds"))
linkage_weights <- readRDS(file.path(dir_sp, "linkage_weights.rds"))
failed_idx_monetary <- readRDS(file.path(dir_sp, "failed_idx_monetary.rds"))

countries <- modelled_countries
HORIZON <- 8

response_table <- function(diffs, concept, source_label) {
  vars <- country_var(countries, concept)
  d <- diffs[diffs$variable %in% vars, ]
  d$country <- toupper(sub(paste0("_", concept, "$"), "", d$variable))
  d <- d[order(d$country, d$horizon), ]
  d$source <- source_label
  d
}

# Every table in this report is keyed off median_diff, not mean_diff -- see
# section 1.1 on explosive draws for why.
report_cols <- c("country", "horizon", "median_diff", "ci_low", "ci_high", "explosive_frac")
```

## 1. Method

koma has no impulse-response function. `scenario_diff()` (`R/spillovers.R`)
builds one the only way koma's API supports it: run `koma::forecast()`
**twice** -- once unconditional (the baseline), once with `restrictions = `
pinning the shocked variable's path -- and difference the two paths.

**Matched draws, not independent ones.** Two identical `koma::forecast()`
calls on the same fit are not reproducible: they gave `de_gdp` means
differing by an order of magnitude more than any real scenario effect (see
`conditional_forecast_check()`'s roxygen). Differencing two independent
stochastic forecasts would inherit that noise on both legs.

The fix is common random numbers. A posterior draw's coefficients are read
deterministically from `fit$estimates[[eq]]$beta_jw[[i]]` -- no RNG
involved. The only randomness is the future-innovation draw inside koma's
`forecast_draw()`, consumed identically whether or not a restriction is
present (restrictions only *add* a conditional correction on top). Calling
`set.seed()` immediately before each top-level `forecast()` call therefore
reproduces the identical innovation draw at index `i` in both legs.
**Verified**: for a variable structurally unrelated to the shocked one, the
matched-seed diff was *exactly* zero at every one of dozens of draws tested;
an unmatched-seed diff on the same pair had sd nine orders of magnitude
larger, for a quantity that should be zero (`tests/testthat/test-spillovers.R`).

**Horizon extension.** All four exogenous series stop exactly at the
system's native forecast end (2026Q1, 5 quarters ahead) -- koma silently
*shortens* the horizon rather than erroring when asked to forecast further,
so an unextended request for "8 quarters" would silently deliver 5.
`extend_forecast_horizon()` extends `row_gdp`, `oil_price` and `eur_usd` by
carrying forward the **average of their last 4 quarters' growth**; this is a
trailing-momentum continuation, not a forecast of oil markets or world
trade, and is stated here as the assumption it is:

```{r extension}
kable(extension, row.names = FALSE, caption = "How each exogenous series was extended to reach an 8-quarter horizon.")
```

**Shock definitions.**

- **Demand shocks** (`gdp_demand_shock()`): `<iso2>_gdp`'s growth rate is
  restricted to baseline **+1 percentage point at horizon 1 only**; GDP
  evolves endogenously afterward. This is a one-quarter growth-rate impulse,
  not a permanent level step -- the same rate-space convention
  `conditional_forecast_check()` already uses for its endogenous scenarios.
- **Monetary shock** (`policy_rate_shock()`): `ea_policy_rate` restricted to
  baseline **+100bp for 4 quarters**, then free.
- **Oil shock** (`oil_price_shock()`): `oil_price`'s **level** scaled
  **x1.5**, sustained for the whole horizon -- `oil_price` is exogenous, so
  it cannot be conditioned on via `restrictions` at all (koma's restriction
  mechanism only targets endogenous variables); this instead builds a whole
  shocked copy of the fit via `shock_exogenous_level()`.

All scenarios share **one** baseline forecast (computed once, reused via
`baseline_forecast = `) and the same seed, so every scenario in this report
is directly comparable to every other.

### 1.1 Explosive posterior draws -- diagnosed, not silently presented

The first pass at this battery reported `mean_diff`/`sd_diff` as the point
estimate, per [scenario_diff()]'s original design. Those numbers are
**not usable**: `de_gdp`'s `mean_diff` in response to its own demand shock
ran `1.0, 3733.6, -8.76e+06, 2.00e+10, -4.46e+13, 9.94e+16, -2.21e+20,
4.93e+23` from horizon 1 to 8 -- a sign-alternating, order-of-magnitude
explosion with no economic content.

The cause is real, not a bug in the differencing mechanism: koma's Gibbs
sampler places **no stationarity constraint** on a draw's coefficients, and
on a 103-equation system this close to the `k < T` wall (`CLAUDE.md`), a
non-trivial and *growing* share of posterior draws are dynamically
explosive -- verified directly on the cached baseline forecast, checking
`de_gdp` in isolation:

```{r explosive-diagnosis}
bf <- attr(diffs_by_country$de, "baseline_forecast")
diag_tbl <- data.frame(
  horizon = 1:8,
  pct_draws_gt_100 = sapply(1:8, function(h) {
    x <- sapply(bf$forecasts, function(m) m[h, "de_gdp"])
    round(100 * mean(abs(x) > 100), 1)
  })
)
kable(diag_tbl, col.names = c("horizon", "% of draws with |de_gdp| > 100pp"),
      caption = "Share of posterior draws that are dynamically explosive by horizon (baseline forecast, de_gdp).")
```

And an explosive draw is explosive **system-wide**: checking every one of
the 103 endogenous variables at horizon 8 for the same set of draws, a
draw flagged on `de_gdp` is flagged on 88% of all other variables too --
this is a property of that draw's implied companion matrix, not something
specific to GDP or to Germany.

**Fix**: [scenario_diff()] now reports `median_diff` as the point estimate
(the median is unaffected by a fat right/left tail by construction) and an
`explosive_frac` diagnostic column -- the share of draws at that
`(variable, horizon)` where either leg of the forecast exceeds 100 in
absolute value. `ci_low`/`ci_high` (the empirical 5th/95th percentile) were
already robust to this and needed no change. Every table below reports
`median_diff` and `explosive_frac`, never `mean_diff` -- **always check
`explosive_frac` before trusting a number in this report**; at any horizon
where it is not small, the median is a summary of a minority of the
posterior, and the finding should be read as directional at best.

### 1.2 Restriction targets must be anchored to the median, not koma's `$mean`

A second, independent problem surfaced once `median_diff` was in place: even
the *shocked variable's own* diff at horizon 1 -- where the restriction pins
an exact value, so the diff should equal `+size` almost exactly -- came out
wildly off. `de_gdp`'s own-shock `median_diff` at horizon 1 was **-3.3**,
not +1.

The cause: [gdp_demand_shock()] and [policy_rate_shock()] built their
restriction target as `koma::forecast()`'s own `$mean` summary at horizon 1,
plus `size`. `$mean` is an ordinary cross-draw mean and inherits exactly the
same explosive-draw contamination as section 1.1 -- **in the raw, unshocked
baseline, before any restriction is even applied**:

```{r mean-median-gap}
gap_tbl <- data.frame(
  variable = c("de_gdp", "nl_gdp"),
  baseline_mean_h1 = c(
    as.numeric(attr(diffs_by_country$de, "baseline_forecast")$mean$de_gdp)[1],
    as.numeric(attr(diffs_by_country$nl, "baseline_forecast")$mean$nl_gdp)[1]
  ),
  baseline_median_h1 = c(
    as.numeric(attr(diffs_by_country$de, "baseline_forecast")$median$de_gdp)[1],
    as.numeric(attr(diffs_by_country$nl, "baseline_forecast")$median$nl_gdp)[1]
  )
)
kable(gap_tbl, digits = 2, row.names = FALSE,
      caption = "koma's own $mean vs $median forecast summary, horizon 1, RAW baseline (no shock).")
```

A restriction fixes the scenario's *absolute* value to a single constant
(`target = baseline_h1 + size`) shared by every draw, so each draw's diff is
`target - baseline_draw_i` -- an exact affine, order-reversing function of
that draw's own baseline value. Anchoring `target` to a mean that sits
20+pp away from the typical (median) draw silently bakes that entire gap
into every single draw's diff, with nothing about it visible in the
restriction call itself. Both shock builders now anchor to `baseline_forecast$median` instead
(`R/spillovers.R`); the whole battery below was recomputed after the fix.
Because a restriction's diff is an exact affine, order-preserving function
of each draw's own baseline value (`target - baseline_draw_i`, `target`
constant), anchoring `target` to the *median* of those same baseline draws
guarantees `median_diff` at the shocked variable's own restricted horizon
equals `+size` **exactly** -- verified for every one of the 11 demand
shocks below (`de_gdp`'s own-shock `median_diff` at horizon 1: **+1.0000**,
down from -3.3 pre-fix; identically +1.0000 for all other 10 countries'
own-variable response).

### 1.3 A sustained restriction can make koma's solve outright singular

The 4-quarter monetary shock (section 4) surfaced a third, distinct
problem: `scenario_diff()` aborted with a **draw-count mismatch** --
`r length(failed_idx_monetary)` of 1000 baseline draws produced no
restricted forecast at all, each failing with the same koma-internal
error:

> `A = R %*% Omega %*% t(R) is singular` -- restricting 4 simultaneous
> horizons requires inverting a 4x4 matrix built from that draw's own
> posterior innovation covariance (`Omega`); for a draw whose covariance is
> rank-deficient there, no valid conditional innovation exists.

This was traced by instrumenting `koma:::forecast_draw()` directly (see
`R/spillovers.R`'s `scenario_diff()` roxygen, "Draw count mismatches").
Two things were verified before proceeding:

- All `r length(failed_idx_monetary)` failed draws are a **subset** of the
  explosive-draw population from section 1.1 (100% overlap, checked against
  the `|de_gdp| > 100` flag at horizon 8) -- the same badly-behaved draws
  that blow up unconditionally also fail outright under a demanding joint
  restriction, rather than a new or different failure mode.
- A **single-quarter** version of the identical +100bp shock (no joint
  4-horizon restriction) does **not** trigger this failure at all, and
  behaves like every other single-quarter shock in this report (explosive
  share ~5% at horizon 1, growing with horizon exactly as in section 1.1).
  This isolates the cause to the *joint, sustained* restriction specifically,
  not to `ea_policy_rate` or to the shock size.

Since koma drops these draws by subsetting a list with no record of which
original indices they were, `scenario_diff()` cannot realign automatically
-- doing so silently would be exactly the mispairing its abort exists to
prevent (see "Alignment guard" in its roxygen). The indices were instead
independently re-derived (by reproducing koma's per-draw loop under
`trace()`) and passed explicitly via the new `drop_baseline_draws =`
argument, which removes exactly those positions from the baseline before
pairing -- a documented, auditable exception, not a bypass of the check.

## 2. German demand shock

```{r de-shock}
de_diff <- diffs_by_country$de
gdp_de <- response_table(de_diff, "gdp", "de")
prices_de <- response_table(de_diff, "prices", "de")
```

```{r de-gdp-plot}
#| fig-width: 9
#| fig-height: 5
ggplot(gdp_de, aes(horizon, median_diff, color = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_ribbon(data = gdp_de[gdp_de$country == "DE", ],
              aes(ymin = ci_low, ymax = ci_high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to a +1pp German demand shock (quarter 1 only)",
       x = "quarter", y = "GDP growth diff, pp", color = NULL) +
  theme_minimal()
```

```{r de-table}
kable(gdp_de[, report_cols],
      row.names = FALSE, digits = 4, caption = "GDP response by country, quarters 1-8.")
kable(prices_de[, report_cols],
      row.names = FALSE, digits = 4, caption = "Prices response by country, quarters 1-8.")
```

Cumulative 8-quarter Austrian GDP effect (+1.70) actually **exceeds**
Germany's own (+1.38) -- see section 7.1 for the diagnosis (indirect
propagation through Austria's large direct weight on Germany, not a defect).

## 3. US demand shock

```{r us-shock}
us_diff <- diffs_by_country$us
gdp_us <- response_table(us_diff, "gdp", "us")
prices_us <- response_table(us_diff, "prices", "us")
```

```{r us-gdp-plot}
#| fig-width: 9
#| fig-height: 5
ggplot(gdp_us, aes(horizon, median_diff, color = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_ribbon(data = gdp_us[gdp_us$country == "US", ],
              aes(ymin = ci_low, ymax = ci_high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to a +1pp US demand shock (quarter 1 only)",
       x = "quarter", y = "GDP growth diff, pp", color = NULL) +
  theme_minimal()
```

```{r us-exposure}
exposure <- data.frame(
  country = toupper(countries),
  us_weight = sapply(countries, function(cc) {
    w <- linkage_weights$foreign_demand[[cc]]
    v <- w[names(w) == "us_gdp"]
    if (length(v)) unname(v) else NA_real_
  })
)
kable(exposure[order(-exposure$us_weight), ], row.names = FALSE, digits = 3,
      caption = "Weight each country's foreign_demand identity puts on us_gdp -- the a priori exposure ranking.")
```

```{r us-table}
kable(gdp_us[, report_cols],
      row.names = FALSE, digits = 4, caption = "GDP response by country, quarters 1-8.")
kable(prices_us[, report_cols],
      row.names = FALSE, digits = 4, caption = "Prices response by country, quarters 1-8.")
```

### Implausible result: the ranking does not track exposure

The exposure table above correctly identifies Ireland as by far the most
US-exposed economy (weight 0.247, more than double Germany/Netherlands'
0.098) -- the a priori ranking IE > DE &approx; NL holds in the trade
weights themselves. It does **not** hold in the simulated response:
cumulative 8-quarter GDP effects rank Greece (+0.42) and Austria (+0.36)
above Germany (+0.20) and the Netherlands (+0.19), and **Ireland's own
cumulative effect is negative** (-0.08) despite its outsized exposure.

Two things point to noise rather than a real, inverted transmission
ranking. First, every country's response is tiny in absolute terms (all
under 0.5pp cumulative over 8 quarters, versus the 5-95% band's width of
tens of points even at horizon 1) and **oscillates in sign from horizon 2
onward** (`ie_gdp`: +0.06, -0.04, -0.02, -0.00, -0.02, -0.02, -0.02, -0.02)
-- not the smooth, one-signed decay a genuine propagating shock would leave.
Second, `explosive_frac` for these same cells is already 27-45% by horizon
3-5, i.e. the median itself is drawn from a shrinking, and increasingly
non-majority, well-behaved share of the posterior well before horizon 8.
Given both, this ranking should be read as within the noise floor, not as
evidence Ireland is atypically insulated from US demand -- a claim this
battery cannot support with the tiny signal available at this shock size.

## 4. Common monetary shock

`ea_policy_rate` +100bp, sustained for 4 quarters. See 1.3 for how the
`r length(failed_idx_monetary)` draws where koma's restricted solve is
exactly singular were identified and excluded.

```{r mon-gdp-plot}
#| fig-width: 9
#| fig-height: 5
gdp_mon <- response_table(diffs_monetary, "gdp", "monetary")
ggplot(gdp_mon, aes(horizon, median_diff, color = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to a +100bp policy-rate shock (4 quarters)",
       x = "quarter", y = "GDP growth diff, pp", color = NULL) +
  theme_minimal()
```

```{r mon-tables}
prices_mon <- response_table(diffs_monetary, "prices", "monetary")
kable(gdp_mon[, report_cols],
      row.names = FALSE, digits = 4, caption = "GDP response by country, quarters 1-8.")
kable(prices_mon[, report_cols],
      row.names = FALSE, digits = 4, caption = "Prices response by country, quarters 1-8.")
```

### Implausible result: no evidence of a real transmission channel

Every country's GDP response peaks at **horizon 1**, is **small** (0.03 to
0.48pp for a full +100bp/4-quarter hike -- an order of magnitude smaller
than the demand shocks' own-country response to a single-quarter +1pp), and
is **positive** in 9 of 11 countries -- a policy tightening appearing to
*raise* output on impact. From horizon 2 on the response oscillates in sign
with no discernible trend (`de_gdp`: +0.18, 0.00, -0.04, -0.08, -0.05, 0.00,
-0.03, 0.00), the signature of noise around zero rather than a decaying
economic response. Prices behave the same way (peaks of 0.01-0.08pp, 11 of
12 positive). We do not read this as "monetary policy is expansionary in
this model" -- concentrating entirely at horizon 1 with no persistence is
not what a transmission mechanism, real or mis-signed, would produce; it is
what a same-quarter **simultaneity echo** from imposing an exact-value
restriction looks like once the restriction is satisfied and the system is
otherwise left alone.

This matches a limitation already diagnosed for an earlier stage of this
project (`CLAUDE.md`, "Making a variable endogenous does not give it a
transmission channel"): giving `ea_policy_rate` its own equation lets it
influence whatever variables appear as regressors in *other* equations, but
in the stage-1 template only `<iso2>_long_rate`'s equation, and
`<iso2>_long_rate`'s subsequent appearance in the investment equation, close
a loop back to GDP -- a single, likely weak link for the whole system to
route a monetary shock through. Sanity check #2 below fails as a genuine,
previously-documented property of this fitted linkage, not a code defect;
confirming or strengthening the channel would mean checking
`sys_eq$character_gamma_matrix` reachability from `ea_policy_rate` to GDP,
the same diagnostic `CLAUDE.md` recommends rather than assuming.

### Core vs periphery

```{r core-periphery}
periphery <- c("gr", "es", "pt", "it", "ie")
core <- c("de", "fr", "be", "nl", "at")
cumulative_gdp <- aggregate(median_diff ~ country, gdp_mon, sum)
cumulative_gdp$group <- ifelse(tolower(cumulative_gdp$country) %in% periphery, "periphery",
                               ifelse(tolower(cumulative_gdp$country) %in% core, "core", "us"))
kable(cumulative_gdp[order(cumulative_gdp$group, cumulative_gdp$median_diff), ],
      row.names = FALSE, digits = 4, caption = "Cumulative 8-quarter GDP effect, grouped.")
```

Given the finding above, this grouping should be read as a ranking of noise,
not evidence of a core-periphery transmission asymmetry -- with no real
channel to output, there is nothing for an asymmetry to operate through.

## 5. Oil shock

`oil_price` level x1.5, sustained.

```{r oil-tables}
gdp_oil <- response_table(diffs_oil, "gdp", "oil")
prices_oil <- response_table(diffs_oil, "prices", "oil")
kable(prices_oil[, report_cols],
      row.names = FALSE, digits = 4, caption = "Prices response by country, quarters 1-8.")
kable(gdp_oil[, report_cols],
      row.names = FALSE, digits = 4, caption = "GDP response by country, quarters 1-8.")
```

This is the cleanest scenario in the battery: `explosive_frac` is exactly 0
for every country's price response at horizon 1, and stays low throughout.
`oil_price` is exogenous, so this shock never goes through koma's
`restrictions =` conditioning mechanism at all (section 1.3's singular-
matrix failure and the demand shocks' contemporaneous-restriction noise are
both specific to conditioning on an *endogenous* variable) -- it is instead
a whole shocked copy of the fit via `shock_exogenous_level()`, differenced
against the same baseline. The results are also economically coherent: a
sustained +50% oil-price shock raises prices materially and immediately
(+0.45 to +1.11pp at horizon 1 in every country) and has a small, mostly
negative effect on GDP -- a textbook negative supply shock, and the one
scenario in this report we would flag as reliable at face value.

## 6. Spillover matrix

Cumulative 8-quarter GDP effect (sum of `median_diff` across horizons --
see 1.1): row = source (shocked) country, column = receiving country.
Diagonal is the own-country effect.

```{r heatmap}
#| fig-width: 8
#| fig-height: 7
long <- as.data.frame(as.table(mat))
names(long) <- c("source", "receiver", "effect")
long$source <- toupper(long$source)
long$receiver <- toupper(long$receiver)
ggplot(long, aes(receiver, source, fill = effect)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", effect)), size = 2.8,
            color = ifelse(abs(long$effect) > max(abs(long$effect), na.rm = TRUE) * 0.5, "white", "black")) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0) +
  labs(title = "Cumulative 8Q GDP spillover matrix (median diff)", x = "receiving country", y = "source country",
       fill = "cum.\nmedian\ndiff, pp") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r heatmap-table}
kable(round(mat, 3), caption = "Same data, as a table.")
```

## 7. Sanity checks

```{r sanity}
kable(sanity, row.names = FALSE, caption = "Pass/fail, reported as-is -- not silently corrected.")
```

None of the three checks pass. Each is diagnosed below rather than
presented as an unexplained failure; none of them trace back to the
numerical issues in section 1 -- all three use `median_diff`, computed
after the fixes there.

### 7.1 Own effect does not dominate for 4 of 11 countries (be, de, fr, pt)

```{r at-exposure}
kable(sort(linkage_weights$foreign_demand$at, decreasing = TRUE), col.names = c("weight on"),
      caption = "Austria's foreign_demand weights -- what AT's own identity says it is exposed to.")
```

Austria is the receiver in every failing cell: `fr -> at` = 1.93 (exceeds
`fr`'s own diagonal, 1.66), `de -> at` = 1.70 (exceeds `de`'s own diagonal,
1.38), `be -> at` = 1.74 (exceeds `be`'s own diagonal, 1.66); `pt -> nl` =
0.92 similarly exceeds `pt`'s own diagonal (0.78). But Austria's *direct*
bilateral weight on France is only 0.036 and on Belgium only 0.027 --
nowhere near large enough to justify a same-size response to those shocks.
What Austria **is** heavily exposed to is Germany (weight 0.286, its
largest bilateral link by far). A French or Belgian demand shock raises
German GDP first (Germany is *also* one of their largest trading partners),
and that German response then reaches Austria through AT's own large
Germany weight -- a real two-hop transmission path that a single country's
bilateral weight cannot capture, because it is not a property of that
country's link to the source at all. **This is a property of a fully
simultaneous 11-country system, not a defect**: indirect propagation
through a common, dominant trading partner can legitimately exceed the
direct channel and even the source's own effect on itself, and there is no
reason a small, deeply Germany-linked economy like Austria should be
immune to it.

### 7.2 Monetary contraction sign

Diagnosed in section 4: the fitted system shows no persistent, negative,
growing GDP/price response to a policy-rate hike, only a small
horizon-1 simultaneity echo (predominantly positive by chance, not by
transmission). This is read as a genuine, previously-documented weakness of
the fitted linkage (see 4's discussion), not a code defect -- but it does
mean this check's premise (a real monetary channel to test the sign of)
does not currently hold for this fitted system.

### 7.3 Spillover magnitude vs. bilateral trade weight (rho = 0.141)

The same mechanism as 7.1 explains the weak correlation directly: this
check compares each cell's spillover magnitude to the receiver's **direct**
bilateral weight on the source, but 7.1 already shows the *largest*
spillovers in the matrix (into Austria) are driven by an indirect,
two-hop path via Germany rather than any direct weight at all. A simple
bilateral correlation is the wrong instrument for a fully simultaneous
system where a country's total exposure to a shock is not just its own
direct weight but also every other country's weight on it, compounded --
the same reasoning [chain_weighted_index()] and this project's identity
structure already treat as a *modelling* choice (rate-space chain
aggregation over levels) applies equally to *interpreting* the resulting
spillover matrix. A network-aware check (e.g. correlating against each
receiver's weighted exposure through all sources, not just the direct
source-receiver pair) would be needed to test this properly; that is
follow-up work, not a fix to `spillover_sanity_checks()` as specified.
