---
title: "Stage 2 equations: koma strings, plain language, and fitted coefficients"
subtitle: "The eleven-country linked core system (stage 2b)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

f2b <- readRDS(file.path("data", "cache", "stage2b", "full_tuned.rds"))
spec <- f2b$spec
eq_order <- names(c(spec$stochastic, spec$identities))
ov <- equation_overview(spec)
ct <- readable_coefficient_table(f2b$fit, eq_order)
```

Every stage-1 equation for every country plus the `ea_`/`world_`
aggregation identities, in **one** joint `koma::system_of_equations()`
call, so cross-country simultaneity is estimated rather than assumed:
`r nrow(ov[ov$kind == "stochastic", ])` stochastic equations and
`r nrow(ov[ov$kind == "identity", ])` identities. This is stage 2b (all
eleven economies, its own 2000Q1-2024Q4 estimation window), not the
two-country stage 2a pilot -- see `CLAUDE.md` for why stage 2b needs a
different window and `reports/stage2b_full_system.qmd` for the estimation
diagnostics behind the coefficients below.

Government appears in no equation (`include_government = FALSE`, stage 2b's
own departure from stage 2a, which buys back the exogenous columns the
eleven-country system needs to fit at all); the monetary block adds one
equation (`ea_policy_rate`); `ea_gdp`/`ea_prices` are the two euro-area
aggregation identities on top of each country's own `gdp`/`domestic_demand`/
`foreign_demand`.

## 1. Equations as koma parsed them

```{r koma-strings}
kable(ov[ov$kind == "stochastic", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "Stochastic equations, as passed to koma::system_of_equations().")
kable(ov[ov$kind == "identity", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "Identities.")
```

## 2. The same equations, plain language

```{r readable}
kable(ov[ov$kind == "stochastic", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "Stochastic equations.")
kable(ov[ov$kind == "identity", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "Identities.")
```

## 3. Fitted coefficients

90% credible intervals (`ci_low`-`ci_high`), from `coefficient_table()`
applied to the cached, tau-tuned fit
(`data/cache/stage2b/full_tuned.rds`). Identities carry no estimated
coefficients -- their fixed weights are already shown in full in sections 1
and 2.

```{r coefficients}
#| results: asis
for (eq in intersect(eq_order, unique(ct$equation))) {
  sub <- ct[ct$equation == eq, ]
  print(kable(sub[, c("term_readable", "estimate", "ci_low", "ci_high")],
             digits = 4, row.names = FALSE,
             col.names = c("Term", "Estimate", "CI low (5%)", "CI high (95%)"),
             caption = sub$equation_readable[1]))
  cat("\n\n")
}
```
