---
title: "Stage 3b equations: koma strings, plain language, and fitted coefficients"
subtitle: "Stage 3a plus Germany's external, fiscal and financial blocks"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3b_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

f3a <- readRDS(file.path("data", "cache", "stage3a", "fit_a_final.rds"))
f3b <- readRDS(file.path("data", "cache", "stage3b", "fit_financial.rds"))
spec <- f3b$spec
eq_order <- names(c(spec$stochastic, spec$identities))
ov <- equation_overview(spec)
ct <- readable_coefficient_table(f3b$fit, eq_order)

new_stochastic <- setdiff(names(spec$stochastic), names(f3a$spec$stochastic))
new_identities <- setdiff(names(spec$identities), names(f3a$spec$identities))
common <- intersect(names(spec$stochastic), names(f3a$spec$stochastic))
changed_stochastic <- common[vapply(common, function(nm) !identical(spec$stochastic[[nm]], f3a$spec$stochastic[[nm]]), logical(1))]
touched_stochastic <- c(new_stochastic, changed_stochastic)
dropped_stochastic <- setdiff(names(f3a$spec$stochastic), names(spec$stochastic))
```

This is the cumulative "financial" step: stage 3a plus, in order, the
external block (competitiveness, terms of trade, current account), the
fiscal block (`de_government` re-enters as its own equation, government
debt accumulates via `de_govdebt == de_govdebt.L(1) + de_netborrowing`),
and the financial block (credit, house prices, and the sovereign spread
replacing the long-rate level equation --
`de_long_rate == de_spread + ea_policy_rate` is now an **identity**, not a
regression). `de_investment` and `de_exports` also changed content (a
credit term and a competitiveness/foreign-demand term respectively) without
changing which variable they explain. `k = 90` against `T = 98`, `df = 8`
-- see `reports/stage3b_external_fiscal_financial.qmd` for the
degrees-of-freedom accounting, the fiscal-block redesign that made this
window feasible, and why `de_long_rate` moved from a regression to an
identity.

Full system:
`r nrow(ov[ov$kind == "stochastic", ])` stochastic equations,
`r nrow(ov[ov$kind == "identity", ])` identities. This section again shows
only what stage 3b **added or changed** relative to stage 3a
(`reports/stage3a_equations.qmd`) -- `r length(dropped_stochastic)` equation
(`r paste(dropped_stochastic, collapse = ", ")`) moved from stochastic to
identity, shown in the identities table below instead.

## 1. Equations as koma parsed them

```{r koma-strings}
kable(ov[ov$equation %in% touched_stochastic, c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "New or changed stochastic equations.")
kable(ov[ov$equation %in% new_identities, c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "New identities (including de_long_rate, now an identity).")
```

## 2. The same equations, plain language

```{r readable}
kable(ov[ov$equation %in% touched_stochastic, c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "New or changed stochastic equations.")
kable(ov[ov$equation %in% new_identities, c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "New identities.")
```

## 3. Fitted coefficients

90% credible intervals (`ci_low`-`ci_high`), from `coefficient_table()`
applied to the cached fit (`data/cache/stage3b/fit_financial.rds`).
Identities (including `de_long_rate`, now one) carry no estimated
coefficients -- their fixed weights are shown in full above.

```{r coefficients}
#| results: asis
for (eq in touched_stochastic) {
  sub <- ct[ct$equation == eq, ]
  print(kable(sub[, c("term_readable", "estimate", "ci_low", "ci_high")],
             digits = 4, row.names = FALSE,
             col.names = c("Term", "Estimate", "CI low (5%)", "CI high (95%)"),
             caption = sub$equation_readable[1]))
  cat("\n\n")
}
```
