---
title: "Stage 3a equations: koma strings, plain language, and fitted coefficients"
subtitle: "The stage-2b system plus Germany's labour market and disaggregated-price block"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3a_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

f3a <- readRDS(file.path("data", "cache", "stage3a", "fit_a_final.rds"))
spec <- f3a$spec
eq_order <- names(c(spec$stochastic, spec$identities))
ov <- equation_overview(spec)
ct <- readable_coefficient_table(f3a$fit, eq_order)

de_labour_stochastic <- c("de_employment", "de_unemployment", "de_wages",
                          "de_nonenergy_prices", "de_energy_prices", "de_export_prices")
de_labour_identities <- c("de_productivity", "de_ulc", "de_prices", "de_real_income", "de_foreign_prices")
```

Every equation from stage 2 (`reports/stage2_equations.qmd`) is unchanged
here **except** Germany's `de_prices`, which drops out as a stochastic
equation and reappears below as an identity, and `de_exports`, which gains
a competitiveness term. On top, Germany gets seven new behavioural equations
and five new identities: employment, unemployment (a wage Phillips curve),
wages, the energy/non-energy price split, import/export prices, and the
productivity / unit-labour-cost / real-income / foreign-price identities
built from them. This section shows **only the stage-3a-specific
equations** -- for the full system (all
`r nrow(ov[ov$kind == "stochastic", ])` stochastic equations and
`r nrow(ov[ov$kind == "identity", ])` identities), see
`reports/stage2_equations.qmd` and diff against here. See
`reports/stage3a_labour_prices.qmd` for the HICP-weight sourcing,
identification checks and estimation diagnostics behind these coefficients.

## 1. Equations as koma parsed them

```{r koma-strings}
kable(ov[ov$equation %in% c(de_labour_stochastic, "de_exports"), c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "New or changed stochastic equations.")
kable(ov[ov$equation %in% de_labour_identities, c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = "New identities.")
```

## 2. The same equations, plain language

```{r readable}
kable(ov[ov$equation %in% c(de_labour_stochastic, "de_exports"), c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "New or changed stochastic equations.")
kable(ov[ov$equation %in% de_labour_identities, c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = "New identities.")
```

## 3. Fitted coefficients

90% credible intervals (`ci_low`-`ci_high`), from `coefficient_table()`
applied to the cached fit (`data/cache/stage3a/fit_a_final.rds`). Identities
carry no estimated coefficients -- their fixed weights are shown in full
above.

```{r coefficients}
#| results: asis
for (eq in c(de_labour_stochastic, "de_exports")) {
  sub <- ct[ct$equation == eq, ]
  print(kable(sub[, c("term_readable", "estimate", "ci_low", "ci_high")],
             digits = 4, row.names = FALSE,
             col.names = c("Term", "Estimate", "CI low (5%)", "CI high (95%)"),
             caption = sub$equation_readable[1]))
  cat("\n\n")
}
```
