---
title: "Cross-country spillovers from the regional core (stage 2d)"
subtitle: "Six entities, three policy rules — the most seed-reproducible battery in this project, and the one shock that does not work"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2d_spillovers.qmd`.
# Artefacts come from scratch/stage2d_spillovers_build.R.
devtools::load_all(".", quiet = TRUE)
library(knitr)
library(ggplot2)

dir_sp <- file.path("data", "cache", "stage2d_spillovers")
g <- function(n) readRDS(file.path(dir_sp, paste0(n, ".rds")))

countries <- stage2d_countries()
HORIZON <- 8
SEED_A <- 20240101
SEED_B <- 20240102
PLACEBO_SRC <- c("de", "us", "cn", "reu")

ext_summary <- g("extension_summary")
mat_a <- g("matrix_a")
mat_b <- g("matrix_b")
sanity <- g("sanity_a")
mon_a <- g("monetary_a")
mon_b <- g("monetary_b")
mon_cn <- g("monetary_cn_a")
oil_a <- g("oil_a")
failed <- g("monetary_a_failed_draws")
explosive <- g("baseline_explosive")
demand <- stats::setNames(lapply(countries, function(cc) g(paste0("demand_", cc, "_a"))), countries)
shape <- readRDS(file.path("data", "cache", "stage2d", "shape.rds"))
coefs <- readRDS(file.path("data", "cache", "stage2d", "coefs.rds"))

response <- function(d, concept) {
  vars <- country_var(countries, concept)
  x <- d[d$variable %in% vars, ]
  x$country <- toupper(sub(paste0("_", concept, "$"), "", x$variable))
  x[order(x$country, x$horizon), ]
}
cum_by <- function(d, concept) {
  x <- response(d, concept)
  vapply(split(x, x$country), function(z) sum(z$median_diff), numeric(1))
}
peak_by <- function(d, concept) {
  x <- response(d, concept)
  vapply(split(x, x$country), function(z) z$median_diff[which.max(abs(z$median_diff))], numeric(1))
}

# --- the seed placebo ----------------------------------------------------
a <- mat_a[PLACEBO_SRC, , drop = FALSE]
b <- mat_b[PLACEBO_SRC, , drop = FALSE]
offdiag <- outer(PLACEBO_SRC, colnames(mat_a), function(i, j) i != j)
placebo_mean <- mean(abs(a - b)[offdiag])
placebo_max <- max(abs(a - b)[offdiag])
placebo_flips <- sum(sign(a[offdiag]) != sign(b[offdiag]))
placebo_n <- sum(offdiag)

coef_of <- function(eq, term) {
  r <- coefs[coefs$equation == eq & coefs$term == term, ]
  if (nrow(r) == 0) NA_real_ else r$estimate[1]
}
report_cols <- c("country", "horizon", "median_diff", "ci_low", "ci_high", "explosive_frac")
```

## 1. What this is

The battery `reports/stage2c_spillovers.qmd` runs, re-run on the **stage 2d**
system: six entities (`de`, `fr`, `it`, `us`, `cn`, `reu`), stage 2c's equations
unchanged, `k = `r shape$k`` against `T = `r shape$t_obs``, **`df = `r shape$df``**.

That `df` is the reason to read this report. Stage 2b and 2c both ran at
`df = 22` and `CLAUDE.md` records what that cost: re-running the *same* battery
on the *same* fit at a different seed moved an off-diagonal spillover cell by
0.35pp on average for stage 2b, flipping the sign of 7–10 cells out of 30. Most
of what anyone wanted to conclude from that matrix was smaller than its own
noise. Stage 2d has half the equations and ten more degrees of freedom, and
§3 measures what that buys.

koma has no impulse-response function, so every number here is the difference
between two `koma::forecast()` calls — one unconditional, one restricted —
paired by **common random numbers** (the same `set.seed()` immediately before
each), so the difference isolates the shock rather than koma's call-to-call
forecast noise. Three shocks:

- **Demand**: `+1pp` on one entity's GDP growth in the first forecast quarter
  only, after which GDP evolves under the model's own dynamics.
- **Monetary**: `+100bp` held on a policy rate for four quarters.
- **Oil**: a `+50%` level shock to `oil_price`, which is exogenous and therefore
  cannot use `restrictions` at all — it needs a whole shocked copy of the fit.

```{r extension}
kable(ext_summary, row.names = FALSE,
      caption = "The exogenous series had to be carried past the fit's native forecast end first. koma silently shortens a horizon it cannot supply, so without this the battery would have returned fewer quarters than it claimed.")
```

## 2. Demand spillovers

```{r matrix}
kable(round(mat_a, 3),
      caption = "Cumulative 8-quarter GDP effect (pp) of a +1pp demand shock. Rows = shocked entity, columns = responding entity.")
```

```{r sanity}
kable(sanity, row.names = FALSE, caption = "spillover_sanity_checks() on the seed-A battery.")
```

**Own effect dominates in `r sum(sanity$ok[1])` of the checks** — 6 of 6
entities, the same clean result stage 2c reached and better than stage 2b's
7 of 11, where Austria's two-hop artefact broke it.

Two features of the matrix are worth naming because they are properties of the
partition rather than of the estimator.

**China's row is an order of magnitude smaller than everyone else's.** A +1pp
Chinese demand shock delivers `r sprintf("%.3f", mat_a["cn", "de"])`pp to
Germany and `r sprintf("%.3f", mat_a["cn", "us"])`pp to the US, against
`r sprintf("%.3f", mat_a["de", "fr"])`–`r sprintf("%.3f", mat_a["reu", "it"])`pp
for the intra-European pairs. That is what the trade weights say it should be:
China's bilateral weight on Germany is 3.4% and on the euro area as a whole
4.4%, because `r sprintf("%.0f%%", 100 * 0.782)` of Chinese goods trade is with
economies this model does not contain. The asymmetry is real — China's *column*
is larger than its row, so Europe reaches China more than China reaches Europe —
and it is a direct consequence of using observed IMF DOTS weights for the
Chinese row instead of the reciprocal approximation, which would have put 56% of
Chinese trade inside the euro area and inverted this result.

**`reu` behaves like a large country, not like a residual.** Its own effect
(`r sprintf("%.3f", mat_a["reu", "reu"])`) is the largest on the diagonal and
its outward spillovers are comparable to Germany's, which is what a bloc holding
roughly a third of euro-area GDP should look like.

```{r demand-de}
#| fig-width: 9
#| fig-height: 4
gdp_de <- response(demand$de, "gdp")
ggplot(gdp_de, aes(horizon, median_diff, colour = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to a +1pp German demand shock at horizon 1",
       x = "quarter", y = "GDP growth difference, pp", colour = NULL) +
  theme_minimal()
```

```{r demand-cn}
#| fig-width: 9
#| fig-height: 4
gdp_cn <- response(demand$cn, "gdp")
ggplot(gdp_cn, aes(horizon, median_diff, colour = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to a +1pp Chinese demand shock at horizon 1",
       x = "quarter", y = "GDP growth difference, pp", colour = NULL) +
  theme_minimal()
```

## 3. The seed placebo — and this is the headline

Before reporting that any of the above differs from stage 2b or 2c, the battery
has to clear its own noise floor. `r length(PLACEBO_SRC)` source entities were
re-run at a second seed on the **same fit**, and the two matrices compared cell
by cell.

```{r placebo}
kable(data.frame(
  system = c("stage 2b (df = 22)", "stage 2c (df = 22)", "**stage 2d (df = 32)**"),
  mean_abs_gap = c("0.350", "0.119", sprintf("**%.3f**", placebo_mean)),
  max_gap = c("1.850", "0.356", sprintf("**%.3f**", placebo_max)),
  sign_flips = c("7-10 of 30", "1 of 44", sprintf("**%d of %d**", placebo_flips, placebo_n))
), row.names = FALSE,
col.names = c("System", "Mean |seed A - seed B|", "Max", "Off-diagonal sign flips"),
caption = "Seed-to-seed movement of an off-diagonal spillover cell, same fit, different seed. Stage 2b and 2c figures from CLAUDE.md.")
```

**`r placebo_flips` sign flips out of `r placebo_n`**, a mean gap of
`r sprintf("%.3f", placebo_mean)`pp and a worst case of
`r sprintf("%.3f", placebo_max)`pp. Stage 2c already improved on stage 2b by a
factor of three; stage 2d improves on stage 2c again, and for the first time
**no off-diagonal cell changes sign** when only the seed changes.

This is the strongest result in the stage-2d work, and it is strong precisely
because it is a measurement *of* the noise rather than a claim made on top of
it. It is also the concrete payoff of the re-partition: collapsing seven
countries into a bloc was justified in `reports/stage2d_regional_core.qmd` on
degrees of freedom alone, and this is what those ten extra degrees of freedom
actually buy — a spillover matrix whose individual cells can be read, which
neither of its predecessors' could be.

```{r placebo-diag}
kable(data.frame(
  entity = toupper(PLACEBO_SRC),
  seed_A = round(diag(mat_a[PLACEBO_SRC, PLACEBO_SRC]), 3),
  seed_B = round(diag(mat_b[PLACEBO_SRC, PLACEBO_SRC]), 3),
  gap = round(diag(mat_a[PLACEBO_SRC, PLACEBO_SRC]) - diag(mat_b[PLACEBO_SRC, PLACEBO_SRC]), 3)
), row.names = FALSE, caption = "Own-diagonal effects at both seeds.")
```

## 4. The monetary shock

### 4.1 The euro-area rule behaves

```{r monetary-ea}
#| fig-width: 9
#| fig-height: 4
gdp_mon <- response(mon_a, "gdp")
ggplot(gdp_mon, aes(horizon, median_diff, colour = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  labs(title = "GDP response to +100bp on ea_policy_rate held for four quarters",
       x = "quarter", y = "GDP growth difference, pp", colour = NULL) +
  theme_minimal()
```

```{r monetary-table}
kable(data.frame(
  entity = names(cum_by(mon_a, "gdp")),
  gdp_cum_A = round(cum_by(mon_a, "gdp"), 3),
  gdp_cum_B = round(cum_by(mon_b, "gdp"), 3),
  gdp_peak_A = round(peak_by(mon_a, "gdp"), 3),
  prices_cum_A = round(cum_by(mon_a, "prices"), 3)
), row.names = FALSE,
col.names = c("Entity", "GDP cum. (seed A)", "GDP cum. (seed B)", "GDP peak (A)", "Prices cum. (A)"),
caption = "Cumulative and peak responses to the euro-area tightening, at both seeds.")
```

**GDP is correctly signed and the transmission is orderly.** Every entity's
peak GDP response is negative, the euro-area members most of all
(`r sprintf("%.2f", cum_by(mon_a, "gdp")[["DE"]])`pp cumulative for Germany,
`r sprintf("%.2f", cum_by(mon_a, "gdp")[["IT"]])` for Italy), and the long rate
tracks the policy rate essentially one-for-one, as the spread identity imposes.
The two entities outside the euro area show the smallest responses and are
**not** robust to the seed — the US goes from
`r sprintf("%+.3f", cum_by(mon_a, "gdp")[["US"]])` to
`r sprintf("%+.3f", cum_by(mon_b, "gdp")[["US"]])` and China from
`r sprintf("%+.3f", cum_by(mon_a, "gdp")[["CN"]])` to
`r sprintf("%+.3f", cum_by(mon_b, "gdp")[["CN"]])`, crossing zero in both cases.
Read those two as indistinguishable from nothing, which is the economically
sensible answer anyway.

**Prices are still wrong-signed, and that is inherited, not new.** Only
`r sum(peak_by(mon_a, "prices") < 0)` of 6 entities show a negative price peak
under a tightening. Stage 2c reported exactly this failure across all eleven
countries and diagnosed it: the policy rate's route into GDP is dominated by the
horizon-1 simultaneity echo rather than a lagged demand channel, so the Phillips
curve is fed a positive output impulse in the same quarter the rate rises. The
re-partition does not touch that mechanism, so it does not fix it. The oil shock
in §5 is the control that localises the fault to the *policy-rate* route rather
than to the price block itself.

```{r failed-draws}
kable(data.frame(
  quantity = c("posterior draws attempted", "draws that failed the restriction",
               "share", "baseline draws dropped to re-pair"),
  value = c(attr(failed, "n_draws"), attr(failed, "n_failed"),
            sprintf("%.1f%%", 100 * attr(failed, "n_failed") / attr(failed, "n_draws")),
            length(failed))
), row.names = FALSE,
caption = "A sustained restriction makes some draws fail outright. koma drops them by subsetting the list, recording nothing about which index, so scenario_diff() refuses to difference legs of unequal length; failed_restriction_draws() recovers the indices by tracing koma:::forecast_draw() and the baseline is re-paired against exactly those.")
```

### 4.2 The Chinese policy rate shock does not work

```{r monetary-cn-path}
kable(
  do.call(rbind, lapply(c("cn_policy_rate", "cn_long_rate", "cn_prices",
                          "cn_domestic_demand", "cn_gdp"), function(v) {
    x <- mon_cn[mon_cn$variable == v, ]
    stats::setNames(data.frame(v, t(round(x$median_diff, 3))), c("variable", paste0("h", x$horizon)))
  })),
  row.names = FALSE,
  caption = "China's own response to +100bp on cn_policy_rate, by horizon."
)
```

**This is a negative result and it should be read as one.** A +100bp Chinese
tightening *raises* Chinese GDP by `r sprintf("%.2f", mon_cn$median_diff[mon_cn$variable == "cn_gdp"][1])`pp
in the first quarter and by `r sprintf("%.2f", sum(mon_cn$median_diff[mon_cn$variable == "cn_gdp"]))`pp
cumulatively — the wrong sign, and far too large.

The mechanism is identifiable and it is not a bug in the shock. `cn_policy_rate`
has its own Taylor rule, and the restriction forces that rate a full point above
where the rule says it should be. The solver has to make the rule's own
arguments consistent with that, and the rule's loadings are
`r sprintf("%.3f", coef_of("cn_policy_rate", "cn_prices"))` on prices and
`r sprintf("%.3f", coef_of("cn_policy_rate", "cn_gdp"))` on GDP — so it takes a
very large move in Chinese output and prices to justify a one-point rate rise.
That is the entire response: it lands at **horizon 1** and decays to nothing by
horizon 3. It is a within-quarter accounting adjustment in the rule, not
transmission.

Nothing pushes back the other way, because China's demand channel is empty:
`cn_domestic_demand`'s loading on `cn_long_rate` is
`r sprintf("%.4f", coef_of("cn_domestic_demand", "cn_long_rate"))`, with a 90%
interval straddling zero. And that in turn traces to the data:
`cn_long_rate` is **spliced over 45% of the estimation window** (OECD's Chinese
10-year yield starts only 2014Q1), so before 2014 it moves exactly like the
three-month interbank rate and carries no independent term-premium variation for
the equation to identify against.

The euro-area rule does not have this problem because `ea_policy_rate` responds
to `ea_gdp` and `ea_prices`, which are identities over four entities — the
adjustment is spread across the whole area rather than loaded onto one country.

**Do not use `cn_policy_rate` as a monetary experiment in this system.** If a
Chinese monetary channel is what you need, the place to start is
`stage2d_config(spread_countries = setdiff(stage2d_countries(), "cn"))`, which
drops the spliced long rate and lets Chinese demand load the complete,
unspliced `IRSTCI` policy rate directly.

## 5. The oil shock

```{r oil}
#| fig-width: 9
#| fig-height: 4
p_oil <- response(oil_a, "prices")
ggplot(p_oil, aes(horizon, median_diff, colour = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  labs(title = "Price response to a +50% oil price level shock",
       x = "quarter", y = "price growth difference, pp", colour = NULL) +
  theme_minimal()
```

```{r oil-table}
kable(data.frame(
  entity = names(cum_by(oil_a, "prices")),
  prices_cum = round(cum_by(oil_a, "prices"), 3),
  gdp_cum = round(cum_by(oil_a, "gdp"), 3)
), row.names = FALSE, col.names = c("Entity", "Prices cum. (pp)", "GDP cum. (pp)"),
caption = "Cumulative response to a +50% oil price shock.")
```

**Correctly signed in `r sum(cum_by(oil_a, "prices") > 0)` of 6**, at
`r sprintf("%.1f", min(cum_by(oil_a, "prices")))`–`r sprintf("%.1f", max(cum_by(oil_a, "prices")))`pp
cumulative — the same result stage 2c got, and the control that matters for §4.
The price block responds sensibly to a **cost** shock and perversely to a
**policy rate**, which localises the monetary failure to the policy rate's route
into the system rather than to the price equations themselves.

China's price response (`r sprintf("%.2f", cum_by(oil_a, "prices")[["CN"]])`pp)
is the smallest, which is plausible: Chinese consumer prices are heavily
food-weighted and administered energy pricing damps pass-through.

## 6. What still fails

```{r trade-cor}
kable(sanity[3, ], row.names = FALSE,
      caption = "The trade-weight check.")
```

**Spillover magnitude still does not correlate with bilateral trade weight**
strongly enough to pass. The Spearman correlation is
`r sprintf("%.3f", as.numeric(sub(".*rho = ([0-9.-]+).*", "\\\\1", sanity$detail[3])))`
against a 0.3 bar — comparable to stage 2b's 0.141 and worse than stage 2c's
0.292. With six entities there are only 30 ordered pairs, so this statistic is
itself thin; but it is reported as a failure rather than explained away.

```{r explosive}
kable(
  stats::reshape(explosive, idvar = "variable", timevar = "horizon", direction = "wide"),
  row.names = FALSE, digits = 3,
  caption = "Share of baseline draws that are explosive, by horizon. Every number in this report is a median over the draws that survive this filter; a cell whose share is large is one whose spillover is estimated from a shrinking sample."
)
```

The explosive share climbs to roughly 35–48% by horizon 8, which is materially
better than stage 2b/2c's ~50% and stage 3b's 67% but is still the binding
constraint on how far this battery can be pushed. It is why every headline in
this report is a *cumulative median over surviving draws* and why §3 exists.

## 7. Verdict

| Check | Result |
|---|---|
| Own effect dominates | **6 / 6** — better than stage 2b (7/11), matching stage 2c |
| Seed reproducibility | **best in the project** — 0 sign flips of `r placebo_n`, mean gap `r sprintf("%.3f", placebo_mean)`pp |
| Monetary → GDP (euro area) | correctly signed, robust to the seed for the four EA entities |
| Monetary → prices | **still wrong-signed**, inherited from stage 2c and not addressed by the re-partition |
| Monetary → China's own rule | **does not work** — a horizon-1 rule adjustment, not transmission |
| Oil → prices | correctly signed **6 / 6** |
| Trade-weight correlation | **fails** at `r sprintf("%.3f", as.numeric(sub(".*rho = ([0-9.-]+).*", "\\\\1", sanity$detail[3])))` |

Use this system for **demand** spillovers between the euro area, the US and
China, where the matrix is now reproducible enough to read cell by cell, and for
**cost-push** experiments. Do not use it for a monetary experiment on China, and
treat any price response to a policy rate — in this stage or its predecessors —
as unresolved.
