---
title: "Stage 3b: external, fiscal and financial blocks"
subtitle: "Three blocks added one at a time — what each cost, which one broke things, and a fourth failed attempt at the import price"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3b_external_fiscal_financial.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir3a <- file.path("data", "cache", "stage3a")
dir3b <- file.path("data", "cache", "stage3b")

base <- readRDS(file.path(dir3a, "fit_a_final.rds"))
steps <- list(external = "fit_external.rds", fiscal = "fit_fiscal.rds",
              financial = "fit_financial.rds")
fits <- lapply(steps, function(f) {
  p <- file.path(dir3b, f)
  if (file.exists(p)) readRDS(p) else NULL
})
fits <- Filter(Negate(is.null), fits)

flags <- list(
  external = list(labour = TRUE, external = TRUE),
  fiscal = list(labour = TRUE, external = TRUE, fiscal = TRUE),
  financial = list(labour = TRUE, external = TRUE, fiscal = TRUE, financial = TRUE)
)

coefs_of <- function(x) coefficient_table(x$fit)
ci_width <- function(ct) {
  y <- ct[!grepl("^covid_|^constant$", ct$term), ]
  mean(y$ci_high - y$ci_low)
}
signs_of <- function(nm) {
  a <- flags[[nm]]
  do.call(sign_checks, c(list(coefs_of(fits[[nm]]), "de"), a))
}

base_coefs <- coefficient_table(base$fit)
base_signs <- sign_checks(base_coefs, "de", labour = TRUE)
```

## 1. What each block cost

Blocks were added cumulatively, with the full diagnostic battery after each,
so it is visible which one degrades what. The baseline is stage 3a phase A.

```{r cost-table}
rows <- list(data.frame(
  step = "stage 3a (baseline)",
  k = length(base$sys_eq$total_exogenous_variables),
  df = 98 - length(base$sys_eq$total_exogenous_variables),
  signs = paste0(sum(base_signs$ok), "/", nrow(base_signs)),
  acceptance_flagged = sum(check_acceptance_rates(base$fit)$flagged),
  mean_ci_width = ci_width(base_coefs)
))
for (nm in names(fits)) {
  ct <- coefs_of(fits[[nm]]); sc <- signs_of(nm)
  k <- length(fits[[nm]]$sys_eq$total_exogenous_variables)
  rows[[length(rows) + 1]] <- data.frame(
    step = paste("+", nm), k = k, df = 98 - k,
    signs = paste0(sum(sc$ok), "/", nrow(sc)),
    acceptance_flagged = sum(check_acceptance_rates(fits[[nm]]$fit)$flagged),
    mean_ci_width = ci_width(ct)
  )
}
kable(do.call(rbind, rows), row.names = FALSE, digits = 3,
      caption = "Cumulative cost of each block. T = 98 throughout.")
```

The degrees-of-freedom arithmetic was predicted before estimating and matched
exactly. The fiscal block is the expensive one: its accumulation identity
carries a lagged term, and its government-consumption equation takes a lagged
GDP term, so it costs four columns where the others cost one and two.

## 2. The fiscal block had to be reduced, and the reason is data, not design

The request specified revenue and expenditure equations with
`de_deficit == de_expenditure - de_revenue`. That block **cannot be
estimated**: German quarterly revenue and expenditure (Eurostat
`gov_10q_ggnfa`) begin **2002Q1**, and `T` is system-wide, so pulling them in
shortens the estimation window for *every* equation from 98 to 90 while `k`
rises to 90 — leaving **zero** residual degrees of freedom.

| fiscal variant | k | T | df |
|---|---|---|---|
| as specified (revenue + expenditure) | 90 | **90** | **0** |
| as built (Maastricht debt) | 88 | 98 | 10 |

The block is therefore built on `gov_10q_ggdebt` (`GD`, percent of GDP), which
is clean 2000Q1–2026Q1 for all eleven economies. What is lost is the
revenue/expenditure decomposition. What is kept is the mechanism the request
called out as mattering: debt accumulates, and the sovereign spread responds
to it.

**`de_netborrowing` is not the headline deficit**, and no coefficient on it
should be read as a deficit elasticity. It is the quarterly change in the debt
ratio, which is the deficit *plus* the growth-denominator effect *plus*
stock-flow adjustments. It is derived that way deliberately: it makes the
accumulation identity exact by construction (verified below), and it sets the
carry weight to exactly 1, which matters because koma's injected weights are
**not** time-varying — they are annualised, lagged a year and collapsed to a
single scalar — so a genuine `(1+i)/(1+g)` snowball factor could not have been
expressed as a weight at all.

## 3. Two koma limitations found on the way

Both would have caused silent damage rather than an error.

**A lagged identity could not be built.** koma supports one — its own Klein
vignette ships `capital_stock == 1.001*capital_stock.L(1) + 0.0002*investment`
— but `stage2_exogenous_variables()` collected identity component names raw,
so `de_govdebt.L(1)` was declared exogenous while koma strips the suffix on
its side and rejects the leftover as "Redundant exogenous variables detected".
Fixed by stripping `.L(...)`, as the preflight already did.

**Lagged endogenous names must not prefix one another.** koma's
`construct_phi()` locates a variable's lag columns with an *unanchored*
`grepl(paste0("^", name), ...)`. With both `de_debt.L(1)` and
`de_debt_ratio.L(1)` present, the shorter name matches both, the inner loop
runs twice for the same lag order, and the second pass overwrites the first —
pointing the companion matrix at the wrong row of B, with no error and no
warning. The names here (`de_govdebt`, `de_netborrowing`) share no prefix, and
`stage2_preflight()` gained an eighth check so it cannot be reintroduced.

A third, smaller trap: an identity term written without an explicit weight
(`b == b.L(1) + d`) is stored with `character(0)` weights — the same silent
corruption `CLAUDE.md` records for `+ -0.4*x`. `identity_equation()` always
emits `1*x`, so building identities through it stays safe.

## 4. Block 1 — external

```{r external-detail}
ct <- coefs_of(fits$external)
kable(ct[ct$equation %in% c("de_exports", "de_imports", "de_current_account") &
           !grepl("^covid_|^constant$", ct$term), ],
      row.names = FALSE, digits = 3, caption = "The external block's equations.")
```

The competitiveness restriction worked as intended. Stage 3a estimated the two
export-price terms separately and got −0.92 and +1.26 with both intervals wide
and spanning zero — the data did not reject equal-and-opposite, and did not
identify them apart either. Imposing the difference trades two
poorly-identified coefficients for one, and that one comes back correctly
signed. The current account's own equation behaves: it rises with exports,
falls with imports, and improves with the terms of trade.

### 4.1 A fourth failed attempt at the import price

The one substantive failure is `de_imports ← de_terms_of_trade`, and it is the
fourth specification to fail in the same direction.

```{r imports-four}
a_con <- readRDS(file.path(dir3a, "coefs_a.rds"))
a_rel <- coefficient_table(readRDS(file.path(dir3a, "fit_a_relative.rds"))$fit)
a_lag <- readRDS(file.path(dir3a, "coefs_a_lagged.rds"))
a_fin <- readRDS(file.path(dir3a, "coefs_a_final.rds"))
pick <- function(ct, term) {
  x <- ct[ct$equation == "de_imports" & ct$term == term, ]
  if (nrow(x) == 0) NA_real_ else x$estimate[1]
}
tot <- pick(ct, "de_terms_of_trade")
kable(data.frame(
  specification = c("no price term (adopted in 3a)", "own price, contemporaneous",
                    "own price + domestic price", "own price, lagged",
                    "inside terms of trade (3b)"),
  domestic_demand = c(pick(a_fin, "de_domestic_demand"), pick(a_con, "de_domestic_demand"),
                      pick(a_rel, "de_domestic_demand"), pick(a_lag, "de_domestic_demand"),
                      pick(ct, "de_domestic_demand")),
  implied_import_price = c(NA, pick(a_con, "de_import_prices"), pick(a_rel, "de_import_prices"),
                           pick(a_lag, "de_import_prices.L(1)"), -tot)
), digits = 3,
  caption = "Every attempt to give German import volumes a price term. The last row's implied coefficient is -1 times the terms-of-trade coefficient, since ToT enters import prices negatively.")
```

Terms of trade is `export_prices − import_prices`, so a coefficient of
`r round(tot, 3)` on it implies **`r sprintf("%+.3f", -tot)`** on import
prices — the same positive sign as the direct contemporaneous term, and the
same sign as every other attempt. Four specifications, one contamination:
within a quarter, German import volumes and German import prices both respond
to the same global impulse, and the equation contains no control for world
activity (`foreign_demand` sits on the export side) to separate them.

The terms-of-trade version does damage the demand elasticity less than the bare
own price did (`r round(pick(ct, "de_domestic_demand"), 3)` against 0.08), so
it is the least bad of the four — but it is still wrong-signed, and being
least-bad is not a reason to keep a coefficient that says import demand rises
when imports get expensive.

**Recommendation: drop the terms-of-trade term from the imports equation**, as
was done in stage 3a for the direct price. Terms of trade would remain in the
system for the current-account equation, where it is correctly signed and where
a relative price genuinely belongs. This is a change to the block as approved,
so it is reported rather than applied.

## 5. Block 2 — fiscal

```{r fiscal-detail}
ctf <- coefs_of(fits$fiscal)
kable(ctf[ctf$equation %in% c("de_netborrowing", "de_government") &
            !grepl("^covid_", ctf$term), ],
      row.names = FALSE, digits = 3, caption = "The fiscal block's equations.")
```

Both fiscal sign rules pass. Net borrowing falls when output grows
(automatic stabilisers) and rises with the long rate (debt service), though
both intervals span zero — at this sample length the signs are right and the
magnitudes are not pinned down.

`de_government` correctly has **no Metropolis step**: both its regressors are
lagged, so its gamma block is empty and `count_accepted` is `NA`. That is
deliberate. Letting government consumption respond to *contemporaneous* output
would build in a within-quarter automatic stabiliser this data cannot
identify, and would put a policy variable inside the simultaneous block.

### 5.1 The accumulation identity is exact, and does not destabilise the lags

```{r fiscal-identity}
sp <- fits$fiscal$panel
m <- stats::na.omit(cbind(koma::rate(sp$de_govdebt),
                          stats::lag(koma::rate(sp$de_govdebt), -1),
                          koma::rate(sp$de_netborrowing)))
lagf <- check_lag_stability(ctf)
kable(data.frame(
  check = c("govdebt == govdebt.L(1) + netborrowing, max |residual|",
            "own lags with |coefficient| >= 1",
            "largest own lag"),
  value = c(format(max(abs(m[, 1] - m[, 2] - m[, 3])), digits = 3),
            as.character(sum(lagf$flagged)),
            paste0(round(max(abs(lagf$estimate)), 4), " (", lagf$equation[which.max(abs(lagf$estimate))], ")"))
), caption = "Debt accumulation and lag stability.")
```

The debt ratio's law of motion holds exactly, and **nothing became explosive**.
This was the stated worry — a near-unit-root accumulation entering a system
whose largest own lag was already 0.9999 — and it did not materialise. The
reason is structural rather than lucky: `de_govdebt` is an *identity* with a
fixed unit weight, not an estimated equation, so the unit root is imposed
exactly at 1 and cannot be drawn above it. What could still destabilise the
system is the *loop* the accumulation sits in, which is section 6's business.

### 5.2 On its own, the fiscal block transmits nothing

```{r fiscal-terminal}
se <- fits$fiscal$sys_eq
users <- Filter(function(e) {
  col <- se$character_gamma_matrix[, e]
  "de_govdebt" %in% rownames(se$character_gamma_matrix)[nzchar(col) & col != "0" & col != "1"]
}, se$endogenous_variables)
kable(data.frame(
  question = "equations taking de_govdebt contemporaneously, at the fiscal step",
  answer = if (length(users)) paste(users, collapse = ", ") else "none"
), caption = "The debt stock is terminal until the financial block arrives.")
```

This is worth stating plainly because it is the honest reading of the
per-block table: **the fiscal block is pure cost until the financial block
lands.** It spends four columns of `k` and widens every credible interval in
the system by about a quarter, and until `de_long_rate` takes `de_govdebt`
there is no equation for the debt stock to influence — the same
terminal-variable pathology `CLAUDE.md` records for `long_rate` in stage 2 and
for `export_prices` in stage 3a. The two blocks have to be judged together,
not separately.

## 6. Block 3 — financial

The financial block was estimated twice. The first attempt modelled the long
rate as a level and broke the system; the second models the **spread** and
mostly repairs it. Both are shown, because the failure is the more instructive
half.

### 6.1 The level specification broke the long-rate equation

```{r long-rate-degradation}
srcs <- list(`stage 3a` = file.path(dir3a, "coefs_a_final.rds"),
             `+ external` = file.path(dir3b, "coefs_external.rds"),
             `+ fiscal` = file.path(dir3b, "coefs_fiscal.rds"))
tab <- NULL
for (nm in names(srcs)) {
  if (!file.exists(srcs[[nm]])) next
  x <- readRDS(srcs[[nm]]); x <- x[x$equation == "de_long_rate", ]
  g <- function(t) { v <- x$estimate[x$term == t]; if (!length(v)) NA_real_ else v }
  tab <- rbind(tab, data.frame(step = nm, own_lag = g("de_long_rate.L(1)"),
                               policy_rate = g("ea_policy_rate"), govdebt = g("de_govdebt")))
}
old <- readRDS(file.path(dir3b, "prefix_coefs_financial.rds"))
go <- function(t) { v <- old$estimate[old$equation == "de_long_rate" & old$term == t]
                    if (!length(v)) NA_real_ else v }
tab <- rbind(tab, data.frame(step = "+ financial (LEVEL)", own_lag = go("de_long_rate.L(1)"),
                             policy_rate = go("ea_policy_rate"), govdebt = go("de_govdebt")))
kable(tab, row.names = FALSE, digits = 4,
      caption = "de_long_rate under the level specification. The own lag is stable for three systems, then crosses 1 exactly when de_govdebt is added.")
```

The own lag sat at 0.92–0.94 for three consecutive systems, then jumped to
**1.004** — explosive — the moment `de_govdebt` became its fifth regressor at
`df = 8`, with the policy-rate loading flipping negative at the same instant.
At that point the equation was no longer identified: the lag ran to a unit root
and absorbed the structural content, leaving the debt coefficient at **−0.006**
and dragging down everything that loads on the long rate. Seven sign checks
failed.

### 6.2 Modelling the spread repairs most of it

```
de_spread    ~ de_govdebt + de_gdp + de_spread.L(1)
de_long_rate == 1*de_spread + 1*ea_policy_rate
```

This imposes the policy-rate coefficient at 1 rather than estimating it,
removes the unit root (a spread is stationary where a rate level is not), and
cuts the competing regressors from five to three. It costs **nothing** in
degrees of freedom — the long rate loses its own lag column and the spread
gains one.

```{r spread-vs-level}
ctf2 <- coefs_of(fits$financial)
sp_now <- ctf2[ctf2$equation == "de_spread" & !grepl("^covid_|^constant$", ctf2$term), ]
kable(data.frame(
  quantity = c("own lag of the modelled rate", "coefficient on de_govdebt",
               "policy-rate coefficient", "failing sign checks"),
  level_spec = c(sprintf("%.4f  (explosive)", go("de_long_rate.L(1)")),
                 sprintf("%+.4f  (wrong sign)", go("de_govdebt")),
                 sprintf("%+.4f  (estimated, wrong sign)", go("ea_policy_rate")),
                 "7"),
  spread_spec = c(sprintf("%.4f", sp_now$estimate[sp_now$term == "de_spread.L(1)"]),
                  sprintf("%+.4f  (correct sign)", sp_now$estimate[sp_now$term == "de_govdebt"]),
                  "1.0000  (imposed by the identity)",
                  as.character(sum(!signs_of("financial")$ok)))
), caption = "Level versus spread.")
```

The explosive lag is gone, the policy-rate coefficient is imposed rather than
mis-estimated, and **the debt coefficient changes sign to the one the economics
requires**. `sovereign_spread_widens_with_debt` now passes.

### 6.3 But the debt channel is still weak, and the credit channel is wrong

```{r spread-detail}
kable(sp_now, row.names = FALSE, digits = 4, caption = "The spread equation.")
scf <- signs_of("financial")
kable(scf[!scf$ok, c("check", "equation", "term", "estimate", "expected")],
      row.names = FALSE, digits = 4, caption = "Remaining failures.")
```

Two things should temper the improvement:

- **The debt coefficient's interval still straddles zero.** The sign is right;
  the magnitude is not identified. "Correctly signed but indistinguishable from
  zero" is an honest description of the sovereign-spread channel in this fit,
  and it is weaker than the block's premise requires.
- **The credit channel does not work.** Credit falls in output rather than
  rising, and investment falls in credit rather than rising. These are the
  financial block's own economics, not collateral damage from the long rate,
  and they do not have a specification fix waiting the way the spread did.

The other two failures — wage indexation at −0.006 and the price Phillips curve
at +0.004 — are the same near-zero pair that trades places between runs. Both
have intervals an order of magnitude wider than their estimates; they are
sampling noise on weak coefficients rather than findings.

## 7. Verdict

```{r verdict}
kable(data.frame(
  block = c("external", "fiscal", "financial"),
  cost_df = c(1, 4, 2),
  verdict = c(
    "Keep. The competitiveness restriction works and the current account behaves; near-zero cost in precision.",
    "Keep, but only with the financial block. Correct signs, exact identity, no destabilisation -- yet it costs 24% of the system's precision and on its own de_govdebt reaches no equation at all.",
    "Weakest of the three, and improved rather than fixed. The spread reformulation removed the explosive lag and corrected the debt sign, but the debt channel stays indistinguishable from zero and the credit channel is wrong-signed."
  )
), caption = "Per-block verdict.")
```

**Which block breaks things: the financial one** — and the reason was
identification, not economics. The level specification put a fifth regressor on
an equation already carrying a near-unit-root lag at `df = 8`; the lag won.
Modelling the spread instead recovered most of what was lost at zero cost in
degrees of freedom, which is worth remembering as a general lesson: when a
coefficient goes null, check whether a neighbouring lag has gone to a unit root
and swallowed it before concluding the channel is absent.

**What survives.** The external block is close to free and works. The fiscal
block is sound in isolation and delivers an exact accumulation identity without
destabilising anything — the stated worry about a near-unit-root stock did not
materialise, because an identity with a fixed unit weight cannot be drawn above
one. The sovereign-spread channel exists and points the right way, but is not
sharply identified.

**What does not.** The credit channel: credit falls in output, investment falls
in credit. Unlike the spread, this has no obvious specification fix, and at
`df = 8` with intervals 51% wider than the stage-3a baseline it is not clear the
data can support the block at all.

**On the import price, now 4 for 4.** Direct (+0.70), with a domestic
counterpart (+0.80), lagged (+0.40), and inside terms of trade (implied +0.91).
Every specification puts a positive coefficient on import prices in the volume
equation. The term is dropped; terms of trade remains in the current-account
equation, where it is correctly signed.
