---
title: "Stage 2a: two-country linked pilot"
subtitle: "Germany and France in one koma system -- proving the linkage mechanism before scaling to eleven"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2a_pilot.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir2a <- file.path("data", "cache", "stage2a")
pilot <- readRDS(file.path(dir2a, "pilot.rds"))
tuned <- readRDS(file.path(dir2a, "pilot_tuned.rds"))
acceptance <- readRDS(file.path(dir2a, "acceptance.rds"))
tuning <- readRDS(file.path(dir2a, "acceptance_tuning.rds"))
comparison <- readRDS(file.path(dir2a, "comparison.rds"))
sys_eq <- pilot$sys_eq
```

## 1. The system

Everything below is a single `koma::system_of_equations()` call: 21 equations,
13 stochastic and 8 identities, no external packages and no per-country loop
over `estimate()`.

```{r equations}
#| comment: ""
cat(paste(sys_eq$equations, collapse = ",\n"))
```

Three things differ from stage 1, and each is deliberate:

1. **Exports are driven by `<iso2>_foreign_demand`**, a fixed-weight identity
   over the *other country's endogenous GDP*. Stage 1 used the exogenous
   `row_gdp`, which excludes every modelled economy — so intra-EA export
   demand was missing entirely.
2. **`ea_policy_rate` is endogenous**, with a Taylor-type rule over
   GDP-weighted `ea_gdp` / `ea_prices` identities.
3. **`<iso2>_long_rate` enters investment.** This is what actually gives the
   policy rate a route to GDP — see §4.

### Endogenous vs exogenous, as koma parses them

```{r vars}
kable(data.frame(
  n = c(length(sys_eq$endogenous_variables), length(sys_eq$stochastic_equations),
        length(sys_eq$identities), length(sys_eq$exogenous_variables),
        length(sys_eq$predetermined_variables)),
  variables = c(
    paste(sys_eq$endogenous_variables, collapse = ", "),
    paste(sys_eq$stochastic_equations, collapse = ", "),
    paste(names(sys_eq$identities), collapse = ", "),
    paste(sys_eq$exogenous_variables, collapse = ", "),
    paste(sub("\\.L\\(1\\)", ".L1", sys_eq$predetermined_variables), collapse = ", ")),
  row.names = c("endogenous", "stochastic equations", "identities",
                "exogenous", "predetermined")
), caption = "How koma parses the system. Note identity left-hand sides are endogenous.")
```

The exogenous set is **five**, not the three truly-global variables. `<iso2>_government`
has no equation of its own (fiscal policy is a given, as in stage 1) but does
appear in the domestic-demand identities. koma's `validate_completeness()` is
an **exact-set** check — an undeclared variable aborts, and so does a
declared-but-unused one — so this set is neither a subset nor a superset of
what is used.

### Linkage weights

```{r weights}
lw <- pilot$linkage_weights
kable(data.frame(
  identity = c("de_foreign_demand", "fr_foreign_demand", "ea_gdp / ea_prices"),
  weights = c(
    paste(sprintf("%s %.3f", names(lw$foreign_demand$de), lw$foreign_demand$de), collapse = ", "),
    paste(sprintf("%s %.3f", names(lw$foreign_demand$fr), lw$foreign_demand$fr), collapse = ", "),
    paste(sprintf("%s %.3f", names(lw$ea), lw$ea), collapse = ", "))
), caption = "Each modelled partner keeps its true W_trade weight; everything else lumps into row_gdp.")
```

**Documented approximation.** `build_row_gdp()` excludes all eleven modelled
economies by construction, so in a two-country pilot the 0.935 weight on
`row_gdp` is proxying the nine *absent modelled* partners (0.374 of German
trade) as well as genuine rest-of-world. This shrinks monotonically as
countries are added — each new country claims its own `W_trade` weight out of
the `row` residual, and no other weight is rescaled.

The immediate consequence is that **the DE↔FR channel is small by
construction**: France is only 6.5% of German trade. §5 shows that this is
exactly what the estimates say.

### The four constructed series

`koma::estimate()` requires a series for *every* endogenous variable, identity-
defined ones included, and only says so from inside `estimate()`. Four have no
observed counterpart: `de_foreign_demand`, `fr_foreign_demand`, `ea_gdp`,
`ea_prices`.

They are built with `chain_weighted_index()`, **not** `apply_weights()`,
because koma estimates on growth rates: an identity is a statement about
`diff(log())`, and `log(0.6a + 0.4b) != 0.6*log(a) + 0.4*log(b)`. A level-space
aggregate would leave the estimator reading one relationship out of the data
while the identity's `Gamma` column encodes another — silently, since koma
performs no identity-consistency check. For `de_foreign_demand` the two
constructions differ by up to **13.7 percentage points** of quarterly growth.
The chained version reproduces its identity to 8.5e-14.

## 2. Pre-flight

```{r preflight}
kable(pilot$preflight, caption = "All checks run before committing to estimation.")
```

`model_identification()` fills free coefficients with `rnorm` draws, so a
single pass proves very little — it is repeated across 10 seeds, and passes
the order **and** rank conditions on all of them.

The "identities declared last" check is not cosmetic. koma assumes it
positionally in two places (`model_identification()` loops
`for (j in seq(1, n_endogenous - n_identities))` over columns, and
`estimate_sem()` indexes `y_matrix[, jx]` by the same index), so an identity
declared anywhere else makes koma estimate the wrong columns and mislabel the
results **with no error**. This is undocumented; every example koma ships
obeys it.

## 3. Runtime and acceptance rates

```{r runtime}
#| results: asis
cat(sprintf("Estimated in **%.1f s** (%d stochastic equations, %d observations, koma defaults: `ndraws = 2000`, `burnin_ratio = 0.5`, `tau = 1.1`). koma fans out one future per stochastic equation.\n",
            attr(pilot$fit, "runtime_s"), length(sys_eq$stochastic_equations), nrow(pilot$fit$y_matrix)))
```

```{r acceptance}
a <- acceptance
a$acceptance_rate <- ifelse(is.na(a$acceptance_rate), NA, round(a$acceptance_rate, 4))
kable(a, caption = "Acceptance rates at the default tau = 1.1. Band is 20-60% (koma:::get_default_acceptance_prob()).")
```

`de_prices` and `fr_prices` show `NA`: they have **no contemporaneous
endogenous regressor**, so there is no `gamma` block and no Metropolis step at
all. They are correctly not flagged.

Four equations sat marginally over the 60% ceiling. One tuning round at
`tau = 2.2` (koma's rule of thumb: doubling `tau` roughly halves acceptance)
cleared all four:

```{r tuning}
t <- tuning
t$acceptance_rate_before <- round(t$acceptance_rate_before, 4)
t$acceptance_rate_after <- round(t$acceptance_rate_after, 4)
kable(t[, c("equation", "tau", "acceptance_rate_before", "flagged_before",
            "acceptance_rate_after", "flagged_after")],
      row.names = FALSE,
      caption = "Before/after one tau round: 4 flagged -> 0.")
```

Worth noting for its own sake: the nine untouched equations' acceptance rates
are **bit-identical** before and after, confirming koma's documented claim that
a per-equation `[tau = ]` override leaves the rest of the system alone.

## 4. Does the policy rate reach GDP?

The stage-1 diagnostics report closed on a structural finding: a policy-rate
shock could not move GDP, because `long_rate` was a terminal satellite
equation. Making `ea_policy_rate` endogenous does **not** on its own fix
that — traced on the contemporaneous `Gamma` graph, a policy rate with no
investment channel reaches only `de_long_rate` and `fr_long_rate`, and stops.

Adding `<iso2>_long_rate` to the investment equation is what closes the loop:

```{r reachability}
g <- sys_eq$character_gamma_matrix
v <- sys_eq$endogenous_variables
adj <- g != "0" & g != "1"
diag(adj) <- FALSE
reachable <- function(from) {
  seen <- from
  repeat {
    nxt <- unique(c(seen, v[which(adj[v %in% seen, , drop = FALSE], arr.ind = TRUE)[, 2]]))
    if (setequal(nxt, seen)) break
    seen <- nxt
  }
  setdiff(seen, from)
}
kable(data.frame(
  from = c("ea_policy_rate", "de_gdp"),
  reaches = c(length(reachable("ea_policy_rate")), length(reachable("de_gdp"))),
  includes_fr_gdp = c("fr_gdp" %in% reachable("ea_policy_rate"), "fr_gdp" %in% reachable("de_gdp")),
  detail = c(paste(sort(reachable("ea_policy_rate")), collapse = ", "), "(full cycle)")
), caption = "Contemporaneous reachability on the Gamma matrix.")
```

The cross-country cycle asked for is present and closed:
`de_gdp → fr_foreign_demand → fr_exports → fr_gdp → de_foreign_demand →
de_exports → de_gdp`.

**Is that well-posed?** Yes, and estimation and forecasting stress different
things:

- **Estimation** is limited-information and equation-by-equation
  (`draw_parameters_j`). Contemporaneous endogenous regressors are read from
  the *observed* `y_matrix`; koma never inverts `Gamma` while sampling. A cycle
  is the ordinary simultaneous-equations case. The binding constraint is
  identification, checked symbolically before any data is touched — and it
  passes.
- **Forecasting** is where a bad loop would bite: the reduced form is
  `B %*% solve(Gamma)`, so `Gamma` must be invertible. Measured `det(Gamma)`
  stays in 0.31–1.11 across rate→investment coefficients of −10…+5 and Taylor
  gains 0…5. The economically-correct *negative* rate→investment sign moves it
  **away** from singularity, since a negative feedback loop is stabilising.

Had it not been well-posed there would have been three distinct failure modes:
under-identification aborts up front (cheap, loud); a singular `Gamma` throws a
bare base-R `solve()` error at forecast time; and a *near*-singular `Gamma`
throws nothing at all and silently returns explosive forecasts. koma has no
cycle, stability or eigenvalue check anywhere, so the third is the dangerous
one. Here the cross-country loop gain is ≈0.369 × 0.065 × 0.291 × 0.139 ≈
**0.001** — three orders of magnitude from 1.

## 5. Comparison against the stage-1 standalone DE and FR fits

Three of six equations per country are structurally identical to stage 1 and
three were respecified. Only the identical ones isolate what *joint estimation*
did, so they are reported separately — treating the others as comparable would
attribute a respecification to the linkage.

```{r structural}
kable(data.frame(
  equation = c("consumption", "imports", "prices", "investment", "exports", "long_rate"),
  `vs stage 1` = c("identical", "identical", "identical",
                   "gained <iso2>_long_rate", "row_gdp -> <iso2>_foreign_demand",
                   "gained <iso2>_gdp"),
  comparable = c("yes", "yes", "yes", "no", "no", "no"),
  check.names = FALSE
), caption = "Which equations can be compared at all.")
```

### The clean comparison

```{r clean}
d <- comparison[comparison$structurally_identical, ]
d <- d[order(d$equation, d$term), ]
kable(data.frame(
  equation = d$equation, term = sub("\\.L\\(1\\)", ".L1", d$term),
  stage1 = sprintf("%.3f [%.2f, %.2f]", d$estimate_s1, d$ci_low_s1, d$ci_high_s1),
  stage2a = sprintf("%.3f [%.2f, %.2f]", d$estimate_s2, d$ci_low_s2, d$ci_high_s2),
  delta = round(d$delta, 3),
  `outside stage-1 90% CI` = ifelse(d$outside_s1_ci, "YES", "-"),
  check.names = FALSE
), row.names = FALSE,
caption = "Structurally identical equations: posterior mean and 90% interval.")
```

```{r clean-summary}
#| results: asis
clean <- comparison[comparison$structurally_identical, ]
cat(sprintf("**%d of %d** coefficients in the identical equations moved outside their stage-1 90%% interval.\n",
            sum(clean$outside_s1_ci), nrow(clean)))
```

### Where they differ, and why

Three distinct patterns, and the mechanism for each is visible in koma's
sampler rather than inferred:

**`prices` does not move at all** (deltas 0.000–0.006, i.e. Monte Carlo noise).
It is the one equation with no contemporaneous endogenous regressor, so
`draw_omega_j` takes its `number_endogenous_in_j == 0` branch, which uses only
the equation's *own* restricted regressor columns. Nothing about the wider
system can reach it. This is a useful control: it confirms the differences
elsewhere are a real estimation effect and not a change of sample, vintage or
seed handling.

**`consumption` moves slightly** (`de_gdp` +0.007, `fr_gdp` +0.059).

**`imports` moves most** — `de_domestic_demand` 0.059 → 0.743, `fr_domestic_demand`
1.725 → 2.633.

The mechanism is the **instrument set, not the trade loop**. koma's sampler is
limited-information, but the `gamma` block of any equation with a
contemporaneous endogenous regressor is drawn from a target that projects on
the **full** `x_matrix`. Stage 1 gave Germany 12 columns; stage 2a gives 19:

```{r instruments}
#| comment: ""
de1 <- readRDS(file.path("data", "cache", "stage1", "de.rds"))
cat("added to the projection for every equation:\n  ",
    paste(sub("\\.L\\(1\\)", ".L1", setdiff(colnames(pilot$fit$x_matrix), colnames(de1$x_matrix))),
          collapse = ", "), "\n")
cat("\nestimation sample unchanged: stage 1 n =", nrow(de1$y_matrix),
    " stage 2a n =", nrow(pilot$fit$y_matrix), "\n")
```

Imports move most because `<iso2>_domestic_demand` is the most weakly
identified contemporaneous regressor in the stage-1 system — Germany's stage-1
estimate was 0.059 with a 90% interval of [−1.05, 1.18], essentially
uninformative — so it is the coefficient with the most room to move when the
instrument set widens. Note the interval does not *tighten* (it widens
slightly, to [−0.53, 2.03]): this is a shift in location, not a gain in
precision, and the two estimates remain mutually consistent.

**The headline is the null result.** No coefficient in a structurally
identical equation moved outside its stage-1 90% interval. That is the honest
reading of a two-country pilot with `w_de_fr = 0.065`: the mechanism works,
the cycle is closed and identified, and it changes almost nothing yet — which
is what a loop gain of 0.001 predicts. The linkage becomes economically
material only as countries are added and the `row_gdp` residual is replaced by
modelled partners.

### The new coefficients

```{r new}
ct2 <- readRDS(file.path(dir2a, "coef_stage2a.rds"))
new_terms <- ct2[ct2$term %in% c("de_foreign_demand", "fr_foreign_demand",
                                 "de_long_rate", "fr_long_rate", "ea_gdp", "ea_prices") |
                   ct2$equation == "ea_policy_rate", ]
kable(data.frame(
  equation = new_terms$equation, term = new_terms$term,
  estimate = round(new_terms$estimate, 3),
  `90% interval` = sprintf("[%.2f, %.2f]", new_terms$ci_low, new_terms$ci_high),
  check.names = FALSE
), row.names = FALSE, caption = "Coefficients that exist only in stage 2a.")
```

All signed as theory expects: export elasticities on foreign demand are
positive (DE 0.96, FR 1.67); the rate→investment channel is negative in both
countries (DE −0.111, FR −0.076), though both intervals straddle zero; and the
Taylor rule loads positively on **both** `ea_prices` (0.268) and `ea_gdp`
(0.203) with a near-unit-root lag (0.995). The `ea_policy_rate` equation's own
acceptance rate, 39.3%, sits comfortably mid-band.

## 6. Verdict

The mechanism works. The cross-country contemporaneous cycle is expressible in
one `system_of_equations()`, is identified on every seed tried, and yields an
invertible `Gamma` with a large margin. Acceptance rates are in band after one
tuning round. Every new coefficient is correctly signed.

Two caveats to carry into stage 2b:

- **The pilot's economic content is small.** With France at 6.5% of German
  trade and 93.5% of German foreign demand riding on a `row_gdp` series that
  excludes the nine other modelled economies, the linkage cannot do much yet.
  Nothing here validates the *magnitude* of cross-country transmission — only
  the mechanism.
- **The rate→investment channel is not sharply estimated.** Both coefficients
  are correctly signed but their 90% intervals include zero. The channel exists
  structurally; whether it is strong enough to make a policy-rate conditional
  forecast meaningful is a question for the conditional-forecast checks, not
  this report.
