---
title: "Stage 3a: the German labour market and disaggregated prices"
subtitle: "A wage–price loop and a second cross-country channel — what the data allowed, what the identification assessment found, and which coefficient came out wrong"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3a_labour_prices.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir3a <- file.path("data", "cache", "stage3a")
fit_a <- readRDS(file.path(dir3a, "fit_a_final.rds"))
untuned <- readRDS(file.path(dir3a, "fit_a.rds"))
hicp <- readRDS(file.path(dir3a, "hicp_weights.rds"))
panel3a <- readRDS(file.path(dir3a, "panel_stage2.rds"))
base_panel <- readRDS(file.path(dir3a, "panel.rds"))
stage2b <- readRDS(file.path("data", "cache", "stage2b", "full_tuned.rds"))

sys_eq <- fit_a$sys_eq
coefs <- coefficient_table(fit_a$fit)
signs <- sign_checks(coefs, "de", labour = TRUE)
acceptance <- check_acceptance_rates(fit_a$fit)
lag_stab <- check_lag_stability(coefs)
loop <- wage_price_loop_gain(fit_a$fit, "de", hicp$weights[["nonenergy_prices"]])
collin <- block_collinearity(panel3a, sys_eq, stage2b_dates())

n_stoch <- length(sys_eq$stochastic_equations)
n_id <- length(sys_eq$identities)
k <- length(sys_eq$total_exogenous_variables)
Tn <- estimation_length(panel3a, stage2b_dates())
```

## 1. What was built, and why it looks like this

Stage 2b linked eleven economies through **one** channel: trade-weighted
foreign demand over partners' GDP. Stage 3a adds a second — price
competitiveness — and gives Germany a labour block so that wages, unit
labour costs and prices form a loop rather than a chain.

Germany's block, as estimated (`covid_*` dummies elided):

```
de_consumption      ~ de_gdp + de_real_income + de_consumption.L(1)
de_investment       ~ de_gdp + de_long_rate + de_investment.L(1)
de_exports          ~ de_foreign_demand + de_export_prices + de_foreign_prices + de_exports.L(1)
de_imports          ~ de_domestic_demand + de_imports.L(1)
de_long_rate        ~ de_prices + ea_policy_rate + de_gdp + de_long_rate.L(1)
de_employment       ~ de_gdp + de_employment.L(1)
de_unemployment     ~ de_gdp + de_unemployment.L(1)
de_wages            ~ de_unemployment + de_prices + de_wages.L(1)
de_nonenergy_prices ~ de_ulc + de_import_prices + de_unemployment + de_nonenergy_prices.L(1)
de_energy_prices    ~ oil_price + eur_usd + de_energy_prices.L(1)
de_import_prices    ~ eur_usd + oil_price + de_foreign_prices + de_import_prices.L(1)
de_export_prices    ~ de_ulc + de_foreign_prices + de_export_prices.L(1)

de_gdp             == <expenditure shares>
de_domestic_demand == <expenditure shares>
de_foreign_demand  == <trade-weighted partners' gdp + row_gdp>
de_productivity    == de_gdp - de_employment
de_ulc             == de_wages - de_productivity
de_prices          == 0.891*de_nonenergy_prices + 0.109*de_energy_prices
de_real_income     == de_wages + de_employment - de_prices
de_foreign_prices  == <trade-weighted partners' export_prices>   [exogenous in phase A]
```

Three things in that listing differ from the block as originally specified.
Each is a data finding, not a preference.

### 1.1 `de_wages` is a wage *rate*, because otherwise the identities are wrong

EA-MD/QD publishes `WS`, "Wages and salaries" in current prices — a
whole-economy wage **bill**, not a wage rate. Used directly it breaks the
block in two places: `real_income == wages + employment - prices` would
double-count employment (the bill is already rate × employment), and a wage
Phillips curve run on the bill would largely re-estimate Okun's law.

`de_wages` is therefore `WS / TEMP`, labour income per worker. That makes
the whole block internally exact: `ulc == wages - productivity` telescopes
through `productivity == gdp - employment` to `WS - GDP`, nominal wage bill
over real output — the textbook definition of unit labour costs. This
matters because EA-MD/QD carries **no whole-economy ULC series at all**,
only seven sectoral ones, so ULC has to be derived either way; deriving it
this way makes it consistent with the rest of the block by construction.

The stated assumption: `WS` covers employees while `TEMP` counts total
employment, so this imputes employee compensation to the self-employed.
That is the standard construction for compensation per worker, and it is
preferred to `WS / EMP` because mixing employment concepts between `wages`
and `employment` would break the exactness above.

### 1.2 The price identity splits energy from non-energy, not core from energy

The specified identity was `prices == 0.85*core + 0.15*energy`. It cannot be
made exact: HICP "core" excludes energy **and** food, so core plus energy
covers only about 81% of the German basket and the missing food weight would
have to be renormalised away — silently attributing food inflation to core,
inside the very loop the block exists to model.

Eurostat's `NRG` and `TOT_X_NRG` do partition the basket. Verified against
`prc_hicp_inw` for every year from 1996 to 2025, they sum to `CP00` = 1000
exactly, with no residual. The identity uses those, and the weights come
from the data rather than the example:

```{r hicp-weights}
kable(data.frame(
  component = c("non-energy (TOT_X_NRG)", "energy (NRG)"),
  weight = round(as.numeric(hicp$weights[c("nonenergy_prices", "energy_prices")]), 4),
  specified = c(0.85, 0.15)
), caption = sprintf(
  "HICP weights for Germany, averaged over %d-%d (Eurostat prc_hicp_inw). Exact partition: %s.",
  min(hicp$years), max(hicp$years), hicp$exact
))
```

The energy weight is **`r round(hicp$weights[["energy_prices"]], 3)`**, not
the 0.15 assumed — a 4-point difference in the weight on the most volatile
component of the basket.

Weights are re-based annually, so a fixed identity weight is an
approximation. Over this window the German energy weight ranges from 88.4 to
125.5 per mille; the largest single-year departure from the average used
here is **`r round(hicp$max_deviation, 4)`** in weight units. That is the
size of the approximation, stated rather than buried.

Because "core" no longer appears in the identity, the block uses a new
concept `nonenergy_prices` rather than redefining the existing
`core_prices`. Nothing was silently repurposed; `de_core_prices` is still in
the panel, still `HICPNEF`, and is now simply unused by any equation.

### 1.3 The price variables feed back into exports, but not imports

As originally specified, `de_export_prices` appeared on no other equation's
right-hand side inside a German-only system — it would have been estimated
and then transmitted nothing, the same terminal-variable pathology
`CLAUDE.md` already records for `long_rate`. Exports therefore take a
relative-price pair (own and competitors'), which identifies well and comes
back nearly equal and opposite.

Imports take **no** price term. The symmetric choice was estimated four ways
and rejected every time; section 5.1 is the evidence. This is the one
substantive specification change stage 3a makes relative to the block as
designed, and it is the reason the competitiveness channel is one-sided.

## 2. Degrees of freedom: what fits and what does not

```{r df-table}
kable(data.frame(
  system = c("stage 2b", "stage 3a phase A", "stage 3a phase B", "full block, all 11"),
  stochastic = c(68, 74, 84, 134),
  identities = c(35, 39, 40, 90),
  k = c(76, 83, 92, 142),
  df = c(98 - 76, 98 - 83, 98 - 92, 98 - 142),
  status = c("estimated", "estimated (this report)", "planned", "impossible")
), caption = "k = 1 + one lag column per stochastic equation + exogenous variables; T = 98.")
```

The last row is the important one. Rolling this block out to all eleven
countries needs 142 predetermined columns against 98 observations — not
slow, not badly conditioned, but **arithmetically impossible**: koma
projects every equation on the full `k`-column `x_matrix` on every draw and
draws `Omega` from `riwish(T - k, .)`. Even three countries would leave
`df ≈ 4`. The block is a German pilot in the strict sense that the
architecture cannot support it everywhere.

This report covers **phase A** (`k = `r k``, `T = `r Tn``, `df = `r Tn - k``),
where `de_foreign_prices` is an exogenous series constructed from partners'
export-price data. Phase B makes it endogenous by giving the other ten
countries a one-equation export-price satellite, at `df = 6`.

## 3. Identification

`koma::model_identification()` checks the classical order and rank
conditions per stochastic equation, symbolically, and **aborts** on failure.
`stage2_preflight()` repeats it across ten seeds, because the function fills
free coefficients with `rnorm` draws and a single pass proves little.

```{r preflight}
kable(fit_a$preflight, row.names = FALSE, caption = "stage2_preflight(), phase A.")
```

**No equation is under-identified.** The order condition requires each
equation to exclude at least `n_endogenous - 1` = `r length(sys_eq$endogenous_variables) - 1`
variables; the stacked `[Gamma; B]` has
`r length(sys_eq$endogenous_variables) + length(sys_eq$predetermined_variables)`
rows and no equation includes more than five regressors, so every equation
excludes on the order of 200. Adding endogenous variables raises the bar for
every equation but adds rows faster than it raises it. Rank is the binding
condition and it passes on all ten seeds.

Two caveats on that result, neither of which the check itself would show:

- koma's `gamma_vectorization()` locates parameters with an **unanchored**
  `grep`, so `"gamma1_6"` also matches `"gamma1_60"`. With
  `r length(sys_eq$endogenous_variables)` endogenous variables, index
  collisions are possible. Treat a pass as necessary, not sufficient.
- Identification is symbolic: it sees which variables are *excluded*, never
  how correlated the *included* ones are. That is a different question, and
  it is the one asked next.

### 3.1 Collinearity: the `ulc` / `unemployment` question, answered

The specific concern was that `de_nonenergy_prices` takes both `de_ulc` and
`de_unemployment`, while `de_ulc == de_wages - de_productivity` and
`de_wages` responds to unemployment — so ULC is partly a function of
unemployment by construction. That would be weak identification (imprecise,
trading-off coefficients), not rank failure.

Measured directly from the panel, in the rate space koma estimates in:

```{r collinearity}
c2 <- collin[!grepl("^covid_", collin$term), ]
kable(head(c2[, c("equation", "term", "vif", "max_abs_cor", "worst_partner")], 12),
      row.names = FALSE, digits = 3,
      caption = "Variance inflation factors, worst first. Flagged above 10.")
```

**The concern does not materialise.** `de_ulc` has a VIF of
`r round(c2$vif[c2$equation == "de_nonenergy_prices" & c2$term == "de_ulc"], 2)`
and `de_unemployment` `r round(c2$vif[c2$equation == "de_nonenergy_prices" & c2$term == "de_unemployment"], 2)`
in the price equation, with a correlation of
`r round(c2$max_abs_cor[c2$equation == "de_nonenergy_prices" & c2$term == "de_ulc"], 3)`
between them. Nothing in the block exceeds a VIF of
`r round(max(c2$vif), 2)`, well under the conventional threshold of 10.

The weakest spot is elsewhere: `de_export_prices` and `de_foreign_prices`
in the exports equation correlate at
`r round(c2$max_abs_cor[c2$equation == "de_exports" & c2$term == "de_export_prices"], 3)`.
Both are price indices in a common inflation environment, so separating an
own-price from a competitor-price effect is the least well-identified thing
the block asks for. Their estimated coefficients are nearly equal and
opposite, which is what a relative price should look like, but the credible
intervals are correspondingly wide.

Note also that had `de_wages` and `de_ulc` *both* entered the price equation
— as the original sketch suggested — that would be algebraically identical
to including `de_wages` and `de_productivity`. Legitimate and full rank, but
it should then be written that way.

## 4. Sampler behaviour

### 4.1 Acceptance rates

```{r acceptance}
new_eqs <- c("de_employment", "de_unemployment", "de_wages", "de_nonenergy_prices",
             "de_energy_prices", "de_import_prices", "de_export_prices")
kable(acceptance[acceptance$equation %in% new_eqs, ], row.names = FALSE, digits = 3,
      caption = "The seven new behavioural equations, after tau tuning.")
```

`de_energy_prices` and `de_import_prices` show `has_mh_step = FALSE` and an
`NA` rate. That is correct, not a failure: in phase A every regressor in
both equations (`oil_price`, `eur_usd`, and the exogenous
`de_foreign_prices`) is exogenous, so neither has a contemporaneous
endogenous regressor, koma gives them no Metropolis step, and
`check_acceptance_rates()` never flags them. **This changes in phase B**,
when `de_foreign_prices` becomes endogenous and `de_import_prices` acquires
a Metropolis step for the first time.

An untuned stage-3a fit flagged
`r sum(check_acceptance_rates(untuned$fit)$flagged)` equations against a
tuned stage-2b baseline of zero — but only three of those were new
equations, and the rest were pre-existing equations pushed just over 60% by
the system-wide change in the residual covariance. Acceptance rates are not
a per-equation property in a simultaneous system: adding equations anywhere
moves every equation's rate. Re-tuning from the same doubling rule stage 2b
used (`tune_tau_system()`) is what the table above reflects; attributing the
untuned flags to the labour block would have been wrong.

### 4.2 The wage–price loop is not explosive

Stage 3a introduces a contemporaneous feedback loop that no earlier stage
had: `wages → ulc → nonenergy_prices → prices → wages`. Its round-trip gain
is `beta(nonenergy ← ulc) × w_nonenergy × beta(wages ← prices)`, and a draw
with gain at or above 1 is one where a wage rise more than pays for itself
through prices. koma constrains no draw to be stationary, so what matters is
the share of the posterior above 1, not the average.

```{r loop-gain}
kable(data.frame(
  statistic = c("mean", "median", "5th pct", "95th pct", "share |gain| >= 1"),
  value = c(loop$mean, loop$median, loop$q05, loop$q95, loop$share_ge_1)
), digits = 5, caption = "Posterior distribution of the wage-price loop gain.")
```

The gain is centred near zero and **no draw** reaches 1. The loop exists but
is weak, and it does not add to the explosive-draw problem documented in
`R/spillovers.R`.

### 4.3 Own-lag persistence

```{r lag-stability}
kable(head(lag_stab, 8), row.names = FALSE, digits = 3,
      caption = "Most persistent own lags. Flagged at |coefficient| >= 1.")
```

Nothing is flagged, but `de_unemployment` sits **exactly at the boundary**:
its own-lag coefficient is
`r round(lag_stab$estimate[lag_stab$equation == "de_unemployment"], 4)`
with a credible interval reaching
`r round(lag_stab$ci_high[lag_stab$equation == "de_unemployment"], 3)`.
That is a unit root in all but name: the equation is a random walk with a
GDP term, so an unemployment shock never dies out. For a rate modelled in
levels (`series_type = "rate"`, `method = "none"`) this is common and not by
itself wrong — the pre-existing `ea_policy_rate` and several `long_rate`
equations sit at 0.96-0.98 for the same reason — but it does mean the Okun
channel in this fit transmits GDP to a *permanently* shifted unemployment
level, and anything reading unemployment at long horizons should treat that
as a modelling artefact rather than a finding.

### 4.4 Precision: the block is estimable, but not sharply

```{r new-coefs}
new_terms <- coefs[coefs$equation %in% new_eqs & !grepl("^covid_|^constant$", coefs$term), ]
new_terms$spans_zero <- new_terms$ci_low < 0 & new_terms$ci_high > 0
kable(new_terms, row.names = FALSE, digits = 3,
      caption = "Every coefficient in the seven new equations.")
```

`r sum(new_terms$spans_zero)` of `r nrow(new_terms)` coefficients in the new
equations have credible intervals spanning zero. The signs are right, but at
`df = `r Tn - k`` the magnitudes are not sharply identified — most visibly
the wage-indexation term `de_wages ← de_prices`
(`r round(coefs$estimate[coefs$equation == "de_wages" & coefs$term == "de_prices"], 3)`,
interval
[`r round(coefs$ci_low[coefs$equation == "de_wages" & coefs$term == "de_prices"], 2)`,
`r round(coefs$ci_high[coefs$equation == "de_wages" & coefs$term == "de_prices"], 2)`]),
which is the return leg of the wage-price loop and is why the loop gain in
4.2 is so close to zero. The loop is present in the structure and weak in
the estimates; do not read the near-zero gain as evidence that wage-price
feedback is *absent* in the German economy, only that this window and this
specification do not pin it down.

## 5. Coefficient plausibility

```{r signs}
kable(signs, row.names = FALSE, digits = 4,
      caption = "Sign checks. Posterior means only; a check that could not be evaluated counts as a failure.")
```

**All `r nrow(signs)` pass**, in the adopted specification. Every central
economic claim the block makes holds:

- **The wage Phillips curve has the right sign.** `de_wages ← de_unemployment`
  = `r round(signs$estimate[signs$check == "wage_phillips_curve_negative"], 4)`.
  This was the designated red flag, and it is not raised.
- The price Phillips curve is also negative, ULC pass-through into
  non-energy prices is positive, and Okun's law holds in both the employment
  and unemployment equations.
- Both euro-appreciation rules pass with the correct (negative) sign, which
  is worth noting because `eur_usd` is quoted USD per EUR — a rise is a euro
  appreciation, and the intuitive sign is backwards.
- Exports fall in their own price and rise in competitors', nearly
  symmetrically, which is what makes the new competitiveness channel behave
  like a relative price rather than a price level.

### 5.1 The import-price term: tried four ways, then dropped

The imports equation above carries **no price term**. It did originally, and
removing it is the one substantive specification change stage 3a makes
relative to the block as designed. This section is the evidence for that
decision, because "we tried it and it did not work" is only worth acting on
if the trying is on the record.

In the first fit, `de_imports ← de_import_prices` came back at **+0.65**
— positive, when import demand should *fall* as imports get more expensive.

**A first hypothesis, tested and rejected.** The obvious explanation was a
specification asymmetry: exports received a relative-price *pair* (own and
competitors'), while imports received only their own price, with no domestic
counterpart to be relative *to*. That predicts the fix
`de_imports ~ de_domestic_demand + de_import_prices + de_prices + ...`,
which costs no degrees of freedom. It was estimated and it **does not work**
— the coefficient moves further the wrong way, and the added domestic-price
term takes the wrong sign too:

```{r imports-variants}
# The three rejected specifications, from their archived fits. The live fit
# (`coefs`) no longer contains an import-price term at all.
ccon <- readRDS(file.path(dir3a, "coefs_a.rds"))
crel <- readRDS(file.path(dir3a, "fit_a_relative.rds")) |> (\(x) coefficient_table(x$fit))()
clag <- readRDS(file.path(dir3a, "coefs_a_lagged.rds"))
b2b <- readRDS(file.path("data", "cache", "stage2b", "coef_full.rds"))
pick <- function(ct, term) {
  x <- ct[ct$equation == "de_imports" & ct$term == term, ]
  if (nrow(x) == 0) NA_real_ else x$estimate[1]
}
ci <- function(ct, term) {
  x <- ct[ct$equation == "de_imports" & ct$term == term, ]
  if (nrow(x) == 0) NA_character_ else sprintf("[%.2f, %.2f]", x$ci_low[1], x$ci_high[1])
}
kable(data.frame(
  specification = c("stage 2b (no price term)", "own price, contemporaneous",
                    "own price + domestic price", "own price, LAGGED one quarter",
                    "ADOPTED: no price term"),
  domestic_demand = c(pick(b2b, "de_domestic_demand"), pick(ccon, "de_domestic_demand"),
                      pick(crel, "de_domestic_demand"), pick(clag, "de_domestic_demand"),
                      pick(coefs, "de_domestic_demand")),
  import_price = c(NA, pick(ccon, "de_import_prices"), pick(crel, "de_import_prices"),
                   pick(clag, "de_import_prices.L(1)"), NA),
  import_price_ci = c(NA, ci(ccon, "de_import_prices"), ci(crel, "de_import_prices"),
                      ci(clag, "de_import_prices.L(1)"), NA)
), digits = 3, caption = "The imports equation across five specifications.")
```

**The revised diagnosis is simultaneity, and the table above is the
evidence.** Three things point the same way:

1. Adding a contemporaneous import price **collapses the demand elasticity**,
   from `r round(pick(b2b, "de_domestic_demand"), 2)` in stage 2b to
   `r round(pick(ccon, "de_domestic_demand"), 2)` with the price term and
   `r round(pick(crel, "de_domestic_demand"), 2)` in the variant. The price
   term is absorbing the scale variation that domestic demand should be
   carrying, which is what an omitted common driver looks like.
2. In the variant the two price coefficients come out as roughly
   `+0.7 × (import_prices − domestic_prices)` — a *relative* price with the
   wrong sign, not a mis-specified level. Adding the counterpart did not
   reveal a relative price; it reproduced the same contamination twice.
3. `de_import_prices` is itself driven by `oil_price` and `eur_usd` — global
   variables. The imports equation contains no control for world activity
   (`row_gdp` and `de_foreign_demand` appear in *exports*, not imports), so
   within a quarter the import deflator is the best available proxy for a
   global impulse that also raises German import volumes.

In other words the coefficient is not a demand elasticity at all: it is the
within-quarter correlation between a commodity-driven deflator and
commodity-driven volumes, with nothing in the equation to separate them.

**The decisive test: lag the import price.** That breaks the within-quarter
channel while keeping the variable. The prediction was that the sign would
turn negative. It **half** worked, and the half that failed is as
informative as the half that succeeded:

- **The demand elasticity is fully restored.** It returns to
  `r round(pick(clag, "de_domestic_demand"), 3)`, against
  `r round(pick(b2b, "de_domestic_demand"), 3)` in stage 2b — the
  contamination is gone, and this is the clearest confirmation that the
  contemporaneous term was stealing the scale variation.
- **The price coefficient does not flip.** It halves, to
  `r round(pick(clag, "de_import_prices.L(1)"), 3)`, and its interval now
  **spans zero** (`r ci(clag, "de_import_prices.L(1)")`) where the
  contemporaneous one did not.

So simultaneity is confirmed as the contaminating mechanism, but there is no
negative import-price elasticity underneath it waiting to be recovered.
Breaking the simultaneity does not reveal the textbook relationship; it
reveals that this window and this specification contain no identifiable
import-price effect at all.

**Decision: the import price is dropped from the imports equation** — not
lagged, not paired. Contemporaneously it actively damages the demand
elasticity; lagged it contributes nothing distinguishable from zero while
still costing a predetermined column. There is no version of the term that
earns its place.

The last row of the table is the adopted specification, and it behaves:
the demand elasticity sits at
`r round(pick(coefs, "de_domestic_demand"), 3)`, against
`r round(pick(b2b, "de_domestic_demand"), 3)` in stage 2b — the labour block
no longer disturbs an equation it was never meant to change.

Two consequences worth being explicit about:

- **The price block reaches trade volumes through exports only.** That is
  the side that identifies, with a genuine relative-price pair whose
  coefficients come back nearly equal and opposite. The competitiveness
  channel is therefore one-sided: German export volumes respond to relative
  prices, German import volumes do not.
- **`de_import_prices` remains in the system**, and remains useful — it is
  still an equation of its own and still feeds `de_nonenergy_prices`, which
  is where import-price pass-through into domestic inflation is supposed to
  act, and where it comes back correctly signed
  (`r round(signs$estimate[signs$check == "import_price_passthrough_positive"], 4)`).
  What was dropped is only its direct effect on import *volumes*.

There is a general lesson here that is not about imports. Three of the four
specifications produced a coefficient that a sign check would have reported
as a clean pass or a clean fail, with no indication that the *neighbouring*
coefficient had been gutted. The tell was never the sign — it was watching
`de_domestic_demand` move from 0.38 to 0.08 to 0.03 and back to 0.39.
A sign check on one coefficient cannot see that; only comparing
specifications can.

## 6. Verification

Every constructed identity reproduces its own definition in rate space to
machine precision — the check that catches a silently corrupted identity,
since koma has no identity-consistency check of its own:

```{r identity-exactness}
# Compare in RATE space, using the ts koma::rate() itself returns: it drops
# the first observation and any trailing NA, so re-dating it by hand shifts a
# series by a quarter and makes an exact identity look broken.
maxdiff <- function(lhs, rhs, w) {
  m <- stats::na.omit(do.call(cbind, lapply(c(lhs, rhs), function(n) koma::rate(panel3a[[n]]))))
  max(abs(m[, 1] - as.numeric(m[, -1, drop = FALSE] %*% w)))
}
wts <- hicp$weights
fpw <- foreign_price_weights(fit_a$linkage_weights, "de")
kable(data.frame(
  identity = c("productivity == gdp - employment", "ulc == wages - productivity",
               "prices == w1*nonenergy + w2*energy", "real_income == wages + employment - prices",
               "foreign_prices == trade-weighted partners"),
  max_abs_deviation = c(
    maxdiff("de_productivity", c("de_gdp", "de_employment"), c(1, -1)),
    maxdiff("de_ulc", c("de_wages", "de_productivity"), c(1, -1)),
    maxdiff("de_prices", c("de_nonenergy_prices", "de_energy_prices"),
            c(wts[["nonenergy_prices"]], wts[["energy_prices"]])),
    maxdiff("de_real_income", c("de_wages", "de_employment", "de_prices"), c(1, 1, -1)),
    maxdiff("de_foreign_prices", names(fpw), as.numeric(fpw))
  )
), caption = "Identity reproduction in rate space.")
```

`de_prices` is built from the sub-indices rather than taken from observed
headline HICP, deliberately. Observed HICP satisfies a *fixed*-weight
identity only approximately, because the real weights are re-based every
year, and koma would have enforced the identity anyway — silently. The gap
between the constructed series and observed headline HICP is the size of
that approximation:

```{r prices-vs-observed}
m <- stats::na.omit(cbind(koma::rate(panel3a$de_prices), koma::rate(base_panel$de_prices)))
kable(data.frame(
  statistic = c("max |difference|, pp", "mean |difference|, pp", "correlation"),
  value = c(max(abs(m[, 1] - m[, 2])), mean(abs(m[, 1] - m[, 2])), stats::cor(m[, 1], m[, 2]))
), digits = 4, caption = "Constructed de_prices vs observed headline HICP, quarterly growth.")
```

## 7. Phase B: closing the loop costs more than it buys

Phase B gives the other ten countries a one-equation export-price satellite,
which is what makes `de_foreign_prices` a genuine endogenous identity rather
than a fixed series. It passes preflight — `k = 92`, `T = 98`, `df = 6` — and
the sampler converges: `tune_tau_system()` settles in three rounds with
**zero** equations outside the acceptance band.

It should still not be preferred to phase A, but the case against it is
narrower than it first appeared, and the reason that changed is worth
recording.

```{r phase-b-load}
#| include: false
# Defined here rather than in the `phase-b` chunk below, because the prose
# between this point and that chunk cites `signs_b` in inline `r` expressions.
# knitr evaluates those in document order, so defining them later left the
# report unrenderable with "object 'signs_b' not found".
fit_b <- readRDS(file.path(dir3a, "fit_b_final.rds"))
coefs_b <- coefficient_table(fit_b$fit)
signs_b <- sign_checks(coefs_b, "de", labour = TRUE)
cmp <- merge(signs[, c("check", "estimate", "ok")], signs_b[, c("check", "estimate", "ok")],
             by = "check", suffixes = c("_A", "_B"))
```

**A correction to an earlier version of this report.** Run against the
*contaminated* imports specification of section 5.1, phase B looked far
worse: it lost three sign checks including `long_rate_loads_on_policy_rate`
in a stage-2b equation it never touched, which read as strong evidence that
`df = 6` had dissolved the system's economic content. Re-run against the
corrected specification, **that flip does not occur** — the policy-rate
loading stays positive at
`r round(signs_b$estimate[signs_b$check == "long_rate_loads_on_policy_rate"], 4)`,
against
`r round(signs$estimate[signs$check == "long_rate_loads_on_policy_rate"], 4)`
in phase A. The dramatic reading was an artefact of the bad import-price
term interacting with thin degrees of freedom, not a property of `df = 6`
alone. What survives is real but smaller, and is set out below.

```{r phase-b}
kable(cmp[!cmp$ok_B, ], row.names = FALSE, digits = 4,
      caption = "Sign checks that fail in phase B (phase A passes all of them).")
```

**1. One sign is lost, and it is a marginal one.** Phase A passes
`r sum(signs$ok)`/`r nrow(signs)`; phase B passes
`r sum(signs_b$ok)`/`r nrow(signs_b)`. The casualty is the price Phillips
curve, whose loading on unemployment goes from
`r round(signs$estimate[signs$check == "price_phillips_curve_negative"], 4)`
to
`r round(signs_b$estimate[signs_b$check == "price_phillips_curve_negative"], 4)`
— not a strong wrong-signed result so much as a coefficient that has lost
whatever grip it had on zero. On its own this would be a caution, not a
verdict.

**2. Precision degrades system-wide, not just where the equations were
added.** This is the substantial cost.

```{r ci-widths}
wid <- function(ct) {
  x <- ct[!grepl("^covid_|^constant$", ct$term), ]
  mean(x$ci_high - x$ci_low)
}
kable(data.frame(
  phase = c("A (df = 15)", "B (df = 6)"),
  mean_ci_width = c(wid(coefs), wid(coefs_b))
), digits = 3, caption = "Mean 90% credible-interval width across all coefficients.")
```

Intervals are **`r round(wid(coefs_b) / wid(coefs), 2)`x wider** in phase B.
Ten extra equations cost roughly half the precision of every coefficient in
the system, including the ones the labour block cares about. That is the
honest price of closing the loop, and it is paid everywhere, not just where
the equations were added.

**3. The wage–price loop develops an explosive tail.**

```{r loop-b}
loop_b <- wage_price_loop_gain(fit_b$fit, "de", hicp$weights[["nonenergy_prices"]])
kable(data.frame(
  phase = c("A (df = 15)", "B (df = 6)"),
  mean_gain = c(loop$mean, loop_b$mean),
  q05 = c(loop$q05, loop_b$q05), q95 = c(loop$q95, loop_b$q95),
  share_ge_1 = c(loop$share_ge_1, loop_b$share_ge_1)
), digits = 4, caption = "Wage-price loop gain, phase A vs phase B.")
```

In phase A **no** draw has a self-sustaining wage–price spiral. In phase B
**`r sprintf("%.1f%%", 100 * loop_b$share_ge_1)`** do. This is precisely the
mechanism `R/spillovers.R` documents — koma constrains no draw to be
stationary — arriving through the new loop as soon as the degrees of freedom
get thin enough.

One phase-B change is genuine and worth recording: `de_import_prices`
acquires a Metropolis step for the first time (`de_foreign_prices` is now
endogenous), taking the system from
`r sum(check_acceptance_rates(fit_a$fit)$has_mh_step)` to
`r sum(check_acceptance_rates(fit_b$fit)$has_mh_step)` equations with one.
`de_energy_prices` still has none, correctly — all its regressors remain
exogenous in both phases.

**Verdict: phase A is the usable artifact, and phase B is a defensible
second choice rather than a broken one.** Phase B closes the
price-competitiveness loop as designed, converges, and keeps
`r sum(signs_b$ok)`/`r nrow(signs_b)` of its economics. What it buys —
Germany's foreign prices responding to partners' costs instead of sitting
fixed — is bought with roughly half the precision on every coefficient in
the system and a
`r sprintf("%.1f%%", 100 * loop_b$share_ge_1)` chance that any given
posterior draw contains a self-sustaining wage–price spiral. For anything
that reads the posterior draw by draw — which is exactly what
`scenario_diff()` does — that tail is the binding objection.

The methodological point stands regardless of how the numbers came out.
`stage2_preflight()` passed, tau converged, no acceptance rate was out of
band: **nothing in the pipeline's own checks distinguishes phase A from
phase B.** The difference is only visible by estimating both and comparing,
and the first time this comparison was run it gave a much more alarming
answer that turned out to be driven by a specification error elsewhere in
the system. Both of those are arguments for keeping the cheaper fit around
as a reference point rather than trusting any single fit's diagnostics.

## 8. Where this leaves the block

What works: the identities are exact, the system is identified on every
seed, the labour block's economics come out with the right signs including
the wage Phillips curve, the collinearity concern that motivated the whole
identification assessment does not materialise, and the new wage–price loop
is weak enough that no posterior draw is self-sustaining.

What had to be given up: the import side of the competitiveness channel. The
import-price term is contaminated by within-quarter simultaneity, and once
that is broken there is no identifiable price elasticity underneath, so it
was dropped (5.1). German export volumes respond to relative prices; German
import volumes do not. Import-price pass-through survives where it belongs —
into domestic inflation, via `de_nonenergy_prices` — just not into volumes.

What phase B settled: `df = 6` is expensive but not fatal. The system
estimates, converges, and keeps almost all of its signs; what it loses is
precision — roughly half of it, everywhere — and the guarantee that no
posterior draw contains a runaway wage–price spiral. The competitiveness
channel can be closed on this window; it just costs more than it returns,
and the cost lands hardest on exactly the draw-by-draw use
(`scenario_diff()`) the next stage would want it for.

The rollout arithmetic is unchanged and remains the harder limit. The full
block cannot be rolled out at all (`k = 142` vs `T = 98`), and phase B is
already the cheapest partial rollout that closes the loop. Anything further
in this direction needs a longer sample or a smaller system, not a better
sampler: the Eurostat deflators reach back to 1991–1996 and EA-MD/QD to
2000, so extending the window is a real option worth costing before
extending the model.

One assumption carries more weight than its footnote suggests. Germany's
foreign-price index is renormalised over modelled partners only, because
there is no rest-of-world export-price series — and the weight this
reallocates is **`r round(attr(fpw, "row_weight_dropped"), 3)`**,
over half of Germany's trade weight. Foreign *demand* keeps that share on
`row_gdp` and accounts for the whole world; foreign *prices* cannot. The
index therefore assumes the unmodelled half of the world prices like the
modelled half. That is not a rounding detail, and it bounds how much of the
competitiveness channel this system can actually see.

