---
title: "Stage 2c equations: the template, the full system, and the fitted coefficients"
subtitle: "Eight equation forms, applied to eleven countries — what each one says economically, and every equation in full"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2c_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

# `noic_full_tuned.rds` is the RECOMMENDED stage-2c fit (four refinements);
# the cache name predates the decision to drop refinement 5. See
# reports/stage2c_refined_core.qmd Sec. 4.
f2c <- readRDS(file.path("data", "cache", "stage2c", "noic_full_tuned.rds"))
f2b <- readRDS(file.path("data", "cache", "stage2b", "full_tuned.rds"))

spec <- f2c$spec
spec2b <- f2b$spec
countries <- modelled_countries
eq_order <- names(c(spec$stochastic, spec$identities))
ov <- equation_overview(spec)
ov2b <- equation_overview(spec2b)
ct <- readable_coefficient_table(f2c$fit, eq_order)

# --- the country-generic template, derived from the spec, not asserted ---
tg <- template_groups(spec, countries)
ig <- identity_template_groups(spec, countries)
form <- function(concept) {
  x <- tg$form[tg$concept == concept]
  if (length(x) == 0) NA_character_ else x[1]
}
istruct <- function(concept) {
  x <- ig$structure[ig$concept == concept & ig$n_countries == max(ig$n_countries[ig$concept == concept])]
  if (length(x) == 0) NA_character_ else x[1]
}
ea_rule <- koma_string_stochastic("ea_policy_rate", spec$stochastic[["ea_policy_rate"]])

# --- what changed relative to stage 2b ---
new_stochastic <- setdiff(names(spec$stochastic), names(spec2b$stochastic))
dropped_stochastic <- setdiff(names(spec2b$stochastic), names(spec$stochastic))
common_stoch <- intersect(names(spec$stochastic), names(spec2b$stochastic))
changed_stochastic <- common_stoch[vapply(common_stoch, function(nm)
  !identical(spec$stochastic[[nm]], spec2b$stochastic[[nm]]), logical(1))]
touched_stochastic <- c(new_stochastic, changed_stochastic)
new_identities <- setdiff(names(spec$identities), names(spec2b$identities))
common_ident <- intersect(names(spec$identities), names(spec2b$identities))
changed_identities <- common_ident[vapply(common_ident, function(nm)
  !identical(spec$identities[[nm]], spec2b$identities[[nm]]), logical(1))]

n_stoch <- nrow(ov[ov$kind == "stochastic", ])
n_ident <- nrow(ov[ov$kind == "identity", ])

wrange <- function(concept, component) {
  r <- identity_weight_range(spec, countries, concept, component)
  sprintf("%.3f to %.3f (median %.3f)", r$min, r$max, r$median)
}
```

Stage 2c is the refined eleven-country linked core: `r n_stoch` stochastic
equations and `r n_ident` identities in one `koma::system_of_equations()` call,
`k = 76` against `T = 98`, `df = 22` — the same size as stage 2b, because every
refinement is `k`-neutral.

**`r n_stoch` equations is not `r n_stoch` different models.** There are eight
distinct behavioural forms, six of which are applied identically to all eleven
countries. Section 1 gives those forms generically, with what each one is
claiming economically. Sections 3–5 then list every equation individually, in
full, including the ones stage 2c did not change.

Two conventions apply throughout and are easy to trip over:

- **Everything is estimated in growth-rate space.** koma applies each series'
  own `diff_log` transform, so a coefficient on a level/index series is an
  **elasticity**, and an identity is a statement about growth rates, not levels.
- **Rate-type series are the exception.** `long_rate`, `spread` and the policy
  rates are tagged `series_type = "rate"`, `method = "none"`, so koma leaves
  them alone: level and rate space coincide, and their coefficients are in
  percentage points per percentage point.

## 1. The equation template

### 1.1 Behavioural equations

For a generic country `cc`, six equations, each estimated separately for all
eleven economies:

```{r template-summary}
kable(data.frame(
  concept = c("Consumption", "Investment", "Exports", "Imports", "Prices",
              "Sovereign spread", "Policy rule (euro area)", "Policy rule (US)"),
  form = c(form("consumption"), form("investment"), form("exports"),
           form("imports"), form("prices"), form("spread"),
           ea_rule, form("policy_rate")),
  applies_to = c(rep("all 11", 6), "one shared equation", "US only")
), row.names = FALSE, col.names = c("Equation", "General form", "Applies to"),
caption = "The eight behavioural forms. 66 country equations + 2 policy rules = 68.")
```

#### Consumption

```
cc_consumption ~ cc_gdp + cc_long_rate + <COVID dummies> + cc_consumption.L(1)
```

A consumption function with two channels and persistence. The loading on
`cc_gdp` is the **marginal propensity to consume** out of current income — the
Keynesian core of the equation, and the one coefficient here with a hard prior:
it must lie in `(0, 1)`, which `sign_checks()` enforces as `mpc_in_0_1`. A
value near 1 says households are liquidity-constrained and spend income as it
arrives; a low value says they smooth.

The loading on `cc_long_rate` is **intertemporal substitution**: a higher
return on saving should defer consumption, so the coefficient must be
**negative**. This term is one of stage 2c's four refinements, and its purpose
is structural rather than cosmetic — it gives monetary policy a *second* route
to the real economy alongside investment. In practice it is the weakest of the
four: correctly signed in 8 of 11 countries and small everywhere.

The own lag captures habit persistence and adjustment costs, and absorbs
whatever serial correlation the two structural terms leave behind.

#### Investment

```
cc_investment ~ cc_gdp + cc_long_rate + <COVID dummies> + cc_investment.L(1)
```

The accelerator plus the user cost of capital. `cc_gdp` is the accelerator
term: firms invest when demand is growing, and the elasticity is typically well
above 1 because investment is the most cyclical expenditure component.
`cc_long_rate` is the **cost of capital** and must be negative — and in stages
1 through 2b this was the *only* channel by which a policy rate could reach
output at all, which is why so much of this project's monetary diagnosis turns
on it. The own lag reflects time-to-build and adjustment costs.

#### Exports

```
cc_exports ~ cc_foreign_demand + <COVID dummies> + cc_exports.L(1)
```

Export demand. Volumes track a trade-weighted index of partner activity — that
index is an identity, defined in §1.2 — and the elasticity is normally above 1,
because trade grows faster than output over the cycle.

Note what is **absent**: there is no relative-price or competitiveness term at
stage 2c. Exports respond to how much partners are buying, not to how cheaply
this country sells. A price-competitiveness channel exists only in the
Germany-only stage-3a/3b external block (`reports/stage3a_labour_prices.qmd`),
which could not be rolled out to eleven countries within the degrees-of-freedom
budget.

#### Imports

```
cc_imports ~ cc_domestic_demand + <COVID dummies> + cc_imports.L(1)
```

Import demand as a function of domestic absorption, with an elasticity near 1 —
imports are roughly proportional to what the economy spends.

Note what is **deliberately** absent, because it is the most-tested negative
result in this project. This equation carries **no price term and no
contemporaneous export term**. An import-price term has been tried four
separate times (stage 3a) and a contemporaneous exports term once (stage 2c's
rejected fifth refinement); in every case the added regressor is highly
correlated with imports, improves fit substantially, and **collapses the
domestic-demand elasticity** — Germany's fell from 0.382 to 0.049 under the
exports term. The improved fit is the symptom, not the reassurance. See
`reports/stage2c_refined_core.qmd` §4.

#### Prices

```
cc_prices ~ eur_usd + oil_price + cc_gdp + <COVID dummies> + cc_prices.L(1)
```

A Phillips curve with two imported cost-push terms and inflation persistence.

`cc_gdp` is **stage 2c's headline refinement**. In growth-rate space this is a
"speed-limit" Phillips curve: faster output growth implies faster price growth,
so the coefficient must be positive. Its importance is structural rather than
quantitative. In stage 2b this equation had **no contemporaneous endogenous
regressor at all** — prices were driven only by their own lag and exogenous
variables — which meant `ea_policy_rate` could reach 91 of 103 endogenous
variables and *not one price*. The Taylor rule was responding to an inflation
rate that nothing it did could influence. Adding this one term closes that loop
and costs nothing, because GDP is already endogenous.

`oil_price` is the direct energy cost-push channel. `eur_usd` is exchange-rate
pass-through: a weaker euro raises import prices and thence the HICP. Both are
exogenous. The own lag is inflation persistence — indexation, staggered
contracts, and expectations anchored on recent inflation — and it is the
largest coefficient in the equation in every country.

One inherited quirk: the **US** price equation also uses `eur_usd`, not a
dollar-effective exchange rate. That is a stage-2b decision (the
three-variable exogenous spec) carried forward unchanged, and it breaks
continuity with the US block in stages 1 and 2a.

#### Sovereign spread

```
cc_spread ~ cc_prices + cc_gdp + cc_spread.L(1)
```

The country's sovereign yield **premium over the common policy rate**. This
equation replaces stage 2b's long-rate *level* regression, and the replacement
is stage 2c's third refinement.

The reason is empirical and stark: in stage 2b the long-rate level equation
regressed `cc_long_rate` on the policy rate and domestic prices, and **all
eleven policy-rate loadings had 90% credible intervals straddling zero** while
the own lag sat at 0.910–0.964. The equation was an AR(1) with decoration, and
the one coefficient it most needed to identify — the pass-through from the
policy rate to the sovereign yield — it could not. Modelling the spread imposes
that pass-through at exactly 1 (see the identity in §1.2) instead of estimating
it into insignificance.

Within the spread equation: `cc_prices` is an **inflation risk premium**, and
should be positive — investors demand compensation for inflation risk on
nominal debt. This is the weakest of the four sign priors a priori, since a
flight-to-quality episode can compress spreads while inflation rises, and it
fails in 3 of 11 countries. `cc_gdp` is a cyclical credit-risk term: stronger
growth improves debt dynamics and should compress the spread. The own lag is
large because risk premia are highly persistent.

One expectation this refinement did **not** deliver: the design argument was
that a spread is stationary where a rate level is not, so its own lag would
have no unit root to chase. Measured, the spread own lags run 0.933–0.983 —
as persistent as the levels they replaced, and for Spain and Portugal more so.

#### Policy rules

```
`r ea_rule`
`r form("policy_rate")`
```

Taylor-type rules with interest-rate smoothing. The central bank raises rates
when inflation and output growth are high; the own lag is the smoothing term
and is the dominant coefficient, as it is in essentially every estimated
policy rule.

There is exactly **one** `ea_policy_rate` equation however many euro-area
countries the system contains — that is what makes it a genuinely common
instrument, and it is defined over the area-wide aggregates `ea_prices` and
`ea_gdp` rather than any member's own. The US runs its own rule over its own
aggregates.

#### A note on the COVID dummies

Four impulse dummies (2020Q1/Q2/Q3 and 2021Q2) enter the **five real-side
equations and prices** — consumption, investment, exports, imports, prices —
and deliberately **not** the spread or either policy rule. Those quarters are
the only ones whose cross-country mean absolute GDP growth exceeds three times
the non-COVID baseline; rates and premia show no comparable mechanical break,
so dummying them would spend degrees of freedom to explain nothing.

Stage 2b uses dummies rather than the conditional-fill trick stages 1 and 2a
use, because its longer estimation window (to 2024Q4) is what buys the degrees
of freedom the eleven-country system needs at all.

### 1.2 Identities

Identities are **enforced during estimation**, not fitted. Their weights are
fixed — expenditure shares, trade shares, GDP shares — and koma has no
identity-consistency check, so this project verifies them itself with
`identity_consistency()` inside `stage2_preflight()`.

```{r identity-summary}
kable(data.frame(
  identity = c("GDP", "Domestic demand", "Foreign demand", "Long rate",
               "Euro-area GDP", "Euro-area prices"),
  form = c(istruct("gdp"), istruct("domestic_demand"),
           "cc_foreign_demand == sum_j w_j*<partner j>_imports + w_row*row_gdp",
           istruct("long_rate"),
           "ea_gdp == sum_i g_i*<member i>_gdp",
           "ea_prices == sum_i g_i*<member i>_prices"),
  count = c("11", "11", "11", "11", "1", "1")
), row.names = FALSE, col.names = c("Identity", "General form", "Count"),
caption = "44 country identities + 2 euro-area aggregates = 46.")
```

**GDP** is the expenditure decomposition — absorption plus net exports —
written in growth-rate space with fixed expenditure shares. It is one of only
two identities in this system that is **not exact**: real GDP is independently
observed, and the decomposition holds only up to the statistical discrepancy
and the fact that the shares are fixed rather than time-varying.
`identity_consistency()` reports its error but never flags it.

**Domestic demand** is absorption, consumption plus investment. Government is
**not in this system at all** — dropping it is what buys back the eleven
exogenous columns without which the eleven-country system cannot be estimated
on this window. Because the raw shares then sum to about 0.8, they are
**renormalised to sum to 1**; the cost is real and worth naming: the identity
now explains roughly 89% of domestic-demand growth rather than ~100%, and the
model has no fiscal channel whatsoever.

**Foreign demand** is the trade-weighted external demand index each country's
exports load on, and it is where stage 2c's second refinement lives. Partners
enter as **imports**, not GDP. The reasoning: partner GDP nets imports *out*
by construction, so a partner importing more mechanically lowers the measured
GDP its neighbours are supposed to be selling into — precisely backwards.
Partner imports are what an exporter actually faces. The partner set is
country-specific: a country never loads on itself, and partners below a 1%
trade weight are folded into the residual. Everything not modelled is
`row_gdp`, the rest-of-world proxy.

**Long rate** is the definition a sovereign yield satisfies: the risk-free
policy path plus a premium, with both weights **imposed at exactly 1**. This is
what makes refinement 3 `k`-neutral — the long rate gives up its own lag column
and the spread gains one. For the ten euro-area countries the policy rate is
`ea_policy_rate`; for the US it is `us_policy_rate`, and keeping those two in
step with the constructed spread series is load-bearing (getting it wrong
violated the US identity by 3.25pp in silence — see
`reports/stage2c_refined_core.qmd` §1.1).

**Euro-area GDP and prices** are nominal-GDP-weighted aggregates over the ten
members. They exist so the single policy rule has area-wide arguments to
respond to, which is what makes it a common instrument rather than eleven
separate ones.

### 1.3 The template is verified, not asserted

Every claim in §1.1 that a form is shared across countries is checked against
the spec rather than taken on faith: each country's equation is rewritten with
its own ISO-2 prefix replaced by `cc`, and identical strings are grouped. A
country that deviated would appear as its own group.

```{r template-verify}
kable(tg[, c("concept", "n_countries", "countries")], row.names = FALSE,
      col.names = c("Concept", "Countries sharing this form", "Which"),
      caption = "Stochastic equations, grouped by country-generic form. One row per concept means the form is unanimous.")
```

```{r identity-verify}
iv <- ig[ig$concept != "foreign_demand", c("concept", "n_countries", "countries")]
kable(iv, row.names = FALSE,
      col.names = c("Identity", "Countries sharing this structure", "Which"),
      caption = "Identities, grouped by country-generic component structure (weights ignored, signs kept). foreign_demand is excluded -- its partner set is country-specific by construction, so eleven distinct structures is the expected result, not a deviation.")
```

The only deviation anywhere is the **US long-rate identity**, which resolves to
`us_policy_rate` rather than `ea_policy_rate` — correct, since the US is not in
the currency union.

### 1.4 What the identity weights actually are

```{r weight-ranges}
kable(data.frame(
  identity = c("GDP", "GDP", "GDP", "Domestic demand", "Domestic demand"),
  weight_on = c("domestic demand", "exports", "imports", "consumption", "investment"),
  range = c(wrange("gdp", "domestic_demand"), wrange("gdp", "exports"),
            wrange("gdp", "imports"), wrange("domestic_demand", "consumption"),
            wrange("domestic_demand", "investment"))
), row.names = FALSE, col.names = c("Identity", "Weight on", "Range across the eleven countries"),
caption = "Fixed expenditure shares, computed over each country's own estimation window.")
```

```{r fd-weights}
fd <- do.call(rbind, lapply(countries, function(cc) {
  w <- spec$identities[[country_var(cc, "foreign_demand")]]
  data.frame(iso2 = cc, partners = sum(names(w) != "row_gdp"),
             row_gdp_weight = unname(w[["row_gdp"]]),
             largest_partner = names(w)[which.max(ifelse(names(w) == "row_gdp", -Inf, unlist(w)))],
             largest_weight = max(unlist(w)[names(w) != "row_gdp"]), stringsAsFactors = FALSE)
}))
kable(fd, row.names = FALSE, digits = 3,
      caption = "Each country's foreign-demand index: how many modelled partners it loads on, how much rides on the unmodelled rest of the world, and its single largest bilateral link.")
```

### 1.5 Two caveats in the weights, both documented rather than hidden

**Roughly half of every euro-area country's foreign demand is `row_gdp`.**
The weights above put 0.38–0.57 on the unmodelled rest of the world. The index
therefore assumes the half of the world this model does not contain behaves
like a GDP aggregate — a real approximation, not a rounding detail, and it caps
how much of any country's external cycle the system can attribute to its
modelled partners.

**The US row is an approximation, and it shows.** The ECB's WTS dataflow has no
US-reporter series, so `W_trade["us", ]` is built from the reciprocal of each
euro-area country's own weight on the US, renormalised. The visible consequence
is in the table above: the US `row_gdp` weight is
`r sprintf("%.3f", fd$row_gdp_weight[fd$iso2 == "us"])`, against 0.38–0.57 for
every euro-area country — i.e. the construction implies US foreign demand is
almost entirely euro-area, which is plainly not true of the actual US export
market. Ireland's cell in that row is additionally flagged unreliable, inflated
by the same multinational and tax-redomiciliation distortion behind the 2015
Irish GDP break. This is warned about at build time by `ecb_trade_weight()`
rather than silently substituted, and it is the main reason US spillover
results in this project should be read with more caution than euro-area ones.

## 2. What stage 2c changed relative to stage 2b

```{r changes}
kable(data.frame(
  refinement = c("1. Phillips curve", "2. Foreign-demand basis", "3. Spread",
                 "4. Consumption rate channel"),
  change = c("`cc_prices` gains `cc_gdp`",
             "`cc_foreign_demand` re-weighted over partner **imports** instead of partner GDP",
             "`cc_long_rate` becomes an identity over `cc_spread` + policy rate; the spread gets the equation",
             "`cc_consumption` gains `cc_long_rate`"),
  equations = c(length(grep("_prices$", touched_stochastic)),
                length(changed_identities),
                length(new_stochastic),
                length(grep("_consumption$", touched_stochastic)))
), row.names = FALSE, col.names = c("Refinement", "Change", "Equations affected"),
caption = "All four are k-neutral, and all four apply to every country.")
```

```{r moved}
kable(data.frame(
  movement = c("stochastic in stage 2b, identity in stage 2c",
               "new stochastic equations", "new identities",
               "existing stochastic equations with a new regressor",
               "existing identities re-weighted"),
  count = c(length(dropped_stochastic), length(new_stochastic),
            length(new_identities), length(changed_stochastic),
            length(changed_identities))
), row.names = FALSE, caption = "Where the equations moved. The stochastic count is unchanged because eleven long-rate regressions left and eleven spread equations arrived.")
```

A fifth refinement — a contemporaneous `cc_exports` term in the imports
equation — was estimated and **rejected**, and is not in the system documented
here. `stage2c_config(import_content = TRUE)` reproduces it.

## 3. Every equation, as koma parsed them

These are the literal strings passed to `koma::system_of_equations()`, for the
complete system — unchanged equations included.

```{r koma-stoch}
kable(ov[ov$kind == "stochastic", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = paste0("All ", n_stoch, " stochastic equations."))
```

```{r koma-ident}
kable(ov[ov$kind == "identity", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = paste0("All ", n_ident, " identities."))
```

## 4. Every equation, plain language

```{r readable-stoch}
kable(ov[ov$kind == "stochastic", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = paste0("All ", n_stoch, " stochastic equations."))
```

```{r readable-ident}
kable(ov[ov$kind == "identity", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = paste0("All ", n_ident, " identities."))
```

## 5. Every fitted coefficient

90% credible intervals (`ci_low`–`ci_high`), from `coefficient_table()` applied
to the cached, tau-tuned fit. All `r n_stoch` stochastic equations appear, in
the order the system declares them.

Identities carry no estimated coefficients — their weights are fixed and are
listed in full in §3 and §4. That is the point of refinement 3: the
policy-rate pass-through into the long rate is imposed at exactly 1 rather than
estimated, because when stage 2b did estimate it, all eleven credible intervals
straddled zero.

```{r coef-all}
#| results: asis
for (eq in intersect(eq_order, unique(ct$equation))) {
  sub <- ct[ct$equation == eq, ]
  print(kable(sub[, c("term_readable", "estimate", "ci_low", "ci_high")],
              digits = 4, row.names = FALSE,
              col.names = c("Term", "Estimate", "CI low (5%)", "CI high (95%)"),
              caption = sub$equation_readable[1]))
  cat("\n\n")
}
```
