---
title: "Stage 3d: the labour block on the regional core"
subtitle: "The rollout stage 3c said could not happen — and what a bloc changes about what a column of `k` buys"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3d_regional_blocks.qmd`.
# Artefacts come from scratch/stage3d_build.R.
devtools::load_all(".", quiet = TRUE)
library(knitr)
library(ggplot2)
source("reports/_forecast_helpers.R")

dir3d <- file.path("data", "cache", "stage3d")
g <- function(n) readRDS(file.path(dir3d, paste0(n, ".rds")))

shape      <- g("shape")
frontier   <- g("frontier")
preflight  <- g("preflight")
identity_chk <- g("identity_check")
acc        <- g("acceptance")
tuning     <- g("tuning_history")
coefs      <- g("coefs")
signs      <- g("signs")
lag_stab   <- g("lag_stability")
reach      <- g("reachability")
insample   <- g("insample")
hicp       <- g("hicp_weights")
fpw        <- g("foreign_price_weights")
sys_eq     <- g("sys_eq")

shape2d <- readRDS(file.path("data", "cache", "stage2d", "shape.rds"))
signs2d <- readRDS(file.path("data", "cache", "stage2d", "signs.rds"))
coefs3a <- readRDS(file.path("data", "cache", "stage3a", "coefs_a_final.rds"))
signs3a <- readRDS(file.path("data", "cache", "stage3a", "signs_a.rds"))

LABOUR <- g("labour_countries")
countries <- stage2d_countries()
n_endog <- length(sys_eq$endogenous_variables)

co <- function(tbl, eq, term) {
  r <- tbl[tbl$equation == eq & tbl$term == term, ]
  if (nrow(r) == 0) NA_real_ else r$estimate[1]
}
ci <- function(tbl, eq, term) {
  r <- tbl[tbl$equation == eq & tbl$term == term, ]
  if (nrow(r) == 0) return("--")
  sprintf("[%.3f, %.3f]", r$ci_low[1], r$ci_high[1])
}
fc <- forecast_artefact("stage3d")
```

## 1. What this is, and why it exists

`reports/stage3c_rollout.qmd` set out to roll the stage-3 blocks across the
euro area and concluded it could not be done:

> Data availability is not the constraint. […] The constraint is the
> degrees-of-freedom budget, and it binds by roughly a factor of two: the
> system is already at `df = 8` with **one** country carrying the blocks, and a
> second fully-loaded country would need `k = 104` against `T = 98`.

**That conclusion was right about the eleven-country partition and does not
bind on the six-entity one.** Stage 2d re-partitions the world — `de`, `fr`,
`it`, `us`, `cn` and the `reu` bloc — and in doing so drops `k` from 76 to 46.
The labour block still costs seven columns per entity; it is simply being added
to a system with thirty columns of room instead of none.

**But the arithmetic is not the interesting part. What a bloc changes is what a
column buys.** `reu` is one entity covering seven economies, so giving it the
block extends the wage–price channel to seven countries for the price of one:

```{r frontier}
kable(
  transform(frontier, feasible = ifelse(feasible, "yes", "no")),
  row.names = FALSE, align = "lrrrrrl",
  col.names = c("Configuration", "Labour entities", "Economies covered",
                "k", "T", "df", "Feasible (df >= 4)"),
  caption = "The affordability frontier on the regional core. Compare stage 3c's, where DE+FR labour alone reached df = 8 and covered two economies."
)
```

This report estimates the **`r paste(LABOUR, collapse = " + ")`** configuration:
**`r shape$economies_covered` of the eleven** modelled economies carry a
wage–price block, at `df = `r shape$df``, against stage 3a's **one** economy at
`df = 15`.

```{r shape}
kable(data.frame(
  System = c("Stage 2d (core)", "Stage 3a (DE on stage 2b)", "**Stage 3d (DE + REU on stage 2d)**"),
  `Economies with the block` = c(0, 1, shape$economies_covered),
  Equations = c(paste0(shape2d$stochastic, " + ", shape2d$identities), "74 + 39",
                paste0(shape$stochastic, " + ", shape$identities)),
  k = c(shape2d$k, 83, shape$k),
  T = c(shape2d$t_obs, 98, shape$t_obs),
  df = c(shape2d$df, 15, shape$df),
  check.names = FALSE
), row.names = FALSE, align = "lrlrrr")
```

Phase A only: `<iso2>_foreign_prices` stays exogenous. Stage 3a's phase B —
giving every partner an export-price equation so foreign prices become an
identity — cost ten columns on eleven countries and would cost five here, but
it is a separate question from the one this stage asks.

## 2. The bloc's labour variables

`reu_wages` is the GDP-weighted average of seven national wage **growth**
rates, not a euro-per-worker level, and `reu_employment` is an index rather
than a headcount. That is the right object for a model estimated in growth
space — every stage-3a identity is a statement about growth rates, and they
reproduce on the bloc exactly as on a country — but the levels are not
quotable.

```{r hicp}
kable(data.frame(
  entity = c("DE", "REU", paste0("  ", toupper(names(hicp$members)))),
  nonenergy = round(c(hicp$de[["nonenergy_prices"]], hicp$reu[["nonenergy_prices"]],
                      vapply(hicp$members, function(w) w[["nonenergy_prices"]], numeric(1))), 4),
  energy = round(c(hicp$de[["energy_prices"]], hicp$reu[["energy_prices"]],
                   vapply(hicp$members, function(w) w[["energy_prices"]], numeric(1))), 4)
), row.names = FALSE, align = "lrr",
caption = "HICP basket split. A bloc has no published basket, so bloc_hicp_weights() averages its members' with the same weights the bloc's series were built from; the members are shown indented.")
```

The bloc's energy share (`r sprintf("%.1f%%", 100 * hicp$reu[["energy_prices"]])`)
sits inside its members' range and just below Germany's
(`r sprintf("%.1f%%", 100 * hicp$de[["energy_prices"]])`). The weights are
renormalised to sum to exactly 1, which `labour_block()` requires: the split
*partitions* the basket, and seven rounded weights can otherwise leave a
1e-4 residual that would abort the block for no real reason.

### The foreign-price index is better here than in stage 3a

```{r fpw}
kable(
  do.call(rbind, lapply(names(fpw), function(cc) {
    w <- fpw[[cc]]
    data.frame(entity = toupper(cc),
               partners = paste(sprintf("%s %.3f", sub("_export_prices", "", names(w)),
                                        as.numeric(w)), collapse = ", "),
               row_dropped = round(attr(w, "row_weight_dropped"), 3),
               stringsAsFactors = FALSE)
  })),
  row.names = FALSE, col.names = c("Entity", "Partner weights (renormalised)", "row_gdp weight dropped"),
  caption = "foreign_price_weights(): the trade-weighted index each entity's export and import prices load on."
)
```

There is no rest-of-world export-price series, so the `row_gdp` residual is
dropped and the remainder renormalised — the index therefore assumes the
unmodelled world prices like the modelled partners. **In stage 3a that
reallocated 0.572 of German trade; here it reallocates
`r sprintf("%.3f", attr(fpw$de, "row_weight_dropped"))`.** The improvement is
not a modelling choice, it is coverage: the US and China are now partners with
export-price series of their own (FRED's implicit deflator and IMF IFS
`PXP_IX`), so a fifth of German trade that stage 3a had to assume away is now
observed.

## 3. Does it estimate?

```{r preflight}
kable(preflight, caption = "stage2_preflight(): all nine checks pass.")
```

```{r tuning}
flagged <- tapply(tuning$flagged, tuning$iteration, function(x) sum(x, na.rm = TRUE))
kable(data.frame(iteration = names(flagged), flagged = as.integer(flagged)),
      row.names = FALSE, align = "lr",
      caption = "tune_tau_system(): equations outside the 20-60% acceptance band, by round.")
```

`tune_tau_system()` converges to **zero** flagged equations — which stage 3a
also managed, but at `df = 15` rather than `r shape$df`. Acceptance rates run
`r sprintf("%.1f%%", 100 * min(acc$acceptance_rate[acc$has_mh_step]))`–`r sprintf("%.1f%%", 100 * max(acc$acceptance_rate[acc$has_mh_step]))`.

**`r sum(acc$has_mh_step)` of `r nrow(acc)` equations have a Metropolis step.**
The `r sum(!acc$has_mh_step)` that do not are exactly the ones with no
contemporaneous endogenous regressor, and they are the expected ones:

```{r no-mh}
kable(data.frame(equation = acc$equation[!acc$has_mh_step]), row.names = FALSE,
      caption = "Equations with no Metropolis step -- energy prices (oil and FX only) and import prices (whose only non-exogenous regressor, foreign_prices, is exogenous under phase A). Documented behaviour, not a failure.")
```

```{r lag}
kable(head(lag_stab[order(-lag_stab$estimate), ], 6), row.names = FALSE, digits = 3,
      caption = "The six most persistent own lags. None is flagged.")
```

**No own lag is explosive**, which is the check stage 3b failed: at `df = 8`
`de_long_rate` crossed to 1.004 the moment debt became its fifth regressor. At
`df = `r shape$df`` the most persistent equation here is
`r lag_stab$equation[which.max(lag_stab$estimate)]` at
`r sprintf("%.3f", max(lag_stab$estimate))`.

```{r identity}
kable(head(identity_chk[order(-identity_chk$max_abs_error), ], 8), row.names = FALSE, digits = 4,
      caption = "identity_consistency(). The exact identities hold to 9.6e-14; the ones reported but never flagged are the GDP and domestic-demand identities.")
```

::: {.callout-note}
## A bug this stage exposed

`reu_prices` initially failed the exact-identity check by **1.47e-07** — small,
but the identity is supposed to be exact. The cause was in
`identity_equation()`, which rendered weights with `format()`'s default **7
significant digits** while the identity's left-hand side series is built from
the full-precision numeric weight. Every weight in the project until now was a
rounded expenditure or trade share, so the two always agreed; the bloc's HICP
split is a seven-member average and is not a round number. koma has no
identity-consistency check, so this would have been enforced in silence.
Fixed by rendering 15 digits, which leaves every already-rounded weight
identical — verified against the cached stage-2b and 2c systems.
:::

## 4. What the bloc identifies that Germany alone does not

```{r wage-phillips}
kable(data.frame(
  Equation = c("Wage Phillips curve", "", "Okun (employment)", "",
               "Export-price pass-through", ""),
  Entity = c("DE (stage 3a)", "DE / REU (stage 3d)", "DE", "REU", "DE", "REU"),
  Term = c("de_unemployment", "de_unemployment / reu_unemployment", "de_gdp", "reu_gdp",
           "de_foreign_prices", "reu_foreign_prices"),
  Estimate = c(
    sprintf("%.3f", co(coefs3a, "de_wages", "de_unemployment")),
    sprintf("%.3f / %.3f", co(coefs, "de_wages", "de_unemployment"), co(coefs, "reu_wages", "reu_unemployment")),
    sprintf("%.3f", co(coefs, "de_employment", "de_gdp")),
    sprintf("%.3f", co(coefs, "reu_employment", "reu_gdp")),
    sprintf("%.3f", co(coefs, "de_export_prices", "de_foreign_prices")),
    sprintf("%.3f", co(coefs, "reu_export_prices", "reu_foreign_prices"))),
  `90% interval` = c(
    ci(coefs3a, "de_wages", "de_unemployment"),
    paste(ci(coefs, "de_wages", "de_unemployment"), "/", ci(coefs, "reu_wages", "reu_unemployment")),
    ci(coefs, "de_employment", "de_gdp"), ci(coefs, "reu_employment", "reu_gdp"),
    ci(coefs, "de_export_prices", "de_foreign_prices"), ci(coefs, "reu_export_prices", "reu_foreign_prices")),
  check.names = FALSE
), row.names = FALSE, align = "lllll")
```

**The wage Phillips curve is identified on the bloc and not on Germany.**
Germany's wage–unemployment coefficient is
`r sprintf("%.3f", co(coefs, "de_wages", "de_unemployment"))` with an interval
of `r ci(coefs, "de_wages", "de_unemployment")` — correctly signed but
straddling zero, which is also what stage 3a found on its own window. The
bloc's is `r sprintf("%.3f", co(coefs, "reu_wages", "reu_unemployment"))`,
interval `r ci(coefs, "reu_wages", "reu_unemployment")` — **the interval
excludes zero.**

That is the aggregation working *for* identification rather than against it,
and it is worth being precise about why. Germany's unemployment rate moves in a
narrow band over this window; the bloc's is a GDP-weighted average that
includes Spain, Greece and Ireland, whose unemployment rates swung by tens of
percentage points across the sovereign-debt crisis. The bloc has far more
variation in the regressor, and a wage equation can see the relationship
through it. `reu_employment`'s Okun coefficient
(`r sprintf("%.3f", co(coefs, "reu_employment", "reu_gdp"))` against Germany's
`r sprintf("%.3f", co(coefs, "de_employment", "de_gdp"))`) says the same thing
from the other side: employment in the bloc is far more responsive to output
than in Germany, which matches everything known about those labour markets.

**This cuts the other way too, and the sign checks say where.**

```{r signs}
kable(table(signs$check, ifelse(signs$ok, "pass", "fail")),
      caption = sprintf("sign_checks() across all six entities: %d of %d pass.",
                        sum(signs$ok), nrow(signs)))
```

```{r sign-fails}
kable(signs[!signs$ok, c("iso2", "check", "term", "estimate")], row.names = FALSE, digits = 4,
      caption = "The failures.")
```

Two of the `r sum(!signs$ok)` are in the bloc's own labour block and both are
about **pass-through**, not about the cycle: `reu_wages` does not index to
`reu_prices` (`r sprintf("%.3f", co(coefs, "reu_wages", "reu_prices"))`) and
`reu_export_prices` does not load on `reu_ulc`
(`r sprintf("%.3f", co(coefs, "reu_export_prices", "reu_ulc"))`). Both are
plausibly the aggregation: indexation arrangements and cost pass-through differ
sharply across seven economies, so averaging their growth rates averages away
the very mechanism the coefficient is trying to measure. The cyclical
relationships — which depend on *co-movement* rather than on a shared
institutional parameter — survive aggregation; the institutional ones do not.

That is a genuine limitation of the bloc, and it is the opposite trade to the
one Germany makes: Germany has the institutions but not the variation.

::: {.callout-note}
## A second bug this stage exposed

`sign_checks()` reported `phillips_curve_positive` as a failure for both labour
entities with `estimate = NA`. Under the labour block `<iso2>_prices` is an
**identity** over the energy and non-energy sub-indices, so there is no
`prices ~ gdp` coefficient to look at — the same claim is made and checked by
`price_phillips_curve_negative` on `nonenergy_prices`. The stage-2c rule is now
dropped for a labour country, exactly as the `spread` and `financial` rules
already were. Two deliberate absences were being reported as failures.
:::

## 5. Structure

```{r reach}
kable(data.frame(
  `policy rate` = names(reach),
  `endogenous reached` = sprintf("%d / %d", lengths(reach), n_endog),
  `price variables reached` = vapply(reach, function(r) sum(grepl("_prices$", r)), integer(1)),
  check.names = FALSE
), row.names = FALSE,
caption = "contemporaneous_reachability() from each policy rate.")
```

All three policy rates reach every price variable in the system, including the
disaggregated ones the labour block introduces. Stage 2c's structural fix
survives both the re-partition and the block.

## 6. Eight-quarter forecast

```{r fc-note}
#| results: asis
cat(fc_window_note(fc), "\n")
```

```{r fc-gdp}
#| fig-width: 9
#| fig-height: 5.2
v <- fc_vars(fc$paths, countries, "gdp")
plot_forecast_fan(fc$paths, v, space = "rate", ncol = 3, labeller = fc_entity_labels(v),
                  title = "Real GDP growth, quarterly %",
                  subtitle = "inner 25-75%, outer 10-90% clipped at the panel edge; grey = realised")
```

```{r fc-labour}
#| fig-width: 9
#| fig-height: 5.2
lv <- intersect(c("de_wages", "reu_wages", "de_unemployment", "reu_unemployment",
                  "de_nonenergy_prices", "reu_nonenergy_prices"), unique(fc$paths$variable))
plot_forecast_fan(fc$paths, lv, space = "rate", ncol = 3, labeller = fc_full_labels(lv),
                  title = "The block variables",
                  subtitle = "inner 25-75%, outer 10-90% clipped; grey = realised")
```

```{r fc-expl}
d <- fc$paths[fc$paths$kind == "forecast", ]
d$concept <- sub("^[a-z]{2,3}_", "", d$variable)
kable(
  do.call(rbind, lapply(c("gdp", "prices", "wages", "unemployment"), function(cn) {
    x <- d[d$concept == cn, ]
    if (!nrow(x)) return(NULL)
    data.frame(concept = cn,
               h1 = round(mean(x$explosive_frac[x$horizon == 1]), 4),
               h8 = round(mean(x$explosive_frac[x$horizon == max(x$horizon)]), 4))
  })),
  row.names = FALSE, col.names = c("Concept", "explosive share h1", "h8"),
  caption = "Share of draws discarded as explosive. Compare stage 2d (GDP 0.40 at h8, df 32) and stage 3a (0.70, df 15): this system sits between them, where df 18 says it should."
)
```

```{r fc-extension}
kable(fc$extension, row.names = FALSE,
      caption = "How each exogenous series was carried to the horizon.")
```

**One caveat the extension table makes visible.** `de_foreign_prices` and
`reu_foreign_prices` are extended by **eight** quarters, not three — the whole
forecast window — because they are chained indices over partners' export
prices and stop where the shortest partner does, which is China's IFS
export-price index at 2024Q4. Their extrapolated path is flat (trailing-4Q
average growth of `r sprintf("%.2f%%", 0)`/q, export prices having been flat),
so this is a conservative assumption rather than a wild one, but the price
channel's external driver is assumption rather than data across the whole
horizon.

## 7. Verdict

| | |
|---|---|
| Economies with a wage–price channel | **`r shape$economies_covered` of 11**, against stage 3a's 1 |
| `df` | **`r shape$df`**, against stage 3a's 15 and stage 3b's 8 |
| Acceptance tuning | converges to **0** flagged equations |
| Explosive own lags | **none** |
| Sign checks | `r sum(signs$ok)` of `r nrow(signs)` |
| Wage Phillips curve | **identified on the bloc**, not on Germany |
| Wage indexation, ULC pass-through | **lost** to aggregation on the bloc |

**Stage 3c's finding was not wrong; it was partition-specific.** "The rollout
cannot happen" was a true statement about eleven countries and a false one
about the world. What changed is not the estimator, the data or the block — it
is that a bloc makes one column of `k` buy seven economies instead of one.

**What this does not do.** It does not give Spain or Greece their own wage
equation — they have a share in one. An institution-specific question (national
indexation, a country's own ULC pass-through) still needs a country-level
block, and stage 3c's arithmetic still applies to that. The bloc answers
*regional* questions: what a euro-area-wide wage–price spiral does to the core,
and how the labour channel changes a spillover.
