---
title: "Stage 2d: the regional core with China"
subtitle: "Stage 2c's equations on a re-partitioned world — six entities, a euro-area bloc, and the first economy this project had to assemble from scratch"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2d_regional_core.qmd`.
# Artefacts come from scratch/stage2d_build.R.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir2d <- file.path("data", "cache", "stage2d")
g <- function(name) readRDS(file.path(dir2d, paste0(name, ".rds")))

shape        <- g("shape")
preflight    <- g("preflight")
identity_chk <- g("identity_check")
acc          <- g("acceptance")
acc_untuned  <- g("acceptance_untuned")
tuning       <- g("tuning_history")
tau          <- g("tau")
coefs        <- g("coefs")
signs        <- g("signs")
lag_stab     <- g("lag_stability")
reach        <- g("reachability")
insample     <- g("insample")
W            <- g("trade_weights")
lw           <- g("linkage_weights")
shares       <- g("shares")
bloc_w       <- g("bloc_weights")
cn_shares    <- g("cn_shares")
row_w        <- g("row_weights")
row_w_2b     <- g("row_weights_2b")
implied      <- g("implied_demand_elasticity")
sys_eq       <- g("sys_eq")

cfg <- stage2d_config()
countries <- stage2d_countries()
pct <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
n_mh <- sum(acc$has_mh_step)
```

## What this stage is

Stage 2d changes **who** the model is about, and nothing else. Every equation is
the one stage 2c settled on — the speed-limit Phillips curve, foreign demand
over partner *imports*, the sovereign spread in place of the long-rate level,
and the intertemporal-substitution term in consumption. What changes is the
partition of the world:

| | stage 2c | stage 2d |
|---|---|---|
| entities | 11 countries | `de`, `fr`, `it`, `us`, `cn`, `reu` |
| estimation window | 2000Q1–2024Q4 (`T = 98`) | 2005Q1–2024Q4 (`T = `r shape$t_obs``) |
| stochastic equations | 68 | `r shape$stochastic` |
| identities | 46 | `r shape$identities` |
| policy rules | `ea_policy_rate`, `us_policy_rate` | those plus `cn_policy_rate` |
| `k` | 76 | **`r shape$k`** |
| `df` | 22 | **`r shape$df`** |

Two of the six entities are not ordinary countries.

- **`reu`** is a *bloc*: the seven modelled euro-area economies other than
  Germany, France and Italy, aggregated into one pseudo-country with its own
  equations, its own row and column in the trade-weight matrix, and its own
  `foreign_demand` identity. Its members contribute no equations of their own.
- **`cn`** is China, the first economy this project has had to assemble rather
  than fetch. It is also the only one whose block departs from the stage-2c
  template, and it departs for a data reason, not a modelling preference.

**The headline is the degrees of freedom.** Every earlier structural change in
this project bought detail at the cost of `df` — stage 3a went from 22 to 15 to
6, stage 3b to 8. Stage 2d runs the other way: collapsing seven countries into
one bloc removes enough equations that `k` falls from 76 to `r shape$k`, and
even after giving up eighteen quarters at the start of the sample `df` rises
from 22 to **`r shape$df`**. This is the most comfortably identified joint
system the project has estimated.

## The two constructed entities

### `reu`: the rest of the euro area

```{r bloc-weights}
kable(
  data.frame(member = toupper(names(bloc_w)), weight = round(as.numeric(bloc_w), 4)),
  caption = "Nominal-GDP weights within the reu bloc (bloc_gdp_weights()).",
  align = "lr"
)
```

Every `reu_<concept>` series is built from those weights, with the aggregator
chosen by the series' own attributes: a **level** concept (every volume, every
price index) is chained in *growth* space by `chain_weighted_index()`, and a
**rate** concept (`unemployment`, `long_rate`) is a weighted average of levels.
The distinction is the one that governs `ea_gdp` and `<iso2>_foreign_demand`
already — `log(sum_i w_i x_i) != sum_i w_i log(x_i)`, and koma differences
everything before it estimates.

**The bloc's identity weights are its members' shares averaged, and getting
that wrong is easy.** `expenditure_shares()` forms each share as the mean of a
*level* ratio, which works for a real country because its series share a
currency and a scale. Every bloc series is a base-100 chain index instead, so
the ratio is not a share at all — it is the two indices' relative growth since
the base period. Measured on this panel, `mean(reu_exports / reu_gdp)` comes out
at **1.337**: `reu` exports grew 2.5x since 2000 while its GDP grew 1.6x, and
the ratio reports that rather than an export share. koma has no
identity-consistency check, so the wrong weight would have been enforced in
silence. `bloc_expenditure_shares()` averages the members' own shares instead:

```{r reu-shares}
kable(
  data.frame(
    component = names(shares$reu$gdp),
    weight = round(as.numeric(shares$reu$gdp), 3)
  ),
  caption = "reu_gdp identity weights, from the members' shares. Note reu_exports at 0.598, not 1.337 -- Ireland and the Netherlands carry very large trade shares, so the bloc is more open than the euro area as a whole.",
  align = "lr"
)
```

The name is **rest of the euro *area***, not rest of the EU. The bloc covers the
seven modelled EA economies, which is what the panel has data for; it is not the
EU, and it is not the whole euro area either (the six smallest members were never
in `modelled_countries`). Everything genuinely outside the six stage-2d entities
still goes to `row_gdp`, so nothing is double-counted — but `reu_gdp` is not a
published aggregate and should not be compared against one.

### `cn`: China, and the one thing the public data does not have

China publishes **no quarterly expenditure-side national accounts**, and that is
a fact about the data rather than a gap in the search. Verified absent from:

- **OECD Quarterly National Accounts** — carries `B1GQ` for China and nothing
  else, and only from 2011Q1;
- **OECD Economic Outlook** — annual for everything except CPI and two interest
  rates, and even annually it has no `CPV`/`IGV` for China;
- **IMF International Financial Statistics** — real GDP from 2011Q1, no
  expenditure split;
- **World Bank GEM**, **FRED** (China coverage discontinued between 2019 and
  2023) and the **NBS** itself.

So `cn_consumption` and `cn_investment` do not exist. The alternative to
inventing them from indicator proxies — retail sales, fixed-asset investment —
is to merge them, and that is what `merged_demand_countries` does:

```
cn_domestic_demand ~ cn_gdp + cn_long_rate + cn_domestic_demand.L(1)
cn_exports         ~ cn_foreign_demand + cn_exports.L(1)
cn_imports         ~ cn_domestic_demand + cn_imports.L(1)
cn_prices          ~ eur_usd + oil_price + cn_gdp + cn_prices.L(1)
cn_spread          ~ cn_prices + cn_gdp + cn_spread.L(1)
cn_policy_rate     ~ cn_prices + cn_gdp + cn_policy_rate.L(1)

cn_gdp            == 0.958*cn_domestic_demand + 0.243*cn_exports - 0.201*cn_imports
cn_long_rate      == 1*cn_spread + 1*cn_policy_rate
cn_foreign_demand == (trade-weighted partner imports)
```

The rate term survives the merge unconditionally, because it was investment's
term as well as consumption's and it is China's only monetary channel. What is
given up is the *split*: there is no separate MPC or accelerator to read off.
Net cost is one equation and one column of `k`.

**Sources**, all free and keyless:

| series | source | span (2026-09 vintage) |
|---|---|---|
| `cn_gdp` | WB GEM `NYGDPMKTPSAKN`, real GDP, SA | 1995Q1–2025Q4 |
| `cn_prices` | WB GEM `CPTOTSAXN`, CPI, SA, monthly | 1995M01–2026M02 |
| `cn_exports` | WB GEM `DXGSRMRCHSACD / DXGSRMRCHSAXD` | 2005Q1–2025Q3 |
| `cn_imports` | WB GEM `DMGSRMRCHSACD / DMGSRMRCHSAXD` | 1995Q1–2025Q3 |
| `cn_policy_rate` | OECD `DSD_STES@DF_FINMARK` `IRSTCI` | 1990Q1–2025Q2 |
| `cn_long_rate` | OECD `IRLT` spliced onto `IR3TIB` | see below |

#### Three constructions, each stated because each is one

**Trade volumes are deflated, not taken as published.** GEM's constant-price
China trade series are missing **every first quarter from 2020Q1 to 2025Q1** —
six of the eighty quarters in the window, all in the middle of the sample, where
koma cannot estimate across them at all. (China's customs administration stopped
publishing a separate January figure, merging January and February to smooth the
moving Lunar New Year.) The current-price and price-index series are complete, so
the volume is recovered as `value / price` — the same quotient GEM computes
internally, reproducing the published series where both exist to a mean ratio of
**0.99985** (sd 4.4e-4) for exports and **0.99974** (sd 9.4e-4) for imports. That
recovers six real quarters instead of interpolating them.

**Units are reconciled by a constant rescaling.** GDP is constant local currency
and the trade volumes are constant US dollars; the three cannot be added. Both
volume series are rescaled by a single constant so their window mean equals the
mean *nominal* trade share of GDP times mean real GDP, the shares coming from
GEM's own current-dollar series so numerator and denominator share a currency:
**`r sprintf("%.3f", cn_shares$exports)`** (exports) and
**`r sprintf("%.3f", cn_shares$imports)`** (imports). A constant multiplicative
rescaling changes no growth rate, so koma still differences exactly the published
volume index.

**Domestic demand is the accounting residual**, `cn_gdp - cn_exports +
cn_imports`, which makes the GDP identity exact in level space (verified to
3.7e-9). Because the trade series are **merchandise only**, the residual absorbs
the services balance — for China a persistent services *deficit* — so
`cn_domestic_demand` is a little larger than true domestic demand and grows a
little differently. Same class of approximation as `<iso2>_gdp`'s statistical
discrepancy, and `identity_consistency()` treats it the same way: reported,
never flagged.

#### The long rate is the weakest series here

```{r cn-splice}
sp <- cn_shares$long_rate_splice
kable(
  data.frame(
    quantity = c("mean IRLT - IR3TIB", "sd of that gap", "overlapping quarters",
                 "IRLT begins", "quarters spliced in the window"),
    value = c(sprintf("%.3f pp", sp$offset), sprintf("%.3f pp", sp$offset_sd),
              as.character(sp$overlap), paste0(sp$spliced_before[1], "Q", sp$spliced_before[2]),
              "36 of 80")
  ),
  caption = "The cn_long_rate splice, measured.",
  align = "lr"
)
```

OECD publishes a genuine Chinese 10-year government bond rate, `IRLT`, but only
from **2014Q1** — none of the first 36 quarters of the estimation window. Before
that, `cn_long_rate` is the three-month interbank rate shifted by the mean gap
over the overlap. The shift is additive rather than multiplicative because these
are rates in percentage points that pass through koma untransformed, and a ratio
splice on a series that can approach zero is not safe.

**The standard deviation of that gap is nearly twice its mean**, so the spliced
segment reproduces the *level* of a Chinese long rate but not its independent
variation: before 2014 `cn_long_rate` moves exactly like the three-month
interbank rate. The consequence is specific — `cn_spread` is a genuine term
premium over the second half of the sample and a money-market term spread plus a
constant over the first. If that is not acceptable for a given exercise, the
clean alternative is `stage2d_config(spread_countries = setdiff(stage2d_countries(), "cn"))`,
which lets `cn_domestic_demand` load `cn_policy_rate` directly: `IRSTCI` is
complete over the whole window and needs no splice at all.

One further interpolated observation: `IR3TIB` is missing 2006Q1, which
`fill_internal_gaps()` fills with a warning naming it.

## Trade weights, and a wart that had to be fixed

```{r trade-weights}
kable(round(W, 4), caption = "Stage-2d trade weights. Rows are reporters, columns partners plus the rest-of-world residual; every row sums to 1.")
```

```{r trade-source}
kable(
  data.frame(reporter = names(attr(W, "source")), source = unname(attr(W, "source"))),
  caption = "How each row was built.", align = "ll"
)
```

The euro-area rows come from ECB WTS as everywhere else in this project. The
`reu` row is its members' rows averaged with GDP weights and **intra-bloc trade
renormalised away** — `r pct(attr(W, "intra_bloc")[["reu"]])` of the averaged
row, which has no external counterparty once the seven members are one entity.

**The `us` and `cn` rows are the fix.** ECB WTS computes weights *for*
euro-area reporters only, so stages 1–3 build the US row from the reciprocal of
each euro-area country's own weight on the US, renormalised. Renormalising is
where it goes wrong: it forces the modelled partners to account for the
reporter's entire trade. The stage-2b matrix on disk gives the United States a
rest-of-world weight of **4.6%** and Ireland a **24.7%** share of US trade —
both plainly false, and both already recorded in `CLAUDE.md` as
known-unreliable. With only six entities, and China now among them, applying the
same construction would have given China a rest-of-world weight of **1.0%** and
put 56% of Chinese trade inside the euro area.

Both rows therefore come from observed **IMF Direction of Trade Statistics**
instead (`imf_dots_weights()`), where the world total is published and the
residual is real: the US at `r pct(W["us","row"])` rest-of-world and China at
`r pct(W["cn","row"])`, with the US–China pair at `r pct(W["us","cn"])` /
`r pct(W["cn","us"])`. Two differences from WTS are worth keeping in view, since
the two statistics now coexist in one matrix: DOTS is **goods only** where WTS
covers manufactured products *and services*, and DOTS is a plain trade share
where WTS is a double-weighted effective-exchange-rate weight that also accounts
for third-market competition.

### China leaves `row_gdp`

```{r row-gdp}
all_names <- union(names(row_w_2b), names(row_w))
kable(
  data.frame(
    partner = all_names,
    `stages 1-3` = round(as.numeric(row_w_2b[all_names]), 4),
    `stage 2d` = round(as.numeric(row_w[all_names]), 4),
    check.names = FALSE
  ),
  caption = "row_gdp_weights(): China is modelled now, so it must come out of the rest-of-world aggregate.",
  align = "lrr"
)
```

Passing the stage-2b `row_weights` to stage 2d would not error. It would just
load Chinese demand into every partner's `foreign_demand` identity twice — once
directly and once through `row_gdp` — silently.

## Pre-flight and identities

```{r preflight}
kable(preflight, caption = "stage2_preflight(): every check passes.")
```

```{r identity}
kable(
  head(identity_chk[order(-identity_chk$max_abs_error), ], 12),
  row.names = FALSE, digits = 4,
  caption = "identity_consistency(). The 13 identities meant to be exact hold to 9.8e-14; the ones reported but never flagged are the GDP and domestic-demand identities, which reconcile only up to the statistical discrepancy."
)
```

`reu_gdp`'s worst error, at 5.0 percentage points of quarterly growth, is the
largest in the table and it should be: on top of the statistical discrepancy a
single country carries (Germany 2.0, Italy 1.9), the bloc adds a **dispersion**
term, because its seven members' own expenditure shares differ from the average
the identity uses. It is the price of the aggregation, it is bounded, and it is
of the same order as the country-level slack rather than a different animal.

## Estimation

```{r tuning}
flagged_by_iter <- tapply(tuning$flagged, tuning$iteration, function(x) sum(x, na.rm = TRUE))
max_by_iter <- tapply(tuning$acceptance_rate, tuning$iteration, function(x) round(max(x, na.rm = TRUE), 4))
kable(
  data.frame(iteration = names(flagged_by_iter),
             flagged = as.integer(flagged_by_iter),
             max_acceptance = as.numeric(max_by_iter)),
  caption = "tune_tau_system(): flagged equations and the worst acceptance rate per round.",
  align = "lrr"
)
```

All **`r n_mh` of `r nrow(acc)`** equations have a Metropolis step — every one
of them has a contemporaneous endogenous regressor, which stage 2c's Phillips
curve is what made true. After tuning, acceptance rates run
`r pct(min(acc$acceptance_rate[acc$has_mh_step]))`–`r pct(max(acc$acceptance_rate[acc$has_mh_step]))`
against the 20–60% band, with `r sum(acc$flagged)` equation
(`r paste(acc$equation[acc$flagged], collapse = ", ")`) marginally over at
`r pct(max(acc$acceptance_rate[acc$flagged]))`. The doubling rule oscillates
around the boundary rather than converging on it; three rounds took the count
from `r flagged_by_iter[["0"]]` to `r sum(acc$flagged)` and a fourth would not
help. `r length(tau)` equations carry a `tau` override, all at 2.2.

```{r lag-stability}
kable(
  head(lag_stab[order(-lag_stab$estimate), ], 8), row.names = FALSE, digits = 4,
  caption = "The eight most persistent own lags. None is explosive."
)
```

**No own lag is flagged**, which is worth stating given the history: stage 3b's
`de_long_rate` crossed to 1.004 the moment it gained a fifth regressor at
`df = 8`. At `df = `r shape$df`` the most persistent equation here is
`us_policy_rate` at `r round(max(lag_stab$estimate), 3)`, with the whole
spread family in the 0.93–0.97 range that stage 2c also reported.

## Does the structure work?

### Every policy rate reaches every price

```{r reachability}
n_endog <- length(sys_eq$endogenous_variables)
kable(
  data.frame(
    `policy rate` = names(reach),
    `endogenous reached` = sprintf("%d / %d", lengths(reach), n_endog),
    `price variables reached` = vapply(reach, function(r) sum(grepl("_prices$", r)), integer(1)),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "contemporaneous_reachability() from each of the three policy rates."
)
```

This is the structural claim stage 2c existed to establish, and it survives the
re-partition intact. In stage 2b `ea_policy_rate` reached 91 of 103 endogenous
variables and **not one price**, so the Taylor rule's inflation term responded to
something nothing it did could influence. Here all three policy rates reach all
`r n_endog` endogenous variables and all seven price variables. Adding a third
rule does not weaken it: China's rule closes its own loop through
`cn_prices -> cn_policy_rate -> cn_long_rate -> cn_domestic_demand -> cn_gdp ->
cn_prices`, and reaches the euro area and the US through the trade identities.

Reachability is exact and verifiable. It is **not** a claim that the response has
the right sign or size — stage 2c's own lesson was that closing the loop
structurally did not make the monetary shock work, and only the seed placebo
showed it. A spillover battery on this system is the obvious next step and is
not in this report.

### Sign checks

```{r signs}
kable(
  table(signs$check, ifelse(signs$ok, "pass", "fail")),
  caption = sprintf("sign_checks() across the six entities: %d of %d pass.",
                    sum(signs$ok), nrow(signs))
)
```

```{r sign-failures}
kable(
  signs[!signs$ok, c("iso2", "check", "equation", "term", "estimate")],
  row.names = FALSE, digits = 4, caption = "The failures."
)
```

`r sum(signs$ok)` of `r nrow(signs)` pass. The two Phillips-curve failures are
small and both straddle zero — Italy's at `r round(signs$estimate[signs$iso2 == "it" & signs$check == "phillips_curve_positive"], 5)`
is indistinguishable from nothing at all. `spread_loads_on_prices_positive` was
already flagged in stage 2c as the weakest of the four a priori, since a
flight-to-quality episode compresses spreads while inflation rises. None of the
four is in a China or `reu` equation, which is the useful negative result: the
two constructed entities are not where the specification strains.

### The merged China equation, judged against the right benchmark

```{r implied}
cn_row <- coefs[coefs$equation == "cn_domestic_demand" & coefs$term == "cn_gdp", ]
kable(
  rbind(
    data.frame(iso2 = implied$iso2, consumption = round(implied$consumption, 3),
               investment = round(implied$investment, 3), elasticity = round(implied$implied, 3)),
    data.frame(iso2 = "cn", consumption = NA, investment = NA,
               elasticity = round(cn_row$estimate, 3))
  ),
  row.names = FALSE,
  caption = "Domestic-demand elasticity to GDP: implied from the split for the five entities that have it, estimated directly for China."
)
```

**The first bound put on this rule was wrong, and the data said so.** The
obvious rule for a merged demand equation is the MPC's — a coefficient in
`(0, 1)` — and China's `r round(cn_row$estimate, 3)` fails it. But an MPC below
one is a claim about *consumption*; total domestic absorption also contains
investment, which is far more cyclical than output. Computing the implied
elasticity for the five entities that do have the split gives
`r sprintf("%.2f", min(implied$implied))`–`r sprintf("%.2f", max(implied$implied))`,
with the **United States at `r round(implied$implied[implied$iso2 == "us"], 3)`** —
so a `(0, 1)` bound would have reported the US as a failure too. The rule is
`> 0`, and China's estimate sits inside the range the other five span. Reporting
a correct coefficient as a failure because of a rule invented alongside it is
exactly the trap this project has hit before.

The rate term in that equation is `r round(coefs$estimate[coefs$equation == "cn_domestic_demand" & coefs$term == "cn_long_rate"], 4)`
with a 90% interval of
[`r round(coefs$ci_low[coefs$equation == "cn_domestic_demand" & coefs$term == "cn_long_rate"], 3)`,
`r round(coefs$ci_high[coefs$equation == "cn_domestic_demand" & coefs$term == "cn_long_rate"], 3)`] —
correctly signed and firmly indistinguishable from zero. Given that 36 of the 80
quarters of `cn_long_rate` are spliced, that is the expected result rather than a
surprise, and it is the strongest argument for trying the
`spread_countries`-without-China variant.

### What the bloc costs: `reu_investment`

```{r reu-investment}
inv_sd <- vapply(c("de", "fr", "it", "us", "reu"), function(cc) {
  x <- stats::window(readRDS(file.path(dir2d, "panel.rds"))[[paste0(cc, "_investment")]],
                     start = c(2005, 1), end = c(2024, 4))
  stats::sd(as.numeric(koma::rate(x)), na.rm = TRUE)
}, numeric(1))
inv_rmse <- insample[grepl("_investment$", insample$equation), ]
kable(
  data.frame(
    entity = names(inv_sd),
    `sd of growth` = round(inv_sd, 2),
    `accelerator (on gdp)` = round(vapply(names(inv_sd), function(cc) {
      r <- coefs[coefs$equation == paste0(cc, "_investment") & coefs$term == paste0(cc, "_gdp"), ]
      if (nrow(r) == 0) NA_real_ else r$estimate[1]
    }, numeric(1)), 3),
    `in-sample RMSE` = round(inv_rmse$rmse[match(paste0(names(inv_sd), "_investment"), inv_rmse$equation)], 2),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "The investment equations. reu is the most volatile series and the worst-fitting equation, despite being an average of seven economies."
)
```

**Averaging seven economies should reduce volatility. Here it raises it**, and
that is the tell. `reu_investment` has a quarterly growth standard deviation of
`r round(inv_sd[["reu"]], 2)`, against `r round(inv_sd[["de"]], 2)`–`r round(inv_sd[["it"]], 2)`
for the four real countries. The cause is the one this project has already
documented for Ireland: `ie_investment` has a growth sd of **43.0** on this
window, against 2.2 for Austria and 3.0 for Portugal, and Ireland carries a
`r pct(bloc_w[["ie"]])` weight in the bloc — so Ireland alone contributes more
to `reu_investment`'s variance than every other member combined. The
Netherlands (sd 14.0) adds most of the rest.

The consequence is that **this one equation is not identified**. Its accelerator
is `r round(coefs$estimate[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 3)`
with a 90% interval of
[`r round(coefs$ci_low[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 2)`,
`r round(coefs$ci_high[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 2)`] —
wrong-signed, but so wide that the sign carries no information — and its own lag
has gone to `r round(coefs$estimate[coefs$equation == "reu_investment" & coefs$term == "reu_investment.L(1)"], 3)`,
i.e. the equation has collapsed into mean-reversion around noise rather than an
accelerator. Its in-sample RMSE is `r round(inv_rmse$rmse[inv_rmse$equation == "reu_investment"], 2)`
against `r round(min(inv_rmse$rmse), 2)`–`r round(max(inv_rmse$rmse[inv_rmse$equation != "reu_investment"]), 2)`
for the others.

There is deliberately no sign rule for the investment accelerator in
`base_sign_rules()`, so nothing flagged this — it surfaced only from computing
the implied demand elasticity above, which came out lowest for `reu` at
`r round(implied$implied[implied$iso2 == "reu"], 3)`. That is the familiar
lesson in a new place: **a neighbouring coefficient moving is the first symptom**,
and the check that would have caught it directly does not exist.

Stages 2b and 2c work around Irish measured GDP with `ireland_proxy`, which
swaps the *foreign-demand basis* for a better-behaved Irish series. That switch
is unavailable here, because the distortion is now inside an **estimated**
equation rather than in an identity's components. Two things would actually
help, neither of them attempted in this report: weighting the bloc by something
less Ireland-heavy than nominal GDP, or excluding Ireland from `reu` entirely
and carrying it as its own entity or in `row_gdp`. Until then, read
`reu_investment` as unidentified, and treat `reu_domestic_demand` — which
inherits it through the identity — as correspondingly soft.

## What this stage is and is not for

**Use it for** anything where the euro area's three largest economies, the US
and China need to interact inside one estimated system, and where `df = `r shape$df``
matters more than country detail: cross-bloc spillovers, a China demand or
policy shock, the transmission of a US rate move into euro-area activity.

**Do not use it for** a statement about Austria, Spain, Ireland or any other
`reu` member — they have no equations here, only a weight. Stage 2b or 2c is
where those live. And do not compare its coefficients directly against stage
2c's: the sample is eighteen quarters shorter, and four of the six entities are
different objects.

**The three things to watch** are stated above rather than buried:
`cn_long_rate` is spliced over 45% of the window, `reu_investment` is not
identified because Ireland dominates the bloc's investment variance, and `reu`'s
identities carry a dispersion term on top of the usual statistical discrepancy. Neither is hidden from the
diagnostics — the splice is measured in `cn_long_rate_series()` and the
dispersion shows up in `identity_consistency()` as the largest error in the
table.


## Eight-quarter forecast

```{r fc-load}
#| include: false
source("reports/_forecast_helpers.R")
library(ggplot2)
fc <- forecast_artefact("stage2d")
fc_entities <- stage2d_countries()
gdp_vars <- fc_vars(fc$paths, fc_entities, "gdp")
price_vars <- fc_vars(fc$paths, fc_entities, "prices")
```

`r fc_window_note(fc)`

Every fan below is the **filtered** predictive interval: koma constrains no
draw to be stationary, so a draw whose quarterly growth exceeds 100% in
absolute value is dropped before the quantiles are taken, exactly as
`score_forecast()` does when it scores a backtest. The solid band is the
25th-75th percentile of what survives, the faint band the 10th-90th, the dark
line their median, and the grey line the realised path where one exists.
`r fc_explosive_note(fc, c(gdp_vars, price_vars))`
A variable whose discarded share climbs steeply is one whose fan
**understates** the model's true uncertainty, so read the tail of a wide fan
as an ordering rather than a number.

**The panels are clipped, and that is worth saying plainly.** Eight quarters
out, the 10-90 interval on *quarterly* growth spans tens of percentage points
in every stage of this project -- an honest number, and one that makes the
median invisible if the axis is scaled to contain it. Each panel is therefore
scaled to its inner band, and the outer band runs to the panel edge rather
than being drawn in full. Nothing is concealed: the table at the end of this
section reports the untruncated bounds.

`r fc_anker_note(fc)`

### GDP

```{r fc-gdp}
#| fig-width: 9
#| fig-height: 5.2
plot_forecast_fan(fc$paths, gdp_vars, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(gdp_vars),
                  title = "Real GDP growth, quarterly %",
                  subtitle = "inner 25-75%, outer 10-90% clipped at the panel edge; grey = realised")
```

```{r fc-gdp-table}
kable(fc_median_table(fc, gdp_vars, space = "rate"),
      row.names = FALSE,
      caption = "Median GDP growth forecast by quarter, percent.")
```

### Prices

```{r fc-prices}
#| fig-width: 9
#| fig-height: 5.2
plot_forecast_fan(fc$paths, price_vars, space = "rate", ncol = 3,
                  labeller = fc_entity_labels(price_vars),
                  title = "Consumer price growth, quarterly %",
                  subtitle = "inner 25-75%, outer 10-90% clipped at the panel edge; grey = realised")
```

```{r fc-prices-table}
kable(fc_median_table(fc, price_vars, space = "rate"),
      row.names = FALSE,
      caption = "Median price growth forecast by quarter, percent.")
```

### Policy rates

A `method = "none"` series passes through koma untouched, so for a policy rate
the forecast **is** the level -- there is no rate-versus-level distinction to
make, and no anchoring question either.

```{r fc-rates}
#| fig-width: 9
#| fig-height: 3.2
rate_vars <- intersect(c("ea_policy_rate", "us_policy_rate", "cn_policy_rate"), unique(fc$paths$variable))
plot_forecast_fan(fc$paths, rate_vars, space = "level", ncol = 3,
                  labeller = fc_full_labels(rate_vars),
                  title = "Policy rates, percent",
                  subtitle = "inner 25-75%, outer 10-90%; grey = realised")
```

```{r fc-rates-table}
kable(forecast_table(fc$paths, rate_vars, space = "level"), row.names = FALSE,
      caption = "Median policy-rate path, percent.")
```

### What was discarded, and what the exogenous path assumes

```{r fc-explosive}
kable(
  transform(forecast_explosive_table(fc$paths, c(gdp_vars, price_vars)),
            h1 = round(h1, 3), h_last = round(h_last, 3), max = round(max, 3)),
  row.names = FALSE,
  col.names = c("variable", "explosive share, h1", paste0("h", fc$horizon), "max"),
  caption = "Share of posterior draws discarded as explosive, by variable."
)
```

```{r fc-uncertainty}
kable(forecast_uncertainty_table(fc$paths, c(gdp_vars, price_vars), space = "rate"),
      row.names = FALSE,
      caption = paste("The untruncated predictive band at horizon", fc$horizon,
                      "-- quarterly growth, percent. This is what the charts clip."))
```

A forecast this long needs the **exogenous** series to reach the end of the
horizon, and they do not: every one of them stops at the fit's own native
forecast end. `extend_forecast_horizon()` carries each one forward on its
trailing four-quarter average growth -- a momentum continuation, not a view on
oil markets or world trade -- and koma would otherwise have silently returned a
shorter horizon than the one requested.

```{r fc-extension}
kable(fc$extension, row.names = FALSE,
      caption = "How each exogenous series was carried past the fit's native forecast end.")
```
