---
title: "Stage 1 equations: koma strings, plain language, and fitted coefficients"
subtitle: "Eleven independent per-country satellite models"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage1_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

panel <- readRDS("_targets/objects/panel")
dates <- readRDS("_targets/objects/estimation_dates")

specs <- stats::setNames(lapply(modelled_countries, function(cc) {
  stage1_spec(cc, expenditure_shares(panel, cc, dates))
}), modelled_countries)

fits <- stats::setNames(
  lapply(modelled_countries, function(cc) readRDS(file.path("data", "cache", "stage1", paste0(cc, ".rds")))),
  modelled_countries
)
```

Each of the eleven economies is its own small `koma` system: six behavioural
equations (consumption, investment, exports, imports, prices, long rate) plus
two accounting identities (GDP, domestic demand), all taking shared/world
variables as exogenous. The United States carries a seventh equation, its
own policy rate. See `CLAUDE.md` and `reports/stage1_diagnostics.qmd` for the
estimation window, acceptance-rate band and tau-tuning that produced the
fitted coefficients below.

## 1. Equations as koma parsed them

```{r koma-strings}
#| results: asis
for (cc in modelled_countries) {
  cat("###", country_names[[cc]], "\n\n")
  ov <- equation_overview(specs[[cc]])
  cat("**Stochastic:**\n\n")
  cat(paste0("- `", ov$koma_string[ov$kind == "stochastic"], "`"), sep = "\n")
  cat("\n\n**Identities:**\n\n")
  cat(paste0("- `", ov$koma_string[ov$kind == "identity"], "`"), sep = "\n")
  cat("\n\n")
}
```

## 2. The same equations, plain language

```{r readable}
#| results: asis
for (cc in modelled_countries) {
  cat("###", country_names[[cc]], "\n\n")
  ov <- equation_overview(specs[[cc]])
  cat("**Behavioural equations:**\n\n")
  cat(paste0("- ", ov$readable[ov$kind == "stochastic"]), sep = "\n")
  cat("\n\n**Identities:**\n\n")
  cat(paste0("- ", ov$readable[ov$kind == "identity"]), sep = "\n")
  cat("\n\n")
}
```

## 3. Fitted coefficients

90% credible intervals (`ci_low`-`ci_high`), from `coefficient_table()`
applied to each country's own fit (`data/cache/stage1/<iso2>.rds`).

```{r coefficients}
#| results: asis
for (cc in modelled_countries) {
  cat("###", country_names[[cc]], "\n\n")
  eq_order <- names(c(specs[[cc]]$stochastic, specs[[cc]]$identities))
  ct <- readable_coefficient_table(fits[[cc]], eq_order)
  print(kable(ct[, c("equation_readable", "term_readable", "estimate", "ci_low", "ci_high")],
             digits = 4, col.names = c("Equation", "Term", "Estimate", "CI low (5%)", "CI high (95%)"),
             caption = paste(country_names[[cc]], "-- fitted coefficients")))
  cat("\n\n")
}
```
