---
title: "Stage 2d equations: the same template, a different world"
subtitle: "Stage 2c's equation forms applied to six entities — a euro-area bloc, China, and the one block that had to change"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage2d_equations.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)
source("reports/_equation_overview_helpers.R")

dir2d <- file.path("data", "cache", "stage2d")
spec <- readRDS(file.path(dir2d, "spec.rds"))
sys_eq <- readRDS(file.path(dir2d, "sys_eq.rds"))
shape <- readRDS(file.path(dir2d, "shape.rds"))
coefs <- readRDS(file.path(dir2d, "coefs.rds"))
tuned <- readRDS(file.path(dir2d, "tuned.rds"))
lw <- readRDS(file.path(dir2d, "linkage_weights.rds"))
W <- readRDS(file.path(dir2d, "trade_weights.rds"))

f2c <- readRDS(file.path("data", "cache", "stage2c", "noic_full_tuned.rds"))
spec2c <- f2c$spec

countries <- stage2d_countries()
eq_order <- names(c(spec$stochastic, spec$identities))
ov <- equation_overview(spec)
ct <- readable_coefficient_table(tuned$fit, eq_order)

tg <- template_groups(spec, countries)
ig <- identity_template_groups(spec, countries)
form <- function(concept) {
  x <- tg$form[tg$concept == concept]
  if (length(x) == 0) NA_character_ else x[1]
}
istruct <- function(concept) {
  rows <- ig[ig$concept == concept, ]
  if (nrow(rows) == 0) return(NA_character_)
  rows$structure[which.max(rows$n_countries)]
}
ea_rule <- koma_string_stochastic("ea_policy_rate", spec$stochastic[["ea_policy_rate"]])

n_stoch <- nrow(ov[ov$kind == "stochastic", ])
n_ident <- nrow(ov[ov$kind == "identity", ])

# --- what the re-partition did, relative to stage 2c ---------------------
concepts_of <- function(sp) unique(sub("^[a-z]{2,3}_", "", names(sp$stochastic)))
same_forms <- intersect(concepts_of(spec), concepts_of(spec2c))

wrange <- function(concept, component) {
  r <- identity_weight_range(spec, countries, concept, component)
  sprintf("%.3f to %.3f (median %.3f)", r$min, r$max, r$median)
}
merged <- stage2d_config()$merged_demand_countries
```

Stage 2d is stage 2c's **equation template on a different partition of the
world**: `r n_stoch` stochastic equations and `r n_ident` identities in one
`koma::system_of_equations()` call, `k = `r shape$k`` against
`T = `r shape$t_obs``, `df = `r shape$df``.

Nothing about the *forms* changed. The speed-limit Phillips curve, foreign
demand over partner **imports**, the sovereign spread in place of the long-rate
level, and the intertemporal-substitution term in consumption are all exactly
as `reports/stage2c_equations.qmd` sets them out, and that document remains the
reference for what each one claims economically. This one covers what is
different: **who** the equations are written for, and the single block that had
to depart from the template because the data does not exist.

| | stage 2c | stage 2d |
|---|---|---|
| entities | 11 countries | `de`, `fr`, `it`, `us`, `cn`, `reu` |
| stochastic equations | 68 | `r n_stoch` |
| identities | 46 | `r n_ident` |
| policy rules | `ea_`, `us_` | those plus `cn_policy_rate` |
| `k` / `T` / `df` | 76 / 98 / 22 | **`r shape$k`** / `r shape$t_obs` / **`r shape$df`** |

The two conventions from stage 2c still apply and are still easy to trip over:
everything is estimated in **growth-rate space**, so a coefficient on a
level/index series is an elasticity and an identity is a statement about growth
rates; and **rate-type series are the exception** — `long_rate`, `spread` and
the three policy rates are `series_type = "rate"`, `method = "none"`, so koma
leaves them alone and their coefficients are percentage points per percentage
point.

## 1. The entity set

Four of the six are ordinary countries. Two are not, and each carries
consequences for how its equations should be read.

```{r entities}
kable(data.frame(
  entity = c("de", "fr", "it", "us", "cn", "reu"),
  what = c("Germany", "France", "Italy", "United States", "China",
           "Rest of the euro area: AT, BE, ES, GR, IE, NL, PT"),
  equations = vapply(c("de", "fr", "it", "us", "cn", "reu"), function(cc)
    sum(grepl(paste0("^", cc, "_"), names(spec$stochastic))), integer(1)),
  identities = vapply(c("de", "fr", "it", "us", "cn", "reu"), function(cc)
    sum(grepl(paste0("^", cc, "_"), names(spec$identities))), integer(1)),
  note = c("", "", "", "own policy rule", "own policy rule; merged demand equation",
           "bloc pseudo-country")
), row.names = FALSE,
col.names = c("Entity", "What it is", "Stochastic", "Identities", "Note"),
caption = "The six entities. reu occupies a country slot rather than being an ea_-style aggregate: it has a series for every concept, its own block, and its own row and column in the trade-weight matrix.")
```

**`reu` is a pseudo-country, not an aggregate.** `ea_gdp` and `ea_prices` are
identities computed *alongside* the countries they sum; `reu` **replaces** its
seven members, which contribute no equations at all. Its series are built by
`aggregate_bloc_panel()` — chained in growth space for volumes and indices,
weighted-averaged in levels for rates — and its identity weights are its
members' own expenditure shares averaged, not ratios of its base-100 indices.
That distinction matters: taking the naive route gives `reu_exports` a weight
of 1.337 instead of 0.598, and koma would have enforced it in silence.

**`cn` is the one entity whose block departs from the template**, and it
departs for a data reason. China publishes no quarterly consumption or
investment anywhere in the public record — verified absent from OECD Quarterly
National Accounts, the OECD Economic Outlook, IMF IFS, World Bank GEM, FRED and
the NBS itself — so those two equations collapse into a single estimated
`cn_domestic_demand`, and the `cn_domestic_demand` **identity is dropped**
because the variable is now estimated rather than defined. See §3.

## 2. The equation template

### 2.1 Behavioural equations

```{r template-summary}
kable(data.frame(
  concept = c("Consumption", "Investment", "Merged domestic demand", "Exports",
              "Imports", "Prices", "Sovereign spread", "Policy rules"),
  form = c(form("consumption"), form("investment"), form("domestic_demand"),
           form("exports"), form("imports"), form("prices"), form("spread"),
           ea_rule),
  applies_to = c("de, fr, it, us, reu", "de, fr, it, us, reu", "cn only",
                 "all 6", "all 6", "all 6", "all 6",
                 "ea_policy_rate, us_policy_rate, cn_policy_rate")
), row.names = FALSE, col.names = c("Equation", "General form", "Applies to"),
caption = "Every behavioural form in the system. The first three are the same equation family: cn merges the first two.")
```

Each of these is the stage-2c form unchanged; `reports/stage2c_equations.qmd`
§1.1 explains what each one claims and what the sign priors are. Two things are
worth restating here because the re-partition changes their reading.

**The `reu` equations are estimated on an aggregate, and one of them does not
survive it.** `reu_investment`'s accelerator comes out at
`r round(coefs$estimate[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 3)`
with a 90% interval of
[`r round(coefs$ci_low[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 2)`,
`r round(coefs$ci_high[coefs$equation == "reu_investment" & coefs$term == "reu_gdp"], 2)`]
— wrong-signed but far too wide for the sign to mean anything — and its own lag
has gone to
`r round(coefs$estimate[coefs$equation == "reu_investment" & coefs$term == "reu_investment.L(1)"], 3)`,
i.e. mean reversion around noise rather than an accelerator. Ireland's
investment growth has a standard deviation of 43.0 on this window against 2.2
for Austria, and Ireland carries 10.9% of the bloc weight, so it contributes
more variance than every other member combined. Read that equation as
unidentified, and `reu_domestic_demand` — which inherits it through the
identity — as correspondingly soft. `reports/stage2d_regional_core.qmd` sets
out the numbers.

**There are three policy rules, not two.** China is outside both currency
unions, so without its own rule the `cn_long_rate` identity would price Chinese
sovereign debt off the ECB's refinancing rate. `stage2_policy_rule_countries()`
resolves the country set in one place precisely so that `country_block()`,
`policy_rate_map()` and the constructed `<iso2>_spread` series cannot disagree
about which rate a given country's spread is measured against.

```{r policy-rules}
rules <- grep("_policy_rate$", names(spec$stochastic), value = TRUE)
kable(data.frame(
  rule = rules,
  form = vapply(rules, function(r) koma_string_stochastic(r, spec$stochastic[[r]]), character(1))
), row.names = FALSE, col.names = c("Rule", "koma string"),
caption = "The three Taylor-type rules. The euro-area rule responds to area-wide aggregates; the US and Chinese rules to their own.")
```

### 2.2 Identities

```{r identity-summary}
kable(data.frame(
  identity = c("GDP", "Domestic demand", "Foreign demand", "Long rate",
               "Euro-area GDP", "Euro-area prices"),
  form = c(istruct("gdp"), istruct("domestic_demand"),
           "cc_foreign_demand == sum_j w_j*<partner j>_imports + w_row*row_gdp",
           istruct("long_rate"),
           "ea_gdp == sum_i g_i*<member i>_gdp",
           "ea_prices == sum_i g_i*<member i>_prices"),
  count = c(
    sum(grepl("_gdp$", names(spec$identities)) & !grepl("^ea_", names(spec$identities))),
    sum(grepl("_domestic_demand$", names(spec$identities))),
    sum(grepl("_foreign_demand$", names(spec$identities))),
    sum(grepl("_long_rate$", names(spec$identities))), 1L, 1L)
), row.names = FALSE, col.names = c("Identity", "General form", "Count"),
caption = "The identity set. Note there are five domestic-demand identities, not six: China's is dropped because the variable is estimated.")
```

**`ea_gdp` and `ea_prices` now aggregate four entities, one of which is the
bloc.** Handing `stage2_linkage_weights()` the eleven-country `W_gdp` alongside
the stage-2d entity set would have matched only Germany, France and Italy —
the aggregation identities would then have described three economies while
claiming to describe the currency union, with no error anywhere.
`collapse_gdp_weights()` folds the seven members into `reu` first.

```{r ea-weights}
kable(data.frame(member = toupper(names(lw$ea)), weight = as.numeric(lw$ea)),
      row.names = FALSE, digits = 3,
      caption = "The ea_gdp / ea_prices weights: nominal GDP shares over the four euro-area entities, the bloc counted once.")
```

### 2.3 The template is verified, not asserted

Each entity's equation is rewritten with its own prefix replaced by `cc`, and
identical strings are grouped. A row per concept means the form is unanimous;
more than one row means something deviates, and here exactly one thing does.

```{r template-verify}
kable(tg[, c("concept", "n_countries", "countries")], row.names = FALSE,
      col.names = c("Concept", "Entities sharing this form", "Which"),
      caption = "Stochastic equations grouped by entity-generic form.")
```

The `consumption` and `investment` rows cover five entities rather than six,
and `domestic_demand` appears as a stochastic form for one — that is China's
merged block, and it is the only departure from the template in the system.

```{r identity-verify}
iv <- ig[ig$concept != "foreign_demand", c("concept", "n_countries", "countries")]
kable(iv, row.names = FALSE,
      col.names = c("Identity", "Entities sharing this structure", "Which"),
      caption = "Identities grouped by entity-generic component structure (weights ignored, signs kept). foreign_demand is excluded: its partner set is country-specific by construction, so six distinct structures is the expected result.")
```

The `long_rate` identity splits three ways rather than one, and correctly so:
each entity's spread is taken against the policy rate it actually faces —
`ea_policy_rate` for the four euro-area entities, `us_policy_rate` for the US,
`cn_policy_rate` for China.

### 2.4 What the identity weights actually are

```{r weight-ranges}
kable(data.frame(
  identity = c("GDP", "GDP", "GDP", "Domestic demand", "Domestic demand"),
  weight_on = c("domestic demand", "exports", "imports", "consumption", "investment"),
  range = c(wrange("gdp", "domestic_demand"), wrange("gdp", "exports"),
            wrange("gdp", "imports"), wrange("domestic_demand", "consumption"),
            wrange("domestic_demand", "investment"))
), row.names = FALSE, col.names = c("Identity", "Weight on", "Range across the entities"),
caption = "Fixed expenditure shares over the stage-2d estimation window. The domestic-demand rows cover the five entities that have the split.")
```

```{r fd-weights}
fd <- do.call(rbind, lapply(countries, function(cc) {
  w <- spec$identities[[country_var(cc, "foreign_demand")]]
  bilateral <- unlist(w)[names(w) != "row_gdp"]
  data.frame(entity = cc, partners = length(bilateral),
             row_gdp_weight = unname(w[["row_gdp"]]),
             largest_partner = names(bilateral)[which.max(bilateral)],
             largest_weight = max(bilateral), stringsAsFactors = FALSE)
}))
kable(fd, row.names = FALSE, digits = 3,
      caption = "Each entity's foreign-demand index: modelled partners, how much rides on the unmodelled rest of the world, and its single largest bilateral link.")
```

**The US and Chinese rows are observed here, not approximated**, and that is a
change from every earlier stage. ECB WTS computes trade weights *for*
euro-area reporters only, so stages 1–3 build the US row from the reciprocal of
each euro-area country's own weight on the US, renormalised — which forces the
modelled partners to account for the reporter's entire trade and gives the US a
4.6% rest-of-world weight in `data/raw/W_trade.csv`. At six entities that error
would have dominated rather than distorted: applied to China it would have put
56% of Chinese trade inside the euro area. Both rows come from observed IMF
Direction of Trade Statistics instead, which is why the table above shows
`r sprintf("%.1f%%", 100 * fd$row_gdp_weight[fd$entity == "us"])` and
`r sprintf("%.1f%%", 100 * fd$row_gdp_weight[fd$entity == "cn"])` on
`row_gdp` — the trade those two economies genuinely do outside this model.

```{r trade-source}
kable(data.frame(reporter = names(attr(W, "source")), source = unname(attr(W, "source"))),
      row.names = FALSE, col.names = c("Reporter", "Weight source"),
      caption = "Where each row of the stage-2d trade-weight matrix comes from. DOTS is goods-only and a plain trade share; ECB WTS covers services and is double-weighted, so a `direct` row and a `dots` row are not quite the same statistic.")
```

## 3. China: the one block that changed

```{r cn-block}
cn_eqs <- grep("^cn_", names(spec$stochastic), value = TRUE)
cn_ids <- grep("^cn_", names(spec$identities), value = TRUE)
kable(rbind(
  data.frame(kind = "stochastic", equation = cn_eqs,
             koma_string = vapply(cn_eqs, function(e)
               koma_string_stochastic(e, spec$stochastic[[e]]), character(1)),
             stringsAsFactors = FALSE),
  data.frame(kind = "identity", equation = cn_ids,
             koma_string = vapply(cn_ids, function(e)
               koma_string_identity(e, spec$identities[[e]]), character(1)),
             stringsAsFactors = FALSE)
), row.names = FALSE, col.names = c("Kind", "Equation", "koma string"),
caption = "China's complete block. Six stochastic equations and three identities -- one fewer of each than a full country, because consumption and investment merge and the domestic-demand identity goes with them.")
```

The merged equation keeps the rate term **unconditionally**, and that is
deliberate: `cn_long_rate` was investment's regressor as well as consumption's,
so dropping it under the merge would have removed China's only monetary channel
rather than one of two.

**The bound on that equation is `> 0`, not `(0, 1)`.** The obvious rule for a
demand equation is the marginal propensity to consume's, and China's estimate
of
`r round(coefs$estimate[coefs$equation == "cn_domestic_demand" & coefs$term == "cn_gdp"], 3)`
fails it. But an MPC below one is a claim about *consumption*; total absorption
also contains investment, which is far more cyclical than output. The five
entities that do have the split imply domestic-demand elasticities of 0.43
(`reu`), 0.63 (`de`), 0.96 (`fr`), 0.98 (`it`) and **1.26 (`us`)** — so a
`(0, 1)` bound would have reported the United States as a failure too.
`sign_checks(merged_demand = TRUE)` restates the rule rather than leaving a
permanent red mark against an equation the system deliberately does not have.

## 4. Every equation, as koma parsed them

```{r koma-stoch}
kable(ov[ov$kind == "stochastic", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = paste0("All ", n_stoch, " stochastic equations."))
```

```{r koma-ident}
kable(ov[ov$kind == "identity", c("equation", "koma_string")],
      row.names = FALSE, col.names = c("Equation", "koma string"),
      caption = paste0("All ", n_ident, " identities."))
```

## 5. Every equation, plain language

```{r readable-stoch}
kable(ov[ov$kind == "stochastic", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = paste0("All ", n_stoch, " stochastic equations."))
```

```{r readable-ident}
kable(ov[ov$kind == "identity", c("equation", "readable")],
      row.names = FALSE, col.names = c("Equation", "Plain language"),
      caption = paste0("All ", n_ident, " identities."))
```

## 6. Every fitted coefficient

90% credible intervals, from `coefficient_table()` applied to the cached,
tau-tuned fit. Identities carry no estimated coefficients — their weights are
fixed and are listed in full in §4 and §5.

```{r coef-all}
#| results: asis
for (eq in intersect(eq_order, unique(ct$equation))) {
  sub <- ct[ct$equation == eq, ]
  print(kable(sub[, c("term_readable", "estimate", "ci_low", "ci_high")],
              digits = 4, row.names = FALSE,
              col.names = c("Term", "Estimate", "CI low (5%)", "CI high (95%)"),
              caption = sub$equation_readable[1]))
  cat("\n\n")
}
```
