---
title: "Stage 3a: the German labour market and disaggregated prices"
subtitle: "A wage–price loop and a second cross-country channel — what the data allowed, what the identification assessment found, and which coefficient came out wrong"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3a_labour_prices.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir3a <- file.path("data", "cache", "stage3a")
fit_a <- readRDS(file.path(dir3a, "fit_a_tuned.rds"))
untuned <- readRDS(file.path(dir3a, "fit_a.rds"))
hicp <- readRDS(file.path(dir3a, "hicp_weights.rds"))
panel3a <- readRDS(file.path(dir3a, "panel_stage2.rds"))
base_panel <- readRDS(file.path(dir3a, "panel.rds"))
stage2b <- readRDS(file.path("data", "cache", "stage2b", "full_tuned.rds"))

sys_eq <- fit_a$sys_eq
coefs <- coefficient_table(fit_a$fit)
signs <- sign_checks(coefs, "de", labour = TRUE)
acceptance <- check_acceptance_rates(fit_a$fit)
lag_stab <- check_lag_stability(coefs)
loop <- wage_price_loop_gain(fit_a$fit, "de", hicp$weights[["nonenergy_prices"]])
collin <- block_collinearity(panel3a, sys_eq, stage2b_dates())

n_stoch <- length(sys_eq$stochastic_equations)
n_id <- length(sys_eq$identities)
k <- length(sys_eq$total_exogenous_variables)
Tn <- estimation_length(panel3a, stage2b_dates())
```

## 1. What was built, and why it looks like this

Stage 2b linked eleven economies through **one** channel: trade-weighted
foreign demand over partners' GDP. Stage 3a adds a second — price
competitiveness — and gives Germany a labour block so that wages, unit
labour costs and prices form a loop rather than a chain.

Germany's block, as estimated (`covid_*` dummies elided):

```
de_consumption      ~ de_gdp + de_real_income + de_consumption.L(1)
de_investment       ~ de_gdp + de_long_rate + de_investment.L(1)
de_exports          ~ de_foreign_demand + de_export_prices + de_foreign_prices + de_exports.L(1)
de_imports          ~ de_domestic_demand + de_import_prices + de_imports.L(1)
de_long_rate        ~ de_prices + ea_policy_rate + de_gdp + de_long_rate.L(1)
de_employment       ~ de_gdp + de_employment.L(1)
de_unemployment     ~ de_gdp + de_unemployment.L(1)
de_wages            ~ de_unemployment + de_prices + de_wages.L(1)
de_nonenergy_prices ~ de_ulc + de_import_prices + de_unemployment + de_nonenergy_prices.L(1)
de_energy_prices    ~ oil_price + eur_usd + de_energy_prices.L(1)
de_import_prices    ~ eur_usd + oil_price + de_foreign_prices + de_import_prices.L(1)
de_export_prices    ~ de_ulc + de_foreign_prices + de_export_prices.L(1)

de_gdp             == <expenditure shares>
de_domestic_demand == <expenditure shares>
de_foreign_demand  == <trade-weighted partners' gdp + row_gdp>
de_productivity    == de_gdp - de_employment
de_ulc             == de_wages - de_productivity
de_prices          == 0.891*de_nonenergy_prices + 0.109*de_energy_prices
de_real_income     == de_wages + de_employment - de_prices
de_foreign_prices  == <trade-weighted partners' export_prices>   [exogenous in phase A]
```

Three things in that listing differ from the block as originally specified.
Each is a data finding, not a preference.

### 1.1 `de_wages` is a wage *rate*, because otherwise the identities are wrong

EA-MD/QD publishes `WS`, "Wages and salaries" in current prices — a
whole-economy wage **bill**, not a wage rate. Used directly it breaks the
block in two places: `real_income == wages + employment - prices` would
double-count employment (the bill is already rate × employment), and a wage
Phillips curve run on the bill would largely re-estimate Okun's law.

`de_wages` is therefore `WS / TEMP`, labour income per worker. That makes
the whole block internally exact: `ulc == wages - productivity` telescopes
through `productivity == gdp - employment` to `WS - GDP`, nominal wage bill
over real output — the textbook definition of unit labour costs. This
matters because EA-MD/QD carries **no whole-economy ULC series at all**,
only seven sectoral ones, so ULC has to be derived either way; deriving it
this way makes it consistent with the rest of the block by construction.

The stated assumption: `WS` covers employees while `TEMP` counts total
employment, so this imputes employee compensation to the self-employed.
That is the standard construction for compensation per worker, and it is
preferred to `WS / EMP` because mixing employment concepts between `wages`
and `employment` would break the exactness above.

### 1.2 The price identity splits energy from non-energy, not core from energy

The specified identity was `prices == 0.85*core + 0.15*energy`. It cannot be
made exact: HICP "core" excludes energy **and** food, so core plus energy
covers only about 81% of the German basket and the missing food weight would
have to be renormalised away — silently attributing food inflation to core,
inside the very loop the block exists to model.

Eurostat's `NRG` and `TOT_X_NRG` do partition the basket. Verified against
`prc_hicp_inw` for every year from 1996 to 2025, they sum to `CP00` = 1000
exactly, with no residual. The identity uses those, and the weights come
from the data rather than the example:

```{r hicp-weights}
kable(data.frame(
  component = c("non-energy (TOT_X_NRG)", "energy (NRG)"),
  weight = round(as.numeric(hicp$weights[c("nonenergy_prices", "energy_prices")]), 4),
  specified = c(0.85, 0.15)
), caption = sprintf(
  "HICP weights for Germany, averaged over %d-%d (Eurostat prc_hicp_inw). Exact partition: %s.",
  min(hicp$years), max(hicp$years), hicp$exact
))
```

The energy weight is **`r round(hicp$weights[["energy_prices"]], 3)`**, not
the 0.15 assumed — a 4-point difference in the weight on the most volatile
component of the basket.

Weights are re-based annually, so a fixed identity weight is an
approximation. Over this window the German energy weight ranges from 88.4 to
125.5 per mille; the largest single-year departure from the average used
here is **`r round(hicp$max_deviation, 4)`** in weight units. That is the
size of the approximation, stated rather than buried.

Because "core" no longer appears in the identity, the block uses a new
concept `nonenergy_prices` rather than redefining the existing
`core_prices`. Nothing was silently repurposed; `de_core_prices` is still in
the panel, still `HICPNEF`, and is now simply unused by any equation.

### 1.3 The price variables feed back into trade volumes

As originally specified, `de_export_prices` appeared on no other equation's
right-hand side inside a German-only system — it would have been estimated
and then transmitted nothing, the same terminal-variable pathology
`CLAUDE.md` already records for `long_rate`. Exports therefore take a
relative-price pair (own and competitors'), and imports take their own
price. Section 5.1 is about what that asymmetry did to the imports equation.

## 2. Degrees of freedom: what fits and what does not

```{r df-table}
kable(data.frame(
  system = c("stage 2b", "stage 3a phase A", "stage 3a phase B", "full block, all 11"),
  stochastic = c(68, 74, 84, 134),
  identities = c(35, 39, 40, 90),
  k = c(76, 83, 92, 142),
  df = c(98 - 76, 98 - 83, 98 - 92, 98 - 142),
  status = c("estimated", "estimated (this report)", "planned", "impossible")
), caption = "k = 1 + one lag column per stochastic equation + exogenous variables; T = 98.")
```

The last row is the important one. Rolling this block out to all eleven
countries needs 142 predetermined columns against 98 observations — not
slow, not badly conditioned, but **arithmetically impossible**: koma
projects every equation on the full `k`-column `x_matrix` on every draw and
draws `Omega` from `riwish(T - k, .)`. Even three countries would leave
`df ≈ 4`. The block is a German pilot in the strict sense that the
architecture cannot support it everywhere.

This report covers **phase A** (`k = `r k``, `T = `r Tn``, `df = `r Tn - k``),
where `de_foreign_prices` is an exogenous series constructed from partners'
export-price data. Phase B makes it endogenous by giving the other ten
countries a one-equation export-price satellite, at `df = 6`.

## 3. Identification

`koma::model_identification()` checks the classical order and rank
conditions per stochastic equation, symbolically, and **aborts** on failure.
`stage2_preflight()` repeats it across ten seeds, because the function fills
free coefficients with `rnorm` draws and a single pass proves little.

```{r preflight}
kable(fit_a$preflight, row.names = FALSE, caption = "stage2_preflight(), phase A.")
```

**No equation is under-identified.** The order condition requires each
equation to exclude at least `n_endogenous - 1` = `r length(sys_eq$endogenous_variables) - 1`
variables; the stacked `[Gamma; B]` has
`r length(sys_eq$endogenous_variables) + length(sys_eq$predetermined_variables)`
rows and no equation includes more than five regressors, so every equation
excludes on the order of 200. Adding endogenous variables raises the bar for
every equation but adds rows faster than it raises it. Rank is the binding
condition and it passes on all ten seeds.

Two caveats on that result, neither of which the check itself would show:

- koma's `gamma_vectorization()` locates parameters with an **unanchored**
  `grep`, so `"gamma1_6"` also matches `"gamma1_60"`. With
  `r length(sys_eq$endogenous_variables)` endogenous variables, index
  collisions are possible. Treat a pass as necessary, not sufficient.
- Identification is symbolic: it sees which variables are *excluded*, never
  how correlated the *included* ones are. That is a different question, and
  it is the one asked next.

### 3.1 Collinearity: the `ulc` / `unemployment` question, answered

The specific concern was that `de_nonenergy_prices` takes both `de_ulc` and
`de_unemployment`, while `de_ulc == de_wages - de_productivity` and
`de_wages` responds to unemployment — so ULC is partly a function of
unemployment by construction. That would be weak identification (imprecise,
trading-off coefficients), not rank failure.

Measured directly from the panel, in the rate space koma estimates in:

```{r collinearity}
c2 <- collin[!grepl("^covid_", collin$term), ]
kable(head(c2[, c("equation", "term", "vif", "max_abs_cor", "worst_partner")], 12),
      row.names = FALSE, digits = 3,
      caption = "Variance inflation factors, worst first. Flagged above 10.")
```

**The concern does not materialise.** `de_ulc` has a VIF of
`r round(c2$vif[c2$equation == "de_nonenergy_prices" & c2$term == "de_ulc"], 2)`
and `de_unemployment` `r round(c2$vif[c2$equation == "de_nonenergy_prices" & c2$term == "de_unemployment"], 2)`
in the price equation, with a correlation of
`r round(c2$max_abs_cor[c2$equation == "de_nonenergy_prices" & c2$term == "de_ulc"], 3)`
between them. Nothing in the block exceeds a VIF of
`r round(max(c2$vif), 2)`, well under the conventional threshold of 10.

The weakest spot is elsewhere: `de_export_prices` and `de_foreign_prices`
in the exports equation correlate at
`r round(c2$max_abs_cor[c2$equation == "de_exports" & c2$term == "de_export_prices"], 3)`.
Both are price indices in a common inflation environment, so separating an
own-price from a competitor-price effect is the least well-identified thing
the block asks for. Their estimated coefficients are nearly equal and
opposite, which is what a relative price should look like, but the credible
intervals are correspondingly wide.

Note also that had `de_wages` and `de_ulc` *both* entered the price equation
— as the original sketch suggested — that would be algebraically identical
to including `de_wages` and `de_productivity`. Legitimate and full rank, but
it should then be written that way.

## 4. Sampler behaviour

### 4.1 Acceptance rates

```{r acceptance}
new_eqs <- c("de_employment", "de_unemployment", "de_wages", "de_nonenergy_prices",
             "de_energy_prices", "de_import_prices", "de_export_prices")
kable(acceptance[acceptance$equation %in% new_eqs, ], row.names = FALSE, digits = 3,
      caption = "The seven new behavioural equations, after tau tuning.")
```

`de_energy_prices` and `de_import_prices` show `has_mh_step = FALSE` and an
`NA` rate. That is correct, not a failure: in phase A every regressor in
both equations (`oil_price`, `eur_usd`, and the exogenous
`de_foreign_prices`) is exogenous, so neither has a contemporaneous
endogenous regressor, koma gives them no Metropolis step, and
`check_acceptance_rates()` never flags them. **This changes in phase B**,
when `de_foreign_prices` becomes endogenous and `de_import_prices` acquires
a Metropolis step for the first time.

An untuned stage-3a fit flagged
`r sum(check_acceptance_rates(untuned$fit)$flagged)` equations against a
tuned stage-2b baseline of zero — but only three of those were new
equations, and the rest were pre-existing equations pushed just over 60% by
the system-wide change in the residual covariance. Acceptance rates are not
a per-equation property in a simultaneous system: adding equations anywhere
moves every equation's rate. Re-tuning from the same doubling rule stage 2b
used (`tune_tau_system()`) is what the table above reflects; attributing the
untuned flags to the labour block would have been wrong.

### 4.2 The wage–price loop is not explosive

Stage 3a introduces a contemporaneous feedback loop that no earlier stage
had: `wages → ulc → nonenergy_prices → prices → wages`. Its round-trip gain
is `beta(nonenergy ← ulc) × w_nonenergy × beta(wages ← prices)`, and a draw
with gain at or above 1 is one where a wage rise more than pays for itself
through prices. koma constrains no draw to be stationary, so what matters is
the share of the posterior above 1, not the average.

```{r loop-gain}
kable(data.frame(
  statistic = c("mean", "median", "5th pct", "95th pct", "share |gain| >= 1"),
  value = c(loop$mean, loop$median, loop$q05, loop$q95, loop$share_ge_1)
), digits = 5, caption = "Posterior distribution of the wage-price loop gain.")
```

The gain is centred near zero and **no draw** reaches 1. The loop exists but
is weak, and it does not add to the explosive-draw problem documented in
`R/spillovers.R`.

### 4.3 Own-lag persistence

```{r lag-stability}
kable(head(lag_stab, 8), row.names = FALSE, digits = 3,
      caption = "Most persistent own lags. Flagged at |coefficient| >= 1.")
```

Nothing is flagged, but `de_unemployment` sits **exactly at the boundary**:
its own-lag coefficient is
`r round(lag_stab$estimate[lag_stab$equation == "de_unemployment"], 4)`
with a credible interval reaching
`r round(lag_stab$ci_high[lag_stab$equation == "de_unemployment"], 3)`.
That is a unit root in all but name: the equation is a random walk with a
GDP term, so an unemployment shock never dies out. For a rate modelled in
levels (`series_type = "rate"`, `method = "none"`) this is common and not by
itself wrong — the pre-existing `ea_policy_rate` and several `long_rate`
equations sit at 0.96-0.98 for the same reason — but it does mean the Okun
channel in this fit transmits GDP to a *permanently* shifted unemployment
level, and anything reading unemployment at long horizons should treat that
as a modelling artefact rather than a finding.

### 4.4 Precision: the block is estimable, but not sharply

```{r new-coefs}
new_terms <- coefs[coefs$equation %in% new_eqs & !grepl("^covid_|^constant$", coefs$term), ]
new_terms$spans_zero <- new_terms$ci_low < 0 & new_terms$ci_high > 0
kable(new_terms, row.names = FALSE, digits = 3,
      caption = "Every coefficient in the seven new equations.")
```

`r sum(new_terms$spans_zero)` of `r nrow(new_terms)` coefficients in the new
equations have credible intervals spanning zero. The signs are right, but at
`df = `r Tn - k`` the magnitudes are not sharply identified — most visibly
the wage-indexation term `de_wages ← de_prices`
(`r round(coefs$estimate[coefs$equation == "de_wages" & coefs$term == "de_prices"], 3)`,
interval
[`r round(coefs$ci_low[coefs$equation == "de_wages" & coefs$term == "de_prices"], 2)`,
`r round(coefs$ci_high[coefs$equation == "de_wages" & coefs$term == "de_prices"], 2)`]),
which is the return leg of the wage-price loop and is why the loop gain in
4.2 is so close to zero. The loop is present in the structure and weak in
the estimates; do not read the near-zero gain as evidence that wage-price
feedback is *absent* in the German economy, only that this window and this
specification do not pin it down.

## 5. Coefficient plausibility

```{r signs}
kable(signs, row.names = FALSE, digits = 4,
      caption = "Sign checks. Posterior means only; a check that could not be evaluated counts as a failure.")
```

`r sum(signs$ok)` of `r nrow(signs)` pass. The block's central economic
claims all hold:

- **The wage Phillips curve has the right sign.** `de_wages ← de_unemployment`
  = `r round(signs$estimate[signs$check == "wage_phillips_curve_negative"], 4)`.
  This was the designated red flag, and it is not raised.
- The price Phillips curve is also negative, ULC pass-through into
  non-energy prices is positive, and Okun's law holds in both the employment
  and unemployment equations.
- Both euro-appreciation rules pass with the correct (negative) sign, which
  is worth noting because `eur_usd` is quoted USD per EUR — a rise is a euro
  appreciation, and the intuitive sign is backwards.
- Exports fall in their own price and rise in competitors', nearly
  symmetrically, which is what makes the new competitiveness channel behave
  like a relative price rather than a price level.

### 5.1 Implausible result: imports rise in their own price

```{r imports-fail}
kable(coefs[coefs$equation == "de_imports" & !grepl("^covid_", coefs$term), ],
      row.names = FALSE, digits = 3,
      caption = "The imports equation.")
```

`de_imports ← de_import_prices` is
**`r round(signs$estimate[signs$check == "imports_fall_in_own_price"], 3)`**,
positive, when import demand should fall as imports get more expensive. The
credible interval excludes zero, so this is not noise that can be waved
through — it is a robustly wrong-signed coefficient.

**Diagnosis: a specification asymmetry introduced with the relative-price
terms, not a data or estimation problem.** Exports received a *pair* —
`de_export_prices` and `de_foreign_prices` — so the equation can form a
genuine relative price, and the estimates bear that out: they come back
nearly equal and opposite. Imports received only `de_import_prices`, with no
domestic-price counterpart. There is nothing in that equation for the import
price to be relative *to*, so the coefficient is not measuring a
relative-price elasticity at all. What it can pick up instead is the common
component: quarters when imported goods are expensive are quarters when
world demand is strong, and German import volumes are high for reasons the
equation does not otherwise contain.

The symmetric fix costs nothing in degrees of freedom, since `de_prices` is
already endogenous:

```
de_imports ~ de_domestic_demand + de_import_prices + de_prices + de_imports.L(1)
```

This is a departure from the block as specified, so it is reported here
rather than applied silently. It is the natural completion of the
relative-price treatment that was approved for exports, and it is the first
thing to try before reading anything into the import channel.

Until then: **the import-price elasticity in this fit should not be used**,
and any spillover routed through German import demand inherits the problem.
The export side, which does have a relative-price pair, is unaffected.

## 6. Verification

Every constructed identity reproduces its own definition in rate space to
machine precision — the check that catches a silently corrupted identity,
since koma has no identity-consistency check of its own:

```{r identity-exactness}
# Compare in RATE space, using the ts koma::rate() itself returns: it drops
# the first observation and any trailing NA, so re-dating it by hand shifts a
# series by a quarter and makes an exact identity look broken.
maxdiff <- function(lhs, rhs, w) {
  m <- stats::na.omit(do.call(cbind, lapply(c(lhs, rhs), function(n) koma::rate(panel3a[[n]]))))
  max(abs(m[, 1] - as.numeric(m[, -1, drop = FALSE] %*% w)))
}
wts <- hicp$weights
fpw <- foreign_price_weights(fit_a$linkage_weights, "de")
kable(data.frame(
  identity = c("productivity == gdp - employment", "ulc == wages - productivity",
               "prices == w1*nonenergy + w2*energy", "real_income == wages + employment - prices",
               "foreign_prices == trade-weighted partners"),
  max_abs_deviation = c(
    maxdiff("de_productivity", c("de_gdp", "de_employment"), c(1, -1)),
    maxdiff("de_ulc", c("de_wages", "de_productivity"), c(1, -1)),
    maxdiff("de_prices", c("de_nonenergy_prices", "de_energy_prices"),
            c(wts[["nonenergy_prices"]], wts[["energy_prices"]])),
    maxdiff("de_real_income", c("de_wages", "de_employment", "de_prices"), c(1, 1, -1)),
    maxdiff("de_foreign_prices", names(fpw), as.numeric(fpw))
  )
), caption = "Identity reproduction in rate space.")
```

`de_prices` is built from the sub-indices rather than taken from observed
headline HICP, deliberately. Observed HICP satisfies a *fixed*-weight
identity only approximately, because the real weights are re-based every
year, and koma would have enforced the identity anyway — silently. The gap
between the constructed series and observed headline HICP is the size of
that approximation:

```{r prices-vs-observed}
m <- stats::na.omit(cbind(koma::rate(panel3a$de_prices), koma::rate(base_panel$de_prices)))
kable(data.frame(
  statistic = c("max |difference|, pp", "mean |difference|, pp", "correlation"),
  value = c(max(abs(m[, 1] - m[, 2])), mean(abs(m[, 1] - m[, 2])), stats::cor(m[, 1], m[, 2]))
), digits = 4, caption = "Constructed de_prices vs observed headline HICP, quarterly growth.")
```

## 7. Where this leaves the block

What works: the identities are exact, the system is identified on every
seed, the labour block's economics come out with the right signs including
the wage Phillips curve, the collinearity concern that motivated the whole
identification assessment does not materialise, and the new wage–price loop
is weak enough that no posterior draw is self-sustaining.

What does not: the imports equation, for the reason diagnosed in 5.1, with a
concrete fix to try.

What to watch in phase B: `df` falls from `r Tn - k` to 6, well below stage
2b's only known-good operating point of 22. `de_import_prices` gains a
Metropolis step for the first time, so its acceptance rate becomes
meaningful and will need tuning. And the rest-of-world substitution below
gets more load-bearing.

One assumption carries more weight than its footnote suggests. Germany's
foreign-price index is renormalised over modelled partners only, because
there is no rest-of-world export-price series — and the weight this
reallocates is **`r round(attr(fpw, "row_weight_dropped"), 3)`**,
over half of Germany's trade weight. Foreign *demand* keeps that share on
`row_gdp` and accounts for the whole world; foreign *prices* cannot. The
index therefore assumes the unmodelled half of the world prices like the
modelled half. That is not a rounding detail, and it bounds how much of the
competitiveness channel this system can actually see.
