---
title: "Stage 3c: rolling the extended blocks across the euro area"
subtitle: "Why the rollout cannot happen, what the data would have allowed, and whether the extended system is better or merely bigger"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r load}
#| include: false
# Render from the project root: `quarto render reports/stage3c_rollout.qmd`.
devtools::load_all(".", quiet = TRUE)
library(knitr)

dir3c <- file.path("data", "cache", "stage3c")
dir3b <- file.path("data", "cache", "stage3b")

avail <- readRDS(file.path(dir3c, "availability.rds"))
frontier <- readRDS(file.path(dir3c, "k_frontier.rds"))
room <- readRDS(file.path(dir3c, "holdout_room.rds"))
insample <- readRDS(file.path(dir3c, "insample_rmse.rds"))
bench <- readRDS(file.path(dir3c, "bench_summary.rds"))

spill_new  <- readRDS(file.path(dir3c, "spillover_matrix.rds"))
spill_diff <- readRDS(file.path(dir3c, "spillover_diff.rds"))
sanity_cmp <- readRDS(file.path(dir3c, "sanity_cmp.rds"))
horizon    <- readRDS(file.path(dir3c, "horizon_profile.rds"))
placebo      <- readRDS(file.path(dir3c, "placebo.rds"))
placebo_diag <- readRDS(file.path(dir3c, "placebo_diag.rds"))
```

## Summary

The brief was to roll the stage-3 blocks across all eleven economies, expecting
data availability to be the constraint — Greece and Ireland were flagged as
likely to fall short.

**Data availability is not the constraint.** Most countries can support most
blocks. The constraint is the degrees-of-freedom budget, and it binds by
roughly a factor of two: the system is already at `df = 8` with **one** country
carrying the blocks, and a second fully-loaded country would need `k = 104`
against `T = 98`.

The rollout therefore does not happen, and this report is about what was
learned instead: a country × block availability matrix that would drive the
rollout if it were affordable, the exact affordability frontier, and — most
usefully — an answer to whether the extended system is actually better than
stage 2b, or merely bigger.

It is merely bigger. It fits no better in sample, cannot be validated out of
sample, and its cross-country spillover matrix — the reason the eleven countries
are estimated jointly at all — loses the sign coherence stage 2b mostly had
without gaining interpretable structure in its place. The one genuine gain is
Germany's own response to its own shock, which the new wage–price and credit
blocks amplify by ~50% over eight quarters with the lagged onset those channels
imply; a seed-placebo confirms it reproduces for Germany and not for the two
control countries.

A separate finding falls out of that placebo and applies to stage 2b too:
**individual spillover cells are not reliable in this project at 1000 draws**,
in either system. See §4.4.

## 1. Data availability: country × block

Built by auditing every source series per country: EA-MD/QD codes from the
vintage on disk, and four Eurostat datasets fetched and checked directly. The
column that matters most is not "does the series exist" but **"when does it
start"** — `T` is system-wide, so one late-starting series moves the estimation
start for every equation in the system. That is what forced the fiscal block's
redesign in stage 3b.

```{r availability}
kable(avail[, c("iso2", "labour", "external", "fiscal", "financial")],
      row.names = FALSE, caption = "Which blocks each country's data can support.")
```

```{r availability-reasons}
kable(avail[avail$blocker != "-", c("iso2", "blocker")], row.names = FALSE,
      caption = "Why a country falls short.")
```

Reading the matrix:

- **The fiscal block is universally available** — `gov_10q_ggdebt` starts 2000Q1
  or earlier for all ten euro-area countries with no internal gaps. The 2002Q1
  problem that forced stage 3b's redesign was specific to `gov_10q_ggnfa`
  (revenue and expenditure), which the reduced design does not use. Ten of
  eleven.
- **The external block is the real data constraint**, not Greece and Ireland
  alone: Belgium's and the Netherlands' current accounts start 2003, Greece's
  and Ireland's 2002. Switching the block on for all four would move the whole
  system's start to 2004Q1 and cost every country sixteen quarters.
- **The financial block is 7 of 11.** Greece and Portugal have no BIS house
  price series; Ireland has none of the four loan components used to build
  private credit. Ireland's EA-MD/QD file is the thinnest of the eleven.
- **Spain is the one price-block exception**, and cheaply — its non-energy HICP
  starts 2001Q1, a four-quarter cost.
- **The United States supports nothing beyond the core.** No FRED series for
  employment, wages, energy CPI, credit or house prices is currently fetched.

So the brief's expectation was half right: Greece and Ireland do fall short, but
so do Portugal, Spain, Belgium, the Netherlands and the US, in different blocks
and for different reasons — and *none of that is what stops the rollout*.

## 2. What stops the rollout: the degrees-of-freedom budget

```{r frontier}
kable(frontier[, c("configuration", "k", "df", "feasible")], row.names = FALSE,
      caption = "Measured against the real specification, not estimated. T = 98 throughout.")
```

Three rows carry the argument:

- **`DE + FR, all blocks` needs `k = 104` against `T = 98`.** A second fully
  loaded country is not affordable. Each country costs about fourteen columns
  (labour 7, external 1, fiscal 4, financial 2), so eleven would need roughly
  `k = 230` — more than twice the observations available.
- **`DE full + 10 partner export prices` needs `k = 100`.** This is the
  *cheapest possible* way to satisfy the brief's requirement that the
  competitiveness channel be active for every country, and it overshoots by two
  columns.
- The largest configurations that do fit are `DE full + 7 partner export prices`
  (`df = 1`) and `DE + FR labour blocks` (`df = 8`) — neither of which is a
  rollout, and the first of which has no usable degrees of freedom.

### 2.1 The bidirectional price channel degenerates before it becomes affordable

The brief asks for `<iso2>_foreign_prices` as a trade-weighted identity over
partners' export prices, making the channel two-way. That requires each partner
to have an *endogenous* export price, i.e. its own labour block. With only a
subset of countries carrying one, the "trade-weighted index" is renormalised
over whoever is left:

| labour countries | partners in DE's index | share of DE's modelled trade weight captured |
|---|---|---|
| DE, FR | 1 | 15.2% — the index *is* France |
| DE, FR, IT | 2 | 27.3% |
| DE, FR, IT, NL | 3 | 44.6% (but `df = -2`) |

And Germany's *modelled* trade weight already excludes a 57% rest-of-world
residual, so a two-country index captures roughly **6% of Germany's actual
trade**. The channel becomes affordable only in a form too degenerate to mean
anything, and meaningful only in a form that does not fit.

The one-way channel — `de_foreign_prices` exogenous, constructed from all eight
partners' observed export prices — is what stage 3a already built, and it does
capture the full modelled weight. That remains the honest version of the
competitiveness channel in this system.

## 3. Is the extended system better, or just bigger?

### 3.1 It cannot be validated out of sample

```{r holdout}
kable(room, row.names = FALSE,
      caption = "How much data each system can hold out before k >= T. 'df at that point' is what remains at the maximum hold-out.")
```

The intended test was a hold-out: estimate both systems on a truncated window,
score the held-out quarters. **The extended system cannot sit that test.** At a
2000Q1–2021Q4 estimation window (`T = 86`) stage 2b has `df = 10` and estimates
fine; the extended system needs `k = 90` and has `df = -4`.

Its maximum possible hold-out is seven quarters, and only by running at `df = 1`.
A model with no degrees of freedom to spare for validation is a model whose
generalisation cannot be checked — and that is a property of the extended system
that stage 2b does not share.

### 3.2 In-sample, it is not better either

Both systems were estimated on the identical 2000Q1–2024Q4 window, so their
in-sample RMSEs are directly comparable on the equations they share.

```{r insample}
de <- insample[grepl("^de_", insample$equation), ]
kable(de[order(de$ratio), c("equation", "rmse_stage2b", "rmse_extended", "ratio")],
      row.names = FALSE, digits = 4,
      caption = "German equations -- the only ones the blocks touch. Ratio below 1 favours the extended system.")
```

```{r insample-summary}
non <- insample[!grepl("^de_", insample$equation), ]
kable(data.frame(
  measure = c("equations where extended fits better", "median RMSE ratio (all shared equations)",
              "median ratio, non-German equations"),
  value = c(sprintf("%d of %d (%.0f%%)", sum(insample$better == "extended"), nrow(insample),
                    100 * mean(insample$better == "extended")),
            sprintf("%.4f", stats::median(insample$ratio)),
            sprintf("%.4f", stats::median(non$ratio)))
), caption = "In-sample fit across all shared equations.")
```

This is the damning number: the extended system adds nineteen equations and
fourteen predetermined columns, and the **median equation fits marginally
worse**. In-sample fit almost always improves when parameters are added — that
is what parameters do — so a median ratio slightly *above* one, with only 21% of
equations improving, says the extra structure is not buying fit even where it
cannot be penalised for complexity.

What it does buy is confined and small: German exports fit about 5% better and
consumption 2.6% better, partly offset by investment fitting 2% worse. Every
non-German equation is unchanged, as expected — the blocks do not touch them.

### 3.3 What it costs

```{r bench}
kable(bench, row.names = FALSE, digits = 4, caption = "System comparison.")
```

Fourteen degrees of freedom, nineteen equations, and about 36% more runtime.
Acceptance rates are clean in both after tuning, so the sampler is not the
problem — the budget is.

## 4. The spillover battery, re-run on the extended system

The brief asks whether the new channels changed the spillover answer. The
battery is the same one stage 2 ran: an eleven-country sweep of +1% one-quarter
GDP demand shocks, each a matched-seed `scenario_diff()`, cumulated over eight
quarters into a source × receiver matrix.

```{r spill-extended}
kable(round(spill_new, 3),
      caption = "Extended system: cumulative 8-quarter GDP response, median across draws.")
```

```{r spill-diff}
kable(round(spill_diff, 3), caption = "Change from stage 2b (extended minus stage 2b).")
```

The matrix moves a great deal: mean absolute change **0.589 percentage points**
against cross-country effects that in stage 2b were mostly below 1.0, and a
largest single change of **2.33** (`de -> ie`, from +0.05 to −2.28). Taken at
face value that is a different answer. It is not, and the rest of this section
is why.

### 4.1 The changes are only half where the blocks are

The blocks are German and only German. Every non-German equation is
structurally untouched, and §3.2 confirms it empirically — their in-sample RMSE
ratios are 1.0000. So the German row should move, and little else should.

| cells | mean abs. change |
|---|---|
| Germany as source (the blocks' own row) | 1.275 |
| Germany as receiver | 0.377 |
| **all other cells — no changed equation on either side** | **0.538** |

Cells with an unchanged shock, unchanged transmission and unchanged receiving
equation move by 0.54, **42% as much as the German row**. §4.4 measures the
noise floor at 0.35–0.39, which is most of that 0.54: those cells are moving
about as much as they move when nothing changes at all. The German row, at
1.275, is roughly three times the floor and is the only band clearly above it.

### 4.2 The sign structure collapses

```{r spill-signs}
kable(sanity_cmp, row.names = FALSE, digits = 3,
      caption = "The stage-2 sanity checks, recomputed identically on both matrices.")
```

Stage 2b's off-diagonal cells were **95% positive**: whatever its ordering
problems, a demand expansion raised partners' GDP essentially everywhere, which
is the one thing a trade-linked system must get right. The extended system is
**55% positive** — a coin flip. Fifty-one of 110 off-diagonal cells change sign,
and negative cells go from 5 to 50.

A euro-area demand expansion contracting GDP in 45% of partner countries is not
a competitiveness channel. A competitiveness channel is *selective*: it should
bite where the shocked country competes with the receiver in third markets, and
should not touch pairs with no such relationship. Sign flips distributed across
nearly half of all pairs, uncorrelated with trade weight (Spearman rho falls
from 0.141 to 0.050), have no such structure.

Two caveats, both of which the placebo in §4.4 forces. First, stage 2b's own
sign coherence is less solid than the single cached seed implies: on the three
placebo rows it is 97% positive at one seed and **70% at another**. The contrast
is real but the baseline is weaker than "95%" suggests. Second, sanity check 1
*improves*, 7/11 to 10/11 — which is not good news either, because own effects
dominate more comprehensively when the cross-country cells shrink toward noise
around zero, not because domestic transmission got sharper.

### 4.3 The disorder tracks the explosive-draw share

koma's Gibbs sampler has no stationarity constraint, so some posterior draws
explode; `scenario_diff()` reports the share. Cutting the degrees of freedom
from 22 to 8 widens the posterior, and more draws diverge:

```{r spill-horizon}
kable(horizon, row.names = FALSE, digits = 3,
      caption = "Every measure of disorder rises with the explosive-draw share.")
```

At horizon 1, where only 8% of draws have exploded, magnitudes barely move
(mean change 0.217). By horizon 8, at 67%, the change is 0.599 and the positive
share has fallen from 95% to 55%. The median at horizon 8 is being taken over a
*third* of the draws — and which third survives is itself shock-dependent.

Both series are monotone in horizon, so this correlation is close to mechanical
and is suggestive rather than decisive. That is what §4.4 is for.

### 4.4 Placebo: how much does the matrix move when nothing changes?

The cross-system difference is only evidence of a changed answer if it exceeds
the sampling noise in the estimand. So three source countries — Germany (has
the blocks), France (does not), Ireland (does not, and has the thinnest data)
— were re-run on **the same two fits**, changing nothing but the Monte Carlo
seed. Anything the cross-system difference does not clear is noise.

```{r spill-placebo}
kable(placebo, row.names = FALSE, digits = 3,
      caption = "Rows de/fr/ie, 30 off-diagonal cells. The first two lines are pure noise: identical fit, identical shock, different seed.")
```

This is the most important table in the report, and it cuts both ways.

**The cross-system difference is real, not noise.** At 0.916 it is about 2.4
times the 0.35–0.39 noise floor. Something did change.

**But no individual cell is interpretable, in either system.** Changing only the
seed moves cells by 0.35–0.39 on average, by as much as 1.85, and flips the sign
of 7 to 10 of 30. Stage 2b is barely better than the extended system on this
measure (0.346 vs 0.392). Cell-level spillover readings were never safe here;
the extended system makes a pre-existing problem worse rather than creating it.

**The sign collapse survives the placebo.** Off-diagonal positive share is
40%/30% for the extended system across the two seeds and 97%/70% for stage 2b —
overlapping nowhere. That separation is a property of the systems, not of the
draw.

### 4.5 What survives: the German diagonal, and only Germany's

The placebo doubles as a controlled experiment. If the new blocks genuinely
amplify domestic propagation, the extended-minus-stage-2b gap on a country's own
diagonal should reproduce across seeds **for Germany and for no one else** —
Germany is the only country carrying the blocks.

```{r spill-diag-placebo}
kable(placebo_diag, row.names = FALSE, digits = 3,
      caption = "Own-diagonal gap (extended minus stage 2b), measured at two independent seeds.")
```

That is exactly what happens. Germany's gap is +0.717 and +0.612 — same sign,
similar size, and comfortably above the diagonal's own seed noise of 0.22–0.33.
France's and Ireland's flip sign between seeds (+0.165/−0.143 and
+1.204/−0.168), which is what a null result looks like. Ireland's +1.204 at the
first seed would have been the report's second-largest "finding" had the placebo
not been run.

The horizon shape confirms the mechanism:

```{r spill-diag}
kable(horizon[, c("horizon", "de_diag_stage2b", "de_diag_extended")], row.names = FALSE,
      digits = 3, caption = "German own-effect, cumulative.")
```

The two systems are identical on impact (both exactly 1.000 — the restriction
pins it) and near-identical at horizon 2 (0.997 vs 0.964). They separate from
horizon 3 and the gap widens to +52% by horizon 8. A wage–price loop and a
credit–investment accelerator are slow channels: they work through wages feeding
consumption and credit feeding investment, neither of which moves in the impact
quarter. Stronger, more persistent domestic amplification of a domestic demand
shock, arriving with a lag, is precisely what those blocks were built to produce.

That is a claim about **Germany's response to its own shock** — a within-country
propagation result. It is not a spillover result, and it does not rescue the
cross-country matrix.

### 4.6 Where the new channels changed the answer

Directly, as asked: **only in Germany's own response to its own shock.** That
one result is placebo-controlled, reproduces across seeds, appears in the one
country that has the blocks and not in the two that do not, and has the right
horizon shape.

Everywhere else the matrix changed by more than noise but not in any
interpretable way. The changes are barely larger in structurally untouched cells
than the seed noise itself, they destroy a sign regularity stage 2b mostly got
right, they are uncorrelated with trade weight, and they scale with the
explosive-draw share rather than with any competitiveness linkage.

The specific figure most likely to be quoted, `de -> ie` at −2.28 (a German
expansion contracting Irish GDP by 2.3% cumulatively), should not be. **At the
placebo seed the same cell is −0.54** — a quarter of the size, from an identical
model. Stage 2b's corresponding cells are +0.05 and +0.16, stable and small.
Ireland has the thinnest data of the eleven (§1) and no equation the German
blocks touch. The largest single change in the matrix is an artefact, and it is
reported here rather than in the summary table precisely so it does not get
quoted as a spillover estimate.

## 5. The verdict

**Merely bigger.** The extended blocks are worth building as *mechanisms* — the
wage–price loop, the competitiveness channel and the debt-to-spread link are
real structures that stage 2b cannot represent at all, several of them come back
correctly signed, and the German own-response in §4.5 is a genuine,
placebo-controlled result: the wage–price and credit channels amplify a domestic
demand shock by about 50% over eight quarters, with the lagged onset those
channels imply.

But as a system the extension does not pay. In-sample gain is negligible and the
median equation fits marginally worse (§3.2). German investment regresses 2%.
Runtime rises 36%. Out-of-sample generalisation cannot be checked at all,
because the degrees of freedom to hold data out no longer exist (§3.1). And the
cross-country spillover matrix — the main product of the linked system, and the
thing eleven countries are jointly estimated *for* — loses the one sign
regularity stage 2b mostly had, without gaining any interpretable structure in
its place (§4).

That last point settles it. The extension buys one real result about Germany at
the cost of the answer the system exists to produce.

§4.4 also leaves a finding that is not about the extension at all: **cell-level
spillover readings were never reliable in this project.** Changing only the
Monte Carlo seed moves a stage-2b cell by 0.35 on average and flips the sign of
a third of them. That is a pre-existing property of a 1000-draw posterior with
explosive draws, which the extended system worsens but did not create. Stage 2b's
spillover matrix should be read as a pattern — mostly positive, ordered roughly
by size — and not cell by cell. Any future spillover work should report the
seed-noise floor alongside the estimates, which `scenario_diff()` already has
the machinery to do.

The right reading is that this project has reached the limit of what 98
quarterly observations will support in a single joint system. Further structure
needs either a longer sample or a smaller system — not a better sampler, and
not more equations. If the German blocks are wanted for their own sake, the
place to run them is a single-country model where `k < T` is not competing with
ten other countries; if the spillover matrix is wanted, stage 2b is the system
that produces it.
